### 2017-07-20 Manuscript.Rmd
Added everything to the repository

---
title: \normalsize{A general approach and tool to evaluate the effectiveness of no-take marine reserves}
header-includes:
- \usepackage{setspace}
- \doublespacing
- \usepackage{lineno}
- \linenumbers
output:
  pdf_document:
    fig_caption: yes
fontsize: 12pt
bibliography:
- exported-references.bib
- packages.bib
csl: plos.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
options(knitr.table.format = "latex")

windowsFonts(Times=windowsFont("TT Times New Roman"))

```

Juan Carlos Villaseñor-Derbez^1\*^, Caio Faro^1^, Melaina Wright^1^, Jael Martínez^1^, Alvin Suárez^2^, Stuart Fulton^2^, Sean Fitzgerald^1^, Fiorenza Micheli^3^, Gavin McDonald^4^, Jorge Torre-Cosio^2^, Christopher Costello^1,4^ \

^1^ Bren School of Environmental Science and Management, University of California Santa Barbara, Santa Barbara, California, United States

^2^ Comunidad y Biodiversidad A.C., Calle Isla del Peruano, Guaymas, Sonora, México

^3^ Hopkins Marine Station, Stanford University, Monterey, California, United States

^4^ Sustainable Fisheries Group, University of California Santa Barbara, Santa Barbara, California, United States \

\*Corresponding author

Email: jvillasenor@bren.ucsb.edu (JCVD) \

\clearpage

**Abstract**

Marine reserves are often implemented to preserve habitat and recover overfished stocks. In Mexico, recent regulations allow fishers to propose legally-recognized no-take marine reserves, established by the fishery management agency (CONAPESCA) for a period of time. While the regulation includes instructions to request their implementation, no guidelines are provided for monitoring, evaluation, or renewal. Since 2012, 35 reserves have been implemented, often managed by the communities and supported by local NGOs. Yearly biological monitorings are conducted in  reserves and their control sites. To date, the evaluation of the reserves relies solely on biological data, ignoring socioeconomic and governance dimensions. In absence of a standardized methodology, existing data is analyzed in different ways, hindering the ability to compare results across communities. We developed a framework to evaluate the effectiveness of the reserves by matching 7 commonly stated management objectives to a set of 9 biophysical, 5 socioeconomic, and 14 governance indicators. This framework includes a standardized survey to collect governance and socioeconomic data. Biophysical indicators are evaluated with a Difference-in-Difference analysis, estimating the net effect of the reserve through time. Linear regression models are fitted to socioeconomic indicators through time, testing for the difference in trends before and after the implementation of the reserve. Governance indicators are analyzed based on literature, identifying common governance structures and their associated effectiveness. To make the framework accessible to fishers and decision makers, and allow replication of results, we developed an open source application. The results produced by the application contain a report for a general audience, and a technical report that can inform the renewal and decision-making process.

```{r load packages}
suppressPackageStartupMessages({
  library(MPAtools)
  library(knitr)
  library(kableExtra)
  library(stargazer)
  library(tmap)
  library(tmaptools)
  library(rgdal)
  library(sp)
  library(magrittr)
  library(tidyverse)
})

# RefManageR::ReadBib("exported-references.bib") %>%
# View()
```

# Introduction

Overfishing and unsustainable fishing practices are some of the major threats to the conservation of marine ecosystems around the world [@pauly_2005-qV; @halpern_2008-dK]. Marine Protected Areas (MPAs) are frequently proposed as fishery management and conservation tools to help fish stocks rebound [@lester_2008-F_; @lester_2009-Ks: @sala_2016-PV] by limitting or restricting fishing effort and gears. No-take marine reserves (marine reserves hereinafter) are a particular type of MPAs, where all fishing effort and extractive activities are off-limits or highly limited. The International Union for the Conservation of Nature [-@iucn_website_2017-Bg] categorizes them as Ia (Strict Natural Reserves) or Ib (Wilderness Area). However, see [@hortaecosta_2016-_3] for a detailed regulation-based classification system.

MPAs have proven to increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience of the bounded region [@micheli_2012-EU], and preserve genetic diversity [@munguavega_2015-yg]. Compared to partially protected MPAs, marine reserves are known to have even higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_]. Often, these effects are meassured as biological changes in the area through time and often lack a control site against which to compare [@betti_2017-lq]. This before-after comparison leaves aside other factors for which one must control [@davies_2017-ml], impeding us to talk about causation with complete certainty. While some studies have used control sites, their analyses do not estimate the net effect of the reserve, and often use a control-impact comparison approach [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya]. A smaller fraction of studies use measurements of some biological atribute of a reserve and control sites before and after the implementation of the reserves, thus having a before-after-control-impact design [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], allowing them to use causal inference techniques to estimate the net effect of the reserve.

The diversity of approaches used to evaluate the effect of protected areas highlights the need to standardize approaches that enable us to truly evaluate the effect of protection. Though biophysical aspects are important to reserve success, the effectiveness of reserves also depends on the socioeconomic status and governance system of the local fishing community [@basurto_2013-oq; @basurto_2012-Mz], dimensions largely ignored until recently. Currently, the only the IUCN framework [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biophysical, socioeconomic, and governance indicators, and insight into how these indicators may be measured. However, it does not expand on how to analyze them, nor provide a user-friendly tool that enable replicability ans scalability of the analysis.

Open science and the development of tools have become ingreasingly popular to make science scalable and reproducible [@lowndes_2017-xh]. The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, has succesfully standardized a way to measure the health and benefits of the oceans. This has been succesfuly implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC]. Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance indicators [@anderson_2015-ND], design territorial use rights for fisheries [TURFs; @oyanedel_2017-TO], or to improve decision making in the hydropower industry [@vilela_2017-Zo], among others.

The lack of a comprehensive framework -or the complexity of existing ones, which alienate non-experts- and user-friendly tools to evaluate the effectiveness of marine reserves calls for the development of a new framework and tool. Here, we present a framework to evaluate no-take marine reserves, which incorporates the biological, socioeconomic, and governance dimensions of these conservation areas. We provide a list of commonly stated management objectives, and match them to appropriate indicators that meassure the effectiveness of the management intervention. We include a simple approach to analyze these indicators, building on causal inference techniques [@moland_2013-VP], largely needed to truly understand the effect of management interventions [@burgess_2016-HN]. At the same time, we introduce the Marine Reserve Evaluation App (MAREA), an open source web-based tool that automates the framework described in this paper. Case studies from three regions of Mexico are presented as proof of concept of the app.

\clearpage

# Materials and methods

Here, we describe the proposed framework to evaluate the effectiveness of marine reserves (**Fig. 1**). We explain how management objectives were identified and matched to appropriate indicators that allow the evaluation of the reserves, and provide guidelines on data collection and formating. Alongside, methodologies to analyze these indicators are presented. Then, we describe the development of MAREA and provide guidance on how to use this user-friendly open access tool. Finally, we provide guidelines on how to interpret the results and output generated by MAREA.

![Workflow to evaluate the effectiveness of no-take marine reserves.](./Figures/flowchart.gif){ width=50% }

## Reserve objectives and indicators

Throughout this work, we will refer to the stated goals for which a reserve has been established as objectives. Given that this work was motivated by the need to provide a framework to evaluate mexican marine reserves, we focused on identifying objectives of Mexican marine reserves. However, we group these objectives into seven major categories, which can be applied to other conservation areas in the world. The list of objectives was gathered through a literature review from the reserves’ stated objectives in official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, decrees, and specific legislation [@nom049sagpesc_2014-V6]. Even though each reserve -or set of reserves- has its own goals, seven main categories of objectives were identified:

1. Avoid overexploitation
2. Conserve species under a special protection regime
3. Maintain biological process (reproduction, recruitment, growth, feeding)
4. Improve fishery production in nearby waters
5. Preserve biological diversity and the ecosystem
6. Recover overexploited species
7. Recover species of economic interest

Based on these seven objectives, we determined a set of indicators to evaluate the effectiveness. The list of indicators was built through a review of scientific literature, where we identified indicators that were used to measure similar objectives. The indicators were reviewed at a workshop by members of Mexican fishery management agencies and non-government organizations. Later, these were also presented to fishers from the Ensenada Fishing Cooperative (*S.C.P.P. Ensenada*), which  provided input. Our final list of indicators includes, at least, those identified in review works such as @lester_2009-Ks or @woodcock_2017-Wm.

Indicators are divided into three main types: biophysical, socioeconomic, and governance. Biophysical indicators (n = 7) focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data section for specific sampling design and methodologies). Socioeconomic indicators (n = 3) reflect the performance of the fishery in terms of catches, income from catches and availability of alternative livelihoods. Governance indicators (n = 15) describe the governance structures under which the community operates. These indicators may be numeric (*e.g.* Fish biomass) or descriptive (*e.g.* Reasoning for reserve location). Our list includes indicators that respond to the implementation of the reserve (*i.e.* outcome variables) or that might further the understanding of its performance. In that sense, most biophysical and socioeconomic indicators are outcome variables, while governance indicators might explain poor reserve performance, for example. Whenever an indicator is said to also be applied to “Objective Species”, it means that the indicator will be used for all species (*e.g.* Fish Biomass) and individual species that are either the conservation target of the reserve, or are of particular economic or ecological interest (*e.g.* Grouper Biomass). **Table 1** presents the proposed indicators. **Table 2** shows how objectives are matched with indicators.

\clearpage

```{r, results = "asis"}
read.csv("./Tables/Table_indicators.csv") %>% 
  knitr::kable(caption = "List of indicators to evaluate the effectiveness of no-take marine reserves, grouped by type. Type of data and units are provided.", col.names = c("Code", "Indicator", "Data type", "Unit"), booktabs = T) %>%
  kable_styling(latex_options = c("scale_down")) %>%
  group_rows("Biophysical", 1, 7) %>%
  group_rows("Socioeconomic", 8, 10) %>% 
  group_rows("Governance", 11, 25) %>% 
  add_footnote(c("The indicator is applied to objective species"), notation = "symbol")
```

\clearpage

```{r}
read.csv("./Tables/Objectives_Indicators.csv") %>% 
  knitr::kable(caption = "This table indicates which indicators must be used for each objective.",
               col.names = c("Objective", "B1", "B2", "B3", "B4", "B4*", "B5", "B6", "B7", "B7*", "S1", "S1**", "S2", "S2*", "S3", "G1", "G2", "G3", "G4", "G5", "G6", "G7", "G8", "G9", "G0", "G1", "G2", "G3", "G4", "G5"),
               booktabs = T) %>% 
  kable_styling(latex_options = c("scale_down"))  %>% 
  column_spec(1, width = "4cm") %>%
  landscape()
```

## Data and analyses

Biophysical data is collected via underwater ecological surveys as part of a reserve's monitoring program, often carried out by local fishers with guidance of non-government organizations, such as Comunidad y Biodiversidad or Niparajá. Local fishers,trained in scientific diving, perform SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length. This information is recorded along 30 m transects, with a sampling window of 2 m X 2 m. Between 10 and 24 transects are performed in each site at depths between 1 and 30 m. Ecological surveys are performed yearly in each reserve and its corresponding control site(s), before and after the implementation of the reserve. Control sites are areas where habitat is similar to that of the reserve, but with presence of fishing activity.

Having control sites for biophysical data allows us to use counterfactual approaches to evaluate the net effect of the reserve. The hypothesis that the indicators will respond the the implementation of the reserve is tested by analysing spatial and temporal changes in each numeric biophysical indicator (all but B5) using generalized linear models [@moland_2013-VP]. To account for variations in the environment and survey conditions, covariates that describe the environment and gathered during the underwater ecological surveys are included into a model with form:

\begin{equation}I_{i,t,z}=\beta_0 + \sum_{t = 2}^T\gamma_{i,t}Y_t + \beta_1Z_{i,z} + \beta_2P_{i,t,z}\times Z_{i,t,z} + \beta_3T_{i,t,z} + \beta_4V_{i,t,z} + \beta_5D_{i,t,z} + \epsilon_{i,t,z}\end{equation}

In this model, $i$, $t$, and $z$ are subindices for transect, time, and zone (control or reserve site), respectively. This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), tratment ($Z$), an interaction between a variable that indicates pre- or post-implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$), visibility during the survey ($V$), and depth at which survey was performed ($D$). Here, years are modeled as factors, using the first year as the reference level. The treatment and post variables are modeled by dummy variables coded as Control = 0 and Reserve = 1; and Pre-implementation = 0 and Post-implementation = 1, respectively.

Socioeconomic data is collected by the fishers or fishery management agencies, which record landings, income, and sometimes prices for each species. In order to control for inflation and changes in buying power, income is adjusted with the countries consumer price index:

\begin{equation}I_t = RI\times \frac{CPI_t}{CPI_T}\end{equation}

Where $I_t$ represents the adjusted income for year $t$ as the product between the reported income for year $t$ and the ratio between the consumer price index ($CPI$) in that year to the present year's ($T$) CPI. Since no control sites are available, numeric socioeconomic indicators (G1 and G2) are evaluated with a simplified version of *eq. 1*:

\begin{equation}I_{t}=\beta_0 + \beta_1P_{t} + \epsilon_{t}\end{equation}

While this model does not allow assertion of a causal relation, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input.

Governance data, however, is not readily available nor systematically gathered by the community or any other organizations. In this case, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1-G15). The survey is included as supplementary material (**Appendix 1**). To analyze governance information, we performed a literature review of common governance structures and ther relation to effectiveness in managing fisheries or marine reserves. These approach has prooven to succesfuly allow the evaluation of governance structures [@espinosaromero_2014-PY]. Governance information is not analyzed in a quantitative way, but is presented along with the biophysical and socioeconomic indicators to provide managers and users with a full description of the reserve.

## Marine Reserve Evaluation App (MAREA)

MAREA was developed in R Studio [@rcore_2017] using the shiny framework [@shiny_2017]. This provides the tools to build interactive web applications, hosted on an open server. MAREA can be accessed at [turfeffect.shinyapps.io/marea/](turfeffect.shinyapps.io/marea/). In order to use the tool, a user shoudl have previously identified the management objectives of the reserve and the data. MAREA automatically selects the appropriate inicators, which can be modified by users to adjust to their interests. MAREA produces two types of results, aimed for different audiences.

The first result is a color-coded scorecard, intended to provide a general overview of the effectiveness of the reserve to a general public. The scorecard provides a global score for the reserve, a general score for each category of indicators, and an individual score for each selected indicator. The global and category scores are determined by the percentage of positive indicators, overall and for each category. For numeric biophysical indicators (all but B5), the color is defined by the sign of the interaction term coefficient ($\beta_2$) in *eq. 1*. For socioeconomic indicators, colors are assigned based on the direction of the slope ($\beta_1$). Red, yellow or green are used for $\beta_i<0$, $\beta_i = 0$, and $\beta_i>0$ respectively. The intensity of the color is defined by the significance of coefficient, testing the null hypothesis of no change (*i.e.* $H_0: \beta_i = 0$) with Student t-test. Cutoff values are $p < 0.05$ and $p < 0.1$. Thus, even in a case where $\beta_i > 0$ if the coefficient is not significant (*i.e. * $p>0.1$) the indicator will be assigned a yellow color. A legend (**Fig. 2**) is provided in the app to aid in the interpretation of these results. Governance indicators, however, are evaluated binomially and can be either red or green. The color is defined based on what literature shows to be a positive (green) or negative (red) value for each indicator. However, due to the nature of some governance indicators, which require the user to provide a narrative, only indicators XXX are presented in the scorecard.

The second output from MAREA is a technical report, intended to communicate information and statystical results in a rigurous way. This report also includes a scorecard as a summary of the results, but provides more information for each indicator. For all numeric indicators, the technical report includes graphs of the value of the indicator, for the reserve and control sites through time. It also provides a regression table [@stargazer_2015] that summarizes the value of all coefficients in the regression and their respective standard errors. The summary table also provides information on model fit ($R^2$) and significance of the regression.

The scorecard is produced with functions from the Shinydashboard package [@shinydashboard_2017]. The technical report is produced by a parameterized Rmarkdown document [@rmarkdown_2017] processed by the knitr package [@knitr_2017].

![Legend used to interpret the scorecard produced by MAREA. Colors indicate direction of change, and color intensity is given by the statistical significance.](./Figures/legend2.gif)

## Case study

We apply the proposed framework and tool to evaluate the effectiveness of one no-take marine reserves from Isla Natividad, in Baja California Sur, Mexico. Isla Natividad is located 8 Km off the Pacific Coast of the Baja Peninsula (**Fig. 3**). Fishers in Isla Natividad operate under a fishing cooperative (S.C.P.P. Buzos y Pescadores de la Baja California), which promotes co-management of marine resources [@mccay_2017-1m; @mccay_2014-CN]. Additionaly, fishers have Territorial Use Rights for Fisheries (TURFs), that provide them with exclusive rights to exploit the marine resources for which they hold permits.

In 2006, the community implemented two marine reserves within their TURF. These reserves have prooven to be effective in enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of highly valuable commercial species, like abalone [@munguavega_2015-yg]. These ecological benefits have been translated into economical benefits, enhancing population persistance and supporting abalone fisheries [@rossetto_2015-V0]. For the purpose of this evaluation, we will focus on the "La Plana / Las Cuevas" Marine reserve, located in the southern end of the Island (**Fig. 3**) and its corresponding control site "La Dulce / Babencho".

The reserve was implemented to recover species of economic interest -which have been overexplotied in the last years- in order to enhance fishery production in nearby waters. Fishers were also interested in preserving biological diversity and the ecosystem. Thus, objectives objectives 4-7 were selected. Using **Table 2** to match these objectives with appropriate management indicators, we selected all biophysical, socioeconomic, and governance indicators.

```{r read data}

fish_data <- read.csv("./Data/natividad_fish.csv", stringsAsFactors = F)

invert_data <- read.csv("./Data/natividad_invertebrates.csv", stringsAsFactors = F)

invert_data$Genero[invert_data$GeneroEspecie == "Mesocentrotus franciscanus"] <- "Strongylocentrotus"
invert_data$GeneroEspecie[invert_data$GeneroEspecie == "Mesocentrotus franciscanus"] <- "Strongylocentrotus franciscanus"

# write.csv(invert_data, "./Data/natividad_invertebrates.csv", row.names = F)

landings <- read.csv("Data/natividad_landings_conapesca.csv", stringsAsFactors = F) #%>% 
#   filter(!GeneroEspecie %in% c("Haliotis fulgens", "Haliotis corrugata"))
# 
# landings$GeneroEspecie[landings$GeneroEspecie == "Megastrea undosa"] <- "Megastraea undosa"
# write.csv(landings, "Data/natividad_landings_conapesca.csv", row.names = F)

```

```{r map, fig.cap = "General location of Isla Natividad (left) and map of the island (right). The La Plana / Las Cuevas Marine Reserve polygon is indicated in gray, and the approximate location of the La Dulce / Babencho control site is shown with no fill."}
proj <- CRS("+proj=longlat +datum=NAD27")

baja <- readOGR(dsn = "./Data/Spatial", layer = "muni_2015gw", verbose = F)

stat <- baja[baja$NOM_ENT %in% c("Baja California Sur", "Baja California", "Sonora"),]

baja2 <- baja[baja$NOM_ENT %in% c("Baja California Sur"),]

proj2 <- CRS(proj4string(baja2))

#Small shapefile for region of interest
xy1 <- data.frame(X = mean(-115.24, -115.14), Y = mean(27.85, 27.90))
coordinates(xy1) <- c("X", "Y")
proj4string(xy1) <- proj
res1 <- spTransform(xy1, proj)

# Reserve polygons
reserves <- read.csv("./Data/Spatial/baja_coordinates.csv", stringsAsFactors = F) %>% 
  filter(Location == "PlanaCuevas") %>% 
  select(lon = Longitude, lat = Latitude, id = Location) %>% 
  split(.$id) %>% 
  lapply(function(x) { x["id"] <- NULL; x }) %>% 
  sapply(Polygon) %>% 
  Polygons(ID = 1) %>% 
  list() %>% 
  SpatialPolygons(proj4string = proj2)

# Transects
transects <- data.frame(X = -115.187, Y = 27.865)
coordinates(transects) <- c("X", "Y")
proj4string(transects) <- proj2
transects <- spTransform(transects, proj2)

# Maps
pen_map <- tm_shape(stat, bbox = bb(xlim = c(-118.36648, -109.41321), ylim = c(22.87195, 32.71863))) +
  tm_polygons() +
  tm_shape(res1) +
  tm_symbols(shape = 0, size = 2, border.col = "black", col = "black", border.lwd = 2) +
  tm_scale_bar(position = c("left", "bottom"))

study_area <- tm_shape(baja2, bbox = bb(xlim = c(-115.22, -115.15), ylim = c(27.85, 27.90))) +
  tm_polygons() +
  tm_shape(reserves) +
  tm_polygons(border.col = "black", col = "gray", border.lwd = 0) +
  tm_shape(transects) +
  tm_symbols(shape = 0, size = 4, border.col = "black", col = "black", border.lwd = 1) +
  tm_compass(position = c("left", "bottom")) +
  tm_scale_bar(position = c("left", "bottom"))

tmap_arrange(pen_map, study_area, outer.margins = 0.005, ncol = 2)
```

```{r, data summaries}
fish_summary <- fish_data %>% 
  filter(RC == "La Plana / Las Cuevas - La Dulce / Babencho") %>% 
  group_by(Ano, Zona, Transecto) %>% 
  summarize(n()) %>% 
  group_by(Ano, Zona) %>% 
  summarize(n = n()) %>% 
  spread(Zona, n)

invert_summary <- invert_data %>% 
  filter(RC == "La Plana / Las Cuevas - La Dulce / Babencho") %>% 
  group_by(Ano, Zona, Transecto) %>% 
  summarize(n()) %>% 
  group_by(Ano, Zona) %>% 
  summarize(n = n()) %>% 
  spread(Zona, n)

landings_summary <- landings %>% 
  group_by(GeneroEspecie) %>% 
  summarize(Valor = sum(Ingresos, na.rm = T)) %>% 
  arrange(Valor) %>% 
  mutate(Valor = Valor/sum(Valor)*100)
```

Biophysical data was gathered by fishers from the community. The reserves have been monitored following a standardized ReefCheck protocol [@suman_2010-ez] to collct data on fish and invertebrate communities inside and outside the reserves between 2016 and 2017. A total of `r sum(fish_summary$Reserva)` and `r sum(invert_summary$Reserva)` transects were performed in the reserve site for fish and invertebrate surveys, respectively. Similar sampling effort was applied to the control site, with `r sum(fish_summary$Control)` fish and `r sum(invert_summary$Control)` invertebrate transects. Between 12 and 27 transects were performed in each site every year.

Socioeconomic data was obtained from the National Comission for Fisheries and Acuaculture (Comisión Nacional de Pesca y Acuacultura; CONAPESCA). The data contains species-level (`r length(unique(landings$GeneroEspecie))+2` spp) information on monthly landings and income from 2000 to 2014. Data on landings and income was aggeregated by year, and the yearly Consumer Price Index was included. From the `r length(unique(landings$GeneroEspecie))+2` species available, we selected as objective species those that contribute the most (88.27%) to historical income: lobster (*Panulirus interruptus*; 71.76%), red sea urchin (*Strongylocentrotus franciscanus*; 9.33%), snail (*Megastraea undosa*; 3.93%), and sea cucumber (*Parastichopus parvimensis*; 3.23%). Abalone species (*Haliotis fulgens*; 4.52 and *Haliotis corrugata*; 6.16) were excluded because the cooperative decided to implement an informal closure on the species to allow its populations to recover. Eliminating all fishing preasure on these species means that control sites receive (for this species) the same treatment.

For govenance data, we constructed the indicators based on our knowledge and familiarity of the area, as well as prototype interviews performed to liders of the fishing cooperative.

```{r source functions}
source("Functions/mpa_plot4.R")
source("Functions/my_theme.R")
source("Functions/bio_results.R")
source("Functions/bio_results_i.R")
```

```{r fish analysis}
splist <- sp_list(fish = fish_data, invert = invert_data, rc = "La Plana / Las Cuevas - La Dulce / Babencho")

data("IndListB")

values <-  list(indB = IndListB$Indicators,
                  comunidad = "Isla Natividad",
                  objsp = NULL,
                  ano.imp = 2006)

fish_results <- bio_results(values = values,
                            data = fish_data,
                            res = "La Plana / Las Cuevas",
                            con = c("La Dulce", "Babencho")) %>% 
  filter(!is.na(e))
```

```{r invert analysis}
objsp <- data.frame(sp = c("Panulirus interruptus","Strongylocentrotus franciscanus", "Megastraea undosa", "Parastichopus parvimensis"), class = "invert", stringsAsFactors = F)

values <-  list(indB = IndListB$Indicators,
                comunidad = "Isla Natividad",
                objsp = objsp,
                ano.imp = 2006)

invert_results <- bio_results_i(values = values,
                                data = invert_data,
                                res = "La Plana / Las Cuevas",
                                con = c("La Dulce", "Babencho")) %>% 
  filter(!is.na(e))
```

```{r socioeco analysis}

data("IndListS")

values <-  list(indS = IndListS$Indicators,
                comunidad = "Isla Natividad",
                objsp = objsp,
                ano.imp = 2006)

socio <- soc_results(values = values, data = landings) %>% 
  filter(!is.na(e))
```

\clearpage

# Results and Discussion

The results shown here intend to highlight the relevance and utility of the framework and tool, which automate the analysis and make it replicable. First, we present the scorecard generated by MAREA to provide an overview of the state of the reserve. Then, we provide (mdoified for style) a set of figures generated by MAREA, showing time-series for biophysical and socioeconomic indicators as well as tables containing the magnitude and significance of the $\beta_2$ and $\beta_1$. A table with the governance indicators is also provided and discussed. While presenting each result, we aknowledge possible limitations that the framework and tool might have and provide possible solutions.

![Scorecard for the "La Plana / Las Cuevas" marine reserve in Isla Natividad, Mexico. This scorecard was produced by MAREA.)](./Figures/scorecard.png){ width=50% }

```{r summary of bio results}
shannon_fish_model <- fish_results$model[[1]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

richness_fish_model <- fish_results$model[[2]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

density_fish_model <- fish_results$model[[3]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

trophic_fish_model <- fish_results$model[[4]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

biomass_fish_model <- fish_results$model[[5]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

shannon_invert_model <- invert_results$model[[1]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

richness_invert_model <- invert_results$model[[2]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

density_invert_model <- invert_results$model[[3]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

lobster_density_model <- invert_results$model[[4]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

urchin_density_model <- invert_results$model[[5]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

snail_density_model <- invert_results$model[[6]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

cucumber_density_model <- invert_results$model[[7]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

bio_results_summary <- rbind(shannon_fish_model, richness_fish_model, density_fish_model, trophic_fish_model, biomass_fish_model, shannon_invert_model, richness_invert_model, density_invert_model, lobster_density_model, urchin_density_model, snail_density_model, cucumber_density_model) %>% 
  filter(term == "ZonaReserva:Post1") %>% 
  mutate(Indicator = c("Shannon fish", "Richness fish", "Density fish", "Trophic fish", "Biomass fish", "Shannon invert", "Richness invert", "Density invert", "Lobster", "Urchin", "Snail", "Cucumber"),
         significance1 = ifelse(p.value < 0.1, "*", ""),
         significance2 = ifelse(p.value < 0.05, "*", ""),
         significance = paste0(significance1, significance2)) %>% 
  select(Indicator, estimate, std.error, statistic, p.value, significance)
```

```{r summary of soc results}
landings_model <- socio$model[[1]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

income_model <- socio$model[[2]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

landings_lobster_model <- socio$model[[3]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

landings_urchin_model <- socio$model[[4]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

landings_snail_model <- socio$model[[5]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

landings_cucumber_model <- socio$model[[6]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

income_lobster_model <- socio$model[[7]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

income_urchin_model <- socio$model[[8]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

income_snail_model <- socio$model[[9]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

income_cucumber_model <- socio$model[[10]] %>% 
  lmtest::coeftest(vcov = sandwich::vcovHC(.,type = "HC1")) %>%
  broom::tidy()

socio_results_summary <- rbind(landings_model, income_model, landings_lobster_model, landings_urchin_model, landings_snail_model, landings_cucumber_model, income_lobster_model, income_urchin_model, income_snail_model, income_cucumber_model) %>% 
  filter(term == "Post1") %>% 
  mutate(Indicator = c("Landings", "Income", "Lobster landings", "Urchin landings", "Snail landings", "Cucumber landings", "Lobster income", "Urchin income", "Snail income", "Cucumber income"),
         significance1 = ifelse(p.value < 0.1, "*", ""),
         significance2 = ifelse(p.value < 0.05, "*", ""),
         significance = paste0(significance1, significance2)) %>% 
  select(Indicator, estimate, std.error, statistic, p.value, significance)
```

The scorecard shows that, overall, the reserve has a good performance achieving a general score of 63.6% of positive indicators. All category-level scores were also high, with values of 66.7%, 60%, and 75% positive indicators for Biophysical, Socioeconomic and Governance, respectively.

Within the biophysical indicators, the greatest effect of the reserve was observed for snail and cucumber densities, with values of $\beta_3 = `r formatC(bio_results_summary$estimate[11], digits = 2, format = "f")`$ (*p* < 0.05) and $\beta_3 = `r formatC(bio_results_summary$estimate[12], digits = 2, format = "f")`$ (*p* < 0.05), respectively. Urchin and lobster densities also increased by $\beta_3 = `r formatC(bio_results_summary$estimate[10], digits = 2, format = "f")`$ and $\beta_3 = `r formatC(bio_results_summary$estimate[9], digits = 2, format = "f")`$ respectively. However, changes in urchin densities were only significant at *p* < 0.1, and not significant for lobster (*p* > 0.1). Shannon's diversity index ($\beta_3 = `r formatC(bio_results_summary$estimate[6], digits = 2, format = "f")`$) and richness $\beta_3 = `r formatC(bio_results_summary$estimate[7], digits = 2, format = "f")`$ showed a significant decrease (*p* < 0.05). Fish indicators showed no significant change (*p* > 0.1), with negative trends for Shannon's diversity index and fish species richness and positive trends for density, biomass, and mean trophic level. Changes through time for these indicators are presented in **Figure 5**, and a summary of $\beta_3$ coefficients is provided in **Table 3**.

In the case of socioeconomic indicators, total landings were on average 64.20 tones higher (*p* > 0.1) after the implementation of the reserves. Total income was \$10,344.85 (*p* < 0.05) thousands of Mexican Pesos (MXP) higher after the implementation of the reserves. On avergae, lobster and cucumber landings and increased, while urchin and snail landings and income decreased. **Figure 6** presents the changes in this indicators through time, and **Table 4** summarizes this information.

All invertebrates show increased densities within the reserve, as compared to the control site, providing evidence that the reserve is effectively protecting these species.  While we are not able to infer causality from the socioeconomic analysis, we are able to identify important trends. For example, snail and urchin density hve significantly increased within the reserve, but their landings and income have decreased. The opposite is observed for lobster and cucumber, which have shown increases in densities, landing, and income. While further information on market behaviour of each species is needed, these results provide insights into the state of the reserve, as well as the associated fisheries.

In terms of governance, it is evident that the community is strongly organized, which is a cause of their success. The first point of success is the existance of a fishing cooeprative that is also affiliated to federation. These polycentric governance structures allow different levels of organization that foster communication and cooperation; federations also provide bargain power with governments [@espinosaromero_2014-PY; @finkbeiner_2015-87]. Fishers also have good management instruments. Access to the fishing resources they exploit is managed through permits and fishing quotas. Along with a stable number of fishers participating in extractive activities, these limit the total fishing effort effort applied. Additionally, their TURF promotes a sense of stewardship of their resources and incentivizes correct resource management [@mccay_2017-1m; @finkbeiner_2015-87]. Together, these structures enabled a participative, bottom-up process during the reserve design phase; Opinions of all fishing members -and often non-fishers, but community members- were included. Participation of community members in reserve survailance and yearly monitoring indicate commitment and interest, and allow informal communication of results to un-involved community members.

While the reserve has not been legaly recognized (as of this moment), the community has started the process to have the reserve be recognized as a fishery replenishment zone (*Zona de Refugio Pesquer*) under the corresponding Mexican norms [@nom049sagpesc_2014-V6]. Furthermore, the reserve is partially isolated from poaching activity and fishers have internal regulations pertaining the reserves. The low level of illegal fishing by members of the community and outsiders both inside and outside the reserve represents another indication of effectiveness. A summary of governance indicators is provided in **Table 5**.

\clearpage

```{r table of bio results, results = "asis"}
bio_results_summary %>% 
  mutate(estimate = paste0(formatC(estimate, digits = 2, format = "f"),
                           " (",
                           formatC(std.error, digits = 2, format = "f"),
                           ")",
                           significance)) %>% 
  select(Indicator, estimate, tscore = statistic) %>% 
  kable(booktabs = T,
        caption = "Summary of average treatment effect of the La Plana / Las Cuevas marine reserve on biophysical indicators. Asterisks indicate significance level, with (*) indicating *p* < 0.1 and (**) *p* < 0.05.",
        digits = 2,
        col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
  kable_styling()
```

```{r table of soc results, results = "asis"}
socio_results_summary %>% 
  mutate(estimate = paste0(formatC(estimate, digits = 2, format = "f"),
                           " (",
                           formatC(std.error, digits = 2, format = "f"),
                           ")",
                           significance)) %>% 
  select(Indicator, estimate, tscore = statistic) %>% 
  kable(booktabs = T,
        caption = "Summary of average treatment effect of the La Plana / Las Cuevas marine reserve on socioeconomic indicators. Asterisks indicate significance level, with (*) indicating *p* < 0.1 and (**) *p* < 0.05.",
        digits = 2,
        col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
  kable_styling()
```

\clearpage

```{r plots, fig.height = 6, fig.width = 6, fig.cap = "Plots for values of each biophysical indicator (y-axis) through time (x-axis). Red and blue correspond to the reserve and control sites, respectively. Black lines indicate yearly mean values, and ribbons indicate +/- 1 standard error. Dots are horizontally jittered to aid visualization."}
shannon_fish_plot <- fish_results$plot[[1]] %>% 
  my_theme(y.lab = "H'")

richness_fish_plot <- fish_results$plot[[2]] %>% 
  my_theme(y.lab = "S (spp/transect)")

density_fish_plot <- fish_results$plot[[3]] %>% 
  my_theme(y.lab = "D (org/transect)")

trophic_fish_plot <- fish_results$plot[[4]] %>% 
  my_theme(y.lab = "TL")

biomass_fish_plot <- fish_results$plot[[5]] %>% 
  my_theme(y.lab = "B (Kg/transect)")

#invert

shannon_invert_plot <- invert_results$plot[[1]] %>% 
  my_theme(y.lab = "H'")

richness_invert_plot <- invert_results$plot[[2]] %>% 
  my_theme(y.lab = "S (spp/transect)")

density_invert_plot <- invert_results$plot[[3]] %>% 
  my_theme(y.lab = "D (org/transect)")

lobster_density_invert_plot <- invert_results$plot[[4]] %>% 
  my_theme(y.lab = "lobsters (org/transect)")

urchin_density_invert_plot <- invert_results$plot[[5]] %>% 
  my_theme(x.lab = "year", y.lab = "urchin (org/transect)")

snail_density_invert_plot <- invert_results$plot[[6]] %>% 
  my_theme(x.lab = "year", y.lab = "snail (org/transect)")

cucumber_density_invert_plot <- invert_results$plot[[7]] %>% 
  my_theme(x.lab = "year", y.lab = "cucumber (org/transect)")

gridExtra::grid.arrange(shannon_fish_plot, richness_fish_plot, density_fish_plot, trophic_fish_plot, biomass_fish_plot, shannon_invert_plot, richness_invert_plot, density_invert_plot, lobster_density_invert_plot, urchin_density_invert_plot, snail_density_invert_plot, cucumber_density_invert_plot, ncol = 3)
```

\clearpage

```{r time series socio, fig.height = 6, fig.width = 6, fig.cap = "Plots for values of each socioeconomic indicator (y-axis) through time (x-axis). Red and blue correspond to before and after the implementation of the reserve, respectively."}
landings_plot <- socio$plot[[1]] %>% 
  my_theme(y.lab = "Landings (Kg)")

income_plot <- socio$plot[[2]] %>% 
  my_theme(y.lab = "Income (MXN)")

landings_lobster_plot <- socio$plot[[3]] %>% 
  my_theme(y.lab = "Lobster landings")

landings_urchin_plot <- socio$plot[[4]] %>% 
  my_theme(y.lab = "Urchin landings")

landings_snail_plot <- socio$plot[[5]] %>% 
  my_theme(y.lab = "Snail landings")

landings_cucumber_plot <- socio$plot[[6]] %>% 
  my_theme(y.lab = "Cucumber landings")

income_lobster_plot <- socio$plot[[7]] %>% 
  my_theme(y.lab = "Lobster income")

income_urchin_plot <- socio$plot[[8]] %>% 
  my_theme(y.lab = "Urchin income")

income_snail_plot <- socio$plot[[9]] %>% 
  my_theme(y.lab = "Snail income")

income_cucumber_plot <- socio$plot[[10]] %>% 
  my_theme(y.lab = "Cucumber income")

gridExtra::grid.arrange(landings_plot, income_plot, landings_lobster_plot, landings_urchin_plot, landings_snail_plot, landings_cucumber_plot, income_lobster_plot, income_urchin_plot, income_snail_plot, income_cucumber_plot, ncol = 3)
```

\clearpage

```{r}
read.csv("Tables/Governance_data.csv", col.names = c("Indicator", "Description"), stringsAsFactors = F) %>% 
  kable(booktabs = T,
        caption = "Summary of governance indicators.") %>%
  kable_styling(full_width = F, latex_options = c("scale_down")) %>% 
  column_spec(2, width = "14cm")
```

\clearpage

# Conclusions

# Acknowledgements

The authors wish to thank the fishers from Isla Natividad, who gathered the data used in this study for over 10 years, and the fishers from El Rosario, who helped us validate our survey and framework. We also thank Olivier Deschenes and Andrew Plantinga, who provided valuable input to design the model that evaluates the biophysical indicators.

## Funding

# Other plots

```{r ATE socio, eval = F, include =F}
socio_results_summary %>% 
  transform(Indicator = reorder(Indicator, -estimate)) %>% 
  ggplot(aes(x = Indicator, y = estimate)) +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0.25, size = 1) +
  geom_point(aes(color = significance), size = 2) +
  theme_bw() +
  coord_flip() +
  geom_hline(yintercept = 0) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "None") +
  labs(x = "Indicator", y = "Average Treatment Effect") +
      theme(text = element_text(family = "Times", size = 10))
```

```{r, ATE plot, fig.width = 6, fig.height = 3, eval = F, include =F}
plot_ATE_all_bio <- bio_results_summary %>% 
  transform(Indicator = reorder(Indicator, -estimate)) %>% 
  ggplot(aes(x = Indicator, y = estimate)) +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0.25, size = 1) +
  geom_point(aes(color = significance), size = 2) +
  theme_bw() +
  coord_flip() +
  geom_hline(yintercept = 0) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "None") +
  labs(x = "Indicator", y = "Average Treatment Effect") +
      theme(text = element_text(family = "Times", size = 10))

plot_ATE_small_bio <- bio_results_summary %>% 
  filter(estimate < 10) %>% 
  transform(Indicator = reorder(Indicator, -estimate)) %>% 
  ggplot(aes(x = Indicator, y = estimate)) +
  geom_errorbar(aes(ymin = estimate - std.error, ymax = estimate + std.error), width = 0.25, size = 1) +
  geom_point(aes(color = significance), size = 2) +
  theme_bw() +
  coord_flip() +
  geom_hline(yintercept = 0) +
  scale_color_brewer(palette = "Set1") +
  theme(legend.justification = c(1,1),
        legend.position = c(0.95,0.95),
        legend.background = element_rect(fill = "transparent", size=.25)) +
  labs(x = "Indicator", y = "Average Treatment Effect") +
      theme(text = element_text(family = "Times", size = 10))

gridExtra::grid.arrange(plot_ATE_all_bio, plot_ATE_small_bio, ncol = 2)
```

# References

### 2017-07-20 Functions/bio_results.R

@@ -1,0 +1,447 @@
+bio_results <- function (values, data, res, con) {
+  library(broom)
+  results <- tibble::tibble(Ind = c("Shannon", "Riqueza", 
+                                    "LT50Obj1", "Densidad", "DensidadObj1", "NivelTrofico", 
+                                    "Biomasa", "BiomasaObj1", "LT50Obj2", "LT50Obj3", "LT50Obj4", 
+                                    "LT50Obj5", "LT50Obj6", "LT50Obj7", "LT50Obj8", "BiomasaObj2", 
+                                    "BiomasaObj3", "BiomasaObj4", "BiomasaObj5", "BiomasaObj6", 
+                                    "BiomasaObj7", "BiomasaObj8", "DensidadObj2", "DensidadObj3", 
+                                    "DensidadObj4", "DensidadObj5", "DensidadObj6", "DensidadObj7", 
+                                    "DensidadObj8"), e = NA, p = NA, string = NA, color = NA, 
+                            model = list(NA), plot = list(NA))
+  if ("Indice de diversidad de Shannon" %in% values$indB) {
+    model <- shannon(data, values$comunidad) %>% turfeffect(res, 
+                                                            con, type = "bio", year.imp = values$ano.imp)
+    TidyModel <- model$TidyModel %>% filter(term == "ZonaReserva:Post1")
+    model <- model$model
+    results$e[1] <- TidyModel$estimate
+    results$p[1] <- TidyModel$p.value
+    results$string[1] <- valueBoxString(TidyModel)
+    results$color[1] <- bio_score(TidyModel)
+    results$model[[1]] <- model
+    results$plot[[1]] <- mpa_plot4(model, y.lab = "Shannon (H'/Transecto)")
+  }
+  if ("Riqueza" %in% values$indB) {
+    model <- richness(data, values$comunidad) %>% turfeffect(res, 
+                                                             con, type = "bio", year.imp = values$ano.imp)
+    TidyModel <- model$TidyModel %>% filter(term == "ZonaReserva:Post1")
+    model <- model$model
+    results$e[2] <- TidyModel$estimate
+    results$p[2] <- TidyModel$p.value
+    results$string[2] <- valueBoxString(TidyModel)
+    results$color[2] <- bio_score(TidyModel)
+    results$model[[2]] <- model
[…]

### 2017-07-20 Functions/bio_results_i.R

@@ -1,0 +1,226 @@
+bio_results_i <- function(values, data, res, con) {
+  library(broom)
+  # Establish a dataframe where to store all values
+  results <- tibble::tibble(
+    Ind = c("Shannon", "Riqueza", "Densidad", "DensidadObj1", "DensidadObj2", "DensidadObj3", "DensidadObj4", "DensidadObj5", "DensidadObj6", "DensidadObj7", "DensidadObj8"),
+    e = NA,
+    p = NA,
+    string = NA,
+    color = NA,
+    model = list(NA),
+    plot = list(NA)
+  )
+  
+  ####
+  if ("Indice de diversidad de Shannon" %in% values$indB) {
+    model <- shannon(data, values$comunidad) %>%
+      turfeffect(res, con, type = "bio", year.imp = values$ano.imp)
+    
+    TidyModel <- model$TidyModel %>%
+      filter(term == "ZonaReserva:Post1")
+    
+    model <- model$model
+    
+    results$e[1] <- TidyModel$estimate
+    results$p[1] <- TidyModel$p.value
+    results$string[1] <- valueBoxString(TidyModel)
+    results$color[1] <- bio_score(TidyModel)
+    results$model[[1]] <- model
+    results$plot[[1]] <- mpa_plot4(model, y.lab = "Shannon (H'/Transecto)")
+  }
+  
+  ####
+  if ("Riqueza" %in% values$indB) {
+    model <- richness(data, values$comunidad) %>%
+      turfeffect(res, con, type = "bio", year.imp = values$ano.imp)
+    
+    TidyModel <- model$TidyModel %>%
+      filter(term == "ZonaReserva:Post1")
+    
+    model <- model$model
+    
+    results$e[2] <- TidyModel$estimate
+    results$p[2] <- TidyModel$p.value
+    results$string[2] <- valueBoxString(TidyModel)
+    results$color[2] <- bio_score(TidyModel)
+    results$model[[2]] <- model
+    results$plot[[2]] <- mpa_plot4(model, y.lab = "Riqueza (Especies/Transecto)")
+  }
+  
+  ####
+  if ("Densidad" %in% values$indB) {
+    model <- density(data, values$comunidad) %>%
+      turfeffect(res, con, type = "bio", year.imp = values$ano.imp)
+    
+    TidyModel <- model$TidyModel %>%
+      filter(term == "ZonaReserva:Post1")
+    
+    model <- model$model
+    
[…]

### 2017-07-20 Functions/mpa_plot4.R

@@ -1,0 +1,16 @@
+mpa_plot4 <- function (model, y.lab = NULL) {
+  library(ggplot2)
+  library(dplyr)
+  library(tidyr)
+  p <- model$model %>%
+    mutate(Ano = as.numeric(as.character(Ano))) %>%
+    ggplot(aes(x = Ano, y = Indicador, color = Zona, fill = Zona)) +
+    geom_jitter(height = 0, width = 0.1, size = 0.25) +
+    stat_summary(fun.data = mean_se, geom = "ribbon", alpha = 0.5) +
+    stat_summary(geom = "line", fun.y = mean, color = "black", alpha = 0.5) +
+    scale_color_brewer(palette = "Set1", direction = -1) +
+    scale_fill_brewer(palette = "Set1",direction = -1) + 
+    theme_bw() + 
+    labs(x = "Ano", y = y.lab)
+  return(p)
+}

### 2017-07-20 Functions/my_theme.R

@@ -1,0 +1,27 @@
+my_theme <- function(my_plot, legend = F, x.lab = "", y.lab = ""){
+  
+  if (legend){
+    my_plot <- my_plot +
+      theme(legend.justification=c(0,1),
+            legend.position=c(0,1),
+            legend.background = element_rect(fill = "transparent", size=.25),
+            legend.key.size = unit(0.5, "cm")) +
+      guides(color = guide_legend(title = NULL),
+             fill = guide_legend(title = NULL)) +
+      labs(x = x.lab, y = y.lab) +
+      scale_x_continuous(breaks = c(2006, 2008, 2010, 2012, 2014, 2016),
+                         labels = c("06", "08", "10", "12", "14", "16")) +
+      theme(text = element_text(family = "Times", size = 10))
+  }
+  
+  if (!legend){
+    my_plot <- my_plot +
+      theme(legend.position = "None") +
+      labs(x = x.lab, y = y.lab) +
+      scale_x_continuous(breaks = c(2006, 2008, 2010, 2012, 2014, 2016),
+                         labels = c("06", "08", "10", "12", "14", "16")) +
+      theme(text = element_text(family = "Times", size = 10))
+  }
+  
+  return(my_plot)
+}

### 2017-07-21 Manuscript.Rmd
Added conclusions and gave some formatting

@@
 
-Juan Carlos Villaseñor-Derbez^1\*^, Caio Faro^1^, Melaina Wright^1^, Jael Martínez^1^, Alvin Suárez^2^, Stuart Fulton^2^, Sean Fitzgerald^1^, Fiorenza Micheli^3^, Gavin McDonald^4^, Jorge Torre-Cosio^2^, Christopher Costello^1,4^ \
+Juan Carlos Villaseñor-Derbez^1\*^, Caio Faro^1^, Melaina Wright^1^, Jael Martínez^1^, Alvin Suárez^2^, Stuart Fulton^2^, Sean Fitzgerald^1^, Gavin McDonald^4^, Jorge Torre-Cosio^2^, Fiorenza Micheli^3^, Christopher Costello^1,4^ \
 
@@
 Marine reserves are often implemented to preserve habitat and recover overfished stocks.
-In Mexico, recent regulations allow fishers to propose legally-recognized no-take marine reserves, established by the fishery management agency (CONAPESCA) for a period of time.
-While the regulation includes instructions to request their implementation, no guidelines are provided for monitoring, evaluation, or renewal.
-Since 2012, 35 reserves have been implemented, often managed by the communities and supported by local NGOs.
-Yearly biological monitorings are conducted in  reserves and their control sites.
+
 To date, the evaluation of the reserves relies solely on biological data, ignoring socioeconomic and governance dimensions.
@@
 The list of indicators was built through a review of scientific literature, where we identified indicators that were used to measure similar objectives.
-The indicators were reviewed at a workshop by members of Mexican fishery management agencies and non-government organizations.
+A first filter eliminated indicators for which baseline data does not exist, or for which data aquisition was expensive or outside of the general scope.
+The initial list of indicators was reviewed at a workshop with participation of members from Mexican fishery management agencies and non-government organizations.
 Later, these were also presented to fishers from the Ensenada Fishing Cooperative (*S.C.P.P.
@@
 This provides the tools to build interactive web applications, hosted on an open server.
-MAREA can be accessed at [turfeffect.shinyapps.io/marea/](turfeffect.shinyapps.io/marea/).
-In order to use the tool, a user shoudl have previously identified the management objectives of the reserve and the data.
-MAREA automatically selects the appropriate inicators, which can be modified by users to adjust to their interests.
-MAREA produces two types of results, aimed for different audiences.
+MAREA can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/).
+While the original version was developed in spanish (aimed mainly at Latinamerican countries), all its content can be translated by a translation tool availabe within the environment.
+
+MAREA is designed into a 6-step process.
+The first part provides an introduction to the app, and provides the user with an overview of the process.
+Then, the user must select management objectives, and MAREA will automatically select appropriate indicators, based on **Table 2**.
+Then, the user can load the data, using standard \*.csv formatting; sample datasets are provided in the environment.
+Once data have been loaded, MAREA identifies all reserves in the data, and lets the user select the reserve to be evaluated.
+At this point, the user can also specify the year of implementation of the reserve, as well as select objective species that are of particular management interest.
+Before presenting the results MAREA provides the user with a section to confirm that all the decisions taken before are correct.
+Finally, the user is taken to the results, which are represented in two ways.
 
@@
 
+![Legend used to interpret the scorecard produced by MAREA. Colors indicate direction of change, and color intensity is given by the statistical significance.](./Figures/legend2.gif)
+
 The second output from MAREA is a technical report, intended to communicate information and statystical results in a rigurous way.
@@
 The technical report is produced by a parameterized Rmarkdown document [@rmarkdown_2017] processed by the knitr package [@knitr_2017].
-
-![Legend used to interpret the scorecard produced by MAREA. Colors indicate direction of change, and color intensity is given by the statistical significance.](./Figures/legend2.gif)
+Another feature of MAREA is that the user can choose to share the data.
+Once the technical report is downloaded, information on the reserve, its management objectives, and all uploaded data is saved into a central repository.
+These data can be accessed at any time by any person interested in aquiring them.
 
 ## Case study
-
-We apply the proposed framework and tool to evaluate the effectiveness of one no-take marine reserves from Isla Natividad, in Baja California Sur, Mexico.
-Isla Natividad is located 8 Km off the Pacific Coast of the Baja Peninsula (**Fig.
-3**).
-Fishers in Isla Natividad operate under a fishing cooperative (S.C.P.P.
-Buzos y Pescadores de la Baja California), which promotes co-management of marine resources [@mccay_2017-1m; @mccay_2014-CN].
-Additionaly, fishers have Territorial Use Rights for Fisheries (TURFs), that provide them with exclusive rights to exploit the marine resources for which they hold permits.
-
-In 2006, the community implemented two marine reserves within their TURF.
-These reserves have prooven to be effective in enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of highly valuable commercial species, like abalone [@munguavega_2015-yg].
-These ecological benefits have been translated into economical benefits, enhancing population persistance and supporting abalone fisheries [@rossetto_2015-V0].
-For the purpose of this evaluation, we will focus on the "La Plana / Las Cuevas" Marine reserve, located in the southern end of the Island (**Fig.
-3**) and its corresponding control site "La Dulce / Babencho".
-
-The reserve was implemented to recover species of economic interest -which have been overexplotied in the last years- in order to enhance fishery production in nearby waters.
-Fishers were also interested in preserving biological diversity and the ecosystem.
-Thus, objectives objectives 4-7 were selected.
-Using **Table 2** to match these objectives with appropriate management indicators, we selected all biophysical, socioeconomic, and governance indicators.
 
@@
 
+We apply the proposed framework and tool to evaluate the effectiveness of one no-take marine reserves from Isla Natividad, in Baja California Sur, Mexico.
+Isla Natividad is located 8 Km off the Pacific Coast of the Baja Peninsula (**Fig.
+3**).
+Fishers in Isla Natividad operate under a fishing cooperative (S.C.P.P.
+Buzos y Pescadores de la Baja California), which promotes co-management of marine resources [@mccay_2017-1m; @mccay_2014-CN].
+Additionaly, fishers have Territorial Use Rights for Fisheries (TURFs), that provide them with exclusive rights to exploit the marine resources for which they hold permits.
+
 ```{r map, fig.cap = "General location of Isla Natividad (left) and map of the island (right).
-The La Plana / Las Cuevas Marine Reserve polygon is indicated in gray, and the approximate location of the La Dulce / Babencho control site is shown with no fill."}
+The marine reserve polygon is indicated in gray, and the approximate location of the control site is shown with no fill."}
 proj <- CRS("+proj=longlat +datum=NAD27")
@@
 
+In 2006, the community implemented two marine reserves within their TURF.
+These reserves have prooven to be effective in enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of highly valuable commercial species, like abalone [@munguavega_2015-yg].
+These ecological benefits have been translated into economical benefits, enhancing population persistance and supporting abalone fisheries [@rossetto_2015-V0].
+For the purpose of this evaluation, we will focus on the "La Plana / Las Cuevas" Marine reserve, located in the southern end of the Island (**Fig.
+3**) and its corresponding control site "La Dulce / Babencho".
+
+The reserve was implemented to recover species of economic interest -which have been overexplotied in the last years- in order to enhance fishery production in nearby waters.
+Fishers were also interested in preserving biological diversity and the ecosystem.
+Thus, objectives objectives 4-7 were selected.
+Using **Table 2** to match these objectives with appropriate management indicators, we selected all biophysical, socioeconomic, and governance indicators.
+
 ```{r, data summaries}
@@
 
-For govenance data, we constructed the indicators based on our knowledge and familiarity of the area, as well as prototype interviews performed to liders of the fishing cooperative.
+For govenance data, we constructed the database based on our knowledge of the area, as well as prototype interviews performed to liders of the fishing cooperative.
 
@@
 
-\clearpage
-
 # Results and Discussion
@@
 First, we present the scorecard generated by MAREA to provide an overview of the state of the reserve.
-Then, we provide (mdoified for style) a set of figures generated by MAREA, showing time-series for biophysical and socioeconomic indicators as well as tables containing the magnitude and significance of the $\beta_2$ and $\beta_1$.
+Then, we provide (mdoified for style) a set of figures generated by MAREA, showing time-series for biophysical and socioeconomic indicators as well as tables containing the magnitude and significance of the $\beta_2$ and $\beta_1$ coefficients.
 A table with the governance indicators is also provided and discussed.
-While presenting each result, we aknowledge possible limitations that the framework and tool might have and provide possible solutions.
+
+The scorecard (**Fig.
+4**) shows that, overall, the reserve has a good performance achieving a general score of 63.6% of positive indicators.
+All category-level scores were also high, with values of 66.7%, 60%, and 75% positive indicators for Biophysical, Socioeconomic and Governance, respectively.
+
+\clearpage
 
@@
 
-The scorecard shows that, overall, the reserve has a good performance achieving a general score of 63.6% of positive indicators.
-All category-level scores were also high, with values of 66.7%, 60%, and 75% positive indicators for Biophysical, Socioeconomic and Governance, respectively.
-
 Within the biophysical indicators, the greatest effect of the reserve was observed for snail and cucumber densities, with values of $\beta_3 = `r formatC(bio_results_summary$estimate[11], digits = 2, format = "f")`$ (*p* < 0.05) and $\beta_3 = `r formatC(bio_results_summary$estimate[12], digits = 2, format = "f")`$ (*p* < 0.05), respectively.
@@
 A summary of governance indicators is provided in **Table 5**.
-
-\clearpage
 
@@
 
+One of the greatest challenges in management measuring the extent to which objectives have been met.
+The present framework provides a simple and straightforward way to align management objectives with performance indicators.
+We aknowledge that these 25 indicators might not fully describe a particular reserve.
+However, we believe that these provide a starting point to perform the evaluation, and urge decision-makers, managers, and users to include other indicators (*e.g.* larval dispersal or connectivity) that are relevant to their reserve.
+
+The proposed methodologies, specially the way in which biophysical indicators are evaluated, provide valuable information for managers.
+The analysis isolates the net effect of the reserve, providing a propper measure of reserve effectiveness.
+We agree there is room for improvement in the way in which socioeconomic and governance data are analized.
+However, we believe that providing a unifying platform where all can be analyzed and comprehensively presented represents a valuable step towards effective management.
+
+The main value of MAREA is that it provides a free, simple, and replicable way to perform rigurous analysis.
+The tool can easily be used by fishers, NGO members, and managers in government agencies, providing transparency of the analysis and results.
+The way in which results are presented allow this information to be interpreted by a wider audience.
+The scorecard is easily understandable by experts and non-experts, and can be used as an effective tool for communicating the results of yearly evaluations.
+On the other side, the technical report can serve as a tool for managers and scientists to rapidly communicate information at a more technical level.
+
+While the first release of MAREAis now available, it will continue to be developed and maintained.
+This will incorporate new features, and enhance current ones, aiming to improve user experience and expand the scope of the analysis.
+Yet, we believe that this first release represents a major step towards effective management of marine reserves.
+
 # Acknowledgements
 
-The authors wish to thank the fishers from Isla Natividad, who gathered the data used in this study for over 10 years, and the fishers from El Rosario, who helped us validate our survey and framework.
+The authors wish to thank the fishers from Isla Natividad, who gathered the data used in this study, and the fishers from El Rosario, who helped us validate our survey and framework.
+Input provided by the participants in the workshop "" was very valuable.
 We also thank Olivier Deschenes and Andrew Plantinga, who provided valuable input to design the model that evaluates the biophysical indicators.
 
-## Funding
-
-# Other plots
+\clearpage
 

### 2017-07-23 Manuscript.Rmd
Finalized manuscript

@@
 - packages.bib
-csl: plos.csl
+# csl: plos.csl
 ---
@@
 
-Marine reserves are often implemented to preserve habitat and recover overfished stocks.
-
-To date, the evaluation of the reserves relies solely on biological data, ignoring socioeconomic and governance dimensions.
-In absence of a standardized methodology, existing data is analyzed in different ways, hindering the ability to compare results across communities.
-We developed a framework to evaluate the effectiveness of the reserves by matching 7 commonly stated management objectives to a set of 9 biophysical, 5 socioeconomic, and 14 governance indicators.
-This framework includes a standardized survey to collect governance and socioeconomic data.
-Biophysical indicators are evaluated with a Difference-in-Difference analysis, estimating the net effect of the reserve through time.
-Linear regression models are fitted to socioeconomic indicators through time, testing for the difference in trends before and after the implementation of the reserve.
-Governance indicators are analyzed based on literature, identifying common governance structures and their associated effectiveness.
-To make the framework accessible to fishers and decision makers, and allow replication of results, we developed an open source application.
-The results produced by the application contain a report for a general audience, and a technical report that can inform the renewal and decision-making process.
+Marine reserves are often implemented to preserve habitat, recover overfished stocks, and secure livelihoods of coastal communities.
+As with any other management interventions, we need to better understand their effectiveness and impacts on coastal communities and the environment.
+To date, the evaluation of the reserves largely relies on analysing ecological data, ignoring socioeconomic and governance dimensions.
+In absence of a standardized methodology, existing data is analyzed in different ways, hindering the ability to compare results across case studies.
+We present a framework and tool to evaluate the effectiveness of marine reserves by matching 7 commonly stated management objectives to a set of 9 biophysical, 5 socioeconomic, and 14 governance indicators.
+We provide guidelines on how to properly colet data that can then be analysied with standardized methods Biophysical indicators are evaluated with causal inference techniques, using a counterfactual approach, to measure the net effect of the reserve on each indicator.
+Linear regression models are fitted to socioeconomic indicators through time, testing for the differences before and after the implementation of the reserve.
+Governance indicators are quantitatively analyzed based on literature, identifying common governance structures and their associated effectiveness.
+To make the framework accessible to fishers and decision makers, and allow replication of results, we developed an open source, web-based Marine Reserve Evaluation App (MAREA).
+Together, this framework and MAREA represent a major step in marine reserve evaluation that can furhter our understanding and simplify our management of marine reserves.
 
@@
 Marine Protected Areas (MPAs) are frequently proposed as fishery management and conservation tools to help fish stocks rebound [@lester_2008-F_; @lester_2009-Ks: @sala_2016-PV] by limitting or restricting fishing effort and gears.
-No-take marine reserves (marine reserves hereinafter) are a particular type of MPAs, where all fishing effort and extractive activities are off-limits or highly limited.
-The International Union for the Conservation of Nature [-@iucn_website_2017-Bg] categorizes them as Ia (Strict Natural Reserves) or Ib (Wilderness Area).
+No-take marine reserves (marine reserves hereinafter) are a particular type of MPAs, where all fishing effort and extractive activities are off-limits or highly regulated.
+The International Union for the Conservation of Nature [-@iucn_website_2017-Bg] categorizes them as Ia (Strict Natural Reserves) or Ib (Wilderness Areas).
 However, see [@hortaecosta_2016-_3] for a detailed regulation-based classification system.
@@
 Compared to partially protected MPAs, marine reserves are known to have even higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_].
-Often, these effects are meassured as biological changes in the area through time and often lack a control site against which to compare [@betti_2017-lq].
-This before-after comparison leaves aside other factors for which one must control [@davies_2017-ml], impeding us to talk about causation with complete certainty.
-While some studies have used control sites, their analyses do not estimate the net effect of the reserve, and often use a control-impact comparison approach [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya].
-A smaller fraction of studies use measurements of some biological atribute of a reserve and control sites before and after the implementation of the reserves, thus having a before-after-control-impact design [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], allowing them to use causal inference techniques to estimate the net effect of the reserve.
+Often, these effects are meassured as biological changes in the area through time and lack a control site against which to compare [@betti_2017-lq].
+This before-after comparison ignores other factors for which one must control [@davies_2017-ml], impeding us to talk about causation with complete certainty.
+While some studies have used control sites, their analyses do not estimate the net effect of the reserve, and often use a control-impact comparison approach [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
+A smaller fraction of studies use measurements of some biological atribute of a reserve and control sites before and after the implementation of the reserves, thus having a before-after-control-impact design [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], allowing them to use causal inference techniques that estimate the net effect of the reserve.
 
-The diversity of approaches used to evaluate the effect of protected areas highlights the need to standardize approaches that enable us to truly evaluate the effect of protection.
-Though biophysical aspects are important to reserve success, the effectiveness of reserves also depends on the socioeconomic status and governance system of the local fishing community [@basurto_2013-oq; @basurto_2012-Mz], dimensions largely ignored until recently.
-Currently, the only the IUCN framework [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biophysical, socioeconomic, and governance indicators, and insight into how these indicators may be measured.
-However, it does not expand on how to analyze them, nor provide a user-friendly tool that enable replicability ans scalability of the analysis.
+As with any other policy intervention, it is important that we can measure its effect in order to adapt and learn.
+The diversity of approaches currently used to evaluate the effect of marine reserve often does not answer the simplest question: What is the net effect of the marine reserve?
+This highlights the need to develop standardized approaches that enable us to truly evaluate the effect of the intervention.
+Furthermore, while biophysical aspects are important to reserve success, the effectiveness of reserves also depends on the socioeconomic status and governance system of the local fishing community [@basurto_2013-oq; @basurto_2012-Mz], dimensions often ignored.
+By excluding these important dimensions.
+the evaluation provides just a partial picture of the state of the reserve.
+Currently, only the IUCN framework "How to evaluate your MPA" [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biophysical, socioeconomic, and governance indicators, and insight into how these indicators may be measured, but does not provide guidelines on how to analyse them.
+A more recent work by @mascia_2017-m_ integrates these three dimensions and uses causal inference techniques to provide a measure of the effect of implemented an MPA.
+However, these two certainly novel approaches do not provide a user-friendly tool that enable replicability ans scalability of the analysis.
 
-Open science and the development of tools have become ingreasingly popular to make science scalable and reproducible [@lowndes_2017-xh].
+An increasingly popular way to mace science reproducible, scalable and replicable is through Open science and the development of tools [@lowndes_2017-xh].
 The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, has succesfully standardized a way to measure the health and benefits of the oceans.
-This has been succesfuly implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC].
-Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance indicators [@anderson_2015-ND], design territorial use rights for fisheries [TURFs; @oyanedel_2017-TO], or to improve decision making in the hydropower industry [@vilela_2017-Zo], among others.
+This has been implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC].
+Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance indicators [@anderson_2015-ND], design territorial use rights for fisheries [TURFs; @oyanedel_2017-TO], or to improve decision making in the hydropower industry [@vilela_2017-Zo], just to list a few.
 
@@
 We provide a list of commonly stated management objectives, and match them to appropriate indicators that meassure the effectiveness of the management intervention.
-We include a simple approach to analyze these indicators, building on causal inference techniques [@moland_2013-VP], largely needed to truly understand the effect of management interventions [@burgess_2016-HN].
-At the same time, we introduce the Marine Reserve Evaluation App (MAREA), an open source web-based tool that automates the framework described in this paper.
-Case studies from three regions of Mexico are presented as proof of concept of the app.
+We include a simple approach to analyze these indicators, building on causal inference techniques [@moland_2013-VP], largely needed to truly understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_].
+We also introduce the Marine Reserve Evaluation App (MAREA), an open source web-based tool that automates the framework described in this paper.
+To demonstrate the potential of MAREA, we perform an evaluation of marine reserve established by fishers from Isla Natividad, Mexico in 2006.
 
@@
 1**).
-We explain how management objectives were identified and matched to appropriate indicators that allow the evaluation of the reserves, and provide guidelines on data collection and formating.
+We explain how management objectives were identified and matched to appropriate indicators that allow the evaluation of the reserves, and provide brief guidelines on data collection.
 Alongside, methodologies to analyze these indicators are presented.
-Then, we describe the development of MAREA and provide guidance on how to use this user-friendly open access tool.
+Then, we describe the development of MAREA and eplain how this user-friendly open access tool can be used by anyone.
 Finally, we provide guidelines on how to interpret the results and output generated by MAREA.
@@
 
-Throughout this work, we will refer to the stated goals for which a reserve has been established as objectives.
-Given that this work was motivated by the need to provide a framework to evaluate mexican marine reserves, we focused on identifying objectives of Mexican marine reserves.
+Throughout this work, we will refer to the stated goals for which a reserve was designed as objectives.
+Given that this work was motivated by the need to provide a framework to evaluate Mexican marine reserves, or focus was on identifying objectives of marine reserves in Mexico.
 However, we group these objectives into seven major categories, which can be applied to other conservation areas in the world.
-The list of objectives was gathered through a literature review from the reserves’ stated objectives in official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, decrees, and specific legislation [@nom049sagpesc_2014-V6].
-Even though each reserve -or set of reserves- has its own goals, seven main categories of objectives were identified:
+The list of objectives was gathered through a literature review from the reserves’ stated objectives in official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, decrees, and specific legislation [@nom049sagpesc_2014-V6] associated to these areas.
+Even though each reserve has its own goals, seven main categories of objectives were identified:
 
@@
 
-Indicators are divided into three main types: biophysical, socioeconomic, and governance.
+Indicators are divided into three main categories: biophysical, socioeconomic, and governance.
 Biophysical indicators (n = 7) focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data section for specific sampling design and methodologies).
-Socioeconomic indicators (n = 3) reflect the performance of the fishery in terms of catches, income from catches and availability of alternative livelihoods.
+Socioeconomic indicators (n = 3) reflect the performance of the fishery in terms of catches, income from catches, and availability of alternative livelihoods.
 Governance indicators (n = 15) describe the governance structures under which the community operates.
@@
 Our list includes indicators that respond to the implementation of the reserve (*i.e.* outcome variables) or that might further the understanding of its performance.
-In that sense, most biophysical and socioeconomic indicators are outcome variables, while governance indicators might explain poor reserve performance, for example.
-Whenever an indicator is said to also be applied to “Objective Species”, it means that the indicator will be used for all species (*e.g.* Fish Biomass) and individual species that are either the conservation target of the reserve, or are of particular economic or ecological interest (*e.g.* Grouper Biomass).
-**Table 1** presents the proposed indicators.
-**Table 2** shows how objectives are matched with indicators.
+In that sense, most biophysical and socioeconomic indicators are outcome variables.
+On the other hand, governance indicators are viewed as explanatory variables that might explain reserve performance.
+Whenever an indicator is said to also be applied to “Objective species”, it means that the indicator can used for all species (*e.g.* Fish Biomass) and individual species that are either the conservation target of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass).
+**Table 1** presents the proposed indicators, and **Table 2** shows how objectives are matched with indicators.
 
@@
 Biophysical data is collected via underwater ecological surveys as part of a reserve's monitoring program, often carried out by local fishers with guidance of non-government organizations, such as Comunidad y Biodiversidad or Niparajá.
-Local fishers,trained in scientific diving, perform SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length.
+Local fishers -trained in scientific diving- perform SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length.
 This information is recorded along 30 m transects, with a sampling window of 2 m X 2 m.
 Between 10 and 24 transects are performed in each site at depths between 1 and 30 m.
-Ecological surveys are performed yearly in each reserve and its corresponding control site(s), before and after the implementation of the reserve.
+Ecological surveys are performed yearly in each reserve and its corresponding control site(s), before and after the implementation of the reserve, allowing us to have a before-after-control-impact sampling design.
 Control sites are areas where habitat is similar to that of the reserve, but with presence of fishing activity.
@@
 The hypothesis that the indicators will respond the the implementation of the reserve is tested by analysing spatial and temporal changes in each numeric biophysical indicator (all but B5) using generalized linear models [@moland_2013-VP].
-To account for variations in the environment and survey conditions, covariates that describe the environment and gathered during the underwater ecological surveys are included into a model with form:
+To account for variations in the environment and survey conditions, covariates that are gathered during the underwater ecological surveys are included into a model with form:
 
@@
 In this model, $i$, $t$, and $z$ are subindices for transect, time, and zone (control or reserve site), respectively.
-This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), tratment ($Z$), an interaction between a variable that indicates pre- or post-implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$), visibility during the survey ($V$), and depth at which survey was performed ($D$).
+This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), a dummy variable that indicates tratment ($Z$; *i.e.* control o reserve), an interaction between a dummy variable that indicates pre- or post-implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$; in °C), horizontal visibility during the survey ($V$; in m), and depth at which survey was performed ($D$; in m).
+$\epsilon$ represents the error term associated to the equation.
 Here, years are modeled as factors, using the first year as the reference level.
-The treatment and post variables are modeled by dummy variables coded as Control = 0 and Reserve = 1; and Pre-implementation = 0 and Post-implementation = 1, respectively.
+By modelling years as factors, we avoid overimposing a linear structure in the way an indicator changes through time (*i.e.* the change in biomass between 2000 and 2001 does not have to be the same as the change between 2006 an 2007).
+The treatment and post variables, modeled as dummy variables, are coded as Control = 0 and Reserve = 1; and Pre-implementation = 0 and Post-implementation = 1, respectively.
 
-Socioeconomic data is collected by the fishers or fishery management agencies, which record landings, income, and sometimes prices for each species.
+Socioeconomic data is often collected by fishers or fishery management agencies, which record landings, income, and sometimes prices for each species.
 In order to control for inflation and changes in buying power, income is adjusted with the countries consumer price index:
@@
 
-Where $I_t$ represents the adjusted income for year $t$ as the product between the reported income for year $t$ and the ratio between the consumer price index ($CPI$) in that year to the present year's ($T$) CPI.
+Where $I_t$ represents the adjusted income for year $t$ as the product between the reported income for that year and the ratio between the consumer price index ($CPI$) in that year to the most recent year's ($T$) CPI.
 Since no control sites are available, numeric socioeconomic indicators (G1 and G2) are evaluated with a simplified version of *eq.
@@
 While this model does not allow assertion of a causal relation, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input.
+For both models (**eq.
+1** and **eq.
+3**), coefficients are estimated via ordinary least squares, and heteroskedastik-robust standard errors are calculated.
 
-Governance data, however, is not readily available nor systematically gathered by the community or any other organizations.
+Governance data, however, is not readily available nor systematically collected by the community or any other organizations.
 In this case, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1-G15).
@@
 
-MAREA was developed in R Studio [@rcore_2017] using the shiny framework [@shiny_2017].
-This provides the tools to build interactive web applications, hosted on an open server.
+MAREA was developed in R Studio [@rcore_2017] using the shiny framework [@shiny_2017], which provides the tools to build interactive web applications, hosted on an open server.
 MAREA can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/).
-While the original version was developed in spanish (aimed mainly at Latinamerican countries), all its content can be translated by a translation tool availabe within the environment.
+While the original version was developed in spanish (aimed at Mexico and other Latinamerican countries), all its content can be translated by a translation widget availabe within the environment.
 
-MAREA is designed into a 6-step process.
-The first part provides an introduction to the app, and provides the user with an overview of the process.
-Then, the user must select management objectives, and MAREA will automatically select appropriate indicators, based on **Table 2**.
-Then, the user can load the data, using standard \*.csv formatting; sample datasets are provided in the environment.
+MAREA is designed into a 6-step process, divided in tabs.
+The first tab provides an introduction to the app and an overview of the evaluation process.
+Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on **Table 2**.
+The user can then load the data, using standard \*.csv files; sample datasets are provided within MAREA.
 Once data have been loaded, MAREA identifies all reserves in the data, and lets the user select the reserve to be evaluated.
-At this point, the user can also specify the year of implementation of the reserve, as well as select objective species that are of particular management interest.
-Before presenting the results MAREA provides the user with a section to confirm that all the decisions taken before are correct.
-Finally, the user is taken to the results, which are represented in two ways.
+At this point, the user can also specify the year of implementation of the reserve, and indicate objective species that are of particular management interest.
+Before presenting the results, MAREA provides the user with a section to confirm that all the decisions taken until then are correct.
+Finally, the user is taken to the results tab, where results can be viewed at a glance or downloaded as a technical report.
 
 The first result is a color-coded scorecard, intended to provide a general overview of the effectiveness of the reserve to a general public.
-The scorecard provides a global score for the reserve, a general score for each category of indicators, and an individual score for each selected indicator.
-The global and category scores are determined by the percentage of positive indicators, overall and for each category.
+The scorecard provides a global score for the reserve, a general score for each category of indicators, and an individual score for each indicator.
+The global and category-level scores are determined by the percentage of positive indicators, overall and for each category.
 For numeric biophysical indicators (all but B5), the color is defined by the sign of the interaction term coefficient ($\beta_2$) in *eq.
@@
 For socioeconomic indicators, colors are assigned based on the direction of the slope ($\beta_1$).
-Red, yellow or green are used for $\beta_i<0$, $\beta_i = 0$, and $\beta_i>0$ respectively.
-The intensity of the color is defined by the significance of coefficient, testing the null hypothesis of no change (*i.e.* $H_0: \beta_i = 0$) with Student t-test.
+Red, yellow and green are used for $\beta_i<0$, $\beta_i = 0$, and $\beta_i>0$, respectively.
+The intensity of the color is defined by the significance of coefficient, testing the null hypothesis of no change (*i.e.* $H_0: \beta_i = 0$) with a Students t-test.
 Cutoff values are $p < 0.05$ and $p < 0.1$.
@@
 A legend (**Fig.
-2**) is provided in the app to aid in the interpretation of these results.
-Governance indicators, however, are evaluated binomially and can be either red or green.
-The color is defined based on what literature shows to be a positive (green) or negative (red) value for each indicator.
-However, due to the nature of some governance indicators, which require the user to provide a narrative, only indicators XXX are presented in the scorecard.
+2**) is provided within the scorecard to aid in the interpretation of these results.
+Governance indicators, however, are represented by red or green.
+The color is defined based on what literature shows to be a positive (green) or negative (red) factor for a reserve.
+However, due to the nature of some governance indicators, which require the user to provide a narrative, only some indicators are presented in the scorecard (although all are included in the techniical report).
 
@@
 
-The second output from MAREA is a technical report, intended to communicate information and statystical results in a rigurous way.
+The second output from MAREA is a technical report, intended to communicate information and statystical results in a more formal and technical way.
 This report also includes a scorecard as a summary of the results, but provides more information for each indicator.
-For all numeric indicators, the technical report includes graphs of the value of the indicator, for the reserve and control sites through time.
-It also provides a regression table [@stargazer_2015] that summarizes the value of all coefficients in the regression and their respective standard errors.
+For all numeric indicators, the report includes a graph of the value of the indicator, for the reserve and control sites, through time.
+It also provides a regression table that summarizes the value of all coefficients in the regression and their respective robust standard errors.
 The summary table also provides information on model fit ($R^2$) and significance of the regression.
@@
 
-We apply the proposed framework and tool to evaluate the effectiveness of one no-take marine reserves from Isla Natividad, in Baja California Sur, Mexico.
+We apply the proposed framework and tool to evaluate the effectiveness of one marine reserve from Isla Natividad, in Baja California Sur, Mexico.
 Isla Natividad is located 8 Km off the Pacific Coast of the Baja Peninsula (**Fig.
-3**).
-Fishers in Isla Natividad operate under a fishing cooperative (S.C.P.P.
-Buzos y Pescadores de la Baja California), which promotes co-management of marine resources [@mccay_2017-1m; @mccay_2014-CN].
-Additionaly, fishers have Territorial Use Rights for Fisheries (TURFs), that provide them with exclusive rights to exploit the marine resources for which they hold permits.
+3**), where fishers operate under a fishing cooperative (S.C.P.P.
+Buzos y Pescadores de la Baja California) that promotes co-management of marine resources [@mccay_2017-1m; @mccay_2014-CN].
+Additionaly, fishers have Territorial Use Rights for Fisheries (TURFs) that provide them with exclusive access rights to exploit the marine resources in a given perimeter.
 
@@
 
-In 2006, the community implemented two marine reserves within their TURF.
-These reserves have prooven to be effective in enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of highly valuable commercial species, like abalone [@munguavega_2015-yg].
+In 2006, the community implemented two marine reserves within their TURF [making them TURF-reserves; @afflerbach_2014-HP].
+The reserves from Isla Natividad have prooven to be effective in enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of highly valuable commercial species, like abalone [@munguavega_2015-yg].
 These ecological benefits have been translated into economical benefits, enhancing population persistance and supporting abalone fisheries [@rossetto_2015-V0].
@@
 
-The reserve was implemented to recover species of economic interest -which have been overexplotied in the last years- in order to enhance fishery production in nearby waters.
+The reserve was implemented to recover species of economic interest -which were overexplotied- in order to enhance fishery production in nearby waters.
 Fishers were also interested in preserving biological diversity and the ecosystem.
@@
 
-All invertebrates show increased densities within the reserve, as compared to the control site, providing evidence that the reserve is effectively protecting these species.
+The results identified by our framework and MAREA are consistent to those previously published [*e.g.* @micheli_2012-EU; @rossetto_2015-V0].
+Invertebrates show increased densities within the reserve, as compared to the control site, providing evidence that the reserve is effectively protecting these species.
 While we are not able to infer causality from the socioeconomic analysis, we are able to identify important trends.
-For example, snail and urchin density hve significantly increased within the reserve, but their landings and income have decreased.
+For example, snail and urchin density have significantly increased within the reserve, but their landings and income have decreased.
 The opposite is observed for lobster and cucumber, which have shown increases in densities, landing, and income.
@@
 A summary of governance indicators is provided in **Table 5**.
+
+\clearpage
 
@@
 The analysis isolates the net effect of the reserve, providing a propper measure of reserve effectiveness.
-We agree there is room for improvement in the way in which socioeconomic and governance data are analized.
-However, we believe that providing a unifying platform where all can be analyzed and comprehensively presented represents a valuable step towards effective management.
+We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analized.
+Despire this, we believe that providing a unifying platform where all can be analyzed and comprehensively presented represents a valuable step towards evidence-based effective management.
 
-The main value of MAREA is that it provides a free, simple, and replicable way to perform rigurous analysis.
+Furthermore, MAREA's value is that it provides a free, simple, and replicable way to perform rigurous analysis.
 The tool can easily be used by fishers, NGO members, and managers in government agencies, providing transparency of the analysis and results.

### 2017-08-17 Manuscript.Rmd
Aded plos csl and updated author list and affiliations

@@
 - packages.bib
-# csl: plos.csl
+csl: plos.csl
 ---
@@
 
-Juan Carlos Villaseñor-Derbez^1\*^, Caio Faro^1^, Melaina Wright^1^, Jael Martínez^1^, Alvin Suárez^2^, Stuart Fulton^2^, Sean Fitzgerald^1^, Gavin McDonald^4^, Jorge Torre-Cosio^2^, Fiorenza Micheli^3^, Christopher Costello^1,4^ \
+Juan Carlos Villaseñor-Derbez^1\*^, Caio Faro^1^, Melaina Wright^1^, Jael Martínez^1^, Sean Fitzgerald^1^, Stuart Fulton^2^, Maria del Mar Mancha-Cisneros^3^, Gavin McDonald^1,4,5^, Fiorenza Micheli^6^, Alvin Suárez^2^, Jorge Torre^2^, Christopher Costello^1,4,5^
 
@@
 
-^3^ Hopkins Marine Station, Stanford University, Monterey, California, United States
+^3^ School of Life Sciences, Arizona State University, Tempe, Arizona, United States
 
-^4^ Sustainable Fisheries Group, University of California Santa Barbara, Santa Barbara, California, United States \
+^4^ Sustainable Fisheries Group, University of California Santa Barbara, Santa Barbara, California, United States
+
+^5^ Marine Science Institute, University of California Santa Barbara, Santa Barbara, California, United States
+
+^6^ Hopkins Marine Station and Center for Ocean Solutions, Stanford University, Pacific Grove, CA 93950, USA
 

### 2017-08-17 Manuscript.Rmd
Updated abstract

@@
 As with any other management interventions, we need to better understand their effectiveness and impacts on coastal communities and the environment.
-To date, the evaluation of the reserves largely relies on analysing ecological data, ignoring socioeconomic and governance dimensions.
-In absence of a standardized methodology, existing data is analyzed in different ways, hindering the ability to compare results across case studies.
-We present a framework and tool to evaluate the effectiveness of marine reserves by matching 7 commonly stated management objectives to a set of 9 biophysical, 5 socioeconomic, and 14 governance indicators.
-We provide guidelines on how to properly colet data that can then be analysied with standardized methods Biophysical indicators are evaluated with causal inference techniques, using a counterfactual approach, to measure the net effect of the reserve on each indicator.
-Linear regression models are fitted to socioeconomic indicators through time, testing for the differences before and after the implementation of the reserve.
-Governance indicators are quantitatively analyzed based on literature, identifying common governance structures and their associated effectiveness.
-To make the framework accessible to fishers and decision makers, and allow replication of results, we developed an open source, web-based Marine Reserve Evaluation App (MAREA).
-Together, this framework and MAREA represent a major step in marine reserve evaluation that can furhter our understanding and simplify our management of marine reserves.
+To date, the evaluation of these reserves largely relies on analyzing ecological data, often ignoring socioeconomic and governance dimensions.
+Existing data are analyzed in different ways, hindering the ability to compare results across case studies.
+Moreover, analysis and evaluation of reserves is generally conducted by academic scientists, not the reserves managers and users, thereby hindering effective local management and rapid response to change.
+We present a framework and tool to evaluate the effectiveness of marine reserves by matching seven commonly stated management objectives to nine biological, five socioeconomic, and 14 governance indicators.
+We provide guidelines on how to properly collect data that can then be analyzed with standardized method.
+Biological indicators are evaluated with causal inference techniques, using a counterfactual approach, to assess the net effect of the reserve on each indicator.
+Linear regression models are fitted to socioeconomic indicators through time to test for differences before and after reserve implementation.
+Governance indicators are qualitatively analyzed using a framework developed through a literature review which identifies common governance structures and their associated effectiveness.
+To make the framework accessible to fishers and decision makers, and allow replication of results, we developed the open source, web-based Marine Reserve Evaluation App (MAREA).
+Together, this new framework and MAREA can further our understanding and support management of marine reserves.
 

### 2017-08-17 Manuscript.Rmd
Updated introduction and first part of methods

@@
 
-Overfishing and unsustainable fishing practices are some of the major threats to the conservation of marine ecosystems around the world [@pauly_2005-qV; @halpern_2008-dK].
-Marine Protected Areas (MPAs) are frequently proposed as fishery management and conservation tools to help fish stocks rebound [@lester_2008-F_; @lester_2009-Ks: @sala_2016-PV] by limitting or restricting fishing effort and gears.
-No-take marine reserves (marine reserves hereinafter) are a particular type of MPAs, where all fishing effort and extractive activities are off-limits or highly regulated.
-The International Union for the Conservation of Nature [-@iucn_website_2017-Bg] categorizes them as Ia (Strict Natural Reserves) or Ib (Wilderness Areas).
-However, see [@hortaecosta_2016-_3] for a detailed regulation-based classification system.
+Overfishing and unsustainable fishing practices are two of the largest threats to the conservation of marine ecosystems around the world [@pauly_2005-qV; @halpern_2008-dK].
+Marine Protected Areas (MPAs) are frequently proposed as fishery management and conservation tools to help fish and invertebrate stocks recover [@lester_2008-F_; @lester_2009-Ks: @sala_2016-PV] by limiting or restricting fishing effort and gears.
+No-take marine reserves (marine reserves hereinafter) are a particular type of MPA, where all fishing effort and extractive activities are off-limits [-@iucn_website_2017-Bg; @hortaecosta_2016-_3].
 
-MPAs have proven to increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience of the bounded region [@micheli_2012-EU], and preserve genetic diversity [@munguavega_2015-yg].
-Compared to partially protected MPAs, marine reserves are known to have even higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_].
-Often, these effects are meassured as biological changes in the area through time and lack a control site against which to compare [@betti_2017-lq].
-This before-after comparison ignores other factors for which one must control [@davies_2017-ml], impeding us to talk about causation with complete certainty.
-While some studies have used control sites, their analyses do not estimate the net effect of the reserve, and often use a control-impact comparison approach [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
-A smaller fraction of studies use measurements of some biological atribute of a reserve and control sites before and after the implementation of the reserves, thus having a before-after-control-impact design [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], allowing them to use causal inference techniques that estimate the net effect of the reserve.
+Empirical studies have shown that MPAs increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience to climatic impacts [@micheli_2012-EU; @roberts_2017-J9], and preserve genetic diversity [@munguavega_2015-yg].
+Compared to partially protected MPAs, marine reserves are known to have even higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO].
+These effects are often measured as biological changes in the area through time and lack a control site against which to compare [@betti_2017-lq].
+This before-after comparison cannot account for other factors for which one must control [@davies_2017-ml] in order to attribute biological change to protection in the reserves.
+While some studies have used control sites, these analyses do not estimate the net effect of the reserve, and often use a control-impact comparison approach that does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
+A smaller fraction of studies have used a before-after-control-impact design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.
 
-As with any other policy intervention, it is important that we can measure its effect in order to adapt and learn.
-The diversity of approaches currently used to evaluate the effect of marine reserve often does not answer the simplest question: What is the net effect of the marine reserve?
-This highlights the need to develop standardized approaches that enable us to truly evaluate the effect of the intervention.
-Furthermore, while biophysical aspects are important to reserve success, the effectiveness of reserves also depends on the socioeconomic status and governance system of the local fishing community [@basurto_2013-oq; @basurto_2012-Mz], dimensions often ignored.
-By excluding these important dimensions.
-the evaluation provides just a partial picture of the state of the reserve.
-Currently, only the IUCN framework "How to evaluate your MPA" [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biophysical, socioeconomic, and governance indicators, and insight into how these indicators may be measured, but does not provide guidelines on how to analyse them.
-A more recent work by @mascia_2017-m_ integrates these three dimensions and uses causal inference techniques to provide a measure of the effect of implemented an MPA.
-However, these two certainly novel approaches do not provide a user-friendly tool that enable replicability ans scalability of the analysis.
+As with any other policy intervention, it is important that we can measure its effect in order to adapt and learn [@ferraro_2006-oW; @burgess_2016-HN].
+The diversity of approaches currently used to evaluate the effect of marine reserves often does not answer the simple question: *What is the effect of a marine reserve on a given attribute?* This gap highlights the need to develop standardized approaches that enable us to evaluate the net effect of the intervention (*i.e.* causes of conservation outcomes; @ferraro_2006-oW).
+Furthermore, while biological aspects are important to reserve success, effectiveness also depends on the socioeconomic status and governance system of the local fishing communities [@basurto_2013-oq; @basurto_2012-Mz], which are often ignored.
+By excluding these important dimensions, the evaluation provides only a partial picture of the impacts of the reserve.
+Currently, only the IUCN framework "How to evaluate your MPA" [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biological, socioeconomic, and governance indicators, and insights into how these indicators may be measured, but does not provide guidelines on how to analyze them.
+Recent work by @mascia_2017-m_ integrates these three dimensions and suggests the use of causal inference techniques to provide a measure of the effect of implementing an MPA.
+However, these two novel approaches do not provide a user-friendly tool that enables replicability and scalability of the analysis, particularly when used by the fishers and decision makers themselves.
 
-An increasingly popular way to mace science reproducible, scalable and replicable is through Open science and the development of tools [@lowndes_2017-xh].
-The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, has succesfully standardized a way to measure the health and benefits of the oceans.
-This has been implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC].
-Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance indicators [@anderson_2015-ND], design territorial use rights for fisheries [TURFs; @oyanedel_2017-TO], or to improve decision making in the hydropower industry [@vilela_2017-Zo], just to list a few.
+An increasingly popular way to make science reproducible, scalable and replicable is through Open Science and the development of open-access tools [@lowndes_2017-xh].
+The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, has successfully standardized a way to measure the health and benefits of the oceans.
+This approach has been implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC].
+Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance [@anderson_2015-ND; @dowling_2016-pO], design territorial use rights for fisheries (TURFs; @oyanedel_2017-TO), and improve decision making in the hydropower industry [@vilela_2017-Zo], just to list a few.
 
-The lack of a comprehensive framework -or the complexity of existing ones, which alienate non-experts- and user-friendly tools to evaluate the effectiveness of marine reserves calls for the development of a new framework and tool.
-Here, we present a framework to evaluate no-take marine reserves, which incorporates the biological, socioeconomic, and governance dimensions of these conservation areas.
-We provide a list of commonly stated management objectives, and match them to appropriate indicators that meassure the effectiveness of the management intervention.
-We include a simple approach to analyze these indicators, building on causal inference techniques [@moland_2013-VP], largely needed to truly understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_].
-We also introduce the Marine Reserve Evaluation App (MAREA), an open source web-based tool that automates the framework described in this paper.
-To demonstrate the potential of MAREA, we perform an evaluation of marine reserve established by fishers from Isla Natividad, Mexico in 2006.
+The lack of a comprehensive framework and user-friendly tools to evaluate the effectiveness of marine reserves ---or the complexity of existing ones, which alienate non--experts--- calls for the development of a new framework and tool.
+The current work presents a framework to evaluate marine reserves, which incorporates the biological, socioeconomic, and governance dimensions of these areas.
+We first provide a list of commonly stated management objectives and match them to appropriate indicators that measure the effectiveness of the management intervention.
+We then include a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_].
+We also introduce the Marine Reserve Evaluation App (MAREA), an open source web--based tool that automates the framework described in this paper and enables its broader use.
+Finally, we present a case study on the evaluation of a marine reserve established by the fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.
 
@@
 
-Here, we describe the proposed framework to evaluate the effectiveness of marine reserves (**Fig.
-1**).
+Here, we describe the proposed framework to evaluate the effectiveness of marine reserves (Fig.
+1).
 We explain how management objectives were identified and matched to appropriate indicators that allow the evaluation of the reserves, and provide brief guidelines on data collection.
 Alongside, methodologies to analyze these indicators are presented.
-Then, we describe the development of MAREA and eplain how this user-friendly open access tool can be used by anyone.
-Finally, we provide guidelines on how to interpret the results and output generated by MAREA.
+Then, we describe the development of MAREA and explain how this user-friendly open access tool can be used by anyone.
+Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA.
 
-![Workflow to evaluate the effectiveness of no-take marine reserves.](./Figures/flowchart.gif){ width=50% }
+![Workflow to evaluate the effectiveness of marine reserves.](./Figures/flowchart.gif){ width=50% }
 
-## Reserve objectives and indicators
+## Marine Reserve objectives and indicators
 
-Throughout this work, we will refer to the stated goals for which a reserve was designed as objectives.
-Given that this work was motivated by the need to provide a framework to evaluate Mexican marine reserves, or focus was on identifying objectives of marine reserves in Mexico.
-However, we group these objectives into seven major categories, which can be applied to other conservation areas in the world.
-The list of objectives was gathered through a literature review from the reserves’ stated objectives in official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, decrees, and specific legislation [@nom049sagpesc_2014-V6] associated to these areas.
+Throughout this study, we will refer to the stated goals for which a marine reserve was designed as objectives.
+This work was motivated by the need to provide a framework to evaluate Mexican marine reserves.
+Thus, our focus was on identifying common objectives of marine reserves in Mexico.
+However, we group these objectives into seven major categories, which can be applied to marine reserves worldwide.
+The list of objectives was developed through a literature review, which compiled stated objectives in and legislation [@nom049sagpesc_2014-V6; @lgeepa_2017-jL] and official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, and decrees associated to these areas.
 Even though each reserve has its own goals, seven main categories of objectives were identified:
@@
 
-Based on these seven objectives, we determined a set of indicators to evaluate the effectiveness.
-The list of indicators was built through a review of scientific literature, where we identified indicators that were used to measure similar objectives.
-A first filter eliminated indicators for which baseline data does not exist, or for which data aquisition was expensive or outside of the general scope.
-The initial list of indicators was reviewed at a workshop with participation of members from Mexican fishery management agencies and non-government organizations.
+Based on these seven objectives, we determined a set of associated indicators to evaluate reserve effectiveness.
+The list of indicators was compiled through a review of scientific literature in which we identified indicators that were used to measure similar objectives.
+A first filter eliminated indicators for which baseline data do not typically exist in Mexico.
+The preliminary list of indicators was reviewed at a workshop with participation of members from Mexican fishery management agencies and non-government organizations.
 Later, these were also presented to fishers from the Ensenada Fishing Cooperative (*S.C.P.P.
-Ensenada*), which provided input.
-Our final list of indicators includes, at least, those identified in review works such as @lester_2009-Ks or @woodcock_2017-Wm.
+Ensenada*), in Baja California, who provided input.
+Our final list of indicators includes those identified in review works such as @lester_2009-Ks or @woodcock_2017-Wm.
 
-Indicators are divided into three main categories: biophysical, socioeconomic, and governance.
-Biophysical indicators (n = 7) focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data section for specific sampling design and methodologies).
+Indicators are divided into three main categories: biological, socioeconomic, and governance (Table 1).
+Biological indicators (n = 7) focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data and Analysis section for specific sampling design and methodologies).
 Socioeconomic indicators (n = 3) reflect the performance of the fishery in terms of catches, income from catches, and availability of alternative livelihoods.
-Governance indicators (n = 15) describe the governance structures under which the community operates.
-These indicators may be numeric (*e.g.* Fish biomass) or descriptive (*e.g.* Reasoning for reserve location).
+Governance indicators (n = 15) describe the governance structures under which the community operates (*e.g.*, access rights to the fishery, number of fishers, legal recognition of the reserve).
+Indicators may be numeric (e.g.
+Fish biomass) or descriptive (*e.g.* Reasoning for reserve location).
 Our list includes indicators that respond to the implementation of the reserve (*i.e.* outcome variables) or that might further the understanding of its performance.
-In that sense, most biophysical and socioeconomic indicators are outcome variables.
-On the other hand, governance indicators are viewed as explanatory variables that might explain reserve performance.
-Whenever an indicator is said to also be applied to “Objective species”, it means that the indicator can used for all species (*e.g.* Fish Biomass) and individual species that are either the conservation target of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass).
-**Table 1** presents the proposed indicators, and **Table 2** shows how objectives are matched with indicators.
+In that sense, most biological and socioeconomic indicators are outcome variables.
+On the other hand, governance indicators are viewed as possible explanatory variables of reserve performance.
+Whenever an indicator is applied to “Target species”, it means that the indicator can be used for all species (*e.g.* Fish Biomass) and/or for individual species that are either the conservation target of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass).
+Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with indicators.
 

### 2017-08-17 Manuscript.Rmd
Updated methods

@@
 
-Biophysical data is collected via underwater ecological surveys as part of a reserve's monitoring program, often carried out by local fishers with guidance of non-government organizations, such as Comunidad y Biodiversidad or Niparajá.
-Local fishers -trained in scientific diving- perform SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length.
-This information is recorded along 30 m transects, with a sampling window of 2 m X 2 m.
-Between 10 and 24 transects are performed in each site at depths between 1 and 30 m.
-Ecological surveys are performed yearly in each reserve and its corresponding control site(s), before and after the implementation of the reserve, allowing us to have a before-after-control-impact sampling design.
+In many coastal marine reserves of Mexico, biological data are collected via underwater ecological surveys as part of a reserve's monitoring program, often carried out by local fishers with guidance from Civil Society Organizations (CSOs).
+Scientific divers record fish and invertebrate richness and abundances, as well as fish total length along belt transects.
+Ecological surveys are performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, allowing us to have a before-after-control-impact (*i.e.* BACI) sampling design.
 Control sites are areas where habitat is similar to that of the reserve, but with presence of fishing activity.
+While transect dimensions (i.e.
+length and width) and sampling methods might vary from study to study, the general idea remains the same: richness, abundances, and sizes of organisms are recorded in a study-specific standardizes way.
+For this reason, MAREA does not assume specific transect dimensions, and pertinent indicators are calculated per transect (Table 1).
 
-Having control sites for biophysical data allows us to use counterfactual approaches to evaluate the net effect of the reserve.
-The hypothesis that the indicators will respond the the implementation of the reserve is tested by analysing spatial and temporal changes in each numeric biophysical indicator (all but B5) using generalized linear models [@moland_2013-VP].
+Having control sites for biological data allows us to use causal inference techniques [@moland_2013-VP; @ferraro_2006-oW] to evaluate the net effect of the reserve.
+The hypothesis that the indicators will respond to implementation of the reserve is tested by analyzing spatial and temporal changes in each numeric biological indicator (all but B5) using generalized linear models [@moland_2013-VP].
 To account for variations in the environment and survey conditions, covariates that are gathered during the underwater ecological surveys are included into a model with form:
@@
 In this model, $i$, $t$, and $z$ are subindices for transect, time, and zone (control or reserve site), respectively.
-This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), a dummy variable that indicates tratment ($Z$; *i.e.* control o reserve), an interaction between a dummy variable that indicates pre- or post-implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$; in °C), horizontal visibility during the survey ($V$; in m), and depth at which survey was performed ($D$; in m).
+This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), a dummy variable that indicates treatment ($Z$; *i.e.* control or reserve), an interaction between a dummy variable that indicates pre-- or post--implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$; in °C), horizontal visibility during the survey ($V$; in m), and depth at which survey was performed ($D$; in m).
 $\epsilon$ represents the error term associated to the equation.
 Here, years are modeled as factors, using the first year as the reference level.
-By modelling years as factors, we avoid overimposing a linear structure in the way an indicator changes through time (*i.e.* the change in biomass between 2000 and 2001 does not have to be the same as the change between 2006 an 2007).
-The treatment and post variables, modeled as dummy variables, are coded as Control = 0 and Reserve = 1; and Pre-implementation = 0 and Post-implementation = 1, respectively.
+By modelling years as factors, we avoid imposing a linear structure in the way an indicator changes through time (*i.e.* the change in biomass between 2006 and 2007 does not have to be the same as the change between 2015 and 2016).
+The treatment and implementation variables, modeled as dummy variables, are coded as Control = 0 and Reserve = 1; and Pre-implementation = 0 and Post-implementation = 1, respectively.
 
-Socioeconomic data is often collected by fishers or fishery management agencies, which record landings, income, and sometimes prices for each species.
-In order to control for inflation and changes in buying power, income is adjusted with the countries consumer price index:
+Socioeconomic data are often collected by fishers, fishery management agencies or OSCs by recording landings, income, and sometimes prices for each species.
+To control for inflation and changes in buying power, income is adjusted with the country’s consumer price index [@oecd_website_2017-VV]:
 
@@
 
-Where $I_t$ represents the adjusted income for year $t$ as the product between the reported income for that year and the ratio between the consumer price index ($CPI$) in that year to the most recent year's ($T$) CPI.
-Since no control sites are available, numeric socioeconomic indicators (G1 and G2) are evaluated with a simplified version of *eq.
-1*:
+Where $I_t$ represents the adjusted income for year $t$ as the product between the reported income for that year and the ratio between the consumer price index ($CPI$) in that year to the most recent year’s ($T$) CPI.
+Since no control sites are typically available for this data type, numeric socioeconomic indicators (G1 and G2) are evaluated with a simplified version of eq.
+1:
 
@@
 
-While this model does not allow assertion of a causal relation, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input.
-For both models (**eq.
-1** and **eq.
-3**), coefficients are estimated via ordinary least squares, and heteroskedastik-robust standard errors are calculated.
+While this model does not allow establishing a causal relation, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input.
+For both models (eq.
+1 and eq.
+3), coefficients are estimated via ordinary least squares, and heteroskedastic-robust standard errors are calculated.
 
-Governance data, however, is not readily available nor systematically collected by the community or any other organizations.
-In this case, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1-G15).
-The survey is included as supplementary material (**Appendix 1**).
-To analyze governance information, we performed a literature review of common governance structures and ther relation to effectiveness in managing fisheries or marine reserves.
-These approach has prooven to succesfuly allow the evaluation of governance structures [@espinosaromero_2014-PY].
-Governance information is not analyzed in a quantitative way, but is presented along with the biophysical and socioeconomic indicators to provide managers and users with a full description of the reserve.
+Governance data are not readily available nor systematically collected by the community or other organizations.
+Therefore, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1-G15).
+The survey is included as supplementary material (Appendix 1).
+To analyze governance information, we develop a framework based on a literature review of common governance structures and their relation to effectiveness in managing fisheries or marine reserves (Appendix 2).
+This approach has been proven to successfully evaluate governance structures [@espinosaromero_2014-PY].
+Governance information is not quantitatively analyzed, but it is presented along with the biological and socioeconomic indicators to provide managers and users with a more complete description of the reserve.
+
+**ME QUEDÉ AQUí**
 

### 2017-08-17 Manuscript.Rmd
Updated legend of socrecard

@@
 
-![Scorecard for the "La Plana / Las Cuevas" marine reserve in Isla Natividad, Mexico. This scorecard was produced by MAREA.)](./Figures/scorecard.png){ width=50% }
+![Scorecard produced by MAREA for the "La Plana / Las Cuevas" marine reserve in Isla Natividad, Mexico.](./Figures/scorecard.png){ width=50% }
 

### 2017-08-18 Manuscript.Rmd
Also us maptools now

@@
   library(stargazer)
+  library(maptools)
   library(tmap)

### 2017-08-18 Manuscript.Rmd
Modifique seccion del case study

@@
 
-**ME QUEDÉ AQUí**
-
 ## Marine Reserve Evaluation App (MAREA)
 
-MAREA was developed in R Studio [@rcore_2017] using the shiny framework [@shiny_2017], which provides the tools to build interactive web applications, hosted on an open server.
+MAREA was developed in R Studio [@rcore_2017] using the Shiny package [@shiny_2017], which provides the tools to build interactive web applications, hosted on an open server.
 MAREA can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/).
-While the original version was developed in spanish (aimed at Mexico and other Latinamerican countries), all its content can be translated by a translation widget availabe within the environment.
+While the original version was developed in Spanish because it was aimed for Mexico and other Latin-American countries, all of its content can be translated by a translation widget available within the app.
 
-MAREA is designed into a 6-step process, divided in tabs.
-The first tab provides an introduction to the app and an overview of the evaluation process.
-Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on **Table 2**.
-The user can then load the data, using standard \*.csv files; sample datasets are provided within MAREA.
-Once data have been loaded, MAREA identifies all reserves in the data, and lets the user select the reserve to be evaluated.
-At this point, the user can also specify the year of implementation of the reserve, and indicate objective species that are of particular management interest.
-Before presenting the results, MAREA provides the user with a section to confirm that all the decisions taken until then are correct.
-Finally, the user is taken to the results tab, where results can be viewed at a glance or downloaded as a technical report.
+MAREA is designed as a 6-step process, divided in tabs.
+The first tab introduces the app and summarizes the evaluation process.
+Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on Table 2.
+Users can also manually select additional indicators or unselect the default ones, based on their interests and data availability.
+The user can then load the data, using standard \*.csv text files; sample datasets are provided within MAREA.
+Once data have been loaded, MAREA identifies all reserves in the data (the uploaded dataset can contain data for more than one reserve), and lets the user select the reserve to be evaluated.
+At this point, the user can also specify the year of implementation of the reserve, reserve dimensions, and indicate target species that are of particular management interest.
+Before presenting the results, MAREA provides the user with a section to confirm that all the decisions made leading up to that point are correct.
+Finally, the user is taken to the results tab where they can be viewed in a simple format.
+The user can also download a more comprehensive technical report produced in \*.pdf format.
 
-The first result is a color-coded scorecard, intended to provide a general overview of the effectiveness of the reserve to a general public.
+The first output is a color-coded scorecard intended to provide a general overview of the effectiveness of the reserve.
 The scorecard provides a global score for the reserve, a general score for each category of indicators, and an individual score for each indicator.
-The global and category-level scores are determined by the percentage of positive indicators, overall and for each category.
-For numeric biophysical indicators (all but B5), the color is defined by the sign of the interaction term coefficient ($\beta_2$) in *eq.
-1*.
+The global and category-level scores are determined by the percentage of positive indicators, overall and for each category, respectively.
+For numeric biological indicators (all but B5), the color is defined by the sign of the interaction term coefficient ($\beta_2$) in eq.
+1.
 For socioeconomic indicators, colors are assigned based on the direction of the slope ($\beta_1$).
@@
 Cutoff values are $p < 0.05$ and $p < 0.1$.
-Thus, even in a case where $\beta_i > 0$ if the coefficient is not significant (*i.e.
-* $p>0.1$) the indicator will be assigned a yellow color.
-A legend (**Fig.
-2**) is provided within the scorecard to aid in the interpretation of these results.
-Governance indicators, however, are represented by red or green.
-The color is defined based on what literature shows to be a positive (green) or negative (red) factor for a reserve.
-However, due to the nature of some governance indicators, which require the user to provide a narrative, only some indicators are presented in the scorecard (although all are included in the techniical report).
+Thus, even in a case where $\beta_i > 0$, if the coefficient is not significant (*i.e.
+* $p>0.1$), the indicator will be assigned a yellow color.
+A legend (Fig.
+2) is provided within the scorecard to aid in the interpretation of these results.
+Governance indicators are represented simply by red or green.
+The color is defined based on what literature shows to be a negative (red) or positive (green) factor for a reserve.
+For example, if the perceived degree of illegal fishing is high, this is indicator will be assigned a red color.
+However, due to the nature of some governance indicators, which require the user to provide a narrative, only some indicators are presented in the scorecard (although all are included in the technical report).
 
@@
 
-The second output from MAREA is a technical report, intended to communicate information and statystical results in a more formal and technical way.
+The second output from MAREA is a technical report intended to communicate information and statistical results in a more comprehensive and technical way.
 This report also includes a scorecard as a summary of the results, but provides more information for each indicator.
-For all numeric indicators, the report includes a graph of the value of the indicator, for the reserve and control sites, through time.
+For all numeric indicators, the report includes a graph of the value of the indicator, for both the reserve and control sites, through time.
 It also provides a regression table that summarizes the value of all coefficients in the regression and their respective robust standard errors.
@@
 Another feature of MAREA is that the user can choose to share the data.
-Once the technical report is downloaded, information on the reserve, its management objectives, and all uploaded data is saved into a central repository.
-These data can be accessed at any time by any person interested in aquiring them.
+Once the technical report is downloaded, the information on the reserve, its management objectives, and all uploaded data is saved into a central repository.
+These data can be accessed at any time by any person interested in acquiring them at [github.com/turfeffect/MAREAdata](github.com/turfeffect/MAREAdata).
 

### 2017-08-18 Manuscript.Rmd
Modify case study data

@@
 
-We apply the proposed framework and tool to evaluate the effectiveness of one marine reserve from Isla Natividad, in Baja California Sur, Mexico.
-Isla Natividad is located 8 Km off the Pacific Coast of the Baja Peninsula (**Fig.
-3**), where fishers operate under a fishing cooperative (S.C.P.P.
+We apply this analytical framework and open access tool to evaluate the effectiveness of one marine reserve from Isla Natividad, in Baja California Sur, Mexico.
+Isla Natividad is located 8 Km off the Pacific Coast of the Baja Peninsula (Fig.
+3), where fishers operate under a fishing cooperative (S.C.P.P.
 Buzos y Pescadores de la Baja California) that promotes co-management of marine resources [@mccay_2017-1m; @mccay_2014-CN].
-Additionaly, fishers have Territorial Use Rights for Fisheries (TURFs) that provide them with exclusive access rights to exploit the marine resources in a given perimeter.
+Additionally, fishers have Territorial Use Rights for Fisheries (TURFs) that provide them with exclusive access rights to exploit the marine resources within a given perimeter [@mccay_2014-CN].
 
 ```{r map, fig.cap = "General location of Isla Natividad (left) and map of the island (right).
-The marine reserve polygon is indicated in gray, and the approximate location of the control site is shown with no fill."}
+The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce)."}
 proj <- CRS("+proj=longlat +datum=NAD27")
@@
 stat <- baja[baja$NOM_ENT %in% c("Baja California Sur", "Baja California", "Sonora"),]
+stat$COUNTRY <- "Mexico"
+stat <- unionSpatialPolygons(stat, stat$COUNTRY)
 

### 2017-08-18 Manuscript.Rmd
Modified map and data sources

@@
 
-# Transects
-transects <- data.frame(X = -115.187, Y = 27.865)
-coordinates(transects) <- c("X", "Y")
-proj4string(transects) <- proj2
-transects <- spTransform(transects, proj2)
+# Labes for first map
+labels1 <- data.frame(X = c(-111, -111, -116), Y = c(32, 30, 25), Label = c("USA", "Mexico", "Pacific Ocean"))
+coordinates(labels1) <- c("X", "Y")
+proj4string(labels1) <- proj2
+labels1 <- spTransform(labels1, proj2)
+
+# controls
+controls <- data.frame(X = c(-115.187, -115.1464), Y = c(27.865, 27.84407), Label = c("B", "D"))
+coordinates(controls) <- c("X", "Y")
+proj4string(controls) <- proj2
+controls <- spTransform(controls, proj2)
+
+# Labels for second map
+labels2 <- data.frame(X = c(-115.17, -115.17, -115.17), Y = c(27.89, 27.8675, 27.8625), Label = c("Pacific Ocean", "Isla", "Natividad"))
+coordinates(labels2) <- c("X", "Y")
+proj4string(labels2) <- proj2
+labels2 <- spTransform(labels2, proj2)
 
 # Maps
-pen_map <- tm_shape(stat, bbox = bb(xlim = c(-118.36648, -109.41321), ylim = c(22.87195, 32.71863))) +
+data(World)
+W <- spTransform(World, proj)
+W <- W[W$name == "United States",]
+
+pen_map <- tm_shape(W, bbox = bb(xlim = c(-118.36648, -109.41321),
+                                 ylim = c(22.87195, 32.71863))) +
+  tm_polygons() +
+  tm_shape(stat) +
   tm_polygons() +
@@
   tm_symbols(shape = 0, size = 2, border.col = "black", col = "black", border.lwd = 2) +
+  tm_shape(labels1) +
+  tm_text("Label") +
   tm_scale_bar(position = c("left", "bottom"))
 
-study_area <- tm_shape(baja2, bbox = bb(xlim = c(-115.22, -115.15), ylim = c(27.85, 27.90))) +
+study_area <- tm_shape(baja2, bbox = bb(xlim = c(-115.22, -115.14), ylim = c(27.85, 27.90))) +
   tm_polygons() +
   tm_shape(reserves) +
-  tm_polygons(border.col = "black", col = "gray", border.lwd = 0) +
-  tm_shape(transects) +
-  tm_symbols(shape = 0, size = 4, border.col = "black", col = "black", border.lwd = 1) +
+  tm_polygons(border.col = "black", col = "red", border.lwd = 0) +
+  tm_shape(controls) +
+  tm_symbols(shape = 22, size = 4, border.col = "black", col = "blue", border.lwd = 1) +
+  tm_text("Label") +
+  tm_shape(labels2) +
+  tm_text("Label") +
   tm_compass(position = c("left", "bottom")) +
@@
 
-In 2006, the community implemented two marine reserves within their TURF [making them TURF-reserves; @afflerbach_2014-HP].
-The reserves from Isla Natividad have prooven to be effective in enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of highly valuable commercial species, like abalone [@munguavega_2015-yg].
-These ecological benefits have been translated into economical benefits, enhancing population persistance and supporting abalone fisheries [@rossetto_2015-V0].
-For the purpose of this evaluation, we will focus on the "La Plana / Las Cuevas" Marine reserve, located in the southern end of the Island (**Fig.
-3**) and its corresponding control site "La Dulce / Babencho".
+In 2006, the community implemented two community-based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh].
+These reserves have proven to be effective in enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of highly valuable commercial species such as abalone [@munguavega_2015-yg].
+These ecological benefits have been translated into economic benefits, enhancing population persistence and bolstering abalone fisheries [@rossetto_2015-V0].
+For the purpose of this evaluation, we focused on the “La Plana / Las Cuevas” marine reserve, located in the southern end of the Island (Fig.
+3) and its corresponding control site “La Dulce / Babencho”.
 
-The reserve was implemented to recover species of economic interest -which were overexplotied- in order to enhance fishery production in nearby waters.
+The reserve was implemented to recover species of economic interest --–which were overexploited--- to enhance fishery production in nearby waters.
 Fishers were also interested in preserving biological diversity and the ecosystem.
-Thus, objectives objectives 4-7 were selected.
-Using **Table 2** to match these objectives with appropriate management indicators, we selected all biophysical, socioeconomic, and governance indicators.
+Thus, objectives 4--7 were selected.
+Using Table 2 to match these objectives with appropriate management indicators, we selected all biological, socioeconomic, and governance indicators included as options in the framework.
 
@@
 
-Biophysical data was gathered by fishers from the community.
-The reserves have been monitored following a standardized ReefCheck protocol [@suman_2010-ez] to collct data on fish and invertebrate communities inside and outside the reserves between 2016 and 2017.
-A total of `r sum(fish_summary$Reserva)` and `r sum(invert_summary$Reserva)` transects were performed in the reserve site for fish and invertebrate surveys, respectively.
+Local fishers ---trained in scientific diving by the NGO Comunidad y Biodiversidad, A.C.
+(COBI; [www.cobi.org](www.cobi.org))--- and personnel from COBI performed SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length.
+Information was recorded along 30 m transects, with a sampling window of 2 m X 2 m following a standardized ReefCheck protocol [@suman_2010-ez].
+Ecological surveys are performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, allowing us to have a before-after-control-impact sampling design.
+Yearly surveys (2006 -– 2016) were carried out in late July -- early August, performing a total of `r sum(fish_summary$Reserva)` and `r sum(invert_summary$Reserva)` transects in the reserve site for fish and invertebrate surveys, respectively.
 Similar sampling effort was applied to the control site, with `r sum(fish_summary$Control)` fish and `r sum(invert_summary$Control)` invertebrate transects.
@@
 
-Socioeconomic data was obtained from the National Comission for Fisheries and Acuaculture (Comisión Nacional de Pesca y Acuacultura; CONAPESCA).
-The data contains species-level (`r length(unique(landings$GeneroEspecie))+2` spp) information on monthly landings and income from 2000 to 2014.
-Data on landings and income was aggeregated by year, and the yearly Consumer Price Index was included.
-From the `r length(unique(landings$GeneroEspecie))+2` species available, we selected as objective species those that contribute the most (88.27%) to historical income: lobster (*Panulirus interruptus*; 71.76%), red sea urchin (*Strongylocentrotus franciscanus*; 9.33%), snail (*Megastraea undosa*; 3.93%), and sea cucumber (*Parastichopus parvimensis*; 3.23%).
-Abalone species (*Haliotis fulgens*; 4.52 and *Haliotis corrugata*; 6.16) were excluded because the cooperative decided to implement an informal closure on the species to allow its populations to recover.
-Eliminating all fishing preasure on these species means that control sites receive (for this species) the same treatment.
+Socioeconomic data was obtained from the National Commission for Fisheries and Aquaculture (Comisión Nacional de Pesca y Acuacultura; CONAPESCA).
+The data contains species-level (`r length(unique(landings$GeneroEspecie))+2` spp.) information on monthly landings and income from 2000 to 2014.
+Data on landings and income was aggregated by year, and the yearly Consumer Price Index was included.
+From the nine species available, we selected as objective species those that contribute the most (88.27%) to the available 2000 to 2014 income: lobster (*Panulirus interruptus*; 71.76%), red sea urchin (*Mesocentrotus franciscanus*; 9.33%), snail (*Megastraea undosa*; 3.93%), and sea cucumber (*Parastichopus parvimensis*; 3.23%).
+Abalone species (*Haliotis fulgens*; 4.52 and *Haliotis corrugata*; 6.16) were excluded because the cooperative has implemented an informal closure of these fisheries to allow its populations to recover since 2010.
+Eliminating all fishing pressure on abalones means that the control site receives (for this species) the same treatment as the reserve.
 
-For govenance data, we constructed the database based on our knowledge of the area, as well as prototype interviews performed to liders of the fishing cooperative.
+For governance data, we constructed the database based on our knowledge of the area and the community.
 

### 2017-08-18 Manuscript.Rmd
Updated results

@@
 
-# Results and Discussion
+# Results
 
 The results shown here intend to highlight the relevance and utility of the framework and tool, which automate the analysis and make it replicable.
-First, we present the scorecard generated by MAREA to provide an overview of the state of the reserve.
-Then, we provide (mdoified for style) a set of figures generated by MAREA, showing time-series for biophysical and socioeconomic indicators as well as tables containing the magnitude and significance of the $\beta_2$ and $\beta_1$ coefficients.
-A table with the governance indicators is also provided and discussed.
+While we highlight some of the general observed trends, we focus on the utility of the tool rather than on the case study.
 
-The scorecard (**Fig.
-4**) shows that, overall, the reserve has a good performance achieving a general score of 63.6% of positive indicators.
-All category-level scores were also high, with values of 66.7%, 60%, and 75% positive indicators for Biophysical, Socioeconomic and Governance, respectively.
+The scorecard (Fig.
+4) shows that, overall, the reserve achieves a general score of 63.6% of positive indicators.
+All category-level scores were also high, with values of 66.7%, 60%, and 75% positive indicators for Biological, Socioeconomic and Governance, respectively.
 
@@
 
-Within the biophysical indicators, the greatest effect of the reserve was observed for snail and cucumber densities, with values of $\beta_3 = `r formatC(bio_results_summary$estimate[11], digits = 2, format = "f")`$ (*p* < 0.05) and $\beta_3 = `r formatC(bio_results_summary$estimate[12], digits = 2, format = "f")`$ (*p* < 0.05), respectively.
-Urchin and lobster densities also increased by $\beta_3 = `r formatC(bio_results_summary$estimate[10], digits = 2, format = "f")`$ and $\beta_3 = `r formatC(bio_results_summary$estimate[9], digits = 2, format = "f")`$ respectively.
-However, changes in urchin densities were only significant at *p* < 0.1, and not significant for lobster (*p* > 0.1).
-Shannon's diversity index ($\beta_3 = `r formatC(bio_results_summary$estimate[6], digits = 2, format = "f")`$) and richness $\beta_3 = `r formatC(bio_results_summary$estimate[7], digits = 2, format = "f")`$ showed a significant decrease (*p* < 0.05).
+Among the biological indicators, the greatest effect of the reserve was observed for snail and cucumber densities, with values of $\beta_2 = `r formatC(bio_results_summary$estimate[11], digits = 2, format = "f")`$ (*p* < 0.05) and $\beta_2 = `r formatC(bio_results_summary$estimate[12], digits = 2, format = "f")`$ (*p* < 0.05), respectively.
 Fish indicators showed no significant change (*p* > 0.1), with negative trends for Shannon's diversity index and fish species richness and positive trends for density, biomass, and mean trophic level.
-Changes through time for these indicators are presented in **Figure 5**, and a summary of $\beta_3$ coefficients is provided in **Table 3**.
+Changes through time for these indicators are presented in Figure 5, and a summary of $\beta_2$ coefficients is provided in Table 3.
 
@@
 Total income was \$10,344.85 (*p* < 0.05) thousands of Mexican Pesos (MXP) higher after the implementation of the reserves.
-On avergae, lobster and cucumber landings and increased, while urchin and snail landings and income decreased.
-**Figure 6** presents the changes in this indicators through time, and **Table 4** summarizes this information.
+On average, lobster and cucumber landings increased, while urchin and snail landings and income decreased.
+Figure 6 presents the changes in these indicators through time, and Table 4 summarizes these results.
 
-The results identified by our framework and MAREA are consistent to those previously published [*e.g.* @micheli_2012-EU; @rossetto_2015-V0].
-Invertebrates show increased densities within the reserve, as compared to the control site, providing evidence that the reserve is effectively protecting these species.
-While we are not able to infer causality from the socioeconomic analysis, we are able to identify important trends.
-For example, snail and urchin density have significantly increased within the reserve, but their landings and income have decreased.
-The opposite is observed for lobster and cucumber, which have shown increases in densities, landing, and income.
-While further information on market behaviour of each species is needed, these results provide insights into the state of the reserve, as well as the associated fisheries.
-
-In terms of governance, it is evident that the community is strongly organized, which is a cause of their success.
-The first point of success is the existance of a fishing cooeprative that is also affiliated to federation.
-These polycentric governance structures allow different levels of organization that foster communication and cooperation; federations also provide bargain power with governments [@espinosaromero_2014-PY; @finkbeiner_2015-87].
+In terms of governance, it is evident that the community is strongly organized, which is likely a driver of their success.
+The first point of success is the existence of a fishing cooperative that is also affiliated to a regional fishing cooperatives federation.
+These polycentric governance structures allow various levels of organization that foster communication and cooperation; federations also provide bargain power with governments [@espinosaromero_2014-PY; @finkbeiner_2015-87].
 Fishers also have good management instruments.
 Access to the fishing resources they exploit is managed through permits and fishing quotas.
-Along with a stable number of fishers participating in extractive activities, these limit the total fishing effort effort applied.
+Along with a stable number of fishers participating in extractive activities, these limit the total fishing effortapplied.
 Additionally, their TURF promotes a sense of stewardship of their resources and incentivizes correct resource management [@mccay_2017-1m; @finkbeiner_2015-87].
-Together, these structures enabled a participative, bottom-up process during the reserve design phase; Opinions of all fishing members -and often non-fishers, but community members- were included.
-Participation of community members in reserve survailance and yearly monitoring indicate commitment and interest, and allow informal communication of results to un-involved community members.
+Together, these structures enabled a participative, bottom--up process during the reserve design phase; Opinions of all fishing members ---and often non-fishers, but community members--- were included.
+Participation of community members in reserve surveillance and yearly monitoring indicate commitment and interest, and allow informal communication of results to un-involved community members.
 
-While the reserve has not been legaly recognized (as of this moment), the community has started the process to have the reserve be recognized as a fishery replenishment zone (*Zona de Refugio Pesquer*) under the corresponding Mexican norms [@nom049sagpesc_2014-V6].
 Furthermore, the reserve is partially isolated from poaching activity and fishers have internal regulations pertaining the reserves.
 The low level of illegal fishing by members of the community and outsiders both inside and outside the reserve represents another indication of effectiveness.
-A summary of governance indicators is provided in **Table 5**.
+A summary of governance indicators is provided in Table 5.
 
@@
   kable(booktabs = T,
-        caption = "Summary of average treatment effect of the La Plana / Las Cuevas marine reserve on biophysical indicators.
-Asterisks indicate significance level, with (*) indicating *p* < 0.1 and (**) *p* < 0.05.",
+        caption = "Summary of average treatment effect of the reserve on biological indicators. Asterisks indicate significance level, with (*) indicating p < 0.1 and (**) p < 0.05.",
         digits = 2,
@@
   kable(booktabs = T,
-        caption = "Summary of average treatment effect of the La Plana / Las Cuevas marine reserve on socioeconomic indicators.
-Asterisks indicate significance level, with (*) indicating *p* < 0.1 and (**) *p* < 0.05.",
+        caption = "Summary of differences in socioeconomic indicators before and after the implementation of the reserve.
+Asterisks indicate significance level, with (*) indicating p < 0.1 and (**) p < 0.05.",
         digits = 2,
@@
 
-```{r plots, fig.height = 6, fig.width = 6, fig.cap = "Plots for values of each biophysical indicator (y-axis) through time (x-axis).
+```{r plots, fig.height = 6, fig.width = 6, fig.cap = "Plots for values of each biological indicator (y-axis) through time (x-axis).
 Red and blue correspond to the reserve and control sites, respectively.
 Black lines indicate yearly mean values, and ribbons indicate +/- 1 standard error.
-Dots are horizontally jittered to aid visualization."}
+Dots are horizontally jittered to aid visualization.
+This figure contains information for fish shannon diversity index (a), fish species richness (b), fish density (c), fish trophic level (d), fish biomass (e), invertebrate shannon diversity index (f), invertebrate species richness (g), invertebrate density (h), lobster density (i), urchin density (j), snail density (k), and cucumber density (l)."}
+
 shannon_fish_plot <- fish_results$plot[[1]] %>% 
-  my_theme(y.lab = "H'")
+  my_theme(y.lab = "H'", text.x = 2005, text.y = 2.75, text.label = "a")
 
 richness_fish_plot <- fish_results$plot[[2]] %>% 
-  my_theme(y.lab = "S (spp/transect)")
+  my_theme(y.lab = "S (spp./transect)", text.x = 2005, text.y = 9.5, text.label = "b")
 
 density_fish_plot <- fish_results$plot[[3]] %>% 
-  my_theme(y.lab = "D (org/transect)")
+  my_theme(y.lab = "D (org/transect)", text.x = 2005, text.y = 175, text.label = "c")
 
 trophic_fish_plot <- fish_results$plot[[4]] %>% 
-  my_theme(y.lab = "TL")
+  my_theme(y.lab = "TL", text.x = 2005, text.y = 4.05, text.label = "d")
 
 biomass_fish_plot <- fish_results$plot[[5]] %>% 
-  my_theme(y.lab = "B (Kg/transect)")
+  my_theme(y.lab = "B (Kg/transect)", text.x = 2005, text.y = 200, text.label = "e")
 
@@
 shannon_invert_plot <- invert_results$plot[[1]] %>% 
-  my_theme(y.lab = "H'")
+  my_theme(y.lab = "H'", text.x = 2005, text.y = 3.2, text.label = "f")
 
 richness_invert_plot <- invert_results$plot[[2]] %>% 
-  my_theme(y.lab = "S (spp/transect)")
+  my_theme(y.lab = "S (spp./transect)", text.x = 2005, text.y = 14, text.label = "g")
 
 density_invert_plot <- invert_results$plot[[3]] %>% 
-  my_theme(y.lab = "D (org/transect)")
+  my_theme(y.lab = "D (org/transect)", text.x = 2005, text.y = 950, text.label = "h")
 
 lobster_density_invert_plot <- invert_results$plot[[4]] %>% 
-  my_theme(y.lab = "lobsters (org/transect)")
+  my_theme(y.lab = "D (org/transect)", text.x = 2005, text.y = 275, text.label = "i")
 
 urchin_density_invert_plot <- invert_results$plot[[5]] %>% 
-  my_theme(x.lab = "year", y.lab = "urchin (org/transect)")
+  my_theme(x.lab = "", y.lab = "D (org/transect)", text.x = 2005, text.y = 50, text.label = "j")
 
 snail_density_invert_plot <- invert_results$plot[[6]] %>% 
-  my_theme(x.lab = "year", y.lab = "snail (org/transect)")
+  my_theme(x.lab = "year", y.lab = "D (org/transect)", text.x = 2005, text.y = 700, text.label = "k")
 
 cucumber_density_invert_plot <- invert_results$plot[[7]] %>% 
-  my_theme(x.lab = "year", y.lab = "cucumber (org/transect)")
+  my_theme(x.lab = "", y.lab = "D (org/transect)", text.x = 2005, text.y = 32, text.label = "l")
 
@@
 ```{r time series socio, fig.height = 6, fig.width = 6, fig.cap = "Plots for values of each socioeconomic indicator (y-axis) through time (x-axis).
-Red and blue correspond to before and after the implementation of the reserve, respectively."}
+Red and blue correspond to before and after the implementation of the reserve, respectively.
+This figure contains information for total landings (a), total income (b), lobster landings (c), urchin landings (d), snail landings (e), cucumber landings (f), lobster income (g), urchin income (h), snail income (i), and cucumber income (j)."}
+
 landings_plot <- socio$plot[[1]] %>% 
-  my_theme(y.lab = "Landings (Kg)")
+  my_theme(y.lab = "Landings (Kg)", text.x = 2001, text.y = 1150, text.label = "a", bio = F)
 
 income_plot <- socio$plot[[2]] %>% 
-  my_theme(y.lab = "Income (MXN)")
+  my_theme(y.lab = "Income (MXP)", text.x = 2001, text.y = 35000, text.label = "b", bio = F)
 
 landings_lobster_plot <- socio$plot[[3]] %>% 
-  my_theme(y.lab = "Lobster landings")
+  my_theme(y.lab = "Landings (Kg)", text.x = 2001, text.y = 175, text.label = "c", bio = F)
 
 landings_urchin_plot <- socio$plot[[4]] %>% 
-  my_theme(y.lab = "Urchin landings")
+  my_theme(y.lab = "Landings (Kg)", text.x = 2001, text.y = 55, text.label = "d", bio = F)
 
 landings_snail_plot <- socio$plot[[5]] %>% 
-  my_theme(y.lab = "Snail landings")
+  my_theme(y.lab = "Landings (Kg)", text.x = 2001, text.y = 300, text.label = "e", bio = F)
 
 landings_cucumber_plot <- socio$plot[[6]] %>% 
-  my_theme(y.lab = "Cucumber landings")
+  my_theme(y.lab = "Landings (Kg)", text.x = 2001, text.y = 58, text.label = "f", bio = F)
 
 income_lobster_plot <- socio$plot[[7]] %>% 
-  my_theme(y.lab = "Lobster income")
+  my_theme(y.lab = "Income (MXP)", text.x = 2001, text.y = 35000, text.label = "g", bio = F)
 
 income_urchin_plot <- socio$plot[[8]] %>% 
-  my_theme(y.lab = "Urchin income")
+  my_theme(y.lab = "Income (MXP)", text.x = 2001, text.y = 10000, text.label = "h", bio = F)
 
 income_snail_plot <- socio$plot[[9]] %>% 
-  my_theme(y.lab = "Snail income")
+  my_theme(y.lab = "Income (MXP)", text.x = 2001, text.y = 2100, text.label = "i", bio = F)
 
 income_cucumber_plot <- socio$plot[[10]] %>% 
-  my_theme(y.lab = "Cucumber income")
+  my_theme(y.lab = "Income (MXP)", text.x = 2001, text.y = 1500, text.label = "j", bio = F)
 

### 2017-08-18 Manuscript.Rmd
Started updating discussions

@@
 
-# Conclusions
+# Discussion and Conclusions
+
+Here we discuss the ways in which results can be interpreted and
+
+The interpretation of the $\beta_2$ coefficient is always relative to the control site.
+A positive value ($\beta_2 > 0$) might not nessesarily indicate an increase in the indicator.
+For example, invertebrate density presents a value of $\beta_2 = `r formatC(bio_results_summary$estimate[8], digits = 2, format = "f")`$ (*p* < 0.05), but Figure 5h shows a decrease in the indicator for the reserve and control sites through time.
+The decrease inside the reserve is not as abrupt as the changes in the control site, indicating that the reserves seem to provide a buffering effect.
+With additional knowledge of environmental variability (*i.e.* indicator B5 -- Natural Disturbance), we can better understand the performance of this indicator.
+As shown by @micheli_2012-EU, decreases in density can be caused by local hypoxic events.
+
+The results identified by our framework and MAREA are consistent to those previously published [*e.g.* @micheli_2012-EU; @rossetto_2015-V0].
+While invertebrates show decreases in density for reserve and control sites, their densities remained higher within the reserve.
+Here, the positive effect indicated by $\beta_2$ does not indicate an increase in density, but rather a slower decrease in the reserve.
+This has occurred likely due to local hypoxic events [@micheli_2012-EU], where the reserves seem to provide a buffering effect.
+
+---
+
+While we are not able to infer causality from the socioeconomic analysis, we are able to identify important trends.
+For example, snail and urchin density have significantly increased within the reserve, but their landings and income have decreased.
+The opposite is observed for lobster and cucumber, which have shown increases in densities, landing, and income.
+While further information on market behavior of each species is needed, these results provide insights into the state of the reserve, as well as the associated fisheries.
+
+-----
 
 One of the greatest challenges in management measuring the extent to which objectives have been met.
-The present framework provides a simple and straightforward way to align management objectives with performance indicators.
-We aknowledge that these 25 indicators might not fully describe a particular reserve.
-However, we believe that these provide a starting point to perform the evaluation, and urge decision-makers, managers, and users to include other indicators (*e.g.* larval dispersal or connectivity) that are relevant to their reserve.
+The present framework provides a simple and replicable way to align management objectives with performance indicators.
+We recognize that these 25 indicators might not fully describe a reserve.
+However, they provide a starting point to perform the evaluation, to which managers and users can add other indicators (*e.g.* larval dispersal or connectivity) that are relevant to their reserve.
 
-The proposed methodologies, specially the way in which biophysical indicators are evaluated, provide valuable information for managers.
-The analysis isolates the net effect of the reserve, providing a propper measure of reserve effectiveness.
-We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analized.
-Despire this, we believe that providing a unifying platform where all can be analyzed and comprehensively presented represents a valuable step towards evidence-based effective management.
+The proposed methodologies, especially the way in which biological indicators are evaluated, provide valuable information for managers.
+The analysis isolates the net effect of the reserve, providing a proper measure of reserve effectiveness.
+We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analyzed.
+Despite this, we believe that providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards evidence-based effective management.
 
-Furthermore, MAREA's value is that it provides a free, simple, and replicable way to perform rigurous analysis.
-The tool can easily be used by fishers, NGO members, and managers in government agencies, providing transparency of the analysis and results.
-The way in which results are presented allow this information to be interpreted by a wider audience.
-The scorecard is easily understandable by experts and non-experts, and can be used as an effective tool for communicating the results of yearly evaluations.
-On the other side, the technical report can serve as a tool for managers and scientists to rapidly communicate information at a more technical level.
+Furthermore, MAREA's value is that it provides a free, simple, and replicable way to perform rigorous impact analysis.
+The tool can easily be used by fishers, SCO members, and managers in government agencies, providing transparency of the analysis and results.
+In addition, it can empower and enable local managers and fishers to respond to local change and adapt by allowing direct and easy access to the information.
+The way in which results are presented allows this information to be interpreted by a wide-ranging audience.
+The scorecard is easily understandable by experts and non-experts, and can be used as an effective tool for communicating the results of annual evaluations.
+Additionally, the technical report can serve as a tool for managers and scientists to rapidly produce and communicate information at a more technical level.
 
-While the first release of MAREAis now available, it will continue to be developed and maintained.
+While the first release of MAREA is now available, it will continue to be developed and maintained.
 This will incorporate new features, and enhance current ones, aiming to improve user experience and expand the scope of the analysis.
-Yet, we believe that this first release represents a major step towards effective management of marine reserves.
+Yet, we believe that this first release represents a major step towards effective evaluation and management of marine reserves.
 
@@
 
-The authors wish to thank the fishers from Isla Natividad, who gathered the data used in this study, and the fishers from El Rosario, who helped us validate our survey and framework.
+We thank Olivier Deschenes and Andrew Plantinga, who provided valuable input to design the model that evaluates the biological indicators.
+Special thanks to the fishers from Isla Natividad, who gathered the data used in this study, and the fishers from El Rosario, who helped us validate our survey and framework, and to Arturo Hernández and Alfonso Romero who provided help with the logistics.
 Input provided by the participants in the workshop "" was very valuable.
-We also thank Olivier Deschenes and Andrew Plantinga, who provided valuable input to design the model that evaluates the biophysical indicators.
+
+# Funding
+
+- CF and JC received support from the Latin American Fisheries Fellowship Program.
+- JC received support from the Consejo Nacional de Ciencia y Tecnología (CONACyT).
+- GM - We would also like to thank the Waitt Foundation for their financial support. Carlos Slim Foundation Alliance.
+- The biological data for Isla Natividad were collected with the support of the Walton Family Foundation and the NSF-CNH program (grant DEB-1212124).
+- We also thank the support of: The Nature Conservancy and World Wildlife Fund-
 

### 2017-08-18 Functions/my_theme.R
added labeling functionality

@@ -1,26 +1,25 @@
-my_theme <- function(my_plot, legend = F, x.lab = "", y.lab = ""){
+my_theme <- function(my_plot, x.lab = "", y.lab = "", text.x = 2006, text.y = 1, text.label = "", bio = T){
   
-  if (legend){
-    my_plot <- my_plot +
-      theme(legend.justification=c(0,1),
-            legend.position=c(0,1),
-            legend.background = element_rect(fill = "transparent", size=.25),
-            legend.key.size = unit(0.5, "cm")) +
-      guides(color = guide_legend(title = NULL),
-             fill = guide_legend(title = NULL)) +
-      labs(x = x.lab, y = y.lab) +
-      scale_x_continuous(breaks = c(2006, 2008, 2010, 2012, 2014, 2016),
-                         labels = c("06", "08", "10", "12", "14", "16")) +
-      theme(text = element_text(family = "Times", size = 10))
-  }
-  
-  if (!legend){
+  if(bio){
     my_plot <- my_plot +
       theme(legend.position = "None") +
       labs(x = x.lab, y = y.lab) +
-      scale_x_continuous(breaks = c(2006, 2008, 2010, 2012, 2014, 2016),
-                         labels = c("06", "08", "10", "12", "14", "16")) +
-      theme(text = element_text(family = "Times", size = 10))
+      scale_x_continuous(limits = c(2005, 2017),
+                         breaks = c(2006, 2008, 2010, 2012, 2014, 2016),
+                         labels = c("'06", "'08", "'10", "'12", "'14", "'16")) +
+      theme(text = element_text(family = "Times", size = 10, face = "plain")) +
+      geom_text(x = text.x, y = text.y, label = text.label, aes(size = 0.5, fontface = "plain"), color = "black")
+  }
+  
+  if(!bio){
+    my_plot <- my_plot +
+      theme(legend.position = "None") +
+      labs(x = x.lab, y = y.lab) +
+      scale_x_continuous(limits = c(2000, 2014),
+                         breaks = c(2000, 2002, 2004, 2006, 2008, 2010, 2012, 2014),
+                         labels = c("'00", "'02", "'04","'06", "'08", "'10", "'12", "'14")) +
+      theme(text = element_text(family = "Times", size = 10, face = "plain")) +
+      geom_text(x = text.x, y = text.y, label = text.label, aes(size = 0.5, fontface = "plain"), color = "black")
   }
   
   return(my_plot)

### 2017-08-19 Manuscript.Rmd
Started updating the dicsussion

@@
 
-Here we discuss the ways in which results can be interpreted and
+Here we focus on the advantages and room for improvement in MAREA and highlight its relevance for management.
+While we do not discuss the case study ---which has been described before [@micheli_2012-EU; @rossetto_2015-V0; @munguavega_2015-yg]-- we draw from it to build on the usefulness of MAREA and discuss the ways in which results can be interpreted and used to better inform management.
 
-The interpretation of the $\beta_2$ coefficient is always relative to the control site.
-A positive value ($\beta_2 > 0$) might not nessesarily indicate an increase in the indicator.
-For example, invertebrate density presents a value of $\beta_2 = `r formatC(bio_results_summary$estimate[8], digits = 2, format = "f")`$ (*p* < 0.05), but Figure 5h shows a decrease in the indicator for the reserve and control sites through time.
-The decrease inside the reserve is not as abrupt as the changes in the control site, indicating that the reserves seem to provide a buffering effect.
-With additional knowledge of environmental variability (*i.e.* indicator B5 -- Natural Disturbance), we can better understand the performance of this indicator.
-As shown by @micheli_2012-EU, decreases in density can be caused by local hypoxic events.
+MAREA uses causal inference techniques to evaluate the effect of the reserve on biological indicators, which allows us to estimate the real effect of the intervention [@ferraro_2006-oW].
+With different approaches, such as a before--after comparision, the trends in an indicator might be ambiguous.
+For example, invertebrate density decreased through time inside and outside of the reserve (Fig.
+5h).
+In this case, a before--after evaluation of the reseve (*i.e.* ignoring the control site) would have incorrectly concluded that the reserve wasn't working.
+However, a control--impact approach (*i.e.* compare reserve vs.
+control site only in 2016) would have seen higher densities inside the reserve, leading to the conclusion that the reserve increases invertebrate density.
+However, MAREA identifies the changes through time and across sites, and estimates the effect of the reserve at $\beta_2 = `r formatC(bio_results_summary$estimate[8], digits = 2, format = "f")`$ (*p* < 0.05).
+What truly happens is that invertebrate densities decrease in both sites through time, but the decrease is faster for the control site, thus yielding a positive value for $\beta_2$.
 
-The results identified by our framework and MAREA are consistent to those previously published [*e.g.* @micheli_2012-EU; @rossetto_2015-V0].
-While invertebrates show decreases in density for reserve and control sites, their densities remained higher within the reserve.
-Here, the positive effect indicated by $\beta_2$ does not indicate an increase in density, but rather a slower decrease in the reserve.
-This has occurred likely due to local hypoxic events [@micheli_2012-EU], where the reserves seem to provide a buffering effect.
+The approach used by MAREA to estimate the effect of the reserve on biological indicators requires cautious interpretation of the results.
+The value of the $\beta_2$ coefficient represents the difference between the temporal trends of the reserve and control sites.
+As exemplified by the case of invertebrate densities, a positive value (*i.e.* $\beta_2 > 0$) does not neccessarily indicate an increase in the indicator through time.
+The same is truth for negative values of $\beta_2$.
 
----
+MAREA provides an overview of the effect of the reserve, allowing the users to rapidly identify trends.
+However, users might need to combine results from multiple indicators to better understand the results.
+For example, with additional knowledge of local environmental variability (*i.e.* indicator B5 -- Natural Disturbance), we can better understand the trends in invertebrate densities.
+As reported before @micheli_2012-EU, hypoxic conditions that have ocurred in Isla Natividad can cause decreases in invertebrate densities and reserves buffer the negative effect.
+MAREA is intended to provide automatization and replicability of the analyses, but a propper interpretation will still depend on the user.
+Results produced by MAREA can only aid in management and decission making after its results have been correctly interpreted.
 
-While we are not able to infer causality from the socioeconomic analysis, we are able to identify important trends.
+Cannot use causal inference techniques in socio eco and governance data due to their nature [@mascia_2017-m_].
+However, we can still extract usefull information.
+Again, by combining results from multiple indicators, MAREA can provide insights into the effect of the reserve.
+**ARREGLAR EL SOCIOECONOMICO --- While we are not able to infer causality from the socioeconomic analysis, we are able to identify important trends.
 For example, snail and urchin density have significantly increased within the reserve, but their landings and income have decreased.
 The opposite is observed for lobster and cucumber, which have shown increases in densities, landing, and income.
-While further information on market behavior of each species is needed, these results provide insights into the state of the reserve, as well as the associated fisheries.
+While further information on market behavior of each species is needed, these results provide insights into the state of the reserve, as well as the associated fisheries.** As for the governance information, there is no clear path to establish causal links between the reserve and the governance systems used.
+However, providing a single platform (scorecard) or document (technical report) where all information is comprehensively included might aid in management.
+Furthermore, this information would be standardly reported across reserves, providing managers with a single method of comapring everything.
 
------
-
-One of the greatest challenges in management measuring the extent to which objectives have been met.
-The present framework provides a simple and replicable way to align management objectives with performance indicators.
 We recognize that these 25 indicators might not fully describe a reserve.
 However, they provide a starting point to perform the evaluation, to which managers and users can add other indicators (*e.g.* larval dispersal or connectivity) that are relevant to their reserve.
-
-The proposed methodologies, especially the way in which biological indicators are evaluated, provide valuable information for managers.
-The analysis isolates the net effect of the reserve, providing a proper measure of reserve effectiveness.
-We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analyzed.
-Despite this, we believe that providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards evidence-based effective management.
 
@@
 Additionally, the technical report can serve as a tool for managers and scientists to rapidly produce and communicate information at a more technical level.
+
+One of the greatest challenges in management measuring the effect of interventions [@ferraro_2006-oW].
+The present framework provides a simple and replicable way to align management objectives with performance indicators.
+The proposed methodologies, especially the way in which biological indicators are evaluated, provide valuable information for managers.
+The analysis isolates the net effect of the reserve, providing a proper measure of reserve effectiveness.
+We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analyzed.
+Despite this, we believe that providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards evidence--based effective management[@ferraro_2006-oW].
 

### 2017-08-21 Manuscript.Rmd
Updated manuscript

--- 2017-08-18 Manuscript.Rmd
@@
 As with any other management interventions, we need to better understand their effectiveness and impacts on coastal communities and the environment.
-To date, the evaluation of these reserves largely relies on analyzing ecological data, often ignoring socioeconomic and governance dimensions.
+To date, their evaluation largely relies on analyzing ecological data, often ignoring socioeconomic and governance dimensions.
 Existing data are analyzed in different ways, hindering the ability to compare results across case studies.
 Moreover, analysis and evaluation of reserves is generally conducted by academic scientists, not the reserves managers and users, thereby hindering effective local management and rapid response to change.
-We present a framework and tool to evaluate the effectiveness of marine reserves by matching seven commonly stated management objectives to nine biological, five socioeconomic, and 14 governance indicators.
+We present a framework and tool to evaluate the effectiveness of marine reserves by matching seven commonly stated management objectives to nine biological, five socioeconomic, and 15 governance indicators.
 We provide guidelines on how to properly collect data that can then be analyzed with standardized method.
-Biological indicators are evaluated with causal inference techniques, using a counterfactual approach, to assess the net effect of the reserve on each indicator.
+Biological indicators are evaluated with causal inference techniques, using a counterfactual approach, to assess the effect of the reserve on each indicator.
 Linear regression models are fitted to socioeconomic indicators through time to test for differences before and after reserve implementation.
-Governance indicators are qualitatively analyzed using a framework developed through a literature review which identifies common governance structures and their associated effectiveness.
+Governance indicators are qualitatively analyzed using a framework developed through a literature review, which identifies common governance structures and their associated effectiveness.
 To make the framework accessible to fishers and decision makers, and allow replication of results, we developed the open source, web-based Marine Reserve Evaluation App (MAREA).
@@
 Overfishing and unsustainable fishing practices are two of the largest threats to the conservation of marine ecosystems around the world [@pauly_2005-qV; @halpern_2008-dK].
-Marine Protected Areas (MPAs) are frequently proposed as fishery management and conservation tools to help fish and invertebrate stocks recover [@lester_2008-F_; @lester_2009-Ks: @sala_2016-PV] by limiting or restricting fishing effort and gears.
+Marine Protected Areas (MPAs) are frequently proposed as fishery management and conservation tools to help fish and invertebrate stocks recover [@lester_2008-F_; @lester_2009-Ks: @sala_2016-PV; @hastings_2017-sm] by limiting or restricting fishing effort and gears.
 No-take marine reserves (marine reserves hereinafter) are a particular type of MPA, where all fishing effort and extractive activities are off-limits [-@iucn_website_2017-Bg; @hortaecosta_2016-_3].
@@
 These effects are often measured as biological changes in the area through time and lack a control site against which to compare [@betti_2017-lq].
-This before-after comparison cannot account for other factors for which one must control [@davies_2017-ml] in order to attribute biological change to protection in the reserves.
+This before--after comparison cannot account for other factors for which one must control [@davies_2017-ml] in order to attribute biological change to protection in the reserves.
 While some studies have used control sites, these analyses do not estimate the net effect of the reserve, and often use a control-impact comparison approach that does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
-A smaller fraction of studies have used a before-after-control-impact design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.
+A smaller fraction of studies have used a before-after-control-impact (*i.e.* BACI) design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.
 
@@
 By excluding these important dimensions, the evaluation provides only a partial picture of the impacts of the reserve.
+
 Currently, only the IUCN framework "How to evaluate your MPA" [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biological, socioeconomic, and governance indicators, and insights into how these indicators may be measured, but does not provide guidelines on how to analyze them.
@@
 
-An increasingly popular way to make science reproducible, scalable and replicable is through Open Science and the development of open-access tools [@lowndes_2017-xh].
+An increasingly popular way to make science accessible, reproducible, scalable, and replicable is through Open Science and the development of open-access tools [@lowndes_2017-xh].
 The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, has successfully standardized a way to measure the health and benefits of the oceans.
 This approach has been implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC].
-Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance [@anderson_2015-ND; @dowling_2016-pO], design territorial use rights for fisheries (TURFs; @oyanedel_2017-TO), and improve decision making in the hydropower industry [@vilela_2017-Zo], just to list a few.
+Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance [@anderson_2015-ND; @dowling_2016-pO], design territorial use rights for fisheries (TURFs; @oyanedel_2017-TO), and improve decision making in the hydro power industry [@vilela_2017-Zo], just to list a few.
 
@@
 The current work presents a framework to evaluate marine reserves, which incorporates the biological, socioeconomic, and governance dimensions of these areas.
-We first provide a list of commonly stated management objectives and match them to appropriate indicators that measure the effectiveness of the management intervention.
+We first provide a list of commonly stated management objectives and match them to appropriate indicators.
 We then include a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_].
-We also introduce the Marine Reserve Evaluation App (MAREA), an open source web--based tool that automates the framework described in this paper and enables its broader use.
+We also introduce the Marine Reserve Evaluation App (MAREA), an open source, web--based tool that automates the framework described in this paper and enables its broader use.
 Finally, we present a case study on the evaluation of a marine reserve established by the fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.
-
-\clearpage
 
@@
 Then, we describe the development of MAREA and explain how this user-friendly open access tool can be used by anyone.
-Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA.
+Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA to inform management.
 
-![Workflow to evaluate the effectiveness of marine reserves.](./Figures/flowchart.gif){ width=50% }
+**Fig. 1. Workflow to evaluate the effectiveness of marine reserves.**
+
+```{r, fig.cap = "Workflow to evaluate the effectiveness of marine reserves."}
+fig1 <- knitr::include_graphics("./Figures/flowchart.gif", dpi = 2380)
+```
 
@@
 However, we group these objectives into seven major categories, which can be applied to marine reserves worldwide.
-The list of objectives was developed through a literature review, which compiled stated objectives in and legislation [@nom049sagpesc_2014-V6; @lgeepa_2017-jL] and official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, and decrees associated to these areas.
+The list of objectives was developed through a literature review, which compiled stated objectives in legislation [@nom049sagpesc_2014-V6; @lgeepa_2017-jL] and official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, and decrees associated to these areas.
 Even though each reserve has its own goals, seven main categories of objectives were identified:
@@
 2.	Conserve species under a special protection regime
-3.	Maintain biological process (reproduction, recruitment, growth, feeding)
+3.	Maintain biological processes (reproduction, recruitment, growth, feeding)
 4.	Improve fishery production in nearby waters
@@
 The preliminary list of indicators was reviewed at a workshop with participation of members from Mexican fishery management agencies and non-government organizations.
-Later, these were also presented to fishers from the Ensenada Fishing Cooperative (*S.C.P.P.
+Later, these were presented to fishers from the Ensenada Fishing Cooperative (*S.C.P.P.
 Ensenada*), in Baja California, who provided input.
@@
 Indicators are divided into three main categories: biological, socioeconomic, and governance (Table 1).
-Biological indicators (n = 7) focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data and Analysis section for specific sampling design and methodologies).
-Socioeconomic indicators (n = 3) reflect the performance of the fishery in terms of catches, income from catches, and availability of alternative livelihoods.
+Biological indicators (n = 9) focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data and Analysis section for specific sampling design and methodologies).
+Socioeconomic indicators (n = 5) reflect the performance of the fishery in terms of landings, income from landings, and availability of alternative livelihoods.
 Governance indicators (n = 15) describe the governance structures under which the community operates (*e.g.*, access rights to the fishery, number of fishers, legal recognition of the reserve).
@@
 Whenever an indicator is applied to “Target species”, it means that the indicator can be used for all species (*e.g.* Fish Biomass) and/or for individual species that are either the conservation target of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass).
-Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with indicators.
+Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with biological and socioeconomic indicators; all governance indicators should always be used.
 
@@
 read.csv("./Tables/Table_indicators.csv") %>% 
-  knitr::kable(caption = "List of indicators to evaluate the effectiveness of no-take marine reserves, grouped by type.
-Type of data and units are provided.", col.names = c("Code", "Indicator", "Data type", "Unit"), booktabs = T) %>%
+  knitr::kable(caption = "List of indicators to evaluate the effectiveness of no-take marine reserves, grouped by type.",
+               col.names = c("Code", "Indicator", "Data type", "Unit")) %>%
   kable_styling(latex_options = c("scale_down")) %>%
-  group_rows("Biophysical", 1, 7) %>%
+  group_rows("Biological", 1, 7) %>%
   group_rows("Socioeconomic", 8, 10) %>% 
   group_rows("Governance", 11, 25) %>% 
-  add_footnote(c("The indicator is applied to objective species"), notation = "symbol")
+  add_footnote(c("The indicator is applied to target species"), notation = "symbol")
 ```
@@
 read.csv("./Tables/Objectives_Indicators.csv") %>% 
-  knitr::kable(caption = "This table indicates which indicators must be used for each objective.",
-               col.names = c("Objective", "B1", "B2", "B3", "B4", "B4*", "B5", "B6", "B7", "B7*", "S1", "S1**", "S2", "S2*", "S3", "G1", "G2", "G3", "G4", "G5", "G6", "G7", "G8", "G9", "G0", "G1", "G2", "G3", "G4", "G5"),
-               booktabs = T) %>% 
+  select(1:15) %>% 
+  knitr::kable(caption = "Management objectives and respective performance indicators.",
+               col.names = c("Objective", "B1", "B2", "B3", "B4", "B4*", "B5", "B6", "B7", "B7*", "S1", "S1*", "S2", "S2*", "S3")) %>% 
   kable_styling(latex_options = c("scale_down"))  %>% 
   column_spec(1, width = "4cm") %>% 
-  landscape()
+  add_footnote(c("The indicator is applied to target species"), notation = "symbol")
 ```
@@
 Scientific divers record fish and invertebrate richness and abundances, as well as fish total length along belt transects.
-Ecological surveys are performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, allowing us to have a before-after-control-impact (*i.e.* BACI) sampling design.
+Ecological surveys are performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, allowing us to have a BACI sampling design.
 Control sites are areas where habitat is similar to that of the reserve, but with presence of fishing activity.
-While transect dimensions (i.e.
-length and width) and sampling methods might vary from study to study, the general idea remains the same: richness, abundances, and sizes of organisms are recorded in a study-specific standardizes way.
+While transect dimensions (*i.e.* length and width) and sampling methods might vary from study to study, the general idea remains the same: richness, abundances, and sizes of organisms are recorded in a study--specific standardized way.
 For this reason, MAREA does not assume specific transect dimensions, and pertinent indicators are calculated per transect (Table 1).
 
-Having control sites for biological data allows us to use causal inference techniques [@moland_2013-VP; @ferraro_2006-oW] to evaluate the net effect of the reserve.
+Having a BACI sampling design for biological data allows us to use causal inference techniques [@moland_2013-VP; @ferraro_2006-oW] to evaluate the effect of the reserve on biological indicators.
 The hypothesis that the indicators will respond to implementation of the reserve is tested by analyzing spatial and temporal changes in each numeric biological indicator (all but B5) using generalized linear models [@moland_2013-VP].
-To account for variations in the environment and survey conditions, covariates that are gathered during the underwater ecological surveys are included into a model with form:
+To account for variations in the environment and survey conditions, covariates that are gathered during the underwater ecological surveys are included into difference-in-differences a model with form:
 
@@
 This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), a dummy variable that indicates treatment ($Z$; *i.e.* control or reserve), an interaction between a dummy variable that indicates pre-- or post--implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$; in °C), horizontal visibility during the survey ($V$; in m), and depth at which survey was performed ($D$; in m).
-$\epsilon$ represents the error term associated to the equation.
+$\epsilon$ represents the error term associated to the regression.
 Here, years are modeled as factors, using the first year as the reference level.
-By modelling years as factors, we avoid imposing a linear structure in the way an indicator changes through time (*i.e.* the change in biomass between 2006 and 2007 does not have to be the same as the change between 2015 and 2016).
-The treatment and implementation variables, modeled as dummy variables, are coded as Control = 0 and Reserve = 1; and Pre-implementation = 0 and Post-implementation = 1, respectively.
+This does not impose a linear structure in the way an indicator changes through time (*i.e.* the change in biomass between 2006 and 2007 does not have to be the same as the change between 2015 and 2016).
+The treatment and implementation variables, modeled as dummy variables, are coded as Control = 0 and Reserve = 1; and Pre--implementation = 0 and Post--implementation = 1, respectively.
 
-Socioeconomic data are often collected by fishers, fishery management agencies or OSCs by recording landings, income, and sometimes prices for each species.
-To control for inflation and changes in buying power, income is adjusted with the country’s consumer price index [@oecd_website_2017-VV]:
+Socioeconomic data are often collected by fishers, fishery management agencies or CSOs by recording landings, income, and sometimes prices for each species.
+To control for inflation, income is adjusted with the country’s consumer price index [@oecd_website_2017-VV]:
 
@@
 1 and eq.
-3), coefficients are estimated via ordinary least squares, and heteroskedastic-robust standard errors are calculated.
+3), coefficients are estimated via ordinary least squares, and heteroskedastic--robust standard errors are calculated.
 
@@
 Therefore, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1-G15).
-The survey is included as supplementary material (Appendix 1).
-To analyze governance information, we develop a framework based on a literature review of common governance structures and their relation to effectiveness in managing fisheries or marine reserves (Appendix 2).
+The survey is included as supplementary material in English (S1 Appendix) and Spanish (S2 Appendix).
+To analyze governance information, we developed a framework based on a literature review of common governance structures and their relation to effectiveness in managing fisheries or marine reserves (S3 Table).
 This approach has been proven to successfully evaluate governance structures [@espinosaromero_2014-PY].
@@
 
-MAREA was developed in R Studio [@rcore_2017] using the Shiny package [@shiny_2017], which provides the tools to build interactive web applications, hosted on an open server.
+MAREA was developed in R version 3.4.1 and R Studio [@rcore_2017] using the Shiny package [@shiny_2017], which provides the tools to build interactive web applications, hosted on an open server.
 MAREA can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/).
@@
 Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on Table 2.
-Users can also manually select additional indicators or unselect the default ones, based on their interests and data availability.
+Users can also manually modify selected indicators based on their interests and data availability.
 The user can then load the data, using standard \*.csv text files; sample datasets are provided within MAREA.
@@
 
-The first output is a color-coded scorecard intended to provide a general overview of the effectiveness of the reserve.
+The first output is a color--coded scorecard intended to provide a general overview of the effectiveness of the reserve.
 The scorecard provides a global score for the reserve, a general score for each category of indicators, and an individual score for each indicator.
-The global and category-level scores are determined by the percentage of positive indicators, overall and for each category, respectively.
+The global and category--level scores are determined by the percentage of positive indicators, overall and for each category, respectively.
 For numeric biological indicators (all but B5), the color is defined by the sign of the interaction term coefficient ($\beta_2$) in eq.
@@
 Red, yellow and green are used for $\beta_i<0$, $\beta_i = 0$, and $\beta_i>0$, respectively.
-The intensity of the color is defined by the significance of coefficient, testing the null hypothesis of no change (*i.e.* $H_0: \beta_i = 0$) with a Students t-test.
+The intensity of the color is defined by the significance of the coefficient, testing the null hypothesis of no change (*i.e.* $H_0: \beta_i = 0$) with a Student's t-test.
 Cutoff values are $p < 0.05$ and $p < 0.1$.
@@
 Governance indicators are represented simply by red or green.
-The color is defined based on what literature shows to be a negative (red) or positive (green) factor for a reserve.
+The color is defined based on what literature shows to be a negative (red) or positive (green) factor for a reserve (S3 Table).
 For example, if the perceived degree of illegal fishing is high, this is indicator will be assigned a red color.
@@
 
-![Legend used to interpret the scorecard produced by MAREA. Colors indicate direction of change, and color intensity is given by the statistical significance.](./Figures/legend2.gif)
+**Fig.
+2.
+Legend used to interpret the scorecard produced by MAREA.** Colors indicate direction of change (red = negative; green = positive), and color intensity is given by the statistical significance.
+
+```{r}
+fig2 <- knitr::include_graphics("./Figures/legend2.gif")
+```
 
@@
 This report also includes a scorecard as a summary of the results, but provides more information for each indicator.
-For all numeric indicators, the report includes a graph of the value of the indicator, for both the reserve and control sites, through time.
+For all numeric indicators, the report includes a graph of the value of the indicatorin the reserve and control sites, through time.
 It also provides a regression table that summarizes the value of all coefficients in the regression and their respective robust standard errors.
@@
 Another feature of MAREA is that the user can choose to share the data.
-Once the technical report is downloaded, the information on the reserve, its management objectives, and all uploaded data is saved into a central repository.
+Once the technical report is downloaded, the information on the reserve, its management objectives, and all uploaded data are saved into a central repository.
 These data can be accessed at any time by any person interested in acquiring them at [github.com/turfeffect/MAREAdata](github.com/turfeffect/MAREAdata).
@@
 
-```{r map, fig.cap = "General location of Isla Natividad (left) and map of the island (right).
-The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce)."}
+**Fig.
+3.
+General location of Isla Natividad (left) and map of the island (right).** The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce).
+
+```{r map}
 proj <- CRS("+proj=longlat +datum=NAD27")
@@
   tm_scale_bar(position = c("left", "bottom"))
-
-tmap_arrange(pen_map, study_area, outer.margins = 0.005, ncol = 2)
 ```
@@
 
-The reserve was implemented to recover species of economic interest --–which were overexploited--- to enhance fishery production in nearby waters.
+The reserve was implemented to recover species of economic interest --–which were overexploited--- and to enhance fishery production in nearby waters.
 Fishers were also interested in preserving biological diversity and the ecosystem.
@@
 Information was recorded along 30 m transects, with a sampling window of 2 m X 2 m following a standardized ReefCheck protocol [@suman_2010-ez].
-Ecological surveys are performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, allowing us to have a before-after-control-impact sampling design.
+Ecological surveys are performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, allowing us to have a BACI sampling design.
 Yearly surveys (2006 -– 2016) were carried out in late July -- early August, performing a total of `r sum(fish_summary$Reserva)` and `r sum(invert_summary$Reserva)` transects in the reserve site for fish and invertebrate surveys, respectively.
@@
 Socioeconomic data was obtained from the National Commission for Fisheries and Aquaculture (Comisión Nacional de Pesca y Acuacultura; CONAPESCA).
-The data contains species-level (`r length(unique(landings$GeneroEspecie))+2` spp.) information on monthly landings and income from 2000 to 2014.
-Data on landings and income was aggregated by year, and the yearly Consumer Price Index was included.
-From the nine species available, we selected as objective species those that contribute the most (88.27%) to the available 2000 to 2014 income: lobster (*Panulirus interruptus*; 71.76%), red sea urchin (*Mesocentrotus franciscanus*; 9.33%), snail (*Megastraea undosa*; 3.93%), and sea cucumber (*Parastichopus parvimensis*; 3.23%).
-Abalone species (*Haliotis fulgens*; 4.52 and *Haliotis corrugata*; 6.16) were excluded because the cooperative has implemented an informal closure of these fisheries to allow its populations to recover since 2010.
+The data contains species-level (n = `r length(unique(landings$GeneroEspecie))+2` spp.) information on monthly landings and income from 2000 to 2014.
+Data on landings and income was aggregated by year, and the yearly Consumer Price Index [@oecd_website_2017-VV] was included.
+From the nine species available, we selected as objective species those that contributed the most (88.27%) income from 2000 to 2014: lobster (*Panulirus interruptus*; 71.76%), red sea urchin (*Mesocentrotus franciscanus*; 9.33%), snail (*Megastraea undosa*; 3.93%), and sea cucumber (*Parastichopus parvimensis*; 3.23%).
+Abalone species (*Haliotis fulgens*; 4.52% and *Haliotis corrugata*; 6.16%) were excluded because the cooperative has implemented an informal closure of these fisheries to allow its populations to recover since 2010.
 Eliminating all fishing pressure on abalones means that the control site receives (for this species) the same treatment as the reserve.
@@
 The scorecard (Fig.
-4) shows that, overall, the reserve achieves a general score of 63.6% of positive indicators.
-All category-level scores were also high, with values of 66.7%, 60%, and 75% positive indicators for Biological, Socioeconomic and Governance, respectively.
+4) shows the reserve achieves a general score of 63.6% positive indicators.
+All category--level scores were also high, with values of 66.7%, 60%, and 75% positive indicators for biological, socioeconomic and governance, respectively.
 
-\clearpage
+**Fig. 4. Scorecard produced by MAREA for the "La Plana / Las Cuevas" marine reserve in Isla Natividad, Mexico.**
 
-![Scorecard produced by MAREA for the "La Plana / Las Cuevas" marine reserve in Isla Natividad, Mexico.](./Figures/scorecard.png){ width=50% }
+```{r}
+fig4 <- include_graphics("./Figures/scorecard.png")
+```
 
@@
 
-In the case of socioeconomic indicators, total landings were on average 64.20 tones higher (*p* > 0.1) after the implementation of the reserves.
-Total income was \$10,344.85 (*p* < 0.05) thousands of Mexican Pesos (MXP) higher after the implementation of the reserves.
-On average, lobster and cucumber landings increased, while urchin and snail landings and income decreased.
-Figure 6 presents the changes in these indicators through time, and Table 4 summarizes these results.
-
-In terms of governance, it is evident that the community is strongly organized, which is likely a driver of their success.
-The first point of success is the existence of a fishing cooperative that is also affiliated to a regional fishing cooperatives federation.
-These polycentric governance structures allow various levels of organization that foster communication and cooperation; federations also provide bargain power with governments [@espinosaromero_2014-PY; @finkbeiner_2015-87].
-Fishers also have good management instruments.
-Access to the fishing resources they exploit is managed through permits and fishing quotas.
-Along with a stable number of fishers participating in extractive activities, these limit the total fishing effortapplied.
-Additionally, their TURF promotes a sense of stewardship of their resources and incentivizes correct resource management [@mccay_2017-1m; @finkbeiner_2015-87].
-Together, these structures enabled a participative, bottom--up process during the reserve design phase; Opinions of all fishing members ---and often non-fishers, but community members--- were included.
-Participation of community members in reserve surveillance and yearly monitoring indicate commitment and interest, and allow informal communication of results to un-involved community members.
-
-Furthermore, the reserve is partially isolated from poaching activity and fishers have internal regulations pertaining the reserves.
-The low level of illegal fishing by members of the community and outsiders both inside and outside the reserve represents another indication of effectiveness.
-A summary of governance indicators is provided in Table 5.
+**Figure 5 Plots for values of each biological indicator (y-axis) through time (x-axis).** Red and blue correspond to the reserve and control sites, respectively.
+Black lines indicate yearly mean values, and ribbons indicate $\pm$ 1 standard error.
+Dots are horizontally jittered to aid visualization.
+This figure contains information for fish Shannon's diversity index (a), fish species richness (b), fish density (c), fish trophic level (d), fish biomass (e), invertebrate Shannon's diversity index (f), invertebrate species richness (g), invertebrate density (h), lobster density (i), urchin density (j), snail density (k), and cucumber density (l).
 
@@
 
-```{r table of bio results, results = "asis"}
-bio_results_summary %>% 
-  mutate(estimate = paste0(formatC(estimate, digits = 2, format = "f"),
-                           " (",
-                           formatC(std.error, digits = 2, format = "f"),
-                           ")",
-                           significance)) %>% 
-  select(Indicator, estimate, tscore = statistic) %>% 
-  kable(booktabs = T,
-        caption = "Summary of average treatment effect of the reserve on biological indicators. Asterisks indicate significance level, with (*) indicating p < 0.1 and (**) p < 0.05.",
-        digits = 2,
-        col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
-  kable_styling()
-```
-
-```{r table of soc results, results = "asis"}
-socio_results_summary %>% 
-  mutate(estimate = paste0(formatC(estimate, digits = 2, format = "f"),
-                           " (",
-                           formatC(std.error, digits = 2, format = "f"),
-                           ")",
-                           significance)) %>% 
-  select(Indicator, estimate, tscore = statistic) %>% 
-  kable(booktabs = T,
-        caption = "Summary of differences in socioeconomic indicators before and after the implementation of the reserve.
-Asterisks indicate significance level, with (*) indicating p < 0.1 and (**) p < 0.05.",
-        digits = 2,
-        col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
-  kable_styling()
-```
-
-\clearpage
-
-```{r plots, fig.height = 6, fig.width = 6, fig.cap = "Plots for values of each biological indicator (y-axis) through time (x-axis).
-Red and blue correspond to the reserve and control sites, respectively.
-Black lines indicate yearly mean values, and ribbons indicate +/- 1 standard error.
-Dots are horizontally jittered to aid visualization.
-This figure contains information for fish shannon diversity index (a), fish species richness (b), fish density (c), fish trophic level (d), fish biomass (e), invertebrate shannon diversity index (f), invertebrate species richness (g), invertebrate density (h), lobster density (i), urchin density (j), snail density (k), and cucumber density (l)."}
+```{r plots"}
 
@@
 
-gridExtra::grid.arrange(shannon_fish_plot, richness_fish_plot, density_fish_plot, trophic_fish_plot, biomass_fish_plot, shannon_invert_plot, richness_invert_plot, density_invert_plot, lobster_density_invert_plot, urchin_density_invert_plot, snail_density_invert_plot, cucumber_density_invert_plot, ncol = 3)
 ```
+
+```{r table of bio results, results = "asis"}
+bio_results_summary %>% 
+  mutate(estimate = paste0(formatC(estimate, digits = 2, format = "f"),
+                           " (",
+                           formatC(std.error, digits = 2, format = "f"),
+                           ")",
+                           significance)) %>% 
+  select(Indicator, estimate, tscore = statistic) %>% 
+  kable(caption = "Summary of average treatment effect of the reserve on biological indicators.",
+        digits = c(NA, NA, 4),
+        col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
+  kable_styling() %>% 
+  add_footnote(c("Asterisks indicate significance level, with (*) indicating p < 0.1 and (**) p < 0.05."),
+               notation = "symbol")
+```
+
+In the case of socioeconomic indicators, total landings were, on average, 64.20 metric tones higher (*p* > 0.1) after the implementation of the reserves.
+Total income was \$10,344.85 (*p* < 0.05) thousands of Mexican Pesos (K MXP) higher after the implementation of the reserves.
+On average, lobster and cucumber landings increased, while urchin and snail landings and income decreased.
+Figure 6 presents the changes in these indicators through time, and Table 4 summarizes these results.
+
+**Figure 6 Plots for values of each socioeconomic indicator (y-axis) through time (x-axis).** Red and blue correspond to before and after the implementation of the reserve, respectively.
+This figure contains information for total landings (a), total income (b), lobster landings (c), urchin landings (d), snail landings (e), cucumber landings (f), lobster income (g), urchin income (h), snail income (i), and cucumber income (j).
 
@@
 
-```{r time series socio, fig.height = 6, fig.width = 6, fig.cap = "Plots for values of each socioeconomic indicator (y-axis) through time (x-axis).
-Red and blue correspond to before and after the implementation of the reserve, respectively.
-This figure contains information for total landings (a), total income (b), lobster landings (c), urchin landings (d), snail landings (e), cucumber landings (f), lobster income (g), urchin income (h), snail income (i), and cucumber income (j)."}
+```{r time series socio, fig.height = 6}
 
@@
 income_plot <- socio$plot[[2]] %>% 
-  my_theme(y.lab = "Income (MXP)", text.x = 2001, text.y = 35000, text.label = "b", bio = F)
+  my_theme(y.lab = "Income (K MXP)", text.x = 2001, text.y = 35000, text.label = "b", bio = F)
 
@@
 income_lobster_plot <- socio$plot[[7]] %>% 
-  my_theme(y.lab = "Income (MXP)", text.x = 2001, text.y = 35000, text.label = "g", bio = F)
+  my_theme(y.lab = "Income (K MXP)", text.x = 2001, text.y = 35000, text.label = "g", bio = F)
 
 income_urchin_plot <- socio$plot[[8]] %>% 
-  my_theme(y.lab = "Income (MXP)", text.x = 2001, text.y = 10000, text.label = "h", bio = F)
+  my_theme(y.lab = "Income (K MXP)", text.x = 2001, text.y = 10000, text.label = "h", bio = F)
 
 income_snail_plot <- socio$plot[[9]] %>% 
-  my_theme(y.lab = "Income (MXP)", text.x = 2001, text.y = 2100, text.label = "i", bio = F)
+  my_theme(y.lab = "Income (K MXP)", text.x = 2001, text.y = 2100, text.label = "i", bio = F)
 
 income_cucumber_plot <- socio$plot[[10]] %>% 
-  my_theme(y.lab = "Income (MXP)", text.x = 2001, text.y = 1500, text.label = "j", bio = F)
+  my_theme(y.lab = "Income (K MXP)", text.x = 2001, text.y = 1500, text.label = "j", bio = F)
 
-gridExtra::grid.arrange(landings_plot, income_plot, landings_lobster_plot, landings_urchin_plot, landings_snail_plot, landings_cucumber_plot, income_lobster_plot, income_urchin_plot, income_snail_plot, income_cucumber_plot, ncol = 3)
+```
+
+```{r table of soc results, results = "asis"}
+socio_results_summary %>% 
+  mutate(estimate = paste0(formatC(estimate, digits = 2, format = "f"),
+                           " (",
+                           formatC(std.error, digits = 2, format = "f"),
+                           ")",
+                           significance)) %>% 
+  select(Indicator, estimate, tscore = statistic) %>% 
+  kable(caption = "Summary of differences in socioeconomic indicators before and after the implementation of the reserve.",
+        digits = c(NA, NA, 4),
+        col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
+  kable_styling() %>% 
+  add_footnote(c("Asterisks indicate significance level, with (*) indicating p < 0.1 and (**) p < 0.05."),
+               notation = "symbol")
+```
+
+In terms of governance, the community is strongly organized, which is likely a driver of their success.
+The first point of success is the existence of a fishing cooperative that is also affiliated to a regional fishing cooperatives federation.
+These polycentric governance structures allow various levels of organization that foster communication and cooperation [@mccay_2014-CN]; federations also provide bargain power with governments [@espinosaromero_2014-PY; @finkbeiner_2015-87].
+Access to fishing resources is managed through a TURF, permits, and fishing quotas (for some species).
+Their TURF promotes a sense of stewardship of their resources and incentivizes sustainable management [@mccay_2017-1m; @finkbeiner_2015-87].
+Together, these structures enabled a participative, bottom--up process during the reserve design phase; opinions of all fishing members ---and often non-fishers, but community members--- were included.
+Participation of community members in reserve surveillance and yearly monitoring indicate commitment and interest, and allow informal communication of results to un-involved community members.
+Furthermore, the reserve is partially isolated from poaching activity and fishers have internal regulations pertaining the reserves.
+The low level of illegal fishing by members of the community and outsiders both inside and outside the reserve represents another indication of effectiveness.
+Governance indicators are summarized in Table 5.
+
+```{r}
+read.csv("Tables/Governance_data.csv", col.names = c("Indicator", "Description"), stringsAsFactors = F) %>% 
+  kable(caption = "Summary of governance indicators.") %>%
+  kable_styling(full_width = F) %>% 
+  column_spec(2, width = "9cm")
 ```
@@
 
-```{r}
-read.csv("Tables/Governance_data.csv", col.names = c("Indicator", "Description"), stringsAsFactors = F) %>% 
-  kable(booktabs = T,
-        caption = "Summary of governance indicators.") %>%
-  kable_styling(full_width = F, latex_options = c("scale_down")) %>% 
-  column_spec(2, width = "14cm")
-```
+# Discussion
 
-\clearpage
+Here we focus on the advantages and room for improvement in MAREA and highlight its utility for evidence--based management.
+The findings from Isla Natividad are used purely to validate the relevance of MAREA rather than to discuss particularities of the marine reserve effectiveness, which has been described before [@micheli_2012-EU; @rossetto_2015-V0; @munguavega_2015-yg].
+We use examples from the case study to build on the utility of MAREA and discuss ways in which results can be interpreted to inform management.
 
-# Discussion and Conclusions
+The causal inference techniques used by MAREA have been suggested [@burgess_2016-HN; @ferraro_2006-oW] and used [@moland_2013-VP] before.
+This approach reduces ambiguity in the interpretation of results.
+For example, invertebrate density decreased through time inside and outside of the reserve (Fig.
+5h).
+In this case, a before--after evaluation the reserve (*i.e.* ignoring the control site) would have incorrectly concluded that the reserve wasn't protecting invertebrates.
+On the other hand, a control--impact approach (*i.e.* compare reserve vs.
+control site only in 2016) would have identified higher densities inside the reserve, concluding that the reserve increases invertebrate density.
+However, MAREA identifies the changes through time and across sites, and estimates the effect of the reserve on density at $\beta_2 = `r formatC(bio_results_summary$estimate[8], digits = 2, format = "f")`$ (*p* < 0.05).
+What really happens is that invertebrate densities decrease in both sites through time, but the decrease is faster for the control site, thus yielding a positive value for $\beta_2$.
 
-Here we discuss the ways in which results can be interpreted and
+The approach used by MAREA to estimate the effect of the reserve on biological indicators requires cautious interpretation of the results.
+The value of the $\beta_2$ coefficient represents the difference between the temporal trends of the reserve and control sites [@moland_2013-VP].
+As exemplified by the case of invertebrate densities, a positive value (*i.e.* $\beta_2 > 0$) does not necessarily indicate an increase in the indicator through time, but rather a positive difference with respect to the temporal trend of the control site.
+The inverse occurs for negative values of $\beta_2$.
 
-The interpretation of the $\beta_2$ coefficient is always relative to the control site.
-A positive value ($\beta_2 > 0$) might not nessesarily indicate an increase in the indicator.
-For example, invertebrate density presents a value of $\beta_2 = `r formatC(bio_results_summary$estimate[8], digits = 2, format = "f")`$ (*p* < 0.05), but Figure 5h shows a decrease in the indicator for the reserve and control sites through time.
-The decrease inside the reserve is not as abrupt as the changes in the control site, indicating that the reserves seem to provide a buffering effect.
-With additional knowledge of environmental variability (*i.e.* indicator B5 -- Natural Disturbance), we can better understand the performance of this indicator.
-As shown by @micheli_2012-EU, decreases in density can be caused by local hypoxic events.
+MAREA provides an overview of the effect of the reserve, allowing users to rapidly identify trends.
+However, users must interpret multiple indicators at a time to better understand the results.
+For example, with additional knowledge of local environmental variability (*i.e.* indicator B5 -- Natural Disturbance), we can better understand the trends in invertebrate densities.
+As reported before @micheli_2012-EU, hypoxic conditions that have occurred in Isla Natividad can cause decreases in invertebrate densities, and reserves buffer the negative effect.
+While MAREA automates the analyzes and makes results replicable, proper interpretation will still depend on the user.
+Results produced by MAREA can only aid in management and decision making when results have been correctly interpreted.
 
-The results identified by our framework and MAREA are consistent to those previously published [*e.g.* @micheli_2012-EU; @rossetto_2015-V0].
-While invertebrates show decreases in density for reserve and control sites, their densities remained higher within the reserve.
-Here, the positive effect indicated by $\beta_2$ does not indicate an increase in density, but rather a slower decrease in the reserve.
-This has occurred likely due to local hypoxic events [@micheli_2012-EU], where the reserves seem to provide a buffering effect.
+Socioeconomic and governance indicators typically lack a control site, which impede us from using causal inference techniques to measure their changes [@mascia_2017-m_].
+However, we can still extract useful information from them.
+Again, by combining results from multiple indicators, MAREA can provide insights into the effect of the reserve.
+For example, lobster and cucumber have shown increases in densities, landings, and income.
+We cannot conclude that landings and income from these species hve increased due to the reserve, but we can at least conclude that landings have not decreased.
+While further information on market behavior of each fishery is needed, these results provide insights into the state of the reserve and its associated fisheries.
 
----
+As for the governance information, it is difficult to establish causal links between the state of the reserve and the governance structures present in the community.
+However, providing a single platform (*i.e.* scorecard) or document (*i.e.* technical report) where biological, socioeconomic, and governance information is comprehensively included can aid in management.
+This information will be standardly reported across reserves, and can help managers identify overarching patterns across sites.
 
-While we are not able to infer causality from the socioeconomic analysis, we are able to identify important trends.
-For example, snail and urchin density have significantly increased within the reserve, but their landings and income have decreased.
-The opposite is observed for lobster and cucumber, which have shown increases in densities, landing, and income.
-While further information on market behavior of each species is needed, these results provide insights into the state of the reserve, as well as the associated fisheries.
-
------
-
-One of the greatest challenges in management measuring the extent to which objectives have been met.
-The present framework provides a simple and replicable way to align management objectives with performance indicators.
-We recognize that these 25 indicators might not fully describe a reserve.
-However, they provide a starting point to perform the evaluation, to which managers and users can add other indicators (*e.g.* larval dispersal or connectivity) that are relevant to their reserve.
-
-The proposed methodologies, especially the way in which biological indicators are evaluated, provide valuable information for managers.
-The analysis isolates the net effect of the reserve, providing a proper measure of reserve effectiveness.
-We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analyzed.
-Despite this, we believe that providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards evidence-based effective management.
-
-Furthermore, MAREA's value is that it provides a free, simple, and replicable way to perform rigorous impact analysis.
-The tool can easily be used by fishers, SCO members, and managers in government agencies, providing transparency of the analysis and results.
-In addition, it can empower and enable local managers and fishers to respond to local change and adapt by allowing direct and easy access to the information.
-The way in which results are presented allows this information to be interpreted by a wide-ranging audience.
+It is important that evidence--based management and conservation outcomes are communicated.
+While stakeholder involvement in the design and implementation phases of marine reserves is important, it is not enough.
+The results of such efforts must be communicated to stakeholders in order to maintain involvement and exchange of ideas [@lundquist_2005-OL; @leslie_2007-yJ].
+The way in which MAREA presents results allows this information to be interpreted by a wide-ranging audience.
 The scorecard is easily understandable by experts and non-experts, and can be used as an effective tool for communicating the results of annual evaluations.
@@
 
-While the first release of MAREA is now available, it will continue to be developed and maintained.
-This will incorporate new features, and enhance current ones, aiming to improve user experience and expand the scope of the analysis.
+We recognize that the 29 indicators used by MAREA might not fully describe a reserve.
+However, they provide a starting point to perform the evaluation, to which managers and users can add other indicators (*e.g.* larval dispersal or connectivity) that are relevant to their reserve.
+Furthermore, MAREA's value is that it provides a free, simple, and replicable way to perform rigorous impact analysis.
+The tool can easily be used by fishers, CSO members, and managers in government agencies, providing transparency of the analysis and results.
+In addition, it can empower and enable local managers and fishers to respond to local change and adapt by allowing direct and easy access to the information.
+
+The effectiveness of marine reserves continues to be a matter of debate [@woodcock_2017-Wm; @edgar_2014-UO].
+With current targets set to increase ocean protection, it is important that we understand the effects of our interventions [@burgess_2016-HN] so we can better inform management [@ferraro_2006-oW].
+It is therefore important that academics, managers, fishers, and CSOs have access to open access tools like MAREA.
+This is particularly relevant for Mexico and other Latin American countries, where management agencies are often shorthanded and underfunded [@lundquist_2005-OL].
+In this context, MAREA provides a simple and replicable way to align management objectives with performance indicators.
+The proposed methodologies, specially the way in which biological indicators are evaluated, provide valuable information for managers.
+We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analyzed.
+Despite this, providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards evidence--based effective management[@ferraro_2006-oW].
+
+While the first release of MAREA is now available, it will continue to be developed and maintained to keep up to date with literature.
+This process will incorporate new features, and enhance current ones, aiming to improve user experience and expand the scope of the analysis.
 Yet, we believe that this first release represents a major step towards effective evaluation and management of marine reserves.
@@
 Special thanks to the fishers from Isla Natividad, who gathered the data used in this study, and the fishers from El Rosario, who helped us validate our survey and framework, and to Arturo Hernández and Alfonso Romero who provided help with the logistics.
-Input provided by the participants in the workshop "" was very valuable.
 
@@
 
-- CF and JC received support from the Latin American Fisheries Fellowship Program.
-- JC received support from the Consejo Nacional de Ciencia y Tecnología (CONACyT).
-- GM - We would also like to thank the Waitt Foundation for their financial support. Carlos Slim Foundation Alliance.
-- The biological data for Isla Natividad were collected with the support of the Walton Family Foundation and the NSF-CNH program (grant DEB-1212124).
-- We also thank the support of: The Nature Conservancy and World Wildlife Fund-
+Funding must not be included in the manuscript.
+There is a special section in the submission that I will manually have to fill in.
+Please, review the following statements and let me know if I should add something else.
+
+- CF and JCVD received support from the Latin American Fisheries Fellowship Program.
+- JCVD received support from the Consejo Nacional de Ciencia y Tecnología (CONACyT).
+- GM received financial support from Waitt Foundation.
+- Biological data for Isla Natividad were collected with the support of the Walton Family Foundation and the NSF-CNH program (grant DEB-1212124).
+- The Nature Conservancy and World Wildlife Fund--Carlos Slim Foundation Alliance.
 
@@
 # References
+
+<div id="refs"></div>
+
+# Supporting information
+
+**S1 Appendix. Survey to collect governance information from fishing communities.** English version
+
+**S2 Appendix. Survey to collect governance information from fishing communities.** Spanish version
+
+**S3 Table. Assigned values and reasoning of socioeconomic and governance indicators used to color-code the scorecard in MAREA**
+
+# Figures
+
+![Workflow to evaluate the effectiveness of marine reserves.](./Figures/flowchart.gif){ width="50%"}
+
+```{r, fig.cap = "Legend used to interpret the scorecard produced by MAREA.
+Colors indicate direction of change (red = negative; green = positive), and color intensity is given by the statistical significance"}
+fig2
+```
+
+```{r, fig.cap = "General location of Isla Natividad (left) and map of the island (right).
+The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce)."}
+tmap_arrange(pen_map, study_area, outer.margins = 0.005, ncol = 2)
+```
+
+![Scorecard produced by MAREA for the "La Plana / Las Cuevas" marine reserve in Isla Natividad, Mexico.](./Figures/scorecard.png){ width=80% }
+
+```{r, fig.height = 6, fig.width = 6, fig.cap = "Plots for values of each biological indicator (y-axis) through time (x-axis).
+Red and blue correspond to the reserve and control sites, respectively.
+Black lines indicate yearly mean values, and ribbons indicate +/- 1 standard error.
+Dots are horizontally jittered to aid visualization.
+This figure contains information for fish shannon diversity index (a), fish species richness (b), fish density (c), fish trophic level (d), fish biomass (e), invertebrate shannon diversity index (f), invertebrate species richness (g), invertebrate density (h), lobster density (i), urchin density (j), snail density (k), and cucumber density (l)."}
+gridExtra::grid.arrange(shannon_fish_plot, richness_fish_plot, density_fish_plot, trophic_fish_plot, biomass_fish_plot, shannon_invert_plot, richness_invert_plot, density_invert_plot, lobster_density_invert_plot, urchin_density_invert_plot, snail_density_invert_plot, cucumber_density_invert_plot, ncol = 3)
+```
+
+```{r, fig.height = 6, fig.width = 6, fig.width = 6, fig.cap = "Plots for values of each socioeconomic indicator (y-axis) through time (x-axis).
+Red and blue correspond to before and after the implementation of the reserve, respectively.
+This figure contains information for total landings (a), total income (b), lobster landings (c), urchin landings (d), snail landings (e), cucumber landings (f), lobster income (g), urchin income (h), snail income (i), and cucumber income (j)."}
+gridExtra::grid.arrange(landings_plot, income_plot, landings_lobster_plot, landings_urchin_plot, landings_snail_plot, landings_cucumber_plot, income_lobster_plot, income_urchin_plot, income_snail_plot, income_cucumber_plot, ncol = 3)
+```

### 2017-08-22 Manuscript.Rmd
Added Giakoumi reference

@@
 Empirical studies have shown that MPAs increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience to climatic impacts [@micheli_2012-EU; @roberts_2017-J9], and preserve genetic diversity [@munguavega_2015-yg].
-Compared to partially protected MPAs, marine reserves are known to have even higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO].
+Compared to partially protected MPAs, marine reserves are known to have even higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO; @giakoumi_2017-V2].
 These effects are often measured as biological changes in the area through time and lack a control site against which to compare [@betti_2017-lq].

### 2017-08-30 Manuscript.Rmd
Incorporated Jorge's and Sean's comments

--- 2017-08-21 Manuscript.Rmd
@@
 Existing data are analyzed in different ways, hindering the ability to compare results across case studies.
-Moreover, analysis and evaluation of reserves is generally conducted by academic scientists, not the reserves managers and users, thereby hindering effective local management and rapid response to change.
+Moreover, analysis and evaluation of reserves is generally conducted by academic scientists, not the reserve managers and users, thereby hindering effective local management and rapid response to change.
 We present a framework and tool to evaluate the effectiveness of marine reserves by matching seven commonly stated management objectives to nine biological, five socioeconomic, and 15 governance indicators.
-We provide guidelines on how to properly collect data that can then be analyzed with standardized method.
+We provide guidelines on how to properly collect data that can then be analyzed with standardized methods.
 Biological indicators are evaluated with causal inference techniques, using a counterfactual approach, to assess the effect of the reserve on each indicator.
@@
 Governance indicators are qualitatively analyzed using a framework developed through a literature review, which identifies common governance structures and their associated effectiveness.
-To make the framework accessible to fishers and decision makers, and allow replication of results, we developed the open source, web-based Marine Reserve Evaluation App (MAREA).
+To make the framework accessible to fishers and decision makers, and allow replication of results, we developed the open source, web-based Marine Reserve Evaluation Application (MAREA).
 Together, this new framework and MAREA can further our understanding and support management of marine reserves.
@@
 Empirical studies have shown that MPAs increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience to climatic impacts [@micheli_2012-EU; @roberts_2017-J9], and preserve genetic diversity [@munguavega_2015-yg].
-Compared to partially protected MPAs, marine reserves are known to have even higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO].
-These effects are often measured as biological changes in the area through time and lack a control site against which to compare [@betti_2017-lq].
-This before--after comparison cannot account for other factors for which one must control [@davies_2017-ml] in order to attribute biological change to protection in the reserves.
-While some studies have used control sites, these analyses do not estimate the net effect of the reserve, and often use a control-impact comparison approach that does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
+Compared to partially protected MPAs, marine reserves have higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO; @giakoumi_2017-V2].
+These effects are often measured as biological changes in the area through time and lack a control site for comparison [@betti_2017-lq].
+This approach does not account for other factors for which one must control [@davies_2017-ml] in order to attribute biological change to protection in the reserves.
+While some studies have used control sites, these analyses do not estimate the effect of the reserve, and often use a control-impact comparison approach that does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
 A smaller fraction of studies have used a before-after-control-impact (*i.e.* BACI) design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.
@@
 This approach has been implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC].
-Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance [@anderson_2015-ND; @dowling_2016-pO], design territorial use rights for fisheries (TURFs; @oyanedel_2017-TO), and improve decision making in the hydro power industry [@vilela_2017-Zo], just to list a few.
+Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance [@anderson_2015-ND; @dowling_2016-pO], design territorial use rights for fisheries (TURFs; @oyanedel_2017-TO), and improve decision making in the hydro power industry [@vilela_2017-Zo].
 
@@
 We then include a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_].
-We also introduce the Marine Reserve Evaluation App (MAREA), an open source, web--based tool that automates the framework described in this paper and enables its broader use.
+We also introduce the Marine Reserve Evaluation Application (MAREA), an open source, web--based tool that automates the framework described in this paper and enables its broader use.
 Finally, we present a case study on the evaluation of a marine reserve established by the fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.
@@
 Fish biomass) or descriptive (*e.g.* Reasoning for reserve location).
-Our list includes indicators that respond to the implementation of the reserve (*i.e.* outcome variables) or that might further the understanding of its performance.
+Our list includes indicators that respond to the implementation of the reserve (*i.e.* outcome variables) or indicators that might further the understanding the reserve's performance.
 In that sense, most biological and socioeconomic indicators are outcome variables.
@@
 
-Socioeconomic data are often collected by fishers, fishery management agencies or CSOs by recording landings, income, and sometimes prices for each species.
+Socioeconomic data are often collected by fishers, natural resource management agencies, or CSOs by recording landings, income, and sometimes prices for each species.
 To control for inflation, income is adjusted with the country’s consumer price index [@oecd_website_2017-VV]:
@@
 This report also includes a scorecard as a summary of the results, but provides more information for each indicator.
-For all numeric indicators, the report includes a graph of the value of the indicatorin the reserve and control sites, through time.
+For all numeric indicators, the report includes a graph of the value of the indicator in the reserve and control sites, through time.
 It also provides a regression table that summarizes the value of all coefficients in the regression and their respective robust standard errors.
@@
 Local fishers ---trained in scientific diving by the NGO Comunidad y Biodiversidad, A.C.
-(COBI; [www.cobi.org](www.cobi.org))--- and personnel from COBI performed SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length.
+(COBI; [www.cobi.org](www.cobi.org), ReefCheck California, and Stanford University )--- and personnel from these institutions performed SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length.
 Information was recorded along 30 m transects, with a sampling window of 2 m X 2 m following a standardized ReefCheck protocol [@suman_2010-ez].
@@
 
-Socioeconomic data was obtained from the National Commission for Fisheries and Aquaculture (Comisión Nacional de Pesca y Acuacultura; CONAPESCA).
+Socioeconomic data was obtained from the National Commission for Aquaculture and Fisheries (Comisión Nacional de Acuacultura y Pesca; CONAPESCA).
 The data contains species-level (n = `r length(unique(landings$GeneroEspecie))+2` spp.) information on monthly landings and income from 2000 to 2014.
@@
 
-In terms of governance, the community is strongly organized, which is likely a driver of their success.
+In terms of governance, the community is strongly organized, which is likely a driver of their success [@gutirrez_2011-0U].
 The first point of success is the existence of a fishing cooperative that is also affiliated to a regional fishing cooperatives federation.
@@
 Their TURF promotes a sense of stewardship of their resources and incentivizes sustainable management [@mccay_2017-1m; @finkbeiner_2015-87].
-Together, these structures enabled a participative, bottom--up process during the reserve design phase; opinions of all fishing members ---and often non-fishers, but community members--- were included.
+Together, these structures enabled a participative, bottom--up process during the reserve design phase; opinions of all fishing members ---and often non--fishers, but community members--- were included.
 Participation of community members in reserve surveillance and yearly monitoring indicate commitment and interest, and allow informal communication of results to un-involved community members.
-Furthermore, the reserve is partially isolated from poaching activity and fishers have internal regulations pertaining the reserves.
+Furthermore, the reserve is partially isolated from poaching activity and fishers have internal regulations pertaining to the reserves.
 The low level of illegal fishing by members of the community and outsiders both inside and outside the reserve represents another indication of effectiveness.
@@
 In this context, MAREA provides a simple and replicable way to align management objectives with performance indicators.
-The proposed methodologies, specially the way in which biological indicators are evaluated, provide valuable information for managers.
+The proposed methodologies, especially the way in which biological indicators are evaluated, provide valuable information for managers.
 We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analyzed.
-Despite this, providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards evidence--based effective management[@ferraro_2006-oW].
+Despite this, providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards effective evidence--based management [@ferraro_2006-oW].
 
-While the first release of MAREA is now available, it will continue to be developed and maintained to keep up to date with literature.
+The first release of MAREA is now available, and it will continue to be developed and maintained to keep up to date with literature.
 This process will incorporate new features, and enhance current ones, aiming to improve user experience and expand the scope of the analysis.
@@
 - GM received financial support from Waitt Foundation.
-- Biological data for Isla Natividad were collected with the support of the Walton Family Foundation and the NSF-CNH program (grant DEB-1212124).
-- The Nature Conservancy and World Wildlife Fund--Carlos Slim Foundation Alliance.
+- Biological data for Isla Natividad were collected with the support of the Walton Family Foundation, the NSF-CNH program (grant DEB-1212124), Marisla Foundation, Sandler Family Supporting Foundation, and Packard Foundation.
+- AS received support from The Nature Conservancy and World Wildlife Fund--Carlos Slim Foundation Alliance
 

### 2017-10-10 Manuscript.Rmd
Changed title

@@
 ---
-title: \normalsize{A general approach and tool to evaluate the effectiveness of no-take marine reserves}
+title: \normalsize{A user-friendly tool to evaluate the effectiveness of no-take marine reserves}
 header-includes:

### 2017-10-10 Manuscript.Rmd
Updated abstract

@@
 
-Marine reserves are often implemented to preserve habitat, recover overfished stocks, and secure livelihoods of coastal communities.
-As with any other management interventions, we need to better understand their effectiveness and impacts on coastal communities and the environment.
-To date, their evaluation largely relies on analyzing ecological data, often ignoring socioeconomic and governance dimensions.
-Existing data are analyzed in different ways, hindering the ability to compare results across case studies.
-Moreover, analysis and evaluation of reserves is generally conducted by academic scientists, not the reserve managers and users, thereby hindering effective local management and rapid response to change.
-We present a framework and tool to evaluate the effectiveness of marine reserves by matching seven commonly stated management objectives to nine biological, five socioeconomic, and 15 governance indicators.
-We provide guidelines on how to properly collect data that can then be analyzed with standardized methods.
-Biological indicators are evaluated with causal inference techniques, using a counterfactual approach, to assess the effect of the reserve on each indicator.
-Linear regression models are fitted to socioeconomic indicators through time to test for differences before and after reserve implementation.
-Governance indicators are qualitatively analyzed using a framework developed through a literature review, which identifies common governance structures and their associated effectiveness.
-To make the framework accessible to fishers and decision makers, and allow replication of results, we developed the open source, web-based Marine Reserve Evaluation Application (MAREA).
-Together, this new framework and MAREA can further our understanding and support management of marine reserves.
+Marine reserves are implemented to achieve a variety of objectives, but are seldom rigorously evaluated to determine whether those objectives are met.
+In the rare cases when evaluations do take place, they typically focus on ecological indicators and ignore other relevant objectives such as socioeconomics and governance.
+And regardless of the objectives, the diversity of locations, monitoring protocols, and analysis approaches hinder the ability to compare results across case studies.
+Moreover, analysis and evaluation of reserves is generally conducted by outside researchers, not the reserve managers or users, plausibly thereby hindering effective local management and rapid response to change.
+We present a framework and tool, called “MAREA”,to overcome these challenges.
+Its purpose is to evaluate the extent to which any given reserve has achieved its stated objectives.
+MAREA provides specific guidance on data collection and formatting, and then conducts rigorous causal inference analysis based on data input by the user, providing real-time outputs about the effectiveness of the reserve.
+MAREA’s ease of use, standardization of state-of-the-art inference methods, and ability to analyze marine reserve effectiveness across ecological, socioeconomic, and governance objectives could dramatically further our understanding and support of effective marine reserve management.
 

### 2017-10-10 Manuscript.Rmd
Updated intro and methods

@@
 
-Overfishing and unsustainable fishing practices are two of the largest threats to the conservation of marine ecosystems around the world [@pauly_2005-qV; @halpern_2008-dK].
-Marine Protected Areas (MPAs) are frequently proposed as fishery management and conservation tools to help fish and invertebrate stocks recover [@lester_2008-F_; @lester_2009-Ks: @sala_2016-PV; @hastings_2017-sm] by limiting or restricting fishing effort and gears.
-No-take marine reserves (marine reserves hereinafter) are a particular type of MPA, where all fishing effort and extractive activities are off-limits [-@iucn_website_2017-Bg; @hortaecosta_2016-_3].
+Unsustainable fishing practices threaten biodiversity, conservation, economic and social outcomes [@pauly_2005-qV; @halpern_2008-dK].
+Marine Protected Areas (MPAs; and marine reserves, in which all extractive efforts are prohibited) are frequently proposed to aid in the recovery of fish and invertebrate stocks [@lester_2008-F_; @lester_2009-Ks; @sala_2016-PV; @hastings_2017-sm] by limiting or restricting fishing effort and gears.
 
-Empirical studies have shown that MPAs increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience to climatic impacts [@micheli_2012-EU; @roberts_2017-J9], and preserve genetic diversity [@munguavega_2015-yg].
-Compared to partially protected MPAs, marine reserves have higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO; @giakoumi_2017-V2].
-These effects are often measured as biological changes in the area through time and lack a control site for comparison [@betti_2017-lq].
-This approach does not account for other factors for which one must control [@davies_2017-ml] in order to attribute biological change to protection in the reserves.
+Available empirical evidence on marine reserve effectiveness is mixed [@padleton_2017-vn].
+Some studies have shown that MPAs increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience to climatic impacts [@micheli_2012-EU; @roberts_2017-J9], and preserve genetic diversity [@munguavega_2015-yg].
+Compared to partially protected MPAs, marine reserves have higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO; @giakoumi_2017-V2; @sala_2017-69].
+These effects are often measured as biological changes in the area through time and many lack a control site for comparison [@betti_2017-lq].
+This approach does not account for other factors for which one must control [@davies_2017-ml] in order to causally attribute a biological change to the reserve.
 While some studies have used control sites, these analyses do not estimate the effect of the reserve, and often use a control-impact comparison approach that does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
@@
 
-As with any other policy intervention, it is important that we can measure its effect in order to adapt and learn [@ferraro_2006-oW; @burgess_2016-HN].
-The diversity of approaches currently used to evaluate the effect of marine reserves often does not answer the simple question: *What is the effect of a marine reserve on a given attribute?* This gap highlights the need to develop standardized approaches that enable us to evaluate the net effect of the intervention (*i.e.* causes of conservation outcomes; @ferraro_2006-oW).
-Furthermore, while biological aspects are important to reserve success, effectiveness also depends on the socioeconomic status and governance system of the local fishing communities [@basurto_2013-oq; @basurto_2012-Mz], which are often ignored.
-By excluding these important dimensions, the evaluation provides only a partial picture of the impacts of the reserve.
+But even when proper causal inference can be drawn, results are often idiosyncratic across reserves.
+Effects of reserves on ecological and economic outcomes are highly heterogeneous, and often depend on the specific ecological, economic, and social context.
+The purpose of this paper is to describe a user-friendly tool, called “MAREA”, to rigorously systematize the evaluation of marine reserve effectiveness.
+The tool is in the form of an open-source application that uses state-of-the-art methods from program evaluation to compare a reserve to control sites along a number of ecological, economic, and governance dimensions.
 
-Currently, only the IUCN framework "How to evaluate your MPA" [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biological, socioeconomic, and governance indicators, and insights into how these indicators may be measured, but does not provide guidelines on how to analyze them.
-Recent work by @mascia_2017-m_ integrates these three dimensions and suggests the use of causal inference techniques to provide a measure of the effect of implementing an MPA.
+The challenge of how to standardize marine reserve evaluation is not a new one.
+The recent IUCN framework “How is your MPA doing?” [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biological, socioeconomic, and governance indicators, and insights into how these indicators may be measured.
+But this framework stops short of analysis, so provides a user with little guidance about establishing causal inference about the reserve.
+Recent work by Mascia *et al.* [@mascia_2017-m_] integrates these three dimensions via the Social Ecological Systems Framework [@ostrom_2009-hg; @basurto_2013-oq] and suggests the use of causal inference techniques to provide a measure of the effect of conservation interventions.
 However, these two novel approaches do not provide a user-friendly tool that enables replicability and scalability of the analysis, particularly when used by the fishers and decision makers themselves.
@@
 An increasingly popular way to make science accessible, reproducible, scalable, and replicable is through Open Science and the development of open-access tools [@lowndes_2017-xh].
-The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, has successfully standardized a way to measure the health and benefits of the oceans.
-This approach has been implemented at global scales, but also at country-level [@selig_2015-F9], and regional scales [@halpern_2014-lQ; @elfes_2014-RC].
-Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance [@anderson_2015-ND; @dowling_2016-pO], design territorial use rights for fisheries (TURFs; @oyanedel_2017-TO), and improve decision making in the hydro power industry [@vilela_2017-Zo].
+The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, successfully standardized a way to measure the health and benefits of the oceans.
+This approach has been implemented at a global scale, but also at country-level [@selig_2015-F9], and regionally [@halpern_2014-lQ; @elfes_2014-RC].
+Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance [@anderson_2015-ND; @dowling_2016-pO], design territorial use rights for fisheries [TURFs; @oyanedel_2017-TO], and improve decision making in the hydro power industry [@vilela_2017-Zo].
 
-The lack of a comprehensive framework and user-friendly tools to evaluate the effectiveness of marine reserves ---or the complexity of existing ones, which alienate non--experts--- calls for the development of a new framework and tool.
-The current work presents a framework to evaluate marine reserves, which incorporates the biological, socioeconomic, and governance dimensions of these areas.
+This paper presents a framework and user-friendly tool to evaluate marine reserve effectiveness, which incorporates the biological, socioeconomic, and governance dimensions of any fishery.
 We first provide a list of commonly stated management objectives and match them to appropriate indicators.
-We then include a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_].
-We also introduce the Marine Reserve Evaluation Application (MAREA), an open source, web--based tool that automates the framework described in this paper and enables its broader use.
+We then develop a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_].
+To implement the analytical approach in a user-friendly format, we introduce the Marine Reserve Evaluation Application (MAREA), an open source, web–based tool that automates the framework described in this paper and enables its broader use.
 Finally, we present a case study on the evaluation of a marine reserve established by the fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.
@@
 Alongside, methodologies to analyze these indicators are presented.
-Then, we describe the development of MAREA and explain how this user-friendly open access tool can be used by anyone.
+We then describe the development of MAREA and explain how this user-friendly open access tool can be used by fishermen, managers, and other stakeholders with little scientific background.
 Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA to inform management.
@@
 
-Throughout this study, we will refer to the stated goals for which a marine reserve was designed as objectives.
-This work was motivated by the need to provide a framework to evaluate Mexican marine reserves.
+Throughout this study, we will refer to the stated goals for which a marine reserve was designed as “objectives.” This work was motivated by the case of Mexico, where 39 reserves have been implemented over the past five years to achieve objectives such as increasing productivity in nearby waters or recover overexploited species; these reserves have never been formally evaluated for effectiveness at meeting those objectives.
 Thus, our focus was on identifying common objectives of marine reserves in Mexico.
-However, we group these objectives into seven major categories, which can be applied to marine reserves worldwide.
-The list of objectives was developed through a literature review, which compiled stated objectives in legislation [@nom049sagpesc_2014-V6; @lgeepa_2017-jL] and official documents such as the Technical Justification Studies (*Estudios Tecnicos Justificativos*), agreements, and decrees associated to these areas.
-Even though each reserve has its own goals, seven main categories of objectives were identified:
+However, a literature review and discussions with marine reserve researchers worldwide suggest that the objectives driving Mexican marine reserve implementation are similar to those in the rest of the world.
+Thus, we group these objectives into seven major categories, which can be applied to marine reserves worldwide.
+The list of objectives includes stated objectives in legislation [@nom049sagpesc_2014-V6; @lgeepa_2017-jL] and official documents such as the Technical Justification Studies (*Estudios Técnicos Justificativos*), agreements, and decrees associated to these areas:
 
@@
 3.	Maintain biological processes (reproduction, recruitment, growth, feeding)
-4.	Improve fishery production in nearby waters
+4.	Improve fishery production in adjacent waters
 5.	Preserve biological diversity and the ecosystem
@@
 Based on these seven objectives, we determined a set of associated indicators to evaluate reserve effectiveness.
+These indicators are specific variables on which data could be collected, and analyzed, to ultimately determine whether the corresponding objective was causally being achieved by the marine reserve.
 The list of indicators was compiled through a review of scientific literature in which we identified indicators that were used to measure similar objectives.
@@
 Later, these were presented to fishers from the Ensenada Fishing Cooperative (*S.C.P.P.
-Ensenada*), in Baja California, who provided input.
-Our final list of indicators includes those identified in review works such as @lester_2009-Ks or @woodcock_2017-Wm.
+Ensenada*), in El Rosario, Baja California, who provided input.
+Our final list of indicators includes those identified in review works such as Lester *et al.* @lester_2009-Ks or Woodcock *et al.* @woodcock_2017-Wm.
 
 Indicators are divided into three main categories: biological, socioeconomic, and governance (Table 1).
-Biological indicators (n = 9) focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data and Analysis section for specific sampling design and methodologies).
-Socioeconomic indicators (n = 5) reflect the performance of the fishery in terms of landings, income from landings, and availability of alternative livelihoods.
-Governance indicators (n = 15) describe the governance structures under which the community operates (*e.g.*, access rights to the fishery, number of fishers, legal recognition of the reserve).
-Indicators may be numeric (e.g.
-Fish biomass) or descriptive (*e.g.* Reasoning for reserve location).
-Our list includes indicators that respond to the implementation of the reserve (*i.e.* outcome variables) or indicators that might further the understanding the reserve's performance.
+The nine biological indicators focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data and Analysis section for specific sampling design and methodologies).
+Five socioeconomic indicators reflect the performance of the fishery in terms of landings, income from landings, and availability of alternative livelihoods.
+Fifteen governance indicators describe the governance structures under which the community operates (*e.g.*, access rights to the fishery, number of fishers, legal recognition of the reserve).
+Some indicators require a numerical entry (*e.g.* Fish biomass) while others are more descriptive (*e.g.* Reasoning for reserve location).
+Many of the indicators specifically measure an outcome of the reserve, though some are designed to further the understanding of the mechanisms driving a reserve’s performance.
 In that sense, most biological and socioeconomic indicators are outcome variables.
@@
 Whenever an indicator is applied to “Target species”, it means that the indicator can be used for all species (*e.g.* Fish Biomass) and/or for individual species that are either the conservation target  of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass).
-Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with biological and socioeconomic indicators; all governance indicators should always be used.
+Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with biological and socioeconomic indicators.
 
@@
   group_rows("Governance", 11, 25) %>% 
-  add_footnote(c("The indicator is applied to target species"), notation = "symbol")
+  add_footnote(c("The indicator is also applied to target species"), notation = "symbol")
 ```
@@
   select(1:15) %>% 
-  knitr::kable(caption = "Management objectives and respective performance indicators.",
+  knitr::kable(caption = "Management objectives and respective performance indicators. All governance indicators should allways be used.",
                col.names = c("Objective", "B1", "B2", "B3", "B4", "B4*", "B5", "B6", "B7", "B7*", "S1", "S1*", "S2", "S2*", "S3")) %>% 
@@
 
-In many coastal marine reserves of Mexico, biological data are collected via underwater ecological surveys as part of a reserve's monitoring program, often carried out by local fishers with guidance from Civil Society Organizations (CSOs).
-Scientific divers record fish and invertebrate richness and abundances, as well as fish total length along belt transects.
-Ecological surveys are performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, allowing us to have a BACI sampling design.
-Control sites are areas where habitat is similar to that of the reserve, but with presence of fishing activity.
-While transect dimensions (*i.e.* length and width) and sampling methods might vary from study to study, the general idea remains the same: richness, abundances, and sizes of organisms are recorded in a study--specific standardized way.
+In many coastal marine reserves of Mexico, biological data are collected via underwater visual censuses as part of a reserve’s monitoring program.
+Scientific divers (which are often local fishermen with guidance from Civil Society Organizations; CSOs) record fish and invertebrate richness and abundances, as well as fish total length along belt transects.
+Ecological surveys are typically performed annually in each reserve and corresponding control site(s), before and after the implementation of the reserve, providing a sampling design that can be used to draw causal inference.
+Control sites are areas where habitat is similar to that of the reserve, but with presence of fishing activity; in principle these are areas that are otherwise observationally identical to the reserve site, but where, for presumably random reasons, a reserve was not implemented.
+While transect dimensions (*i.e.* length and width) and sampling methods might vary from study to study, the general idea remains the same: richness, abundances, and sizes of organisms are recorded in a study–specific standardized way.
 For this reason, MAREA does not assume specific transect dimensions, and pertinent indicators are calculated per transect (Table 1).
 
-Having a BACI sampling design for biological data allows us to use causal inference techniques [@moland_2013-VP; @ferraro_2006-oW] to evaluate the effect of the reserve on biological indicators.
+This sampling design for biological data allows us to use causal inference techniques [@moland_2013-VP; @ferraro_2006-oW] to evaluate the effect of the reserve on biological indicators.
 The hypothesis that the indicators will respond to implementation of the reserve is tested by analyzing spatial and temporal changes in each numeric biological indicator (all but B5) using generalized linear models [@moland_2013-VP].
@@
 
-In this model, $i$, $t$, and $z$ are subindices for transect, time, and zone (control or reserve site), respectively.
+In this model, $i$, $t$, and $z$ are indices for transect, time, and zone (control or reserve site), respectively.
 This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), a dummy variable that indicates treatment ($Z$; *i.e.* control or reserve), an interaction between a dummy variable that indicates pre-- or post--implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$; in °C), horizontal visibility during the survey ($V$; in m), and depth at which survey was performed ($D$; in m).
@@
 Where $I_t$ represents the adjusted income for year $t$ as the product between the reported income for that year and the ratio between the consumer price index ($CPI$) in that year to the most recent year’s ($T$) CPI.
-Since no control sites are typically available for this data type, numeric socioeconomic indicators (G1 and G2) are evaluated with a simplified version of eq.
+Since no control sites are typically available for this data type, numeric socioeconomic indicators (S1 and S2) are evaluated with a simplified version of eq.
 1:
@@
 
-While this model does not allow establishing a causal relation, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input.
+While this model does not formally allow for causal inference, we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input.
 For both models (eq.
 1 and eq.
-3), coefficients are estimated via ordinary least squares, and heteroskedastic--robust standard errors are calculated.
+3), we estimate the model coeficients with ordinary least squares, and calculate heteroskedastic–robust standard errors.
 
-Governance data are not readily available nor systematically collected by the community or other organizations.
+While biological and some economic data are regularly collected, governance data are typically not available nor systematically collected by the community or other organizations.
 Therefore, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1-G15).
@@
 This approach has been proven to successfully evaluate governance structures [@espinosaromero_2014-PY].
-Governance information is not quantitatively analyzed, but it is presented along with the biological and socioeconomic indicators to provide managers and users with a more complete description of the reserve.
+Unlike with biological and socioeconomic objectives (see eqs 1 and 3), MAREA does not quantitatively analyze governance information.
+Rather, it is presented along with the biological and socioeconomic indicators to provide managers and users with a more complete description of the reserve.
 
@@
 
-MAREA was developed in R version 3.4.1 and R Studio [@rcore_2017] using the Shiny package [@shiny_2017], which provides the tools to build interactive web applications, hosted on an open server.
-MAREA can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/).
+We developed MAREAin R version 3.4.2 and R Studio 1.1.383 [@rcore_2017] using the Shiny package [@shiny_2017], to build an interactive web application hosted on an open server; the MAREA app can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/).
 While the original version was developed in Spanish because it was aimed for Mexico and other Latin-American countries, all of its content can be translated by a translation widget available within the app.

### 2017-10-10 Manuscript.Rmd
Updated manuscript

@@
 ---
-title: \normalsize{A user-friendly tool to evaluate the effectiveness of no-take marine reserves}
+title: A user-friendly tool to evaluate the effectiveness of no--take marine reserves
 header-includes:
@@
 
-Juan Carlos Villaseñor-Derbez^1\*^, Caio Faro^1^, Melaina Wright^1^, Jael Martínez^1^, Sean Fitzgerald^1^, Stuart Fulton^2^, Maria del Mar Mancha-Cisneros^3^, Gavin McDonald^1,4,5^, Fiorenza Micheli^6^, Alvin Suárez^2^, Jorge Torre^2^, Christopher Costello^1,4,5^
+Juan Carlos Villaseñor-Derbez^1¶\*^, Caio Faro^1¶^, Melaina Wright^1¶^, Jael Martínez^1¶^, Sean Fitzgerald^1&^, Stuart Fulton^2&^, Maria del Mar Mancha-Cisneros^3&^, Gavin McDonald^1,4,5&^, Fiorenza Micheli^6&^, Alvin Suárez^2&^, Jorge Torre^2&^, Christopher Costello^1,4,5¶^
 
@@
 
-Email: jvillasenor@bren.ucsb.edu (JCVD) \
+Email: jvillasenor@bren.ucsb.edu (JCVD)
+
+¶ These authors contributed equally to this work.
+
+& These authors also contributed equally to this work.
 
@@
 
-**Fig. 1. Workflow to evaluate the effectiveness of marine reserves.**
+**Fig 1. Workflow to evaluate the effectiveness of marine reserves.**
 
@@
 
-\begin{equation}I_{i,t,z}=\beta_0 + \sum_{t = 2}^T\gamma_{i,t}Y_t + \beta_1Z_{i,z} + \beta_2P_{i,t,z}\times Z_{i,t,z} + \beta_3T_{i,t,z} + \beta_4V_{i,t,z} + \beta_5D_{i,t,z} + \epsilon_{i,t,z}\end{equation}
+$$I_{i,t,z}=\beta_0 + \sum_{t = 2}^T\gamma_{i,t}Y_t + \beta_1Z_{i,z} + \beta_2P_{i,t,z}\times Z_{i,t,z} + \beta_3T_{i,t,z} + \beta_4V_{i,t,z} + \beta_5D_{i,t,z} + \epsilon_{i,t,z}$$
 
@@
 
-\begin{equation}I_t = RI\times \frac{CPI_t}{CPI_T}\end{equation}
+$$I_t = RI\times \frac{CPI_t}{CPI_T}$$
 
@@
 
-\begin{equation}I_{t}=\beta_0 + \beta_1P_{t} + \epsilon_{t}\end{equation}
+$$I_{t}=\beta_0 + \beta_1P_{t} + \epsilon_{t}$$
 
@@
 
-MAREA is designed as a 6-step process, divided in tabs.
+MAREA is designed as a 6-step process, divided in tabs that appear upon launching the app.
 The first tab introduces the app and summarizes the evaluation process.
@@
 Users can also manually modify selected indicators based on their interests and data availability.
-The user can then load the data, using standard \*.csv text files; sample datasets are provided within MAREA.
-Once data have been loaded, MAREA identifies all reserves in the data (the uploaded dataset can contain data for more than one reserve), and lets the user select the reserve to be evaluated.
+The user can then load data on one or more reserves, using standard \*.csv text files; sample datasets are provided within MAREA.
+Once data have been loaded, MAREA identifies all reserves in the data, and lets the user select the reserve to be evaluated.
 At this point, the user can also specify the year of implementation of the reserve, reserve dimensions, and indicate target species that are of particular management interest.
-Before presenting the results, MAREA provides the user with a section to confirm that all the decisions made leading up to that point are correct.
-Finally, the user is taken to the results tab where they can be viewed in a simple format.
+MAREA provides the user with a section to confirm that all the decisions made leading up to that point are correct.
+Once the user has confirmed all input data, objectives, and other information, MAREA performs the formal program evaluation analyses discussed above.
+For a typical data set, the automated analysis step takes less than one second --but is dependent on connection speed.
+Finally, the user is taken to the results tab where all results are presented in a simple format.
 The user can also download a more comprehensive technical report produced in \*.pdf  format.
 
-The first output is a color--coded scorecard intended to provide a general overview of the effectiveness of the reserve.
+The first output is a color–coded scorecard intended to provide a general overview of the effectiveness of the reserve.
 The scorecard provides a global score for the reserve, a general score for each category of indicators, and an individual score for each indicator.
-The global and category--level scores are determined by the percentage of positive indicators, overall and for each category, respectively.
+The global and category–level scores are determined by the percentage of positive indicators, overall and for each category, respectively.
 For numeric biological indicators (all but B5), the color is defined by the sign of the interaction term coefficient ($\beta_2$) in eq.
@@
 For socioeconomic indicators, colors are assigned based on the direction of the slope ($\beta_1$).
-Red, yellow and green are used for $\beta_i<0$, $\beta_i = 0$, and $\beta_i>0$, respectively.
-The intensity of the color is defined by the significance of the coefficient, testing the null hypothesis of no change (*i.e.* $H_0: \beta_i = 0$) with a Student's t-test.
+Red, yellow, and green are used for $\beta_i<0$, $\beta_i = 0$, and $\beta_i>0$, respectively.
+The intensity of the color is defined by the significance of the coefficient, testing the null hypothesis of no change (*i.e.* $H_0: \beta_i = 0$) with a Student’s t-test.
 Cutoff values are $p < 0.05$ and $p < 0.1$.
-Thus, even in a case where $\beta_i > 0$, if the coefficient is not significant (*i.e.
+Thus, even in a case where $\beta_i > 0$, if the coefficient is not significant by standard measures (*i.e.
 * $p>0.1$), the indicator will be assigned a yellow color.
@@
 The color is defined based on what literature shows to be a negative (red) or positive (green) factor for a reserve (S3 Table).
-For example, if the perceived degree of illegal fishing is high, this is indicator will be assigned a red color.
+For example, if the perceived degree of illegal fishing is high, this indicator will be assigned a red color.
 However, due to the nature of some governance indicators, which require the user to provide a narrative, only some indicators are presented in the scorecard (although all are included in the technical report).
 
-**Fig.
-2.
+**Fig 2.
 Legend used to interpret the scorecard produced by MAREA.** Colors indicate direction of change (red = negative; green = positive), and color intensity is given by the statistical significance.
@@
 This report also includes a scorecard as a summary of the results, but provides more information for each indicator.
-For all numeric indicators, the report includes a graph of the value of the indicator in the reserve and control sites, through time.
+For all numeric indicators, the report includes a graph of the value of the indicator in the reserve and control sites through time.
 It also provides a regression table that summarizes the value of all coefficients in the regression and their respective robust standard errors.
@@
 The technical report is produced by a parameterized Rmarkdown document [@rmarkdown_2017] processed by the knitr package [@knitr_2017].
-Another feature of MAREA is that the user can choose to share the data.
+Another feature of MAREA is that the user can choose to share the data; the user can also choose not to share the data.
 Once the technical report is downloaded, the information on the reserve, its management objectives, and all uploaded data are saved into a central repository.
@@
 
-We apply this analytical framework and open access tool to evaluate the effectiveness of one marine reserve from Isla Natividad, in Baja California Sur, Mexico.
-Isla Natividad is located 8 Km off the Pacific Coast of the Baja Peninsula (Fig.
-3), where fishers operate under a fishing cooperative (S.C.P.P.
-Buzos y Pescadores de la Baja California) that promotes co-management of marine resources [@mccay_2017-1m; @mccay_2014-CN].
-Additionally, fishers have Territorial Use Rights for Fisheries (TURFs) that provide them with exclusive access rights to exploit the marine resources within a given perimeter [@mccay_2014-CN].
+While MAREA is a general tool that can be easily employed to evaluate the effectiveness of any marine reserve with the required input data, we illustrate its use here by applying it to one marine reserve near Isla Natividad, in Baja California Sur, Mexico.
+Isla Natividad is located 8 Km off the Pacific Coast of the Baja California Peninsula (Fig.
+3), where fishers operate under a fishing cooperative (*S.C.P.P.
+Buzos y Pescadores de la Baja California*) that promotes co-management of marine resources [@mccay_2017-1m; @mccay_2014-CN].
+Additionally, fishers have Territorial Use Rights for Fisheries (TURFs) that provide them with exclusive access rights to exploit the benthic marine resources within a given perimeter [@mccay_2014-CN].
 
-**Fig.
-3.
+**Fig 3.
 General location of Isla Natividad (left) and map of the island (right).** The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce).
@@
 
-In 2006, the community implemented two community-based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh].
-These reserves have proven to be effective in enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of highly valuable commercial species such as abalone [@munguavega_2015-yg].
+In 2006, the Isla Natividad community implemented two community-based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh; @micheli_2012-EU].
+Evidence suggest that these reserves have been effective at enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of high value commercial species such as abalone [@munguavega_2015-yg].
 These ecological benefits have been translated into economic benefits, enhancing population persistence and bolstering abalone fisheries [@rossetto_2015-V0].
-For the purpose of this evaluation, we focused on the “La Plana / Las Cuevas” marine reserve, located in the southern end of the Island (Fig.
+For the purpose of this evaluation, we focused on the “La Plana / Las Cuevas” marine reserve, located at the southern end of the island (Fig.
 3) and its corresponding control site “La Dulce / Babencho”.
 
-The reserve was implemented to recover species of economic interest --–which were overexploited--- and to enhance fishery production in nearby waters.
+The objective of this reserve was to recover species of economic interest -—-which were overexploited--- and to enhance fishery production in nearby waters.
 Fishers were also interested in preserving biological diversity and the ecosystem.
-Thus, objectives 4--7 were selected.
+Thus, objectives 4–-7 were selected.
 Using Table 2 to match these objectives with appropriate management indicators, we selected all biological, socioeconomic, and governance indicators included as options in the framework.
@@
 
-Local fishers ---trained in scientific diving by the NGO Comunidad y Biodiversidad, A.C.
-(COBI; [www.cobi.org](www.cobi.org), ReefCheck California, and Stanford University )--- and personnel from these institutions performed SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length.
-Information was recorded along 30 m transects, with a sampling window of 2 m X 2 m following a standardized ReefCheck protocol [@suman_2010-ez].
-Ecological surveys are performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, allowing us to have a BACI sampling design.
-Yearly surveys (2006 -– 2016) were carried out in late July -- early August, performing a total of `r sum(fish_summary$Reserva)` and `r sum(invert_summary$Reserva)` transects in the reserve site for fish and invertebrate surveys, respectively.
+Local fishers (who were trained in scientific diving by the CSO Comunidad y Biodiversidad, A.C.
+(COBI; [www.cobi.org](www.cobi.org)), ReefCheck California, and Stanford University) and personnel from these institutions performed SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length.
+They recorded information along 30 m transects, with a sampling window of 2 m x 2 m following a standardized ReefCheck protocol [@suman_2010-ez].
+Ecological surveys were performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, providing the requisite time series data inside the reserve and for a suitable control site.
+Annual surveys (2006 -- 2016) were carried out in late July -- early August, performing a total of `r sum(fish_summary$Reserva)` and `r sum(invert_summary$Reserva)` transects in the reserve site for fish and invertebrate surveys, respectively.
 Similar sampling effort was applied to the control site, with `r sum(fish_summary$Control)` fish and `r sum(invert_summary$Control)` invertebrate transects.
@@
 
-Socioeconomic data was obtained from the National Commission for Aquaculture and Fisheries (Comisión Nacional de Acuacultura y Pesca; CONAPESCA).
-The data contains species-level (n = `r length(unique(landings$GeneroEspecie))+2` spp.) information on monthly landings and income from 2000 to 2014.
-Data on landings and income was aggregated by year, and the yearly Consumer Price Index [@oecd_website_2017-VV] was included.
+Socioeconomic data were obtained from the National Commission for Aquaculture and Fisheries (*Comisión Nacional de Acuacultura y Pesca*; CONAPESCA).
+The data contains species-level information on monthly landings and income from nine species from 2000 to 2014.
+Data on landings and income were aggregated by year, and adjusted by the Consumer Price Index [@oecd_website_2017-VV].
 From the nine species available, we selected as objective species those that contributed the most (88.27%) income from 2000 to 2014: lobster (*Panulirus interruptus*; 71.76%), red sea urchin (*Mesocentrotus franciscanus*; 9.33%), snail (*Megastraea undosa*; 3.93%), and sea cucumber (*Parastichopus parvimensis*; 3.23%).
-Abalone species (*Haliotis fulgens*; 4.52% and *Haliotis corrugata*; 6.16%) were excluded because the cooperative has implemented an informal closure of these fisheries to allow its populations to recover since 2010.
+Abalone species (*Haliotis fulgens*; 4.52% and *Haliotis corrugata*; 6.16%) were excluded because the cooperative implemented an informal closure of these fisheries in 2010 to allow the population recover.
 Eliminating all fishing pressure on abalones means that the control site receives (for this species) the same treatment as the reserve.
 
-For governance data, we constructed the database based on our knowledge of the area and the community.
+We constructed the governance data based on local knowledge of the area and the community.
 
@@
 
-# Results
+# Results from illustrative example
 
-The results shown here intend to highlight the relevance and utility of the framework and tool, which automate the analysis and make it replicable.
-While we highlight some of the general observed trends, we focus on the utility of the tool rather than on the case study.
+In this section we show the results of the application of MAREA to the La Plana/Las Cuevas marine reserve in Isla Natividad, Mexico.
+These results are intended to highlight the relevance and utility of the MAREA framework and app, which automate the analysis and make it replicable.
+While we highlight some of the general observed trends, we focus on the utility of the tool rather than on the specific effectiveness of this case study marine reserve.
 
 The scorecard (Fig.
-4) shows the reserve achieves a general score of 63.6% positive indicators.
-All category--level scores were also high, with values of 66.7%, 60%, and 75% positive indicators for biological, socioeconomic and governance, respectively.
+4) shows that this reserve achieves a general score of 64%, suggesting  that  64% of indicators are positive.
+All category–level scores were also high, with values of 67%, 60%, and 75% positive indicators for biological, socioeconomic and governance, respectively.
 
-**Fig. 4. Scorecard produced by MAREA for the "La Plana / Las Cuevas" marine reserve in Isla Natividad, Mexico.**
+**Fig 4. Scorecard produced by MAREA for the "La Plana / Las Cuevas" marine reserve in Isla Natividad, Mexico.**
 
@@
 
-Among the biological indicators, the greatest effect of the reserve was observed for snail and cucumber densities, with values of $\beta_2 = `r formatC(bio_results_summary$estimate[11], digits = 2, format = "f")`$ (*p* < 0.05) and $\beta_2 = `r formatC(bio_results_summary$estimate[12], digits = 2, format = "f")`$ (*p* < 0.05), respectively.
+Among the biological indicators, the greatest effect of the reserve was observed for snail and sea cucumber densities, with values of $\beta_2 = `r formatC(bio_results_summary$estimate[11], digits = 2, format = "f")`$ (*p* < 0.05) and $\beta_2 = `r formatC(bio_results_summary$estimate[12], digits = 2, format = "f")`$ (*p* < 0.05), respectively.
 Fish indicators showed no significant change (*p* > 0.1), with negative trends for Shannon's diversity index and fish species richness and positive trends for density, biomass, and mean trophic level.
@@
 
-**Figure 5 Plots for values of each biological indicator (y-axis) through time (x-axis).** Red and blue correspond to the reserve and control sites, respectively.
+**Fig.
+5 Plots for values of each biological indicator (y-axis) through time (x-axis).** Red and blue correspond to the reserve and control sites, respectively.
 Black lines indicate yearly mean values, and ribbons indicate $\pm$ 1 standard error.
 Dots are horizontally jittered to aid visualization.
-This figure contains information for fish Shannon's diversity index (a), fish species richness (b), fish density (c), fish trophic level (d), fish biomass (e), invertebrate Shannon's diversity index (f), invertebrate species richness (g), invertebrate density (h), lobster density (i), urchin density (j), snail density (k), and cucumber density (l).
+This figure contains information for fish Shannon's diversity index (a), fish species richness (b), fish density (c), fish trophic level (d), fish biomass (e), invertebrate Shannon's diversity index (f), invertebrate species richness (g), invertebrate density (h), lobster density (i), urchin density (j), snail density (k), and sea cucumber density (l).
 
@@
 
-In the case of socioeconomic indicators, total landings were, on average, 64.20 metric tones higher (*p* > 0.1) after the implementation of the reserves.
-Total income was \$10,344.85 (*p* < 0.05) thousands of Mexican Pesos (K MXP) higher after the implementation of the reserves.
-On average, lobster and cucumber landings increased, while urchin and snail landings and income decreased.
+One of the main objectives of this reserve was to increase landings.
+Results of the socioeconomic indicators show that total landings were, on average, 64.20 metric tonnes higher (*p* > 0.1) after the implementation of the reserves, though this cannot necessarily be interpreted as causal, because it relies entirely on a before/after comparison.
+Total income was $10,344.85 (*p* < 0.05) thousands of Mexican Pesos (K MXP) higher after the implementation of the reserves.
+On average, lobster and sea cucumber landings increased, while urchin and snail landings and income decreased.
 Figure 6 presents the changes in these indicators through time, and Table 4 summarizes these results.
 
-**Figure 6 Plots for values of each socioeconomic indicator (y-axis) through time (x-axis).** Red and blue correspond to before and after the implementation of the reserve, respectively.
-This figure contains information for total landings (a), total income (b), lobster landings (c), urchin landings (d), snail landings (e), cucumber landings (f), lobster income (g), urchin income (h), snail income (i), and cucumber income (j).
+**Fig.
+6 Plots for values of each socioeconomic indicator (y-axis) through time (x-axis).** Red and blue correspond to before and after the implementation of the reserve, respectively.
+This figure contains information for total landings (a), total income (b), lobster landings (c), urchin landings (d), snail landings (e), sea cucumber landings (f), lobster income (g), urchin income (h), snail income (i), and sea cucumber income (j).
 
@@
 
-In terms of governance, the community is strongly organized, which is likely a driver of their success [@gutirrez_2011-0U].
-The first point of success is the existence of a fishing cooperative that is also affiliated to a regional fishing cooperatives federation.
-These polycentric governance structures allow various levels of organization that foster communication and cooperation [@mccay_2014-CN]; federations also provide bargain power with governments [@espinosaromero_2014-PY; @finkbeiner_2015-87].
+Recall that the governance objectives are evaluated based on the institutions present, not on a specific quantitative linkage between governance and biological or economic outcomes.
+Data for this reserve suggest that the community is strongly organized, which is a likely driver of the successes reported above [@gutirrez_2011-0U].
+The first point of success is the existence of a fishing cooperative that is also affiliated with a regional federation of cooperatives.
+These polycentric governance structures allow various levels of organization that have been shown to foster communication and cooperation [@mccay_2014-CN]; federations also provide bargain power with governments [@espinosaromero_2014-PY; @finkbeiner_2015-87].
 Access to fishing resources is managed through a TURF, permits, and fishing quotas (for some species).
-Their TURF promotes a sense of stewardship of their resources and incentivizes sustainable management [@mccay_2017-1m; @finkbeiner_2015-87].
-Together, these structures enabled a participative, bottom--up process during the reserve design phase; opinions of all fishing members ---and often non--fishers, but community members--- were included.
-Participation of community members in reserve surveillance and yearly monitoring indicate commitment and interest, and allow informal communication of results to un-involved community members.
+McCay [@mccay_2017-1m] suggests that the TURF promotes a sense of stewardship of their resources and incentivizes sustainable management.
+Together, these structures enabled a participative, bottom–up process during the reserve design phase; opinions of all fishing members ---and often non--fishing community members--- were included.
+Participation of community members in reserve surveillance and yearly monitoring indicate commitment and interest, and allow informal communication of results to un--involved community members.
 Furthermore, the reserve is partially isolated from poaching activity and fishers have internal regulations pertaining to the reserves.
@@
 
-Here we focus on the advantages and room for improvement in MAREA and highlight its utility for evidence--based management.
+We have developed and presented a user--friendly, automated approach for evaluating the effectiveness of marine reserves around the world.
+Here we highlight MAREA’s utility for evidence–-based management, and comment on a few of its shortcomings.
 The findings from Isla Natividad are used purely to validate the relevance of MAREA rather than to discuss particularities of the marine reserve effectiveness, which has been described before [@micheli_2012-EU; @rossetto_2015-V0; @munguavega_2015-yg].
@@
 
-The causal inference techniques used by MAREA have been suggested [@burgess_2016-HN; @ferraro_2006-oW] and used [@moland_2013-VP] before.
+The causal inference techniques used by MAREA have been suggested [@burgess_2016-HN; @ferraro_2006-oW] and used [@moland_2013-VP] before in other ad hoc studies.
 This approach reduces ambiguity in the interpretation of results.
@@
 5h).
-In this case, a before--after evaluation the reserve (*i.e.* ignoring the control site) would have incorrectly concluded that the reserve wasn't protecting invertebrates.
-On the other hand, a control--impact approach (*i.e.* compare reserve vs.
+In this case, a before–after evaluation of the reserve (i.e.
+ignoring the control site) would have incorrectly concluded that the reserve failed to protect invertebrates.
+On the other hand, a control–impact approach (i.e.
+compare reserve vs.
 control site only in 2016) would have identified higher densities inside the reserve, concluding that the reserve increases invertebrate density.
-However, MAREA identifies the changes through time and across sites, and estimates the effect of the reserve on density at $\beta_2 = `r formatC(bio_results_summary$estimate[8], digits = 2, format = "f")`$ (*p* < 0.05).
-What really happens is that invertebrate densities decrease in both sites through time, but the decrease is faster for the control site, thus yielding a positive value for $\beta_2$.
+However, by executing a formal difference-in-differences approach for causal inference, MAREA identifies the changes through time and across sites, and estimates the effect of the reserve on density at $\beta_2 = `r formatC(bio_results_summary$estimate[8], digits = 2, format = "f")`$ (*p* < 0.05).
+This approach reveals that invertebrate densities decrease in both sites through time, but the decrease is faster for the control site, thus yielding a positive value for $\beta_2$.
 
@@
 
-MAREA provides an overview of the effect of the reserve, allowing users to rapidly identify trends.
+MAREA provides in-depth analysis and a convenient snapshot overview of the effect of the reserve, allowing users to rapidly identify trends.
 However, users must interpret multiple indicators at a time to better understand the results.
-For example, with additional knowledge of local environmental variability (*i.e.* indicator B5 -- Natural Disturbance), we can better understand the trends in invertebrate densities.
-As reported before @micheli_2012-EU, hypoxic conditions that have occurred in Isla Natividad can cause decreases in invertebrate densities, and reserves buffer the negative effect.
-While MAREA automates the analyzes and makes results replicable, proper interpretation will still depend on the user.
+For example, with additional knowledge of local environmental variability (*i.e.* indicator B5 – Natural Disturbance), we can better understand the trends in invertebrate densities.
+As reported before [@micheli_2012-EU], hypoxic conditions that have occurred in Isla Natividad can cause decreases in invertebrate densities, and reserves buffer the negative effect.
+While MAREA automates the analysis and makes results replicable, proper interpretation will still depend on the user.
 Results produced by MAREA can only aid in management and decision making when results have been correctly interpreted.
 
-Socioeconomic and governance indicators typically lack a control site, which impede us from using causal inference techniques to measure their changes [@mascia_2017-m_].
+Socioeconomic and governance indicators typically lack a control site, which impede us from using the causal inference techniques employed to measure biological changes [@mascia_2017-m_].
 However, we can still extract useful information from them.
 Again, by combining results from multiple indicators, MAREA can provide insights into the effect of the reserve.
-For example, lobster and cucumber have shown increases in densities, landings, and income.
-We cannot conclude that landings and income from these species hve increased due to the reserve, but we can at least conclude that landings have not decreased.
+For example, lobster and sea cucumber have shown increases in densities, landings, and income.
+We cannot conclude that landings and income from these species have increased due to the reserve, but we can at least conclude that landings have not decreased.
 While further information on market behavior of each fishery is needed, these results provide insights into the state of the reserve and its associated fisheries.
@@
 However, providing a single platform (*i.e.* scorecard) or document (*i.e.* technical report) where biological, socioeconomic, and governance information is comprehensively included can aid in management.
-This information will be standardly reported across reserves, and can help managers identify overarching patterns across sites.
+By using MAREA, this information will be reported across reserves in a standardized way, and can help managers identify overarching patterns across sites.
 
-It is important that evidence--based management and conservation outcomes are communicated.
-While stakeholder involvement in the design and implementation phases of marine reserves is important, it is not enough.
-The results of such efforts must be communicated to stakeholders in order to maintain involvement and exchange of ideas [@lundquist_2005-OL; @leslie_2007-yJ].
-The way in which MAREA presents results allows this information to be interpreted by a wide-ranging audience.
-The scorecard is easily understandable by experts and non-experts, and can be used as an effective tool for communicating the results of annual evaluations.
+By making results straightforward to interpret, MAREA may also assist in communication with a broader stakeholder community.
+While stakeholder involvement in the design and implementation phases of marine reserves is important, that may not be sufficient for ensuring long--run buy--in or success.
+The scorecard is easily understandable by experts and non--experts, and can be used as an effective tool for communicating the results of annual evaluations.
 Additionally, the technical report can serve as a tool for managers and scientists to rapidly produce and communicate information at a more technical level.
@@
 However, they provide a starting point to perform the evaluation, to which managers and users can add other indicators (*e.g.* larval dispersal or connectivity) that are relevant to their reserve.
-Furthermore, MAREA's value is that it provides a free, simple, and replicable way to perform rigorous impact analysis.
+Furthermore, MAREA’s value is that it provides a free, simple, and replicable way to perform rigorous impact analysis.
 The tool can easily be used by fishers, CSO members, and managers in government agencies, providing transparency of the analysis and results.
@@
 
-The effectiveness of marine reserves continues to be a matter of debate [@woodcock_2017-Wm; @edgar_2014-UO].
+The effectiveness of marine reserves continues to be a matter of debate [@woodcock_2017-Wm; @edgar_2014-UO; @padleton_2017-vn].
 With current targets set to increase ocean protection, it is important that we understand the effects of our interventions [@burgess_2016-HN] so we can better inform management [@ferraro_2006-oW].
 It is therefore important that academics, managers, fishers, and CSOs have access to open access tools like MAREA.
-This is particularly relevant for Mexico and other Latin American countries, where management agencies are often shorthanded and underfunded [@lundquist_2005-OL].
+This is particularly relevant for Mexico and other Latin American countries, where management agencies are often understaffed and underfunded [@lundquist_2005-OL], or where materials are often not available in their language.
 In this context, MAREA provides a simple and replicable way to align management objectives with performance indicators.
@@
 We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analyzed.
-Despite this, providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards effective evidence--based management [@ferraro_2006-oW].
+Despite this, providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards effective evidence--–based management [@ferraro_2006-oW].
 
-The first release of MAREA is now available, and it will continue to be developed and maintained to keep up to date with literature.
+The first release of MAREA is now available, and it will continue to be developed and maintained to keep up to date with the literature.
 This process will incorporate new features, and enhance current ones, aiming to improve user experience and expand the scope of the analysis.
-Yet, we believe that this first release represents a major step towards effective evaluation and management of marine reserves.
+Yet, we believe that this first release represents a major step towards effective, replicable evaluation and management of marine reserves.
 
@@
 # Funding
-
-Funding must not be included in the manuscript.
-There is a special section in the submission that I will manually have to fill in.
-Please, review the following statements and let me know if I should add something else.
 
@@
 Dots are horizontally jittered to aid visualization.
-This figure contains information for fish shannon diversity index (a), fish species richness (b), fish density (c), fish trophic level (d), fish biomass (e), invertebrate shannon diversity index (f), invertebrate species richness (g), invertebrate density (h), lobster density (i), urchin density (j), snail density (k), and cucumber density (l)."}
+This figure contains information for fish shannon diversity index (a), fish species richness (b), fish density (c), fish trophic level (d), fish biomass (e), invertebrate shannon diversity index (f), invertebrate species richness (g), invertebrate density (h), lobster density (i), urchin density (j), snail density (k), and sea cucumber density (l)."}
 gridExtra::grid.arrange(shannon_fish_plot, richness_fish_plot, density_fish_plot, trophic_fish_plot, biomass_fish_plot, shannon_invert_plot, richness_invert_plot, density_invert_plot, lobster_density_invert_plot, urchin_density_invert_plot, snail_density_invert_plot, cucumber_density_invert_plot, ncol = 3)
@@
 Red and blue correspond to before and after the implementation of the reserve, respectively.
-This figure contains information for total landings (a), total income (b), lobster landings (c), urchin landings (d), snail landings (e), cucumber landings (f), lobster income (g), urchin income (h), snail income (i), and cucumber income (j)."}
+This figure contains information for total landings (a), total income (b), lobster landings (c), urchin landings (d), snail landings (e), sea cucumber landings (f), lobster income (g), urchin income (h), snail income (i), and sea cucumber income (j)."}
 gridExtra::grid.arrange(landings_plot, income_plot, landings_lobster_plot, landings_urchin_plot, landings_snail_plot, landings_cucumber_plot, income_lobster_plot, income_urchin_plot, income_snail_plot, income_cucumber_plot, ncol = 3)

### 2017-10-11 Manuscript.Rmd
Updated manuscript with minor corrections

@@
 ---
-title: A user-friendly tool to evaluate the effectiveness of no--take marine reserves
+title: A user--friendly tool to evaluate the effectiveness of no--take marine reserves
 header-includes:
@@
 output:
-  pdf_document:
+  word_document:
     fig_caption: yes
@@
 Its purpose is to evaluate the extent to which any given reserve has achieved its stated objectives.
-MAREA provides specific guidance on data collection and formatting, and then conducts rigorous causal inference analysis based on data input by the user, providing real-time outputs about the effectiveness of the reserve.
-MAREA’s ease of use, standardization of state-of-the-art inference methods, and ability to analyze marine reserve effectiveness across ecological, socioeconomic, and governance objectives could dramatically further our understanding and support of effective marine reserve management.
+MAREA provides specific guidance on data collection and formatting, and then conducts rigorous causal inference analysis based on data input by the user, providing real--time outputs about the effectiveness of the reserve.
+MAREA’s ease of use, standardization of state--of--the--art inference methods, and ability to analyze marine reserve effectiveness across ecological, socioeconomic, and governance objectives could dramatically further our understanding and support of effective marine reserve management.
 
@@
 This approach does not account for other factors for which one must control [@davies_2017-ml] in order to causally attribute a biological change to the reserve.
-While some studies have used control sites, these analyses do not estimate the effect of the reserve, and often use a control-impact comparison approach that does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
-A smaller fraction of studies have used a before-after-control-impact (*i.e.* BACI) design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.
+While some studies have used control sites, these analyses do not estimate the effect of the reserve, and often use a control--impact comparison approach that does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
+A smaller fraction of studies have used a before--after--control--impact (*i.e.* BACI) design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.
 
@@
 Effects of reserves on ecological and economic outcomes are highly heterogeneous, and often depend on the specific ecological, economic, and social context.
-The purpose of this paper is to describe a user-friendly tool, called “MAREA”, to rigorously systematize the evaluation of marine reserve effectiveness.
-The tool is in the form of an open-source application that uses state-of-the-art methods from program evaluation to compare a reserve to control sites along a number of ecological, economic, and governance dimensions.
+The purpose of this paper is to describe a user--friendly tool, called “MAREA”, to rigorously systematize the evaluation of marine reserve effectiveness.
+The tool is in the form of an open-source application that uses state--of--the--art methods from program evaluation to compare a reserve to control sites along a number of ecological, economic, and governance dimensions.
 
@@
 
-An increasingly popular way to make science accessible, reproducible, scalable, and replicable is through Open Science and the development of open-access tools [@lowndes_2017-xh].
+An increasingly popular way to make science accessible, reproducible, scalable, and replicable is through Open Science and the development of open--access tools [@lowndes_2017-xh].
 The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, successfully standardized a way to measure the health and benefits of the oceans.
@@
 
-This paper presents a framework and user-friendly tool to evaluate marine reserve effectiveness, which incorporates the biological, socioeconomic, and governance dimensions of any fishery.
+This paper presents a framework and user--friendly tool to evaluate marine reserve effectiveness, which incorporates the biological, socioeconomic, and governance dimensions of any fishery.
 We first provide a list of commonly stated management objectives and match them to appropriate indicators.
@@
 Alongside, methodologies to analyze these indicators are presented.
-We then describe the development of MAREA and explain how this user-friendly open access tool can be used by fishermen, managers, and other stakeholders with little scientific background.
+We then describe the development of MAREA and explain how this user--friendly open--access tool can be used by fishermen, managers, and other stakeholders with little scientific background.
 Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA to inform management.
@@
 
-Throughout this study, we will refer to the stated goals for which a marine reserve was designed as “objectives.” This work was motivated by the case of Mexico, where 39 reserves have been implemented over the past five years to achieve objectives such as increasing productivity in nearby waters or recover overexploited species; these reserves have never been formally evaluated for effectiveness at meeting those objectives.
+Throughout this study, we will refer to the stated goals for which a marine reserve was designed as “objectives.” This work was motivated by the case of Mexico, where 39 reserves have been implemented over the past five years to achieve objectives such as increasing productivity in nearby waters or recover overexploited species; most of these reserves have never been formally evaluated for effectiveness at meeting those objectives.
 Thus, our focus was on identifying common objectives of marine reserves in Mexico.
@@
 Ensenada*), in El Rosario, Baja California, who provided input.
-Our final list of indicators includes those identified in review works such as Lester *et al.* @lester_2009-Ks or Woodcock *et al.* @woodcock_2017-Wm.
+Our final list of indicators includes those identified in review works [@lester_2009-Ks; @woodcock_2017-Wm].
 
@@
 Some indicators require a numerical entry (*e.g.* Fish biomass) while others are more descriptive (*e.g.* Reasoning for reserve location).
-Many of the indicators specifically measure an outcome of the reserve, though some are designed to further the understanding of the mechanisms driving a reserve’s performance.
+Many of them specifically measure an outcome of the reserve, though some are designed to further the understanding of the mechanisms driving a reserve’s performance.
 In that sense, most biological and socioeconomic indicators are outcome variables.
@@
 
-\clearpage
-
 ```{r, results = "asis"}
 read.csv("./Tables/Table_indicators.csv") %>% 
-  knitr::kable(caption = "List of indicators to evaluate the effectiveness of no-take marine reserves, grouped by type.",
+  knitr::kable(caption = "List of indicators to evaluate the effectiveness of no-take marine reserves.",
                col.names = c("Code", "Indicator", "Data type", "Unit")) %>%
@@
 The hypothesis that the indicators will respond to implementation of the reserve is tested by analyzing spatial and temporal changes in each numeric biological indicator (all but B5) using generalized linear models [@moland_2013-VP].
-To account for variations in the environment and survey conditions, covariates that are gathered during the underwater ecological surveys are included into difference-in-differences a model with form:
+To account for variations in the environment and survey conditions, covariates that are gathered during the underwater ecological surveys are included in the difference-in-differences model with form:
 
-$$I_{i,t,z}=\beta_0 + \sum_{t = 2}^T\gamma_{i,t}Y_t + \beta_1Z_{i,z} + \beta_2P_{i,t,z}\times Z_{i,t,z} + \beta_3T_{i,t,z} + \beta_4V_{i,t,z} + \beta_5D_{i,t,z} + \epsilon_{i,t,z}$$
+$$I_{i,t,z}=\beta_0 + \sum_{t = 2}^T\gamma_{t}Y_t + \beta_1Z_{i,z} + \beta_2P_{i,t}\times Z_{i,z} + \beta_3T_{i,t,z} + \beta_4V_{i,t,z} + \beta_5D_{i,t,z} + \epsilon_{i,t,z}$$
 
@@
 While biological and some economic data are regularly collected, governance data are typically not available nor systematically collected by the community or other organizations.
-Therefore, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1-G15).
+Therefore, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1--G15).
 The survey is included as supplementary material in English (S1 Appendix) and Spanish (S2 Appendix).
@@
 
-MAREA is designed as a 6-step process, divided in tabs that appear upon launching the app.
+MAREA is designed as a 6--step process, divided in tabs that appear upon launching the app.
 The first tab introduces the app and summarizes the evaluation process.
@@
 Once the user has confirmed all input data, objectives, and other information, MAREA performs the formal program evaluation analyses discussed above.
-For a typical data set, the automated analysis step takes less than one second --but is dependent on connection speed.
+For a typical data set, the automated analysis step takes less than one second.
 Finally, the user is taken to the results tab where all results are presented in a simple format.
@@
 1.
-For socioeconomic indicators, colors are assigned based on the direction of the slope ($\beta_1$).
+For socioeconomic indicators, colors are assigned based on the direction of the slope ($\beta_1$) in eq.
+3.
 Red, yellow, and green are used for $\beta_i<0$, $\beta_i = 0$, and $\beta_i>0$, respectively.
@@
 This report also includes a scorecard as a summary of the results, but provides more information for each indicator.
-For all numeric indicators, the report includes a graph of the value of the indicator in the reserve and control sites through time.
+For all numeric biological indicators, the report includes a graph of the value of the indicator in the reserve and control sites through time.
 It also provides a regression table that summarizes the value of all coefficients in the regression and their respective robust standard errors.
@@
 The technical report is produced by a parameterized Rmarkdown document [@rmarkdown_2017] processed by the knitr package [@knitr_2017].
-Another feature of MAREA is that the user can choose to share the data; the user can also choose not to share the data.
+Another feature of MAREA is that the user can choose to share the data.
 Once the technical report is downloaded, the information on the reserve, its management objectives, and all uploaded data are saved into a central repository.
@@
 
-invert_data$Genero[invert_data$GeneroEspecie == "Mesocentrotus franciscanus"] <- "Strongylocentrotus"
-invert_data$GeneroEspecie[invert_data$GeneroEspecie == "Mesocentrotus franciscanus"] <- "Strongylocentrotus franciscanus"
-
+# invert_data$Genero[invert_data$GeneroEspecie == "Strongylocentrotus franciscanus"] <- "Mesocentrotus"
+# invert_data$GeneroEspecie[invert_data$GeneroEspecie == "Strongylocentrotus franciscanus"] <- "Mesocentrotus franciscanus"
+# 
 # write.csv(invert_data, "./Data/natividad_invertebrates.csv", row.names = F)
@@
 
-In 2006, the Isla Natividad community implemented two community-based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh; @micheli_2012-EU].
+In 2006, the Isla Natividad community implemented two community--based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh; @micheli_2012-EU].
 Evidence suggest that these reserves have been effective at enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of high value commercial species such as abalone [@munguavega_2015-yg].
@@
 
-The objective of this reserve was to recover species of economic interest -—-which were overexploited--- and to enhance fishery production in nearby waters.
+The objective of this reserve was to recover species of economic interest -—which were overexploited-- and to enhance fishery production in nearby waters.
 Fishers were also interested in preserving biological diversity and the ecosystem.
@@
 Ecological surveys were performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, providing the requisite time series data inside the reserve and for a suitable control site.
-Annual surveys (2006 -- 2016) were carried out in late July -- early August, performing a total of `r sum(fish_summary$Reserva)` and `r sum(invert_summary$Reserva)` transects in the reserve site for fish and invertebrate surveys, respectively.
+Annual surveys (2006--2016) were carried out in late July -- early August, performing a total of `r sum(fish_summary$Reserva)` and `r sum(invert_summary$Reserva)` transects in the reserve site for fish and invertebrate surveys, respectively.
 Similar sampling effort was applied to the control site, with `r sum(fish_summary$Control)` fish and `r sum(invert_summary$Control)` invertebrate transects.
@@
 Socioeconomic data were obtained from the National Commission for Aquaculture and Fisheries (*Comisión Nacional de Acuacultura y Pesca*; CONAPESCA).
-The data contains species-level information on monthly landings and income from nine species from 2000 to 2014.
-Data on landings and income were aggregated by year, and adjusted by the Consumer Price Index [@oecd_website_2017-VV].
+The data contains species--level information on monthly landings and income from nine species from 2000 to 2014.
+Data on landings and income were aggregated by year and species, and adjusted by the Consumer Price Index [@oecd_website_2017-VV].
 From the nine species available, we selected as objective species those that contributed the most (88.27%) income from 2000 to 2014: lobster (*Panulirus interruptus*; 71.76%), red sea urchin (*Mesocentrotus franciscanus*; 9.33%), snail (*Megastraea undosa*; 3.93%), and sea cucumber (*Parastichopus parvimensis*; 3.23%).
@@
 ```{r invert analysis}
-objsp <- data.frame(sp = c("Panulirus interruptus","Strongylocentrotus franciscanus", "Megastraea undosa", "Parastichopus parvimensis"), class = "invert", stringsAsFactors = F)
+objsp <- data.frame(sp = c("Panulirus interruptus","Mesocentrotus franciscanus", "Megastraea undosa", "Parastichopus parvimensis"), class = "invert", stringsAsFactors = F)
 
@@
 The scorecard (Fig.
-4) shows that this reserve achieves a general score of 64%, suggesting  that  64% of indicators are positive.
-All category–level scores were also high, with values of 67%, 60%, and 75% positive indicators for biological, socioeconomic and governance, respectively.
+4) shows that this reserve achieves a general score of 64%, suggesting  that  64% of all indicators are positive.
+All category–level scores were also high, with values of 67%, 60%, and 71% positive indicators for biological, socioeconomic and governance, respectively.
 
@@
 
+\clearpage
+
 **Fig.
@@
 This figure contains information for fish Shannon's diversity index (a), fish species richness (b), fish density (c), fish trophic level (d), fish biomass (e), invertebrate Shannon's diversity index (f), invertebrate species richness (g), invertebrate density (h), lobster density (i), urchin density (j), snail density (k), and sea cucumber density (l).
-
-\clearpage
 
@@
 One of the main objectives of this reserve was to increase landings.
-Results of the socioeconomic indicators show that total landings were, on average, 64.20 metric tonnes higher (*p* > 0.1) after the implementation of the reserves, though this cannot necessarily be interpreted as causal, because it relies entirely on a before/after comparison.
+Results of the socioeconomic indicators show that total landings were, on average, 64.20 metric tonnes higher (*p* > 0.1) after the implementation of the reserves, though this cannot necessarily be interpreted as causal, because it relies entirely on a before--after comparison.
 Total income was $10,344.85 (*p* < 0.05) thousands of Mexican Pesos (K MXP) higher after the implementation of the reserves.
@@
 This figure contains information for total landings (a), total income (b), lobster landings (c), urchin landings (d), snail landings (e), sea cucumber landings (f), lobster income (g), urchin income (h), snail income (i), and sea cucumber income (j).
-
-\clearpage
 
@@
 The first point of success is the existence of a fishing cooperative that is also affiliated with a regional federation of cooperatives.
-These polycentric governance structures allow various levels of organization that have been shown to foster communication and cooperation [@mccay_2014-CN]; federations also provide bargain power with governments [@espinosaromero_2014-PY; @finkbeiner_2015-87].
+These polycentric governance structures allow various levels of organization that have been shown to foster communication and cooperation [@mccay_2014-CN; @espinosaromero_2014-PY]; federations also provide bargain power with governments [@espinosaromero_2014-PY; @finkbeiner_2015-87].
 Access to fishing resources is managed through a TURF, permits, and fishing quotas (for some species).
@@
 5h).
-In this case, a before–after evaluation of the reserve (i.e.
-ignoring the control site) would have incorrectly concluded that the reserve failed to protect invertebrates.
-On the other hand, a control–impact approach (i.e.
-compare reserve vs.
+In this case, a before–after evaluation of the reserve (*i.e.* ignoring the control site) would have incorrectly concluded that the reserve failed to protect invertebrates.
+On the other hand, a control–impact approach (*i.e.* compare reserve vs.
 control site only in 2016) would have identified higher densities inside the reserve, concluding that the reserve increases invertebrate density.
-However, by executing a formal difference-in-differences approach for causal inference, MAREA identifies the changes through time and across sites, and estimates the effect of the reserve on density at $\beta_2 = `r formatC(bio_results_summary$estimate[8], digits = 2, format = "f")`$ (*p* < 0.05).
+However, by executing a formal difference--in--differences approach for causal inference, MAREA identifies the changes through time and across sites, and estimates the effect of the reserve on density at $\beta_2 = `r formatC(bio_results_summary$estimate[8], digits = 2, format = "f")`$ (*p* < 0.05).
 This approach reveals that invertebrate densities decrease in both sites through time, but the decrease is faster for the control site, thus yielding a positive value for $\beta_2$.
@@
 
-MAREA provides in-depth analysis and a convenient snapshot overview of the effect of the reserve, allowing users to rapidly identify trends.
+MAREA provides in--depth analysis and a convenient snapshot overview of the effect of the reserve, allowing users to rapidly identify trends.
 However, users must interpret multiple indicators at a time to better understand the results.
-For example, with additional knowledge of local environmental variability (*i.e.* indicator B5 – Natural Disturbance), we can better understand the trends in invertebrate densities.
+For example, with additional knowledge of local environmental variability (*i.e.* indicator B5: Natural Disturbance), we can better understand the trends in invertebrate densities.
 As reported before [@micheli_2012-EU], hypoxic conditions that have occurred in Isla Natividad can cause decreases in invertebrate densities, and reserves buffer the negative effect.
@@
 We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analyzed.
-Despite this, providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards effective evidence--–based management [@ferraro_2006-oW].
+Despite this, providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards effective evidence-–based management [@ferraro_2006-oW].
 

### 2017-11-06 Manuscript.Rmd
updated the repository with the submitted version

@@
 title: A user--friendly tool to evaluate the effectiveness of no--take marine reserves
+subtitle: Marine reserve evaluation
 header-includes:
@@
 output:
-  word_document:
+  pdf_document:
     fig_caption: yes
+    keep_tex: yes
 fontsize: 12pt
@@
 ---
+<script type="text/x-mathjax-config">
+MathJax.Hub.Config({
+  TeX: { 
+      equationNumbers: { 
+            autoNumber: "all",
+            formatNumber: function (n) {return '9.'+n}
+      } 
+  }
+});
+</script>
 
@@
 Moreover, analysis and evaluation of reserves is generally conducted by outside researchers, not the reserve managers or users, plausibly thereby hindering effective local management and rapid response to change.
-We present a framework and tool, called “MAREA”,to overcome these challenges.
+We present a framework and tool, called “MAREA”, to overcome these challenges.
 Its purpose is to evaluate the extent to which any given reserve has achieved its stated objectives.
@@
 
-$$I_{i,t,z}=\beta_0 + \sum_{t = 2}^T\gamma_{t}Y_t + \beta_1Z_{i,z} + \beta_2P_{i,t}\times Z_{i,z} + \beta_3T_{i,t,z} + \beta_4V_{i,t,z} + \beta_5D_{i,t,z} + \epsilon_{i,t,z}$$
+\begin{equation}I_{i,t,z}=\beta_0 + \sum_{t = 2}^T\gamma_{t}Y_t + \beta_1Z_{i,z} + \beta_2P_{i,t}\times Z_{i,z} + \beta_3T_{i,t,z} + \beta_4V_{i,t,z} + \beta_5D_{i,t,z} + \epsilon_{i,t,z}\end{equation}
 
@@
 
-$$I_t = RI\times \frac{CPI_t}{CPI_T}$$
+\begin{equation}I_t = RI\times \frac{CPI_t}{CPI_T}\end{equation}
 
@@
 
-$$I_{t}=\beta_0 + \beta_1P_{t} + \epsilon_{t}$$
+\begin{equation}I_{t}=\beta_0 + \beta_1P_{t} + \epsilon_{t}\end{equation}
 
@@
 
-# Funding
-
-- CF and JCVD received support from the Latin American Fisheries Fellowship Program.
-- JCVD received support from the Consejo Nacional de Ciencia y Tecnología (CONACyT).
-- GM received financial support from Waitt Foundation.
-- Biological data for Isla Natividad were collected with the support of the Walton Family Foundation, the NSF-CNH program (grant DEB-1212124), Marisla Foundation, Sandler Family Supporting Foundation, and Packard Foundation.
-- AS received support from The Nature Conservancy and World Wildlife Fund--Carlos Slim Foundation Alliance
-
 \clearpage
@@
 **S3 Table. Assigned values and reasoning of socioeconomic and governance indicators used to color-code the scorecard in MAREA**
-
-# Figures
-
-![Workflow to evaluate the effectiveness of marine reserves.](./Figures/flowchart.gif){ width="50%"}
-
-```{r, fig.cap = "Legend used to interpret the scorecard produced by MAREA.
-Colors indicate direction of change (red = negative; green = positive), and color intensity is given by the statistical significance"}
-fig2
-```
-
-```{r, fig.cap = "General location of Isla Natividad (left) and map of the island (right).
-The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce)."}
-tmap_arrange(pen_map, study_area, outer.margins = 0.005, ncol = 2)
-```
-
-![Scorecard produced by MAREA for the "La Plana / Las Cuevas" marine reserve in Isla Natividad, Mexico.](./Figures/scorecard.png){ width=80% }
-
-```{r, fig.height = 6, fig.width = 6, fig.cap = "Plots for values of each biological indicator (y-axis) through time (x-axis).
-Red and blue correspond to the reserve and control sites, respectively.
-Black lines indicate yearly mean values, and ribbons indicate +/- 1 standard error.
-Dots are horizontally jittered to aid visualization.
-This figure contains information for fish shannon diversity index (a), fish species richness (b), fish density (c), fish trophic level (d), fish biomass (e), invertebrate shannon diversity index (f), invertebrate species richness (g), invertebrate density (h), lobster density (i), urchin density (j), snail density (k), and sea cucumber density (l)."}
-gridExtra::grid.arrange(shannon_fish_plot, richness_fish_plot, density_fish_plot, trophic_fish_plot, biomass_fish_plot, shannon_invert_plot, richness_invert_plot, density_invert_plot, lobster_density_invert_plot, urchin_density_invert_plot, snail_density_invert_plot, cucumber_density_invert_plot, ncol = 3)
-```
-
-```{r, fig.height = 6, fig.width = 6, fig.width = 6, fig.cap = "Plots for values of each socioeconomic indicator (y-axis) through time (x-axis).
-Red and blue correspond to before and after the implementation of the reserve, respectively.
-This figure contains information for total landings (a), total income (b), lobster landings (c), urchin landings (d), snail landings (e), sea cucumber landings (f), lobster income (g), urchin income (h), snail income (i), and sea cucumber income (j)."}
-gridExtra::grid.arrange(landings_plot, income_plot, landings_lobster_plot, landings_urchin_plot, landings_snail_plot, landings_cucumber_plot, income_lobster_plot, income_urchin_plot, income_snail_plot, income_cucumber_plot, ncol = 3)
-```

### 2017-12-03 ManuscriptV1.tex
Created a reference document to track changes

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={A user--friendly tool to evaluate the effectiveness of no--take marine reserves},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{A user--friendly tool to evaluate the effectiveness of no--take marine
reserves}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
\subtitle{Marine reserve evaluation}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage[table]{xcolor}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{setspace}
\doublespacing
\usepackage{lineno}
\linenumbers

\begin{document}
\maketitle

Juan Carlos Villaseñor-Derbez\textsuperscript{1¶*}, Caio
Faro\textsuperscript{1¶}, Melaina Wright\textsuperscript{1¶}, Jael
Martínez\textsuperscript{1¶}, Sean Fitzgerald\textsuperscript{1\&},
Stuart Fulton\textsuperscript{2\&}, Maria del Mar
Mancha-Cisneros\textsuperscript{3\&}, Gavin
McDonald\textsuperscript{1,4,5\&}, Fiorenza
Micheli\textsuperscript{6\&}, Alvin Suárez\textsuperscript{2\&}, Jorge
Torre\textsuperscript{2\&}, Christopher Costello\textsuperscript{1,4,5¶}

\textsuperscript{1} Bren School of Environmental Science and Management,
University of California Santa Barbara, Santa Barbara, California,
United States

\textsuperscript{2} Comunidad y Biodiversidad A.C., Calle Isla del
Peruano, Guaymas, Sonora, México

\textsuperscript{3} School of Life Sciences, Arizona State University,
Tempe, Arizona, United States

\textsuperscript{4} Sustainable Fisheries Group, University of
California Santa Barbara, Santa Barbara, California, United States

\textsuperscript{5} Marine Science Institute, University of California
Santa Barbara, Santa Barbara, California, United States

\textsuperscript{6} Hopkins Marine Station and Center for Ocean
Solutions, Stanford University, Pacific Grove, CA 93950, USA

*Corresponding author

Email:
\href{mailto:jvillasenor@bren.ucsb.edu}{\nolinkurl{jvillasenor@bren.ucsb.edu}}
(JCVD)

¶ These authors contributed equally to this work.

\& These authors also contributed equally to this work.

\clearpage

\textbf{Abstract}

Marine reserves are implemented to achieve a variety of objectives, but
are seldom rigorously evaluated to determine whether those objectives
are met. In the rare cases when evaluations do take place, they
typically focus on ecological indicators and ignore other relevant
objectives such as socioeconomics and governance. And regardless of the
objectives, the diversity of locations, monitoring protocols, and
analysis approaches hinder the ability to compare results across case
studies. Moreover, analysis and evaluation of reserves is generally
conducted by outside researchers, not the reserve managers or users,
plausibly thereby hindering effective local management and rapid
response to change. We present a framework and tool, called ``MAREA'',
to overcome these challenges. Its purpose is to evaluate the extent to
which any given reserve has achieved its stated objectives. MAREA
provides specific guidance on data collection and formatting, and then
conducts rigorous causal inference analysis based on data input by the
user, providing real--time outputs about the effectiveness of the
reserve. MAREA's ease of use, standardization of state--of--the--art
inference methods, and ability to analyze marine reserve effectiveness
across ecological, socioeconomic, and governance objectives could
dramatically further our understanding and support of effective marine
reserve management.

\section{Introduction}\label{introduction}

Unsustainable fishing practices threaten biodiversity, conservation,
economic and social outcomes {[}1,2{]}. Marine Protected Areas (MPAs;
and marine reserves, in which all extractive efforts are prohibited) are
frequently proposed to aid in the recovery of fish and invertebrate
stocks {[}3--6{]} by limiting or restricting fishing effort and gears.

Available empirical evidence on marine reserve effectiveness is mixed
{[}7{]}. Some studies have shown that MPAs increase biomass {[}4,8{]},
enhance resilience to climatic impacts {[}9,10{]}, and preserve genetic
diversity {[}11{]}. Compared to partially protected MPAs, marine
reserves have higher levels of biomass, density, richness, and larger
organisms {[}3,12--14{]}. These effects are often measured as biological
changes in the area through time and many lack a control site for
comparison {[}15{]}. This approach does not account for other factors
for which one must control {[}16{]} in order to causally attribute a
biological change to the reserve. While some studies have used control
sites, these analyses do not estimate the effect of the reserve, and
often use a control--impact comparison approach that does not address
temporal variability {[}4,8,17--19{]}. A smaller fraction of studies
have used a before--after--control--impact (\emph{i.e.} BACI) design
comparing reserves to control sites before and after implementation
{[}4,20,21{]}, which allows the use of causal inference techniques that
estimate the effect of the reserve.

But even when proper causal inference can be drawn, results are often
idiosyncratic across reserves. Effects of reserves on ecological and
economic outcomes are highly heterogeneous, and often depend on the
specific ecological, economic, and social context. The purpose of this
paper is to describe a user--friendly tool, called ``MAREA'', to
rigorously systematize the evaluation of marine reserve effectiveness.
The tool is in the form of an open-source application that uses
state--of--the--art methods from program evaluation to compare a reserve
to control sites along a number of ecological, economic, and governance
dimensions.

The challenge of how to standardize marine reserve evaluation is not a
new one. The recent IUCN framework ``How is your MPA doing?''
{[}22,23{]} provides a comprehensive list of biological, socioeconomic,
and governance indicators, and insights into how these indicators may be
measured. But this framework stops short of analysis, so provides a user
with little guidance about establishing causal inference about the
reserve. Recent work by Mascia \emph{et al.} {[}24{]} integrates these
three dimensions via the Social Ecological Systems Framework {[}25,26{]}
and suggests the use of causal inference techniques to provide a measure
of the effect of conservation interventions. However, these two novel
approaches do not provide a user-friendly tool that enables
replicability and scalability of the analysis, particularly when used by
the fishers and decision makers themselves.

An increasingly popular way to make science accessible, reproducible,
scalable, and replicable is through Open Science and the development of
open--access tools {[}27{]}. The Ocean Health Index {[}28,29{]}, for
example, successfully standardized a way to measure the health and
benefits of the oceans. This approach has been implemented at a global
scale, but also at country-level {[}30{]}, and regionally {[}31,32{]}.
Open access tools are not limited to conservation, and have also been
developed to evaluate fishery performance {[}33,34{]}, design
territorial use rights for fisheries {[}35{]}, and improve decision
making in the hydro power industry {[}36{]}.

This paper presents a framework and user--friendly tool to evaluate
marine reserve effectiveness, which incorporates the biological,
socioeconomic, and governance dimensions of any fishery. We first
provide a list of commonly stated management objectives and match them
to appropriate indicators. We then develop a simple approach to
analyzing these indicators building on causal inference techniques
{[}20{]}, which help us understand the effect of management
interventions {[}24,37{]}. To implement the analytical approach in a
user-friendly format, we introduce the Marine Reserve Evaluation
Application (MAREA), an open source, web--based tool that automates the
framework described in this paper and enables its broader use. Finally,
we present a case study on the evaluation of a marine reserve
established by the fishers of Isla Natividad (Mexico) in 2006, to
demonstrate the potential of MAREA.

\section{Materials and methods}\label{materials-and-methods}

Here, we describe the proposed framework to evaluate the effectiveness
of marine reserves (Fig. 1). We explain how management objectives were
identified and matched to appropriate indicators that allow the
evaluation of the reserves, and provide brief guidelines on data
collection. Alongside, methodologies to analyze these indicators are
presented. We then describe the development of MAREA and explain how
this user--friendly open--access tool can be used by fishermen,
managers, and other stakeholders with little scientific background.
Finally, we provide guidelines on how to interpret and use the results
and output generated by MAREA to inform management.

\textbf{Fig 1. Workflow to evaluate the effectiveness of marine
reserves.}

\subsection{Marine Reserve objectives and
indicators}\label{marine-reserve-objectives-and-indicators}

Throughout this study, we will refer to the stated goals for which a
marine reserve was designed as ``objectives.'' This work was motivated
by the case of Mexico, where 39 reserves have been implemented over the
past five years to achieve objectives such as increasing productivity in
nearby waters or recover overexploited species; most of these reserves
have never been formally evaluated for effectiveness at meeting those
objectives. Thus, our focus was on identifying common objectives of
marine reserves in Mexico. However, a literature review and discussions
with marine reserve researchers worldwide suggest that the objectives
driving Mexican marine reserve implementation are similar to those in
the rest of the world. Thus, we group these objectives into seven major
categories, which can be applied to marine reserves worldwide. The list
of objectives includes stated objectives in legislation {[}38,39{]} and
official documents such as the Technical Justification Studies
(\emph{Estudios Técnicos Justificativos}), agreements, and decrees
associated to these areas:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Avoid overexploitation
\item
  Conserve species under a special protection regime
\item
  Maintain biological processes (reproduction, recruitment, growth,
  feeding)
\item
  Improve fishery production in adjacent waters
\item
  Preserve biological diversity and the ecosystem
\item
  Recover overexploited species
\item
  Recover species of economic interest
\end{enumerate}

Based on these seven objectives, we determined a set of associated
indicators to evaluate reserve effectiveness. These indicators are
specific variables on which data could be collected, and analyzed, to
ultimately determine whether the corresponding objective was causally
being achieved by the marine reserve. The list of indicators was
compiled through a review of scientific literature in which we
identified indicators that were used to measure similar objectives. A
first filter eliminated indicators for which baseline data do not
typically exist in Mexico. The preliminary list of indicators was
reviewed at a workshop with participation of members from Mexican
fishery management agencies and non-government organizations. Later,
these were presented to fishers from the Ensenada Fishing Cooperative
(\emph{S.C.P.P. Ensenada}), in El Rosario, Baja California, who provided
input. Our final list of indicators includes those identified in review
works {[}4,40{]}.

Indicators are divided into three main categories: biological,
socioeconomic, and governance (Table 1). The nine biological indicators
focus on fish and invertebrate communities that are evaluated using
underwater ecological surveys performed inside and outside the reserve
(see Data and Analysis section for specific sampling design and
methodologies). Five socioeconomic indicators reflect the performance of
the fishery in terms of landings, income from landings, and availability
of alternative livelihoods. Fifteen governance indicators describe the
governance structures under which the community operates (\emph{e.g.},
access rights to the fishery, number of fishers, legal recognition of
the reserve). Some indicators require a numerical entry (\emph{e.g.}
Fish biomass) while others are more descriptive (\emph{e.g.} Reasoning
for reserve location). Many of them specifically measure an outcome of
the reserve, though some are designed to further the understanding of
the mechanisms driving a reserve's performance. In that sense, most
biological and socioeconomic indicators are outcome variables. On the
other hand, governance indicators are viewed as possible explanatory
variables of reserve performance. Whenever an indicator is applied to
``Target species'', it means that the indicator can be used for all
species (\emph{e.g.} Fish Biomass) and/or for individual species that
are either the conservation target of the reserve or are of particular
economic or ecological interest (\emph{e.g.} Grouper Biomass). Table 1
presents the proposed indicators, and Table 2 shows how objectives are
matched with biological and socioeconomic indicators.

\begin{table}

\caption{\label{tab:unnamed-chunk-2}List of indicators to evaluate the effectiveness of no-take marine reserves.}
\centering
\resizebox{\linewidth}{!}{\begin{tabular}[t]{l|l|l|l}
\hline
Code & Indicator & Data type & Unit\\
\hline
\multicolumn{4}{l}{\textbf{Biological}}\\
\hline
\hspace{1em}B1 & Shannon diversity index & Continuous & \\
\hline
\hspace{1em}B2 & Species richness & Discrete & Number of species/transect\\
\hline
\hspace{1em}B3 & Density of mature organisms & Continuous & Percent points\\
\hline
\hspace{1em}B4 & Density* & Continuous & Organisms/transect\\
\hline
\hspace{1em}B5 & Natural Disturbance & Descriptive & \\
\hline
\hspace{1em}B6 & Mean Trophic Level & Continuous & \\
\hline
\hspace{1em}B7 & Biomass* & Continuous & kg/transect\\
\hline
\multicolumn{4}{l}{\textbf{Socioeconomic}}\\
\hline
\hspace{1em}S1 & Total landings* & Continuous & kg\\
\hline
\hspace{1em}S2 & Income from total landings* & Continuous & \$\\
\hline
\hspace{1em}S3 & Alternative economic opportunities & Ordinal & \\
\hline
\multicolumn{4}{l}{\textbf{Governance}}\\
\hline
\hspace{1em}G1 & Access to the fishery & Categorical & \\
\hline
\hspace{1em}G2 & Number of fishers & Discrete & \\
\hline
\hspace{1em}G3 & Legal recognition of reserve & Binary & \\
\hline
\hspace{1em}G4 & Reserve type & Descriptive & \\
\hline
\hspace{1em}G5 & Illegal harvesting & Ordinal & \\
\hline
\hspace{1em}G6 & Management plan & Binary & \\
\hline
\hspace{1em}G7 & Reserve enforcement & Descriptive & \\
\hline
\hspace{1em}G8 & Size of reserve & Discrete & \\
\hline
\hspace{1em}G9 & Reasoning for reserve location & Descriptive & \\
\hline
\hspace{1em}G10 & Membership to fisher organizations & Binary & \\
\hline
\hspace{1em}G11 & Type of fisheries organizations & Categorical & \\
\hline
\hspace{1em}G12 & Representation & Ordinal & \\
\hline
\hspace{1em}G13 & Internal Regulation & Binary & \\
\hline
\hspace{1em}G14 & Perceived Effectiveness & Categorical & \\
\hline
\hspace{1em}G15 & Social Impact of Reserve & Categorical & \\
\hline
\multicolumn{4}{l}{\textsuperscript{*} The indicator is also applied to target species}\\
\end{tabular}}
\end{table}

\clearpage

\begin{table}

\caption{\label{tab:unnamed-chunk-3}Management objectives and respective performance indicators. All governance indicators should allways be used.}
\centering
\resizebox{\linewidth}{!}{\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
\hline
Objective & B1 & B2 & B3 & B4 & B4* & B5 & B6 & B7 & B7* & S1 & S1* & S2 & S2* & S3\\
\hline
Avoid overexploitation &  &  & x &  & x & x &  &  & x & x & x & x & x & x\\
\hline
Conserve species under a special protection &  &  & x &  & x & x &  &  & x & x &  & x &  & x\\
\hline
Maintain biological process &  &  &  & x & x & x &  & x & x & x & x & x & x & x\\
\hline
Improve fishery production in nearby waters &  &  & x & x & x & x & x & x & x & x & x & x & x & x\\
\hline
Preserve biological diversity and the ecosystem &  &  & x &  & x & x &  &  & x &  & x &  & x & x\\
\hline
Recover overexploited species & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
\hline
Recover species of economic interest & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
\hline
\multicolumn{15}{l}{\textsuperscript{*} The indicator is applied to target species}\\
\end{tabular}}
\end{table}

\subsection{Data and analyses}\label{data-and-analyses}

In many coastal marine reserves of Mexico, biological data are collected
via underwater visual censuses as part of a reserve's monitoring
program. Scientific divers (which are often local fishermen with
guidance from Civil Society Organizations; CSOs) record fish and
invertebrate richness and abundances, as well as fish total length along
belt transects. Ecological surveys are typically performed annually in
each reserve and corresponding control site(s), before and after the
implementation of the reserve, providing a sampling design that can be
used to draw causal inference. Control sites are areas where habitat is
similar to that of the reserve, but with presence of fishing activity;
in principle these are areas that are otherwise observationally
identical to the reserve site, but where, for presumably random reasons,
a reserve was not implemented. While transect dimensions (\emph{i.e.}
length and width) and sampling methods might vary from study to study,
the general idea remains the same: richness, abundances, and sizes of
organisms are recorded in a study--specific standardized way. For this
reason, MAREA does not assume specific transect dimensions, and
pertinent indicators are calculated per transect (Table 1).

This sampling design for biological data allows us to use causal
inference techniques {[}20,41{]} to evaluate the effect of the reserve
on biological indicators. The hypothesis that the indicators will
respond to implementation of the reserve is tested by analyzing spatial
and temporal changes in each numeric biological indicator (all but B5)
using generalized linear models {[}20{]}. To account for variations in
the environment and survey conditions, covariates that are gathered
during the underwater ecological surveys are included in the
difference-in-differences model with form:

\begin{equation}I_{i,t,z}=\beta_0 + \sum_{t = 2}^T\gamma_{t}Y_t + \beta_1Z_{i,z} + \beta_2P_{i,t}\times Z_{i,z} + \beta_3T_{i,t,z} + \beta_4V_{i,t,z} + \beta_5D_{i,t,z} + \epsilon_{i,t,z}\end{equation}

In this model, \(i\), \(t\), and \(z\) are indices for transect, time,
and zone (control or reserve site), respectively. This model allows us
to estimate the change in an indicator (\(I\)) based on the year
(\(Y\)), a dummy variable that indicates treatment (\(Z\); \emph{i.e.}
control or reserve), an interaction between a dummy variable that
indicates pre-- or post--implementation (\(P\)) and treatment (\(Z\)),
and covariates such as bottom temperature (\(T\); in °C), horizontal
visibility during the survey (\(V\); in m), and depth at which survey
was performed (\(D\); in m). \(\epsilon\) represents the error term
associated to the regression. Here, years are modeled as factors, using
the first year as the reference level. This does not impose a linear
structure in the way an indicator changes through time (\emph{i.e.} the
change in biomass between 2006 and 2007 does not have to be the same as
the change between 2015 and 2016). The treatment and implementation
variables, modeled as dummy variables, are coded as Control = 0 and
Reserve = 1; and Pre--implementation = 0 and Post--implementation = 1,
respectively.

Socioeconomic data are often collected by fishers, natural resource
management agencies, or CSOs by recording landings, income, and
sometimes prices for each species. To control for inflation, income is
adjusted with the country's consumer price index {[}42{]}:

\begin{equation}I_t = RI\times \frac{CPI_t}{CPI_T}\end{equation}

Where \(I_t\) represents the adjusted income for year \(t\) as the
product between the reported income for that year and the ratio between
the consumer price index (\(CPI\)) in that year to the most recent
year's (\(T\)) CPI. Since no control sites are typically available for
this data type, numeric socioeconomic indicators (S1 and S2) are
evaluated with a simplified version of eq. 1:

\begin{equation}I_{t}=\beta_0 + \beta_1P_{t} + \epsilon_{t}\end{equation}

While this model does not formally allow for causal inference, we can
still measure changes in mean landings and income before and after the
implementation of the reserve and provide valuable input. For both
models (eq. 1 and eq. 3), we estimate the model coeficients with
ordinary least squares, and calculate heteroskedastic--robust standard
errors.

While biological and some economic data are regularly collected,
governance data are typically not available nor systematically collected
by the community or other organizations. Therefore, we created a survey
specifically designed to collect information needed for the proposed
indicators (B5, S3, and G1--G15). The survey is included as
supplementary material in English (S1 Appendix) and Spanish (S2
Appendix). To analyze governance information, we developed a framework
based on a literature review of common governance structures and their
relation to effectiveness in managing fisheries or marine reserves (S3
Table). This approach has been proven to successfully evaluate
governance structures {[}43{]}. Unlike with biological and socioeconomic
objectives (see eqs 1 and 3), MAREA does not quantitatively analyze
governance information. Rather, it is presented along with the
biological and socioeconomic indicators to provide managers and users
with a more complete description of the reserve.

\subsection{Marine Reserve Evaluation App
(MAREA)}\label{marine-reserve-evaluation-app-marea}

We developed MAREAin R version 3.4.2 and R Studio 1.1.383 {[}44{]} using
the Shiny package {[}45{]}, to build an interactive web application
hosted on an open server; the MAREA app can be accessed at
\href{turfeffect.shinyapps.io/marea/}{turfeffect.shinyapps.io/marea}.
While the original version was developed in Spanish because it was aimed
for Mexico and other Latin-American countries, all of its content can be
translated by a translation widget available within the app.

MAREA is designed as a 6--step process, divided in tabs that appear upon
launching the app. The first tab introduces the app and summarizes the
evaluation process. Then, the user selects management objectives, which
MAREA automatically matches to appropriate indicators, based on Table 2.
Users can also manually modify selected indicators based on their
interests and data availability. The user can then load data on one or
more reserves, using standard *.csv text files; sample datasets are
provided within MAREA. Once data have been loaded, MAREA identifies all
reserves in the data, and lets the user select the reserve to be
evaluated. At this point, the user can also specify the year of
implementation of the reserve, reserve dimensions, and indicate target
species that are of particular management interest. MAREA provides the
user with a section to confirm that all the decisions made leading up to
that point are correct. Once the user has confirmed all input data,
objectives, and other information, MAREA performs the formal program
evaluation analyses discussed above. For a typical data set, the
automated analysis step takes less than one second. Finally, the user is
taken to the results tab where all results are presented in a simple
format. The user can also download a more comprehensive technical report
produced in *.pdf format.

The first output is a color--coded scorecard intended to provide a
general overview of the effectiveness of the reserve. The scorecard
provides a global score for the reserve, a general score for each
category of indicators, and an individual score for each indicator. The
global and category--level scores are determined by the percentage of
positive indicators, overall and for each category, respectively. For
numeric biological indicators (all but B5), the color is defined by the
sign of the interaction term coefficient (\(\beta_2\)) in eq. 1. For
socioeconomic indicators, colors are assigned based on the direction of
the slope (\(\beta_1\)) in eq. 3. Red, yellow, and green are used for
\(\beta_i<0\), \(\beta_i = 0\), and \(\beta_i>0\), respectively. The
intensity of the color is defined by the significance of the
coefficient, testing the null hypothesis of no change (\emph{i.e.}
\(H_0: \beta_i = 0\)) with a Student's t-test. Cutoff values are
\(p < 0.05\) and \(p < 0.1\). Thus, even in a case where
\(\beta_i > 0\), if the coefficient is not significant by standard
measures (\emph{i.e. } \(p>0.1\)), the indicator will be assigned a
yellow color. A legend (Fig. 2) is provided within the scorecard to aid
in the interpretation of these results. Governance indicators are
represented simply by red or green. The color is defined based on what
literature shows to be a negative (red) or positive (green) factor for a
reserve (S3 Table). For example, if the perceived degree of illegal
fishing is high, this indicator will be assigned a red color. However,
due to the nature of some governance indicators, which require the user
to provide a narrative, only some indicators are presented in the
scorecard (although all are included in the technical report).

\textbf{Fig 2. Legend used to interpret the scorecard produced by
MAREA.} Colors indicate direction of change (red = negative; green =
positive), and color intensity is given by the statistical significance.

The second output from MAREA is a technical report intended to
communicate information and statistical results in a more comprehensive
and technical way. This report also includes a scorecard as a summary of
the results, but provides more information for each indicator. For all
numeric biological indicators, the report includes a graph of the value
of the indicator in the reserve and control sites through time. It also
provides a regression table that summarizes the value of all
coefficients in the regression and their respective robust standard
errors. The summary table also provides information on model fit
(\(R^2\)) and significance of the regression.

The scorecard is produced with functions from the Shinydashboard package
{[}46{]}. The technical report is produced by a parameterized Rmarkdown
document {[}47{]} processed by the knitr package {[}48{]}. Another
feature of MAREA is that the user can choose to share the data. Once the
technical report is downloaded, the information on the reserve, its
management objectives, and all uploaded data are saved into a central
repository. These data can be accessed at any time by any person
interested in acquiring them at \url{github.com/turfeffect/MAREAdata}.

\subsection{Case study}\label{case-study}

While MAREA is a general tool that can be easily employed to evaluate
the effectiveness of any marine reserve with the required input data, we
illustrate its use here by applying it to one marine reserve near Isla
Natividad, in Baja California Sur, Mexico. Isla Natividad is located 8
Km off the Pacific Coast of the Baja California Peninsula (Fig. 3),
where fishers operate under a fishing cooperative (\emph{S.C.P.P. Buzos
y Pescadores de la Baja California}) that promotes co-management of
marine resources {[}49,50{]}. Additionally, fishers have Territorial Use
Rights for Fisheries (TURFs) that provide them with exclusive access
rights to exploit the benthic marine resources within a given perimeter
{[}50{]}.

\textbf{Fig 3. General location of Isla Natividad (left) and map of the
island (right).} The marine reserve polygon is indicated in red, and the
approximate location of control sites is indicated by blue squares (B =
Babencho, D = La Dulce).

In 2006, the Isla Natividad community implemented two community--based
marine reserves within their TURF {[}9,51,52{]}. Evidence suggest that
these reserves have been effective at enhancing resilience to climate
variations {[}9{]} and preserving genetic diversity of high value
commercial species such as abalone {[}11{]}. These ecological benefits
have been translated into economic benefits, enhancing population
persistence and bolstering abalone fisheries {[}53{]}. For the purpose
of this evaluation, we focused on the ``La Plana / Las Cuevas'' marine
reserve, located at the southern end of the island (Fig. 3) and its
corresponding control site ``La Dulce / Babencho''.

The objective of this reserve was to recover species of economic
interest ----which were overexploited-- and to enhance fishery
production in nearby waters. Fishers were also interested in preserving
biological diversity and the ecosystem. Thus, objectives 4---7 were
selected. Using Table 2 to match these objectives with appropriate
management indicators, we selected all biological, socioeconomic, and
governance indicators included as options in the framework.

Local fishers (who were trained in scientific diving by the CSO
Comunidad y Biodiversidad, A.C. (COBI; \url{www.cobi.org}), ReefCheck
California, and Stanford University) and personnel from these
institutions performed SCUBA dives to record fish and invertebrate
richness and abundances, as well as fish total length. They recorded
information along 30 m transects, with a sampling window of 2 m x 2 m
following a standardized ReefCheck protocol {[}54{]}. Ecological surveys
were performed yearly in each reserve and corresponding control site(s),
before and after the implementation of the reserve, providing the
requisite time series data inside the reserve and for a suitable control
site. Annual surveys (2006--2016) were carried out in late July -- early
August, performing a total of 242 and 245 transects in the reserve site
for fish and invertebrate surveys, respectively. Similar sampling effort
was applied to the control site, with 221 fish and 222 invertebrate
transects. Between 12 and 27 transects were performed in each site every
year.

Socioeconomic data were obtained from the National Commission for
Aquaculture and Fisheries (\emph{Comisión Nacional de Acuacultura y
Pesca}; CONAPESCA). The data contains species--level information on
monthly landings and income from nine species from 2000 to 2014. Data on
landings and income were aggregated by year and species, and adjusted by
the Consumer Price Index {[}42{]}. From the nine species available, we
selected as objective species those that contributed the most (88.27\%)
income from 2000 to 2014: lobster (\emph{Panulirus interruptus};
71.76\%), red sea urchin (\emph{Mesocentrotus franciscanus}; 9.33\%),
snail (\emph{Megastraea undosa}; 3.93\%), and sea cucumber
(\emph{Parastichopus parvimensis}; 3.23\%). Abalone species
(\emph{Haliotis fulgens}; 4.52\% and \emph{Haliotis corrugata}; 6.16\%)
were excluded because the cooperative implemented an informal closure of
these fisheries in 2010 to allow the population recover. Eliminating all
fishing pressure on abalones means that the control site receives (for
this species) the same treatment as the reserve.

We constructed the governance data based on local knowledge of the area
and the community.

\section{Results from illustrative
example}\label{results-from-illustrative-example}

In this section we show the results of the application of MAREA to the
La Plana/Las Cuevas marine reserve in Isla Natividad, Mexico. These
results are intended to highlight the relevance and utility of the MAREA
framework and app, which automate the analysis and make it replicable.
While we highlight some of the general observed trends, we focus on the
utility of the tool rather than on the specific effectiveness of this
case study marine reserve.

The scorecard (Fig. 4) shows that this reserve achieves a general score
of 64\%, suggesting that 64\% of all indicators are positive. All
category--level scores were also high, with values of 67\%, 60\%, and
71\% positive indicators for biological, socioeconomic and governance,
respectively.

\textbf{Fig 4. Scorecard produced by MAREA for the ``La Plana / Las
Cuevas'' marine reserve in Isla Natividad, Mexico.}

Among the biological indicators, the greatest effect of the reserve was
observed for snail and sea cucumber densities, with values of
\(\beta_2 = 97.17\) (\emph{p} \textless{} 0.05) and \(\beta_2 = 2.31\)
(\emph{p} \textless{} 0.05), respectively. Fish indicators showed no
significant change (\emph{p} \textgreater{} 0.1), with negative trends
for Shannon's diversity index and fish species richness and positive
trends for density, biomass, and mean trophic level. Changes through
time for these indicators are presented in Figure 5, and a summary of
\(\beta_2\) coefficients is provided in Table 3.

\clearpage

\textbf{Fig. 5 Plots for values of each biological indicator (y-axis)
through time (x-axis).} Red and blue correspond to the reserve and
control sites, respectively. Black lines indicate yearly mean values,
and ribbons indicate \(\pm\) 1 standard error. Dots are horizontally
jittered to aid visualization. This figure contains information for fish
Shannon's diversity index (a), fish species richness (b), fish density
(c), fish trophic level (d), fish biomass (e), invertebrate Shannon's
diversity index (f), invertebrate species richness (g), invertebrate
density (h), lobster density (i), urchin density (j), snail density (k),
and sea cucumber density (l).

\begin{table}

\caption{\label{tab:table of bio results}Summary of average treatment effect of the reserve on biological indicators.}
\centering
\begin{tabular}[t]{l|l|r}
\hline
Indicator & Estimate (SD) & t-score\\
\hline
Shannon fish & -0.22 (0.16) & -1.3969\\
\hline
Richness fish & -0.61 (0.43) & -1.4073\\
\hline
Density fish & 0.74 (6.15) & 0.1205\\
\hline
Trophic fish & 0.00 (0.01) & 0.1399\\
\hline
Biomass fish & 0.22 (1.47) & 0.1476\\
\hline
Shannon invert & -0.67 (0.22)** & -3.0481\\
\hline
Richness invert & -2.71 (0.81)** & -3.3519\\
\hline
Density invert & 91.21 (47.11)* & 1.9362\\
\hline
Lobster & 7.66 (8.93) & 0.8583\\
\hline
Urchin & 2.15 (1.23)* & 1.7425\\
\hline
Snail & 97.17 (42.90)** & 2.2652\\
\hline
Cucumber & 2.31 (1.17)** & 1.9782\\
\hline
\multicolumn{3}{l}{\textsuperscript{*} Asterisks indicate significance level, with (*)}\\
\multicolumn{3}{l}{indicating p < 0.1 and (**) p < 0.05.}\\
\end{tabular}
\end{table}

One of the main objectives of this reserve was to increase landings.
Results of the socioeconomic indicators show that total landings were,
on average, 64.20 metric tonnes higher (\emph{p} \textgreater{} 0.1)
after the implementation of the reserves, though this cannot necessarily
be interpreted as causal, because it relies entirely on a before--after
comparison. Total income was \$10,344.85 (\emph{p} \textless{} 0.05)
thousands of Mexican Pesos (K MXP) higher after the implementation of
the reserves. On average, lobster and sea cucumber landings increased,
while urchin and snail landings and income decreased. Figure 6 presents
the changes in these indicators through time, and Table 4 summarizes
these results.

\textbf{Fig. 6 Plots for values of each socioeconomic indicator (y-axis)
through time (x-axis).} Red and blue correspond to before and after the
implementation of the reserve, respectively. This figure contains
information for total landings (a), total income (b), lobster landings
(c), urchin landings (d), snail landings (e), sea cucumber landings (f),
lobster income (g), urchin income (h), snail income (i), and sea
cucumber income (j).

\begin{table}

\caption{\label{tab:table of soc results}Summary of differences in socioeconomic indicators before and after the implementation of the reserve.}
\centering
\begin{tabular}[t]{l|l|r}
\hline
Indicator & Estimate (SD) & t-score\\
\hline
Landings & 64.20 (90.07) & 0.7127\\
\hline
Income & 10344.85 (3982.20)** & 2.5978\\
\hline
Lobster landings & 7.37 (13.95) & 0.5281\\
\hline
Urchin landings & -30.00 (9.49)** & -3.1620\\
\hline
Snail landings & -69.53 (33.82)* & -2.0561\\
\hline
Cucumber landings & 9.34 (6.72) & 1.3906\\
\hline
Lobster income & 14372.85 (3634.64)** & 3.9544\\
\hline
Urchin income & -5800.46 (1867.50)** & -3.1060\\
\hline
Snail income & -404.85 (187.07)** & -2.1641\\
\hline
Cucumber income & 131.49 (185.66) & 0.7082\\
\hline
\multicolumn{3}{l}{\textsuperscript{*} Asterisks indicate significance level, with (*)}\\
\multicolumn{3}{l}{indicating p < 0.1 and (**) p < 0.05.}\\
\end{tabular}
\end{table}

Recall that the governance objectives are evaluated based on the
institutions present, not on a specific quantitative linkage between
governance and biological or economic outcomes. Data for this reserve
suggest that the community is strongly organized, which is a likely
driver of the successes reported above {[}55{]}. The first point of
success is the existence of a fishing cooperative that is also
affiliated with a regional federation of cooperatives. These polycentric
governance structures allow various levels of organization that have
been shown to foster communication and cooperation {[}43,50{]};
federations also provide bargain power with governments {[}43,56{]}.
Access to fishing resources is managed through a TURF, permits, and
fishing quotas (for some species). McCay {[}49{]} suggests that the TURF
promotes a sense of stewardship of their resources and incentivizes
sustainable management. Together, these structures enabled a
participative, bottom--up process during the reserve design phase;
opinions of all fishing members ---and often non--fishing community
members--- were included. Participation of community members in reserve
surveillance and yearly monitoring indicate commitment and interest, and
allow informal communication of results to un--involved community
members. Furthermore, the reserve is partially isolated from poaching
activity and fishers have internal regulations pertaining to the
reserves. The low level of illegal fishing by members of the community
and outsiders both inside and outside the reserve represents another
indication of effectiveness. Governance indicators are summarized in
Table 5.

\begin{table}

\caption{\label{tab:unnamed-chunk-6}Summary of governance indicators.}
\centering
\begin{tabular}[t]{l|>{\raggedright\arraybackslash}p{9cm}}
\hline
Indicator & Description\\
\hline
Access to the fishery & Permits, Territorial Use Rights for Fisheries, Quotas (for some fisheries)\\
\hline
Number of fishers & Stable\\
\hline
Legal recognition of reserve & Not recognized\\
\hline
Reserve type & Community-based Marine Reserve\\
\hline
Illegal harvesting & Due to its relative isolations, neither the reserve or TURF suffer from significant illegal harvesting\\
\hline
Management plan & The reserve does not have a management plan, but written rules exist within the cooperative\\
\hline
Reserve enforcement & Fishers have two land stations equipped with radars and patrol boats 24/7 to patrol the reserves.\\
\hline
Size of reserve & The reserve is big enough to protect the targeted sessile or not highly mobile invertebrates (lobster, urchin, snail, cucumber, and abalone)\\
\hline
Reasoning for reserve location & The reserves were put in place in zones that, according to local knowledge, were once very productive. Habitat heterogeneity and ease of monitoring, surveillance and enforcement were also considered.\\
\hline
Membership to fisher organizations & The fishers are part of fisher organizations.\\
\hline
Type of fisheries organizations & The fishers are part of a cooperative (S.C.P.P. Buzos y Pescadores de la Baja California) and are affiliated to a federation (FEDECOOP).\\
\hline
Representation & Reserves were designed by fishers in a bottom-up approach, incorporating expertise from academics and CSO members. This was a highly inclusive and participatory process.\\
\hline
Internal Regulation & Fishers have stringent internal regulations to control fishing effort throughout their TURF, assigning different fishing zones and gears to different teams. Rules pertaining the marine reserves also exist.\\
\hline
Perceived Effectiveness & The fishers have a positive perception about the effectiveness of their reserve, often stating that they have seen significant economic benefits.\\
\hline
Social Impact of Reserve & The reserves have had a significant positive social impact. Fishers are proud to be an example of successgul marine conservation, allowing them to have increased social capital.\\
\hline
\end{tabular}
\end{table}

\clearpage

\section{Discussion}\label{discussion}

We have developed and presented a user--friendly, automated approach for
evaluating the effectiveness of marine reserves around the world. Here
we highlight MAREA's utility for evidence---based management, and
comment on a few of its shortcomings. The findings from Isla Natividad
are used purely to validate the relevance of MAREA rather than to
discuss particularities of the marine reserve effectiveness, which has
been described before {[}9,11,53{]}. We use examples from the case study
to build on the utility of MAREA and discuss ways in which results can
be interpreted to inform management.

The causal inference techniques used by MAREA have been suggested
{[}37,41{]} and used {[}20{]} before in other ad hoc studies. This
approach reduces ambiguity in the interpretation of results. For
example, invertebrate density decreased through time inside and outside
of the reserve (Fig. 5h). In this case, a before--after evaluation of
the reserve (\emph{i.e.} ignoring the control site) would have
incorrectly concluded that the reserve failed to protect invertebrates.
On the other hand, a control--impact approach (\emph{i.e.} compare
reserve vs.~control site only in 2016) would have identified higher
densities inside the reserve, concluding that the reserve increases
invertebrate density. However, by executing a formal
difference--in--differences approach for causal inference, MAREA
identifies the changes through time and across sites, and estimates the
effect of the reserve on density at \(\beta_2 = 91.21\) (\emph{p}
\textless{} 0.05). This approach reveals that invertebrate densities
decrease in both sites through time, but the decrease is faster for the
control site, thus yielding a positive value for \(\beta_2\).

The approach used by MAREA to estimate the effect of the reserve on
biological indicators requires cautious interpretation of the results.
The value of the \(\beta_2\) coefficient represents the difference
between the temporal trends of the reserve and control sites {[}20{]}.
As exemplified by the case of invertebrate densities, a positive value
(\emph{i.e.} \(\beta_2 > 0\)) does not necessarily indicate an increase
in the indicator through time, but rather a positive difference with
respect to the temporal trend of the control site. The inverse occurs
for negative values of \(\beta_2\).

MAREA provides in--depth analysis and a convenient snapshot overview of
the effect of the reserve, allowing users to rapidly identify trends.
However, users must interpret multiple indicators at a time to better
understand the results. For example, with additional knowledge of local
environmental variability (\emph{i.e.} indicator B5: Natural
Disturbance), we can better understand the trends in invertebrate
densities. As reported before {[}9{]}, hypoxic conditions that have
occurred in Isla Natividad can cause decreases in invertebrate
densities, and reserves buffer the negative effect. While MAREA
automates the analysis and makes results replicable, proper
interpretation will still depend on the user. Results produced by MAREA
can only aid in management and decision making when results have been
correctly interpreted.

Socioeconomic and governance indicators typically lack a control site,
which impede us from using the causal inference techniques employed to
measure biological changes {[}24{]}. However, we can still extract
useful information from them. Again, by combining results from multiple
indicators, MAREA can provide insights into the effect of the reserve.
For example, lobster and sea cucumber have shown increases in densities,
landings, and income. We cannot conclude that landings and income from
these species have increased due to the reserve, but we can at least
conclude that landings have not decreased. While further information on
market behavior of each fishery is needed, these results provide
insights into the state of the reserve and its associated fisheries.

As for the governance information, it is difficult to establish causal
links between the state of the reserve and the governance structures
present in the community. However, providing a single platform
(\emph{i.e.} scorecard) or document (\emph{i.e.} technical report) where
biological, socioeconomic, and governance information is comprehensively
included can aid in management. By using MAREA, this information will be
reported across reserves in a standardized way, and can help managers
identify overarching patterns across sites.

By making results straightforward to interpret, MAREA may also assist in
communication with a broader stakeholder community. While stakeholder
involvement in the design and implementation phases of marine reserves
is important, that may not be sufficient for ensuring long--run buy--in
or success. The scorecard is easily understandable by experts and
non--experts, and can be used as an effective tool for communicating the
results of annual evaluations. Additionally, the technical report can
serve as a tool for managers and scientists to rapidly produce and
communicate information at a more technical level.

We recognize that the 29 indicators used by MAREA might not fully
describe a reserve. However, they provide a starting point to perform
the evaluation, to which managers and users can add other indicators
(\emph{e.g.} larval dispersal or connectivity) that are relevant to
their reserve. Furthermore, MAREA's value is that it provides a free,
simple, and replicable way to perform rigorous impact analysis. The tool
can easily be used by fishers, CSO members, and managers in government
agencies, providing transparency of the analysis and results. In
addition, it can empower and enable local managers and fishers to
respond to local change and adapt by allowing direct and easy access to
the information.

The effectiveness of marine reserves continues to be a matter of debate
{[}7,12,40{]}. With current targets set to increase ocean protection, it
is important that we understand the effects of our interventions
{[}37{]} so we can better inform management {[}41{]}. It is therefore
important that academics, managers, fishers, and CSOs have access to
open access tools like MAREA. This is particularly relevant for Mexico
and other Latin American countries, where management agencies are often
understaffed and underfunded {[}57{]}, or where materials are often not
available in their language. In this context, MAREA provides a simple
and replicable way to align management objectives with performance
indicators. The proposed methodologies, especially the way in which
biological indicators are evaluated, provide valuable information for
managers. We acknowledge there is room for improvement in the way in
which socioeconomic and governance data are analyzed. Despite this,
providing a unifying platform where all indicators can be analyzed and
comprehensively presented represents a valuable step towards effective
evidence---based management {[}41{]}.

The first release of MAREA is now available, and it will continue to be
developed and maintained to keep up to date with the literature. This
process will incorporate new features, and enhance current ones, aiming
to improve user experience and expand the scope of the analysis. Yet, we
believe that this first release represents a major step towards
effective, replicable evaluation and management of marine reserves.

\section{Acknowledgements}\label{acknowledgements}

We thank Olivier Deschenes and Andrew Plantinga, who provided valuable
input to design the model that evaluates the biological indicators.
Special thanks to the fishers from Isla Natividad, who gathered the data
used in this study, and the fishers from El Rosario, who helped us
validate our survey and framework, and to Arturo Hernández and Alfonso
Romero who provided help with the logistics.

\clearpage

\section{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-pauly_2005-qV}{}
1. Pauly D, Watson R, Alder J. Global trends in world fisheries: Impacts
on marine ecosystems and food security. Philos Trans R Soc Lond, B, Biol
Sci. 2005;360: 5--12.
doi:\href{https://doi.org/10.1098/rstb.2004.1574}{10.1098/rstb.2004.1574}

\hypertarget{ref-halpern_2008-dK}{}
2. Halpern BS, Walbridge S, Selkoe KA, Kappel CV, Micheli F, D'Agrosa C,
et al. A global map of human impact on marine ecosystems. Science.
2008;319: 948--952.
doi:\href{https://doi.org/10.1126/science.1149345}{10.1126/science.1149345}

\hypertarget{ref-lester_2008-F_}{}
3. Lester S, Halpern B. Biological responses in marine no-take reserves
versus partially protected areas. Mar Ecol Prog Ser. 2008;367: 49--56.
doi:\href{https://doi.org/10.3354/meps07599}{10.3354/meps07599}

\hypertarget{ref-lester_2009-Ks}{}
4. Lester S, Halpern B, Grorud-Colvert K, Lubchenco J, Ruttenberg B,
Gaines S, et al. Biological effects within no-take marine reserves: A
global synthesis. Mar Ecol Prog Ser. 2009;384: 33--46.
doi:\href{https://doi.org/10.3354/meps08029}{10.3354/meps08029}

\hypertarget{ref-sala_2016-PV}{}
5. Sala E, Costello C, De Bourbon Parme J, Fiorese M, Heal G, Kelleher
K, et al. Fish banks: An economic model to scale marine conservation.
Marine Policy. 2016;73: 154--161.
doi:\href{https://doi.org/10.1016/j.marpol.2016.07.032}{10.1016/j.marpol.2016.07.032}

\hypertarget{ref-hastings_2017-sm}{}
6. Hastings A, Gaines SD, Costello C. Marine reserves solve an important
bycatch problem in fisheries. Proc Natl Acad Sci U S A. 2017;
doi:\href{https://doi.org/10.1073/pnas.1705169114}{10.1073/pnas.1705169114}

\hypertarget{ref-padleton_2017-vn}{}
7. Padleton L, Aghmadia G, Browman H, Thurstand R, Kaplan D, Bartolino
V. Debating the effectiveness of marine protected areas. ICES Journal of
Marine Science. 2017;
doi:\href{https://doi.org/10.1093/icesjms/fsx154}{10.1093/icesjms/fsx154}

\hypertarget{ref-aburtooropeza_2011-ya}{}
8. Aburto-Oropeza O, Erisman B, Galland GR, Mascareñas-Osorio I, Sala E,
Ezcurra E. Large recovery of fish biomass in a no-take marine reserve.
PLoS ONE. 2011;6: e23601.
doi:\href{https://doi.org/10.1371/journal.pone.0023601}{10.1371/journal.pone.0023601}

\hypertarget{ref-micheli_2012-EU}{}
9. Micheli F, Saenz-Arroyo A, Greenley A, Vazquez L, Espinoza Montes JA,
Rossetto M, et al. Evidence that marine reserves enhance resilience to
climatic impacts. PLoS ONE. 2012;7: e40832.
doi:\href{https://doi.org/10.1371/journal.pone.0040832}{10.1371/journal.pone.0040832}

\hypertarget{ref-roberts_2017-J9}{}
10. Roberts CM, O'Leary BC, McCauley DJ, Cury PM, Duarte CM, Lubchenco
J, et al. Marine reserves can mitigate and promote adaptation to climate
change. Proc Natl Acad Sci U S A. 2017;114: 6167--6175.
doi:\href{https://doi.org/10.1073/pnas.1701262114}{10.1073/pnas.1701262114}

\hypertarget{ref-munguavega_2015-yg}{}
11. Munguía-Vega A, Sáenz-Arroyo A, Greenley AP, Espinoza-Montes JA,
Palumbi SR, Rossetto M, et al. Marine reserves help preserve genetic
diversity after impacts derived from climate variability: Lessons from
the pink abalone in baja california. Global Ecology and Conservation.
2015;4: 264--276.
doi:\href{https://doi.org/10.1016/j.gecco.2015.07.005}{10.1016/j.gecco.2015.07.005}

\hypertarget{ref-edgar_2014-UO}{}
12. Edgar GJ, Stuart-Smith RD, Willis TJ, Kininmonth S, Baker SC, Banks
S, et al. Global conservation outcomes depend on marine protected areas
with five key features. Nature. 2014;506: 216--220.
doi:\href{https://doi.org/10.1038/nature13022}{10.1038/nature13022}

\hypertarget{ref-giakoumi_2017-V2}{}
13. Giakoumi S, Scianna C, Plass-Johnson J, Micheli F, Grorud-Colvert K,
Thiriet P, et al. Ecological effects of full and partial protection in
the crowded mediterranean sea: A regional meta-analysis. Sci Rep.
2017;7: 8940.
doi:\href{https://doi.org/10.1038/s41598-017-08850-w}{10.1038/s41598-017-08850-w}

\hypertarget{ref-sala_2017-69}{}
14. Sala E, Giakoumi S. No-take marine reserves are the most effective
protected areas in the ocean. ICES Journal of Marine Science. 2017;
doi:\href{https://doi.org/10.1093/icesjms/fsx059}{10.1093/icesjms/fsx059}

\hypertarget{ref-betti_2017-lq}{}
15. Betti F, Bavestrello G, Bo M, Asnaghi V, Chiantore M, Bava S, et al.
Over 10 years of variation in mediterranean reef benthic communities.
Marine Ecology. 2017;38: e12439.
doi:\href{https://doi.org/10.1111/maec.12439}{10.1111/maec.12439}

\hypertarget{ref-davies_2017-ml}{}
16. Davies TK, Mees CC, Milner-Gulland EJ. Use of a counterfactual
approach to evaluate the effect of area closures on fishing location in
a tropical tuna fishery. PLoS ONE. 2017;12: e0174758.
doi:\href{https://doi.org/10.1371/journal.pone.0174758}{10.1371/journal.pone.0174758}

\hypertarget{ref-guidetti_2014-8Z}{}
17. Guidetti P, Baiata P, Ballesteros E, Di Franco A, Hereu B,
Macpherson E, et al. Large-scale assessment of mediterranean marine
protected areas effects on fish assemblages. PLoS ONE. 2014;9: e91841.
doi:\href{https://doi.org/10.1371/journal.pone.0091841}{10.1371/journal.pone.0091841}

\hypertarget{ref-friedlander_2017-oI}{}
18. Friedlander AM, Golbuu Y, Ballesteros E, Caselle JE, Gouezo M,
Olsudong D, et al. Size, age, and habitat determine effectiveness of
palau's marine protected areas. PLoS ONE. 2017;12: e0174787.
doi:\href{https://doi.org/10.1371/journal.pone.0174787}{10.1371/journal.pone.0174787}

\hypertarget{ref-rodriguez_2017-PD}{}
19. Rodriguez AG, Fanning LM. Assessing marine protected areas
effectiveness: A case study with the tobago cays marine park. OJMS.
2017;07: 379--408.
doi:\href{https://doi.org/10.4236/ojms.2017.73027}{10.4236/ojms.2017.73027}

\hypertarget{ref-moland_2013-VP}{}
20. Moland E, Olsen EM, Knutsen H, Garrigou P, Espeland SH, Kleiven AR,
et al. Lobster and cod benefit from small-scale northern marine
protected areas: Inference from an empirical before-after control-impact
study. Proc Biol Sci. 2013;280: 20122679.
doi:\href{https://doi.org/10.1098/rspb.2012.2679}{10.1098/rspb.2012.2679}

\hypertarget{ref-soykan_2015-nu}{}
21. Soykan CU, Lewison RL. Using community-level metrics to monitor the
effects of marine protected areas on biodiversity. Conserv Biol.
2015;29: 775--783.
doi:\href{https://doi.org/10.1111/cobi.12445}{10.1111/cobi.12445}

\hypertarget{ref-pomeroy_2005-Py}{}
22. Pomeroy RS, Watson LM, Parks JE, Cid GA. How is your mpa doing? A
methodology for evaluating the management effectiveness of marine
protected areas. Ocean Coast Manag. 2005;48: 485--502.
doi:\href{https://doi.org/10.1016/j.ocecoaman.2005.05.004}{10.1016/j.ocecoaman.2005.05.004}

\hypertarget{ref-pomeroy_2004-23}{}
23. Pomeroy RS, Parks JE, Watson LM. How is your mpa doing ? A guidebook
of natural and social indicators for evaluating marine protected areas
management effectiveness {[}Internet{]}. IUCN; 2004.
doi:\href{https://doi.org/10.2305/IUCN.CH.2004.PAPS.1.en}{10.2305/IUCN.CH.2004.PAPS.1.en}

\hypertarget{ref-mascia_2017-m_}{}
24. Mascia MB, Fox HE, Glew L, Ahmadia GN, Agrawal A, Barnes M, et al. A
novel framework for analyzing conservation impacts: Evaluation, theory,
and marine protected areas. Ann N Y Acad Sci. 2017;1399: 93--115.
doi:\href{https://doi.org/10.1111/nyas.13428}{10.1111/nyas.13428}

\hypertarget{ref-ostrom_2009-hg}{}
25. Ostrom E. A general framework for analyzing sustainability of
social-ecological systems. Science. 2009;325: 419--422.
doi:\href{https://doi.org/10.1126/science.1172133}{10.1126/science.1172133}

\hypertarget{ref-basurto_2013-oq}{}
26. Basurto X, Gelcich S, Ostrom E. The social--ecological system
framework as a knowledge classificatory system for benthic small-scale
fisheries. Global Environmental Change. 2013;23: 1366--1380.
doi:\href{https://doi.org/10.1016/j.gloenvcha.2013.08.001}{10.1016/j.gloenvcha.2013.08.001}

\hypertarget{ref-lowndes_2017-xh}{}
27. Lowndes JSS, Best BD, Scarborough C, Afflerbach JC, Frazier MR,
O'Hara CC, et al. Our path to better science in less time using open
data science tools. Nat ecol evol. 2017;1: 0160.
doi:\href{https://doi.org/10.1038/s41559-017-0160}{10.1038/s41559-017-0160}

\hypertarget{ref-halpern_2012-k9}{}
28. Halpern BS, Longo C, Hardy D, McLeod KL, Samhouri JF, Katona SK, et
al. An index to assess the health and benefits of the global ocean.
Nature. 2012;488: 615--620.
doi:\href{https://doi.org/10.1038/nature11397}{10.1038/nature11397}

\hypertarget{ref-halpern_2017-Zi}{}
29. Halpern BS, Frazier M, Afflerbach J, O'Hara C, Katona S, Stewart
Lowndes JS, et al. Drivers and implications of change in global ocean
health over the past five years. PLoS ONE. 2017;12: e0178267.
doi:\href{https://doi.org/10.1371/journal.pone.0178267}{10.1371/journal.pone.0178267}

\hypertarget{ref-selig_2015-F9}{}
30. Selig ER, Frazier M, O'Leary JK, Jupiter SD, Halpern BS, Longo C, et
al. Measuring indicators of ocean health for an island nation: The ocean
health index for fiji. Ecosystem Services. 2015;16: 403--412.
doi:\href{https://doi.org/10.1016/j.ecoser.2014.11.007}{10.1016/j.ecoser.2014.11.007}

\hypertarget{ref-halpern_2014-lQ}{}
31. Halpern BS, Longo C, Scarborough C, Hardy D, Best BD, Doney SC, et
al. Assessing the health of the u.S. west coast with a regional-scale
application of the ocean health index. PLoS ONE. 2014;9: e98995.
doi:\href{https://doi.org/10.1371/journal.pone.0098995}{10.1371/journal.pone.0098995}

\hypertarget{ref-elfes_2014-RC}{}
32. Elfes CT, Longo C, Halpern BS, Hardy D, Scarborough C, Best BD, et
al. A regional-scale ocean health index for brazil. PLoS ONE. 2014;9:
e92589.
doi:\href{https://doi.org/10.1371/journal.pone.0092589}{10.1371/journal.pone.0092589}

\hypertarget{ref-anderson_2015-ND}{}
33. Anderson JL, Anderson CM, Chu J, Meredith J, Asche F, Sylvia G, et
al. The fishery performance indicators: A management tool for triple
bottom line outcomes. PLoS ONE. 2015;10: e0122809.
doi:\href{https://doi.org/10.1371/journal.pone.0122809}{10.1371/journal.pone.0122809}

\hypertarget{ref-dowling_2016-pO}{}
34. Dowling N, Wilson J, Rudd M, Babcock E, Caillaux M, Cope J, et al.
FishPath: A decision support system for assessing and managing data- and
capacity- limited fisheries. In: Quinn II T, Armstrong J, Baker M,
Heifetz J, Witherell D, editors. Assessing and managing data-limited
fish stocks. Alaska Sea Grant, University of Alaska Fairbansk; 2016.
doi:\href{https://doi.org/10.4027/amdlfs.2016.03}{10.4027/amdlfs.2016.03}

\hypertarget{ref-oyanedel_2017-TO}{}
35. Oyanedel R, Macy Humberstone J, Shattenkirk K, Rodriguez Van-Dyck S,
Joye Moyer K, Poon S, et al. A decision support tool for designing
turf-reserves. BMS. 2017;93: 155--172.
doi:\href{https://doi.org/10.5343/bms.2015.1095}{10.5343/bms.2015.1095}

\hypertarget{ref-vilela_2017-Zo}{}
36. Vilela T, Reid J. Improving hydropower choices via an online and
open access tool. PLoS ONE. 2017;12: e0179393.
doi:\href{https://doi.org/10.1371/journal.pone.0179393}{10.1371/journal.pone.0179393}

\hypertarget{ref-burgess_2016-HN}{}
37. Burgess MG, Clemence M, McDermott GR, Costello C, Gaines SD. Five
rules for pragmatic blue growth. Marine Policy. 2016;
doi:\href{https://doi.org/10.1016/j.marpol.2016.12.005}{10.1016/j.marpol.2016.12.005}

\hypertarget{ref-nom049sagpesc_2014-V6}{}
38. NOM-049-SAG/PESC. NORMA oficial mexicana nom-049-sag/pesc-2014, que
determina el procedimiento para establecer zonas de refugio para los
recursos pesqueros en aguas de jurisdicción federal de los estados
unidos mexicanos. DOF. 2014;

\hypertarget{ref-lgeepa_2017-jL}{}
39. LGEEPA. Ley general del equilibrio ecológico y la protección al
ambiente. DOF. 2017; Available:
\url{http://www.diputados.gob.mx/LeyesBiblio/pdf/148/_240117.pdf}

\hypertarget{ref-woodcock_2017-Wm}{}
40. Woodcock P, O'Leary BC, Kaiser MJ, Pullin AS. Your evidence or mine?
Systematic evaluation of reviews of marine protected area effectiveness.
Fish Fish. 2017;18: 668--681.
doi:\href{https://doi.org/10.1111/faf.12196}{10.1111/faf.12196}

\hypertarget{ref-ferraro_2006-oW}{}
41. Ferraro PJ, Pattanayak SK. Money for nothing? A call for empirical
evaluation of biodiversity conservation investments. PLoS Biol. 2006;4:
e105.
doi:\href{https://doi.org/10.1371/journal.pbio.0040105}{10.1371/journal.pbio.0040105}

\hypertarget{ref-oecd_website_2017-VV}{}
42. OECD. Prices - inflation (cpi) - oecd data {[}Internet{]}. 2017.
Available: \url{https://data.oecd.org/price/inflation-cpi.htm}

\hypertarget{ref-espinosaromero_2014-PY}{}
43. Espinosa-Romero MJ, Rodriguez LF, Weaver AH, Villanueva-Aznar C,
Torre J. The changing role of ngos in mexican small-scale fisheries:
From environmental conservation to multi-scale governance. Marine
Policy. 2014;50: 290--299.
doi:\href{https://doi.org/10.1016/j.marpol.2014.07.005}{10.1016/j.marpol.2014.07.005}

\hypertarget{ref-rcore_2017}{}
44. R Core Team. R: A language and environment for statistical computing
{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
2017. Available: \url{https://www.R-project.org/}

\hypertarget{ref-shiny_2017}{}
45. Chang W, Cheng J, Allaire J, Xie Y, McPherson J. Shiny: Web
application framework for r {[}Internet{]}. 2017. Available:
\url{https://CRAN.R-project.org/package=shiny}

\hypertarget{ref-shinydashboard_2017}{}
46. Chang W, Borges Ribeiro B. Shinydashboard: Create dashboards with
'shiny' {[}Internet{]}. 2017. Available:
\url{https://CRAN.R-project.org/package=shinydashboard}

\hypertarget{ref-rmarkdown_2017}{}
47. Allaire J, Cheng J, Xie Y, McPherson J, Chang W, Allen J, et al.
Rmarkdown: Dynamic documents for r {[}Internet{]}. 2017. Available:
\url{https://CRAN.R-project.org/package=rmarkdown}

\hypertarget{ref-knitr_2017}{}
48. Xie Y. Knitr: A general-purpose package for dynamic report
generation in r {[}Internet{]}. 2017. Available:
\url{http://yihui.name/knitr/}

\hypertarget{ref-mccay_2017-1m}{}
49. McCay B. Territorial use rights in fisheries of the northern pacific
coast of mexico. BMS. 2017;93: 69--81.
doi:\href{https://doi.org/10.5343/bms.2015.1091}{10.5343/bms.2015.1091}

\hypertarget{ref-mccay_2014-CN}{}
50. McCay BJ, Micheli F, Ponce-Díaz G, Murray G, Shester G,
Ramirez-Sanchez S, et al. Cooperatives, concessions, and co-management
on the pacific coast of mexico. Marine Policy. 2014;44: 49--59.
doi:\href{https://doi.org/10.1016/j.marpol.2013.08.001}{10.1016/j.marpol.2013.08.001}

\hypertarget{ref-afflerbach_2014-HP}{}
51. Afflerbach JC, Lester SE, Dougherty DT, Poon SE. A global survey of
-reserves, territorial use rights for fisheries coupled with marine
reserves. Global Ecology and Conservation. 2014;2: 97--106.
doi:\href{https://doi.org/10.1016/j.gecco.2014.08.001}{10.1016/j.gecco.2014.08.001}

\hypertarget{ref-lester_2017-nh}{}
52. Lester S, McDonald G, Clemence M, Dougherty D, Szuwalski C. Impacts
of turfs and marine reserves on fisheries and conservation goals:
Theory, empirical evidence, and modeling. BMS. 2017;93: 173--198.
doi:\href{https://doi.org/10.5343/bms.2015.1083}{10.5343/bms.2015.1083}

\hypertarget{ref-rossetto_2015-V0}{}
53. Rossetto M, Micheli F, Saenz-Arroyo A, Montes JAE, De Leo GA, Rochet
M-J. No-take marine reserves can enhance population persistence and
support the fishery of abalone. Can J Fish Aquat Sci. 2015;72:
1503--1517.
doi:\href{https://doi.org/10.1139/cjfas-2013-0623}{10.1139/cjfas-2013-0623}

\hypertarget{ref-suman_2010-ez}{}
54. Suman CS, Saenz-Arroyo A, Dawson C, Luna MC. Manual de instruccion
de reef check california: Guia de instruccion para el monitoreo del
bosque de sargazo en la peninsula de baja california. Pacific Palisades,
CA, USA: Reef Check Foundation; 2010.

\hypertarget{ref-gutirrez_2011-0U}{}
55. Gutiérrez NL, Hilborn R, Defeo O. Leadership, social capital and
incentives promote successful fisheries. Nature. 2011;470: 386--389.
doi:\href{https://doi.org/10.1038/nature09689}{10.1038/nature09689}

\hypertarget{ref-finkbeiner_2015-87}{}
56. Finkbeiner EM, Basurto X. Re-defining co-management to facilitate
small-scale fisheries reform: An illustration from northwest mexico.
Marine Policy. 2015;51: 433--441.
doi:\href{https://doi.org/10.1016/j.marpol.2014.10.010}{10.1016/j.marpol.2014.10.010}

\hypertarget{ref-lundquist_2005-OL}{}
57. Lundquist CJ, Granek EF. Strategies for successful marine
conservation: Integrating socioeconomic, political, and scientific
factors. Conserv Biol. 2005;19: 1771--1778.
doi:\href{https://doi.org/10.1111/j.1523-1739.2005.00279.x}{10.1111/j.1523-1739.2005.00279.x}

\section{Supporting information}\label{supporting-information}

\textbf{S1 Appendix. Survey to collect governance information from
fishing communities.} English version

\textbf{S2 Appendix. Survey to collect governance information from
fishing communities.} Spanish version

\textbf{S3 Table. Assigned values and reasoning of socioeconomic and
governance indicators used to color-code the scorecard in MAREA}

\end{document}

### 2017-12-03 Manuscript.Rmd
Incorporated changes from reviewers

@@
 
-Available empirical evidence on marine reserve effectiveness is mixed [@padleton_2017-vn].
-Some studies have shown that MPAs increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience to climatic impacts [@micheli_2012-EU; @roberts_2017-J9], and preserve genetic diversity [@munguavega_2015-yg].
-Compared to partially protected MPAs, marine reserves have higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO; @giakoumi_2017-V2; @sala_2017-69].
-These effects are often measured as biological changes in the area through time and many lack a control site for comparison [@betti_2017-lq].
-This approach does not account for other factors for which one must control [@davies_2017-ml] in order to causally attribute a biological change to the reserve.
-While some studies have used control sites, these analyses do not estimate the effect of the reserve, and often use a control--impact comparison approach that does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
+Empirical evidence shows that MPAs increase biomass [@aburtooropeza_2011-ya; @lester_2009-Ks], enhance resilience to climatic impacts [@micheli_2012-EU; @roberts_2017-J9], and preserve genetic diversity [@munguavega_2015-yg].
+Compared to MPAs that grant partial protection, marine reserves have higher levels of biomass, density, richness, and larger organisms [@lester_2008-F_; @edgar_2014-UO; @giakoumi_2017-V2; @sala_2017-69].
+However, these effects are often measured as biological changes within the reserves through time, and many lack a control site for comparison [@betti_2017-lq].
+This approach does not account for other factors (*e.g.* system-level changes in productivity ; @szuwalski_2017-jc) for which one must control [@davies_2017-ml] in order to causally attribute a biological change to the reserve.
+Other studies have used a control--impact comparison approach that uses control sites but does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
 A smaller fraction of studies have used a before--after--control--impact (*i.e.* BACI) design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.
+But even when proper causal inference can be drawn, results are often different across reserves.
+Effects of reserves on ecological and economic outcomes are highly heterogeneous, and often depend on the specific ecological, economic, and social context.
 
-But even when proper causal inference can be drawn, results are often idiosyncratic across reserves.
-Effects of reserves on ecological and economic outcomes are highly heterogeneous, and often depend on the specific ecological, economic, and social context.
-The purpose of this paper is to describe a user--friendly tool, called “MAREA”, to rigorously systematize the evaluation of marine reserve effectiveness.
-The tool is in the form of an open-source application that uses state--of--the--art methods from program evaluation to compare a reserve to control sites along a number of ecological, economic, and governance dimensions.
-
-The challenge of how to standardize marine reserve evaluation is not a new one.
-The recent IUCN framework “How is your MPA doing?” [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biological, socioeconomic, and governance indicators, and insights into how these indicators may be measured.
-But this framework stops short of analysis, so provides a user with little guidance about establishing causal inference about the reserve.
-Recent work by Mascia *et al.* [@mascia_2017-m_] integrates these three dimensions via the Social Ecological Systems Framework [@ostrom_2009-hg; @basurto_2013-oq] and suggests the use of causal inference techniques to provide a measure of the effect of conservation interventions.
-However, these two novel approaches do not provide a user-friendly tool that enables replicability and scalability of the analysis, particularly when used by the fishers and decision makers themselves.
+Standardization of marine reserve evaluation is not new.
+The IUCN framework “How is your MPA doing?” [@pomeroy_2005-Py; @pomeroy_2004-23] provides a comprehensive list of biological, socioeconomic, and governance indicators, and insights into how these may be measured or collected.
+But this framework stops short of analysis, and provides a user with little guidance about establishing causal inference about the reserve.
+Recent work [@mascia_2017-m_] integrates these three dimensions via the Social Ecological Systems Framework [@ostrom_2009-hg; @basurto_2013-oq] and suggests the use of causal inference techniques to provide a measure of the effect of conservation interventions.
+However, neither of these approaches provide a user-friendly tool that ensures replicability and scalability of the analysis, particularly when used by the fishers and decision makers themselves.
 
@@
 
-This paper presents a framework and user--friendly tool to evaluate marine reserve effectiveness, which incorporates the biological, socioeconomic, and governance dimensions of any fishery.
+The purpose of this paper is to describe a user--friendly tool, called “MAREA”, to rigorously systematize the evaluation of marine reserve effectiveness in terms of fisheries and marine conservation.
+The tool is in the form of an open--source application that uses state--of--the--art methods from program evaluation to compare a reserve to control sites along a number of ecological, economic, and governance dimensions.
 We first provide a list of commonly stated management objectives and match them to appropriate indicators.
 We then develop a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_].
-To implement the analytical approach in a user-friendly format, we introduce the Marine Reserve Evaluation Application (MAREA), an open source, web–based tool that automates the framework described in this paper and enables its broader use.
+To implement the analytical approach, we introduce the Marine Reserve Evaluation Application (MAREA), an open source, web–based tool that automates the framework described in this paper and enables its broader use.
 Finally, we present a case study on the evaluation of a marine reserve established by the fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.
@@
 Alongside, methodologies to analyze these indicators are presented.
-We then describe the development of MAREA and explain how this user--friendly open--access tool can be used by fishermen, managers, and other stakeholders with little scientific background.
+We then describe the development of MAREA and explain how this open--access tool can be used by fishermen, managers, and other stakeholders with little scientific background.
 Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA to inform management.
@@
 
-Throughout this study, we will refer to the stated goals for which a marine reserve was designed as “objectives.” This work was motivated by the case of Mexico, where 39 reserves have been implemented over the past five years to achieve objectives such as increasing productivity in nearby waters or recover overexploited species; most of these reserves have never been formally evaluated for effectiveness at meeting those objectives.
+Throughout this study, we will refer to the stated goals for which a marine reserve was designed as “objectives.” This work was motivated by the case of Mexico, where 39 reserves have been implemented over the past five years to achieve objectives such as increasing productivity in nearby waters or recovery of overexploited species; most of these reserves have never been formally evaluated for effectiveness at meeting those objectives.
 Thus, our focus was on identifying common objectives of marine reserves in Mexico.
-However, a literature review and discussions with marine reserve researchers worldwide suggest that the objectives driving Mexican marine reserve implementation are similar to those in the rest of the world.
-Thus, we group these objectives into seven major categories, which can be applied to marine reserves worldwide.
-The list of objectives includes stated objectives in legislation [@nom049sagpesc_2014-V6; @lgeepa_2017-jL] and official documents such as the Technical Justification Studies (*Estudios Técnicos Justificativos*), agreements, and decrees associated to these areas:
+However, a literature review and discussions with marine reserve researchers suggest that the objectives driving Mexican marine reserve implementation are similar to those in the rest of the world.
+Thus, we group these objectives into seven major categories, which may be applied to marine reserves worldwide.
+Any given reserve may have been implemented to meet one or more of these.
+The list includes objectives stated in legislation [@nom049sagpesc_2014-V6; @lgeepa_2017-jL] and official documents such as the Technical Justification Studies (*Estudios Técnicos Justificativos*), agreements, and decrees associated to these areas:
 
@@
 These indicators are specific variables on which data could be collected, and analyzed, to ultimately determine whether the corresponding objective was causally being achieved by the marine reserve.
-The list of indicators was compiled through a review of scientific literature in which we identified indicators that were used to measure similar objectives.
+The list of indicators was compiled through a review of scientific literature in which we identified indicators that were used to measure similar objectives[@lester_2017-nh; @sala_2017-69; @chirico_2017-Rz; @edgar_2014-UO; @rodriguez_2017-PD; @rossetto_2015-V0; @betti_2017-lq; @sala_2016-PV; @woodcock_2017-Wm; @friedlander_2017-oI; @moland_2013-VP; @aburtooropeza_2011-ya; @guidetti_2014-8Z; @lester_2008-F_; @lester_2009-Ks; @pomeroy_2005-Py; @pomeroy_2004-23].
 A first filter eliminated indicators for which baseline data do not typically exist in Mexico.
@@
 Fifteen governance indicators describe the governance structures under which the community operates (*e.g.*, access rights to the fishery, number of fishers, legal recognition of the reserve).
-Some indicators require a numerical entry (*e.g.* Fish biomass) while others are more descriptive (*e.g.* Reasoning for reserve location).
+Some indicators are quantitative and require a numerical entry (*e.g.* Fish biomass) while others are qualitative and rely on a descriptive entry (*e.g.* Reasoning for reserve location).
 Many of them specifically measure an outcome of the reserve, though some are designed to further the understanding of the mechanisms driving a reserve’s performance.
@@
 Whenever an indicator is applied to “Target species”, it means that the indicator can be used for all species (*e.g.* Fish Biomass) and/or for individual species that are either the conservation target  of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass).
+Finally, indicators B3 and B4 are different in that B3 look only at the density of organisms above size at first maturity (related to reproductive potential), while B4 measures the density of all fish or of a target species.
+Each indicator targets different plausible desired outcomes, like increased reproductive potential (*i.e.* B3; @carter_2017-Uf) or having more fish -regardless of their size- to atract tourism (*i.e.* B4).
 Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with biological and socioeconomic indicators.
+Governance indicators are excluded from Table 2, as we suggest that all of them should be used in an evaluation, regardless of the objectives.
 
@@
 
-```{r}
+```{r, results = "asis"}
 read.csv("./Tables/Objectives_Indicators.csv") %>% 
   select(1:15) %>% 
-  knitr::kable(caption = "Management objectives and respective performance indicators. All governance indicators should allways be used.",
+  knitr::kable(caption = "Management objectives and respective performance indicators.",
                col.names = c("Objective", "B1", "B2", "B3", "B4", "B4*", "B5", "B6", "B7", "B7*", "S1", "S1*", "S2", "S2*", "S3")) %>% 
@@
 
-In many coastal marine reserves of Mexico, biological data are collected via underwater visual censuses as part of a reserve’s monitoring program.
-Scientific divers (which are often local fishermen with guidance from Civil Society Organizations; CSOs) record fish and invertebrate richness and abundances, as well as fish total length along belt transects.
+In many coastal marine reserves, biological data are often collected via underwater visual censuses as part of a reserve’s monitoring program.
+Scientific divers record fish and invertebrate richness and abundances, as well as fish total length along belt transects.
 Ecological surveys are typically performed annually in each reserve and corresponding control site(s), before and after the implementation of the reserve, providing a sampling design that can be used to draw causal inference.
@@
 For this reason, MAREA does not assume specific transect dimensions, and pertinent indicators are calculated per transect (Table 1).
+More information on data collection and formatting is provided in a guidebook [@villaseorderbez_website_2017-xE], which is available in English and Spanish in MAREA's interface.
 
@@
 In this model, $i$, $t$, and $z$ are indices for transect, time, and zone (control or reserve site), respectively.
-This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), a dummy variable that indicates treatment ($Z$; *i.e.* control or reserve), an interaction between a dummy variable that indicates pre-- or post--implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$; in °C), horizontal visibility during the survey ($V$; in m), and depth at which survey was performed ($D$; in m).
+This model allows us to estimate the change in an indicator ($I$) based on the year ($Y$), a dummy variable that indicates treatment ($Z$; *i.e.* control or reserve), an interaction between a dummy variable that indicates before or after implementation ($P$) and treatment ($Z$), and covariates such as bottom temperature ($T$; in °C), horizontal visibility during the survey ($V$; in m), and depth at which survey was performed ($D$; in m).
 $\epsilon$ represents the error term associated to the regression.
@@
 This does not impose a linear structure in the way an indicator changes through time (*i.e.* the change in biomass between 2006 and 2007 does not have to be the same as the change between 2015 and 2016).
-The treatment and implementation variables, modeled as dummy variables, are coded as Control = 0 and Reserve = 1; and Pre--implementation = 0 and Post--implementation = 1, respectively.
+The treatment and implementation variables, modeled as dummy variables, are coded as Control = 0 and Reserve = 1; and Before implementation = 0 and After implementation = 1, respectively.
 
@@
 1 and eq.
-3), we estimate the model coeficients with ordinary least squares, and calculate heteroskedastic–robust standard errors.
+3), we estimate the model coefficient with ordinary least squares, and calculate heteroskedastic–robust standard errors [@sandwich_2014].
 
@@
 Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on Table 2.
-Users can also manually modify selected indicators based on their interests and data availability.
+Users can also select and un-select indicators based on their interests and data availability by "clicking" on the check-boxes in MAREA.
 The user can then load data on one or more reserves, using standard \*.csv text files; sample datasets are provided within MAREA.
@@
 General location of Isla Natividad (left) and map of the island (right).** The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce).
+Shapefiles for Mexican coastline and the US were obtained from the Instituto Nacional de Estadistica, Geografia e Informatica of Mexico (INEGI: www.inegi.org.mx/geo/contenidos/geoestadistica/m_geoestadistico.aspx) and the tmap R package [@tmap_2017], respectively.
 
@@
 #Small shapefile for region of interest
-xy1 <- data.frame(X = mean(-115.24, -115.14), Y = mean(27.85, 27.90))
-coordinates(xy1) <- c("X", "Y")
-proj4string(xy1) <- proj
-res1 <- spTransform(xy1, proj)
+res1 <- cbind(x_coord = c(-115.32, -115.32, -115.04, -115.04, -115.32), y_coord= c(27.75, 28, 28, 27.75, 27.75)) %>% 
+  Polygon() %>% 
+  list() %>% 
+  Polygons(srl = ., ID = 1) %>% 
+  list() %>% 
+  SpatialPolygons(proj4string = proj)
 
-# Reserve polygons
 reserves <- read.csv("./Data/Spatial/baja_coordinates.csv", stringsAsFactors = F) %>% 
@@
   tm_shape(res1) +
-  tm_symbols(shape = 0, size = 2, border.col = "black", col = "black", border.lwd = 2) +
+  tm_borders(col = "black", lwd = 2) +
   tm_shape(labels1) +
@@
   tm_scale_bar(position = c("left", "bottom"))
+
+# tmap_arrange(pen_map, study_area, outer.margins = 0.005, ncol = 2)
 ```
 
-In 2006, the Isla Natividad community implemented two community--based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh; @micheli_2012-EU].
+In 2006, the Isla Natividad community implemented two community--based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh; @micheli_2012-EU] after establishing a baseline for the soon--to--be reserves and control sites.
 Evidence suggest that these reserves have been effective at enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of high value commercial species such as abalone [@munguavega_2015-yg].
@@
 
-We have developed and presented a user--friendly, automated approach for evaluating the effectiveness of marine reserves around the world.
+We have developed and presented an automated approach for evaluating the effectiveness of marine reserves in Mexico, and perhaps around the world.
 Here we highlight MAREA’s utility for evidence–-based management, and comment on a few of its shortcomings.
@@
 
-We recognize that the 29 indicators used by MAREA might not fully describe a reserve.
-However, they provide a starting point to perform the evaluation, to which managers and users can add other indicators (*e.g.* larval dispersal or connectivity) that are relevant to their reserve.
+We recognize that the seven objectives and 29 indicators used by MAREA might not fully describe a reserve.
+In order to ensure the applicability of the tool to reserves in other countries, further testing in other regions should take place.
+However, the proposed objectives and indicators provide a starting point to perform the evaluation, to which managers and users can add other indicators (*e.g.* larval dispersal or connectivity) that are relevant to their reserve.
 Furthermore, MAREA’s value is that it provides a free, simple, and replicable way to perform rigorous impact analysis.
@@
 In addition, it can empower and enable local managers and fishers to respond to local change and adapt by allowing direct and easy access to the information.
+
+An evident limitation of MAREA is its dependence on data obtained through a BACI design, and the amount of samples needed to estimate coefficients in Eqn.
+1.
+It is not uncommon for control sites or baselines that describe the biological state of the area before the management intervention to be abscent.
+Propperly designing marine reserves by identifying control sites and establishing a baseline before the implementation of the reserve is enough to overcome this issue; reserves for which there is no control site and baseline can not be evaluated with MAREA.
+On the second case, typical underwater surveys require that at least 12 - 16 transects are performed on each site (*i.e.* reserve and control) each year.
+This provides with at least 48 samples (12 samples per site, per year), enough to avoid overfitting Eqn 1.
+On either case, these problems can be easily avoided by accounting for data requirements in the eventual evaluation, an important phase during the design process.
+
+To the best of our knowledge, MAREA is the first tool designed to evaluate marine reserves.
+Previous work [@pomeroy_2005-Py; @mascia_2017-m_] addressed MPA evaluation and provided the foundation for our contribution.
+However, these did not intended to create user--friendly tools to aid in the evaluation.
+As with other conservation management tools, development of tools that automatize complex calculations can have an important impact in management [@ball_2009-qi].
+The use of open data science enables the creation of open-access tools that can address techincal gaps and inprove management [@lowndes_2017-xh].
 
@@
 This process will incorporate new features, and enhance current ones, aiming to improve user experience and expand the scope of the analysis.
+Other modifications may also include addition of more objectives and indicators to ensure applicability in other regions, full translation into other languages to avoid any ambiguities introduced via the automatic translation, or reporting effects over time in percentages to aid interpretation.
 Yet, we believe that this first release represents a major step towards effective, replicable evaluation and management of marine reserves.
@@
 Special thanks to the fishers from Isla Natividad, who gathered the data used in this study, and the fishers from El Rosario, who helped us validate our survey and framework, and to Arturo Hernández and Alfonso Romero who provided help with the logistics.
+Finally, we thank the Editor for a professional handling, and two anonymous reviewers for their suggestions, which significantly improved the quality of this paper.
 

### 2017-12-03 Manuscript.tex
Incorporated changes from reviewers

--- 2017-12-03 ManuscriptV1.tex
@@
 
-Available empirical evidence on marine reserve effectiveness is mixed
-{[}7{]}. Some studies have shown that MPAs increase biomass {[}4,8{]},
-enhance resilience to climatic impacts {[}9,10{]}, and preserve genetic
-diversity {[}11{]}. Compared to partially protected MPAs, marine
+Empirical evidence shows that MPAs increase biomass {[}4,7{]}, enhance
+resilience to climatic impacts {[}8,9{]}, and preserve genetic diversity
+{[}10{]}. Compared to MPAs that grant partial protection, marine
 reserves have higher levels of biomass, density, richness, and larger
-organisms {[}3,12--14{]}. These effects are often measured as biological
-changes in the area through time and many lack a control site for
-comparison {[}15{]}. This approach does not account for other factors
-for which one must control {[}16{]} in order to causally attribute a
-biological change to the reserve. While some studies have used control
-sites, these analyses do not estimate the effect of the reserve, and
-often use a control--impact comparison approach that does not address
-temporal variability {[}4,8,17--19{]}. A smaller fraction of studies
-have used a before--after--control--impact (\emph{i.e.} BACI) design
-comparing reserves to control sites before and after implementation
-{[}4,20,21{]}, which allows the use of causal inference techniques that
-estimate the effect of the reserve.
-
-But even when proper causal inference can be drawn, results are often
-idiosyncratic across reserves. Effects of reserves on ecological and
-economic outcomes are highly heterogeneous, and often depend on the
-specific ecological, economic, and social context. The purpose of this
-paper is to describe a user--friendly tool, called ``MAREA'', to
-rigorously systematize the evaluation of marine reserve effectiveness.
-The tool is in the form of an open-source application that uses
-state--of--the--art methods from program evaluation to compare a reserve
-to control sites along a number of ecological, economic, and governance
-dimensions.
-
-The challenge of how to standardize marine reserve evaluation is not a
-new one. The recent IUCN framework ``How is your MPA doing?''
-{[}22,23{]} provides a comprehensive list of biological, socioeconomic,
-and governance indicators, and insights into how these indicators may be
-measured. But this framework stops short of analysis, so provides a user
-with little guidance about establishing causal inference about the
-reserve. Recent work by Mascia \emph{et al.} {[}24{]} integrates these
-three dimensions via the Social Ecological Systems Framework {[}25,26{]}
-and suggests the use of causal inference techniques to provide a measure
-of the effect of conservation interventions. However, these two novel
-approaches do not provide a user-friendly tool that enables
-replicability and scalability of the analysis, particularly when used by
-the fishers and decision makers themselves.
+organisms {[}3,11--13{]}. However, these effects are often measured as
+biological changes within the reserves through time, and many lack a
+control site for comparison {[}14{]}. This approach does not account for
+other factors (\emph{e.g.} system-level changes in productivity ;
+{[}15{]}) for which one must control {[}16{]} in order to causally
+attribute a biological change to the reserve. Other studies have used a
+control--impact comparison approach that uses control sites but does not
+address temporal variability {[}4,7,17--19{]}. A smaller fraction of
+studies have used a before--after--control--impact (\emph{i.e.} BACI)
+design comparing reserves to control sites before and after
+implementation {[}4,20,21{]}, which allows the use of causal inference
+techniques that estimate the effect of the reserve. But even when proper
+causal inference can be drawn, results are often different across
+reserves. Effects of reserves on ecological and economic outcomes are
+highly heterogeneous, and often depend on the specific ecological,
+economic, and social context.
+
+Standardization of marine reserve evaluation is not new. The IUCN
+framework ``How is your MPA doing?'' {[}22,23{]} provides a
+comprehensive list of biological, socioeconomic, and governance
+indicators, and insights into how these may be measured or collected.
+But this framework stops short of analysis, and provides a user with
+little guidance about establishing causal inference about the reserve.
+Recent work {[}24{]} integrates these three dimensions via the Social
+Ecological Systems Framework {[}25,26{]} and suggests the use of causal
+inference techniques to provide a measure of the effect of conservation
+interventions. However, neither of these approaches provide a
+user-friendly tool that ensures replicability and scalability of the
+analysis, particularly when used by the fishers and decision makers
+themselves.
 
@@
 
-This paper presents a framework and user--friendly tool to evaluate
-marine reserve effectiveness, which incorporates the biological,
-socioeconomic, and governance dimensions of any fishery. We first
-provide a list of commonly stated management objectives and match them
-to appropriate indicators. We then develop a simple approach to
+The purpose of this paper is to describe a user--friendly tool, called
+``MAREA'', to rigorously systematize the evaluation of marine reserve
+effectiveness in terms of fisheries and marine conservation. The tool is
+in the form of an open--source application that uses state--of--the--art
+methods from program evaluation to compare a reserve to control sites
+along a number of ecological, economic, and governance dimensions. We
+first provide a list of commonly stated management objectives and match
+them to appropriate indicators. We then develop a simple approach to
 analyzing these indicators building on causal inference techniques
 {[}20{]}, which help us understand the effect of management
-interventions {[}24,37{]}. To implement the analytical approach in a
-user-friendly format, we introduce the Marine Reserve Evaluation
-Application (MAREA), an open source, web--based tool that automates the
-framework described in this paper and enables its broader use. Finally,
-we present a case study on the evaluation of a marine reserve
-established by the fishers of Isla Natividad (Mexico) in 2006, to
-demonstrate the potential of MAREA.
+interventions {[}24,37{]}. To implement the analytical approach, we
+introduce the Marine Reserve Evaluation Application (MAREA), an open
+source, web--based tool that automates the framework described in this
+paper and enables its broader use. Finally, we present a case study on
+the evaluation of a marine reserve established by the fishers of Isla
+Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.
 
@@
 presented. We then describe the development of MAREA and explain how
-this user--friendly open--access tool can be used by fishermen,
-managers, and other stakeholders with little scientific background.
-Finally, we provide guidelines on how to interpret and use the results
-and output generated by MAREA to inform management.
+this open--access tool can be used by fishermen, managers, and other
+stakeholders with little scientific background. Finally, we provide
+guidelines on how to interpret and use the results and output generated
+by MAREA to inform management.
 
@@
 past five years to achieve objectives such as increasing productivity in
-nearby waters or recover overexploited species; most of these reserves
-have never been formally evaluated for effectiveness at meeting those
-objectives. Thus, our focus was on identifying common objectives of
-marine reserves in Mexico. However, a literature review and discussions
-with marine reserve researchers worldwide suggest that the objectives
+nearby waters or recovery of overexploited species; most of these
+reserves have never been formally evaluated for effectiveness at meeting
+those objectives. Thus, our focus was on identifying common objectives
+of marine reserves in Mexico. However, a literature review and
+discussions with marine reserve researchers suggest that the objectives
 driving Mexican marine reserve implementation are similar to those in
 the rest of the world. Thus, we group these objectives into seven major
-categories, which can be applied to marine reserves worldwide. The list
-of objectives includes stated objectives in legislation {[}38,39{]} and
-official documents such as the Technical Justification Studies
-(\emph{Estudios Técnicos Justificativos}), agreements, and decrees
-associated to these areas:
+categories, which may be applied to marine reserves worldwide. Any given
+reserve may have been implemented to meet one or more of these. The list
+includes objectives stated in legislation {[}38,39{]} and official
+documents such as the Technical Justification Studies (\emph{Estudios
+Técnicos Justificativos}), agreements, and decrees associated to these
+areas:
 
@@
 compiled through a review of scientific literature in which we
-identified indicators that were used to measure similar objectives. A
-first filter eliminated indicators for which baseline data do not
-typically exist in Mexico. The preliminary list of indicators was
-reviewed at a workshop with participation of members from Mexican
-fishery management agencies and non-government organizations. Later,
-these were presented to fishers from the Ensenada Fishing Cooperative
-(\emph{S.C.P.P. Ensenada}), in El Rosario, Baja California, who provided
-input. Our final list of indicators includes those identified in review
-works {[}4,40{]}.
+identified indicators that were used to measure similar
+objectives{[}3--5,7,11,13,14,17--20,22,23,40--43{]}. A first filter
+eliminated indicators for which baseline data do not typically exist in
+Mexico. The preliminary list of indicators was reviewed at a workshop
+with participation of members from Mexican fishery management agencies
+and non-government organizations. Later, these were presented to fishers
+from the Ensenada Fishing Cooperative (\emph{S.C.P.P. Ensenada}), in El
+Rosario, Baja California, who provided input. Our final list of
+indicators includes those identified in review works {[}4,43{]}.
 
@@
 access rights to the fishery, number of fishers, legal recognition of
-the reserve). Some indicators require a numerical entry (\emph{e.g.}
-Fish biomass) while others are more descriptive (\emph{e.g.} Reasoning
-for reserve location). Many of them specifically measure an outcome of
-the reserve, though some are designed to further the understanding of
-the mechanisms driving a reserve's performance. In that sense, most
-biological and socioeconomic indicators are outcome variables. On the
-other hand, governance indicators are viewed as possible explanatory
-variables of reserve performance. Whenever an indicator is applied to
-``Target species'', it means that the indicator can be used for all
-species (\emph{e.g.} Fish Biomass) and/or for individual species that
-are either the conservation target of the reserve or are of particular
-economic or ecological interest (\emph{e.g.} Grouper Biomass). Table 1
+the reserve). Some indicators are quantitative and require a numerical
+entry (\emph{e.g.} Fish biomass) while others are qualitative and rely
+on a descriptive entry (\emph{e.g.} Reasoning for reserve location).
+Many of them specifically measure an outcome of the reserve, though some
+are designed to further the understanding of the mechanisms driving a
+reserve's performance. In that sense, most biological and socioeconomic
+indicators are outcome variables. On the other hand, governance
+indicators are viewed as possible explanatory variables of reserve
+performance. Whenever an indicator is applied to ``Target species'', it
+means that the indicator can be used for all species (\emph{e.g.} Fish
+Biomass) and/or for individual species that are either the conservation
+target of the reserve or are of particular economic or ecological
+interest (\emph{e.g.} Grouper Biomass). Finally, indicators B3 and B4
+are different in that B3 look only at the density of organisms above
+size at first maturity (related to reproductive potential), while B4
+measures the density of all fish or of a target species. Each indicator
+targets different plausible desired outcomes, like increased
+reproductive potential (\emph{i.e.} B3; {[}44{]}) or having more fish
+-regardless of their size- to atract tourism (\emph{i.e.} B4). Table 1
 presents the proposed indicators, and Table 2 shows how objectives are
-matched with biological and socioeconomic indicators.
+matched with biological and socioeconomic indicators. Governance
+indicators are excluded from Table 2, as we suggest that all of them
+should be used in an evaluation, regardless of the objectives.
 
@@
 \hline
-\hspace{1em}B3 & Density of mature organisms & Continuous & Percent points\\
+\hspace{1em}B3 & Density of mature organisms & Continuous & Percent\\
 \hline
@@
 
-\caption{\label{tab:unnamed-chunk-3}Management objectives and respective performance indicators. All governance indicators should allways be used.}
+\caption{\label{tab:unnamed-chunk-3}Management objectives and respective performance indicators.}
 \centering
@@
 \hline
-Avoid overexploitation &  &  & x &  & x & x &  &  & x & x & x & x & x & x\\
+Avoid overexploitation &  &  & x & x & x & x & x & x & x & x & x & x & x & x\\
 \hline
@@
 \hline
-Maintain biological process &  &  &  & x & x & x &  & x & x & x & x & x & x & x\\
-\hline
-Improve fishery production in nearby waters &  &  & x & x & x & x & x & x & x & x & x & x & x & x\\
-\hline
-Preserve biological diversity and the ecosystem &  &  & x &  & x & x &  &  & x &  & x &  & x & x\\
-\hline
-Recover overexploited species & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
-\hline
-Recover species of economic interest & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
+Maintain biological process & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
+\hline
+Improve fishery production in nearby waters &  &  &  & x & x & x &  & x & x & x & x & x & x & x\\
+\hline
+Preserve biological diversity and the ecosystem & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
+\hline
+Recover overexploited species &  &  & x &  & x & x &  &  & x &  & x &  & x & x\\
+\hline
+Recover species of economic interest &  &  & x &  & x & x &  &  & x &  & x &  & x & x\\
 \hline
@@
 
-In many coastal marine reserves of Mexico, biological data are collected
-via underwater visual censuses as part of a reserve's monitoring
-program. Scientific divers (which are often local fishermen with
-guidance from Civil Society Organizations; CSOs) record fish and
-invertebrate richness and abundances, as well as fish total length along
-belt transects. Ecological surveys are typically performed annually in
-each reserve and corresponding control site(s), before and after the
-implementation of the reserve, providing a sampling design that can be
-used to draw causal inference. Control sites are areas where habitat is
-similar to that of the reserve, but with presence of fishing activity;
-in principle these are areas that are otherwise observationally
-identical to the reserve site, but where, for presumably random reasons,
-a reserve was not implemented. While transect dimensions (\emph{i.e.}
-length and width) and sampling methods might vary from study to study,
-the general idea remains the same: richness, abundances, and sizes of
-organisms are recorded in a study--specific standardized way. For this
-reason, MAREA does not assume specific transect dimensions, and
-pertinent indicators are calculated per transect (Table 1).
+In many coastal marine reserves, biological data are often collected via
+underwater visual censuses as part of a reserve's monitoring program.
+Scientific divers record fish and invertebrate richness and abundances,
+as well as fish total length along belt transects. Ecological surveys
+are typically performed annually in each reserve and corresponding
+control site(s), before and after the implementation of the reserve,
+providing a sampling design that can be used to draw causal inference.
+Control sites are areas where habitat is similar to that of the reserve,
+but with presence of fishing activity; in principle these are areas that
+are otherwise observationally identical to the reserve site, but where,
+for presumably random reasons, a reserve was not implemented. While
+transect dimensions (\emph{i.e.} length and width) and sampling methods
+might vary from study to study, the general idea remains the same:
+richness, abundances, and sizes of organisms are recorded in a
+study--specific standardized way. For this reason, MAREA does not assume
+specific transect dimensions, and pertinent indicators are calculated
+per transect (Table 1). More information on data collection and
+formatting is provided in a guidebook {[}45{]}, which is available in
+English and Spanish in MAREA's interface.
 
 This sampling design for biological data allows us to use causal
-inference techniques {[}20,41{]} to evaluate the effect of the reserve
+inference techniques {[}20,46{]} to evaluate the effect of the reserve
 on biological indicators. The hypothesis that the indicators will
@@
 control or reserve), an interaction between a dummy variable that
-indicates pre-- or post--implementation (\(P\)) and treatment (\(Z\)),
+indicates before or after implementation (\(P\)) and treatment (\(Z\)),
 and covariates such as bottom temperature (\(T\); in °C), horizontal
@@
 variables, modeled as dummy variables, are coded as Control = 0 and
-Reserve = 1; and Pre--implementation = 0 and Post--implementation = 1,
+Reserve = 1; and Before implementation = 0 and After implementation = 1,
 respectively.
@@
 sometimes prices for each species. To control for inflation, income is
-adjusted with the country's consumer price index {[}42{]}:
+adjusted with the country's consumer price index {[}47{]}:
 
@@
 implementation of the reserve and provide valuable input. For both
-models (eq. 1 and eq. 3), we estimate the model coeficients with
+models (eq. 1 and eq. 3), we estimate the model coefficient with
 ordinary least squares, and calculate heteroskedastic--robust standard
-errors.
+errors {[}48{]}.
 
@@
 Table). This approach has been proven to successfully evaluate
-governance structures {[}43{]}. Unlike with biological and socioeconomic
+governance structures {[}49{]}. Unlike with biological and socioeconomic
 objectives (see eqs 1 and 3), MAREA does not quantitatively analyze
@@
 
-We developed MAREAin R version 3.4.2 and R Studio 1.1.383 {[}44{]} using
-the Shiny package {[}45{]}, to build an interactive web application
+We developed MAREAin R version 3.4.2 and R Studio 1.1.383 {[}50{]} using
+the Shiny package {[}51{]}, to build an interactive web application
 hosted on an open server; the MAREA app can be accessed at
@@
 MAREA automatically matches to appropriate indicators, based on Table 2.
-Users can also manually modify selected indicators based on their
-interests and data availability. The user can then load data on one or
-more reserves, using standard *.csv text files; sample datasets are
-provided within MAREA. Once data have been loaded, MAREA identifies all
-reserves in the data, and lets the user select the reserve to be
-evaluated. At this point, the user can also specify the year of
-implementation of the reserve, reserve dimensions, and indicate target
-species that are of particular management interest. MAREA provides the
-user with a section to confirm that all the decisions made leading up to
-that point are correct. Once the user has confirmed all input data,
-objectives, and other information, MAREA performs the formal program
-evaluation analyses discussed above. For a typical data set, the
-automated analysis step takes less than one second. Finally, the user is
-taken to the results tab where all results are presented in a simple
-format. The user can also download a more comprehensive technical report
-produced in *.pdf format.
+Users can also select and un-select indicators based on their interests
+and data availability by ``clicking'' on the check-boxes in MAREA. The
+user can then load data on one or more reserves, using standard *.csv
+text files; sample datasets are provided within MAREA. Once data have
+been loaded, MAREA identifies all reserves in the data, and lets the
+user select the reserve to be evaluated. At this point, the user can
+also specify the year of implementation of the reserve, reserve
+dimensions, and indicate target species that are of particular
+management interest. MAREA provides the user with a section to confirm
+that all the decisions made leading up to that point are correct. Once
+the user has confirmed all input data, objectives, and other
+information, MAREA performs the formal program evaluation analyses
+discussed above. For a typical data set, the automated analysis step
+takes less than one second. Finally, the user is taken to the results
+tab where all results are presented in a simple format. The user can
+also download a more comprehensive technical report produced in *.pdf
+format.
 
@@
 The scorecard is produced with functions from the Shinydashboard package
-{[}46{]}. The technical report is produced by a parameterized Rmarkdown
-document {[}47{]} processed by the knitr package {[}48{]}. Another
+{[}52{]}. The technical report is produced by a parameterized Rmarkdown
+document {[}53{]} processed by the knitr package {[}54{]}. Another
 feature of MAREA is that the user can choose to share the data. Once the
@@
 y Pescadores de la Baja California}) that promotes co-management of
-marine resources {[}49,50{]}. Additionally, fishers have Territorial Use
+marine resources {[}55,56{]}. Additionally, fishers have Territorial Use
 Rights for Fisheries (TURFs) that provide them with exclusive access
 rights to exploit the benthic marine resources within a given perimeter
-{[}50{]}.
+{[}56{]}.
 
@@
 approximate location of control sites is indicated by blue squares (B =
-Babencho, D = La Dulce).
+Babencho, D = La Dulce). Shapefiles for Mexican coastline and the US
+were obtained from the Instituto Nacional de Estadistica, Geografia e
+Informatica of Mexico (INEGI:
+www.inegi.org.mx/geo/contenidos/geoestadistica/m\_geoestadistico.aspx)
+and the tmap R package {[}57{]}, respectively.
 
 In 2006, the Isla Natividad community implemented two community--based
-marine reserves within their TURF {[}9,51,52{]}. Evidence suggest that
-these reserves have been effective at enhancing resilience to climate
-variations {[}9{]} and preserving genetic diversity of high value
-commercial species such as abalone {[}11{]}. These ecological benefits
-have been translated into economic benefits, enhancing population
-persistence and bolstering abalone fisheries {[}53{]}. For the purpose
-of this evaluation, we focused on the ``La Plana / Las Cuevas'' marine
-reserve, located at the southern end of the island (Fig. 3) and its
-corresponding control site ``La Dulce / Babencho''.
+marine reserves within their TURF {[}8,40,58{]} after establishing a
+baseline for the soon--to--be reserves and control sites. Evidence
+suggest that these reserves have been effective at enhancing resilience
+to climate variations {[}8{]} and preserving genetic diversity of high
+value commercial species such as abalone {[}10{]}. These ecological
+benefits have been translated into economic benefits, enhancing
+population persistence and bolstering abalone fisheries {[}42{]}. For
+the purpose of this evaluation, we focused on the ``La Plana / Las
+Cuevas'' marine reserve, located at the southern end of the island (Fig.
+3) and its corresponding control site ``La Dulce / Babencho''.
 
@@
 information along 30 m transects, with a sampling window of 2 m x 2 m
-following a standardized ReefCheck protocol {[}54{]}. Ecological surveys
+following a standardized ReefCheck protocol {[}59{]}. Ecological surveys
 were performed yearly in each reserve and corresponding control site(s),
@@
 landings and income were aggregated by year and species, and adjusted by
-the Consumer Price Index {[}42{]}. From the nine species available, we
+the Consumer Price Index {[}47{]}. From the nine species available, we
 selected as objective species those that contributed the most (88.27\%)
@@
 suggest that the community is strongly organized, which is a likely
-driver of the successes reported above {[}55{]}. The first point of
+driver of the successes reported above {[}60{]}. The first point of
 success is the existence of a fishing cooperative that is also
@@
 governance structures allow various levels of organization that have
-been shown to foster communication and cooperation {[}43,50{]};
-federations also provide bargain power with governments {[}43,56{]}.
+been shown to foster communication and cooperation {[}49,56{]};
+federations also provide bargain power with governments {[}49,61{]}.
 Access to fishing resources is managed through a TURF, permits, and
-fishing quotas (for some species). McCay {[}49{]} suggests that the TURF
+fishing quotas (for some species). McCay {[}55{]} suggests that the TURF
 promotes a sense of stewardship of their resources and incentivizes
@@
 
-We have developed and presented a user--friendly, automated approach for
-evaluating the effectiveness of marine reserves around the world. Here
-we highlight MAREA's utility for evidence---based management, and
-comment on a few of its shortcomings. The findings from Isla Natividad
-are used purely to validate the relevance of MAREA rather than to
-discuss particularities of the marine reserve effectiveness, which has
-been described before {[}9,11,53{]}. We use examples from the case study
-to build on the utility of MAREA and discuss ways in which results can
-be interpreted to inform management.
+We have developed and presented an automated approach for evaluating the
+effectiveness of marine reserves in Mexico, and perhaps around the
+world. Here we highlight MAREA's utility for evidence---based
+management, and comment on a few of its shortcomings. The findings from
+Isla Natividad are used purely to validate the relevance of MAREA rather
+than to discuss particularities of the marine reserve effectiveness,
+which has been described before {[}8,10,42{]}. We use examples from the
+case study to build on the utility of MAREA and discuss ways in which
+results can be interpreted to inform management.
 
 The causal inference techniques used by MAREA have been suggested
-{[}37,41{]} and used {[}20{]} before in other ad hoc studies. This
+{[}37,46{]} and used {[}20{]} before in other ad hoc studies. This
 approach reduces ambiguity in the interpretation of results. For
@@
 Disturbance), we can better understand the trends in invertebrate
-densities. As reported before {[}9{]}, hypoxic conditions that have
+densities. As reported before {[}8{]}, hypoxic conditions that have
 occurred in Isla Natividad can cause decreases in invertebrate
@@
 
-We recognize that the 29 indicators used by MAREA might not fully
-describe a reserve. However, they provide a starting point to perform
-the evaluation, to which managers and users can add other indicators
-(\emph{e.g.} larval dispersal or connectivity) that are relevant to
-their reserve. Furthermore, MAREA's value is that it provides a free,
-simple, and replicable way to perform rigorous impact analysis. The tool
-can easily be used by fishers, CSO members, and managers in government
-agencies, providing transparency of the analysis and results. In
-addition, it can empower and enable local managers and fishers to
-respond to local change and adapt by allowing direct and easy access to
-the information.
+We recognize that the seven objectives and 29 indicators used by MAREA
+might not fully describe a reserve. In order to ensure the applicability
+of the tool to reserves in other countries, further testing in other
+regions should take place. However, the proposed objectives and
+indicators provide a starting point to perform the evaluation, to which
+managers and users can add other indicators (\emph{e.g.} larval
+dispersal or connectivity) that are relevant to their reserve.
+Furthermore, MAREA's value is that it provides a free, simple, and
+replicable way to perform rigorous impact analysis. The tool can easily
+be used by fishers, CSO members, and managers in government agencies,
+providing transparency of the analysis and results. In addition, it can
+empower and enable local managers and fishers to respond to local change
+and adapt by allowing direct and easy access to the information.
+
+An evident limitation of MAREA is its dependence on data obtained
+through a BACI design, and the amount of samples needed to estimate
+coefficients in Eqn. 1. It is not uncommon for control sites or
+baselines that describe the biological state of the area before the
+management intervention to be abscent. Propperly designing marine
+reserves by identifying control sites and establishing a baseline before
+the implementation of the reserve is enough to overcome this issue;
+reserves for which there is no control site and baseline can not be
+evaluated with MAREA. On the second case, typical underwater surveys
+require that at least 12 - 16 transects are performed on each site
+(\emph{i.e.} reserve and control) each year. This provides with at least
+48 samples (12 samples per site, per year), enough to avoid overfitting
+Eqn 1. On either case, these problems can be easily avoided by
+accounting for data requirements in the eventual evaluation, an
+important phase during the design process.
+
+To the best of our knowledge, MAREA is the first tool designed to
+evaluate marine reserves. Previous work {[}22,24{]} addressed MPA
+evaluation and provided the foundation for our contribution. However,
+these did not intended to create user--friendly tools to aid in the
+evaluation. As with other conservation management tools, development of
+tools that automatize complex calculations can have an important impact
+in management {[}62{]}. The use of open data science enables the
+creation of open-access tools that can address techincal gaps and
+inprove management {[}27{]}.
 
 The effectiveness of marine reserves continues to be a matter of debate
-{[}7,12,40{]}. With current targets set to increase ocean protection, it
-is important that we understand the effects of our interventions
-{[}37{]} so we can better inform management {[}41{]}. It is therefore
+{[}11,43,63{]}. With current targets set to increase ocean protection,
+it is important that we understand the effects of our interventions
+{[}37{]} so we can better inform management {[}46{]}. It is therefore
 important that academics, managers, fishers, and CSOs have access to
@@
 and other Latin American countries, where management agencies are often
-understaffed and underfunded {[}57{]}, or where materials are often not
+understaffed and underfunded {[}64{]}, or where materials are often not
 available in their language. In this context, MAREA provides a simple
@@
 comprehensively presented represents a valuable step towards effective
-evidence---based management {[}41{]}.
+evidence---based management {[}46{]}.
 
@@
 process will incorporate new features, and enhance current ones, aiming
-to improve user experience and expand the scope of the analysis. Yet, we
-believe that this first release represents a major step towards
-effective, replicable evaluation and management of marine reserves.
+to improve user experience and expand the scope of the analysis. Other
+modifications may also include addition of more objectives and
+indicators to ensure applicability in other regions, full translation
+into other languages to avoid any ambiguities introduced via the
+automatic translation, or reporting effects over time in percentages to
+aid interpretation. Yet, we believe that this first release represents a
+major step towards effective, replicable evaluation and management of
+marine reserves.
 
@@
 validate our survey and framework, and to Arturo Hernández and Alfonso
-Romero who provided help with the logistics.
+Romero who provided help with the logistics. Finally, we thank the
+Editor for a professional handling, and two anonymous reviewers for
+their suggestions, which significantly improved the quality of this
+paper.
 
@@
 6. Hastings A, Gaines SD, Costello C. Marine reserves solve an important
-bycatch problem in fisheries. Proc Natl Acad Sci U S A. 2017;
+bycatch problem in fisheries. Proc Natl Acad Sci U S A. 2017;114:
+8927--8934.
 doi:\href{https://doi.org/10.1073/pnas.1705169114}{10.1073/pnas.1705169114}
 
-\hypertarget{ref-padleton_2017-vn}{}
-7. Padleton L, Aghmadia G, Browman H, Thurstand R, Kaplan D, Bartolino
-V. Debating the effectiveness of marine protected areas. ICES Journal of
-Marine Science. 2017;
-doi:\href{https://doi.org/10.1093/icesjms/fsx154}{10.1093/icesjms/fsx154}
-
 \hypertarget{ref-aburtooropeza_2011-ya}{}
-8. Aburto-Oropeza O, Erisman B, Galland GR, Mascareñas-Osorio I, Sala E,
+7. Aburto-Oropeza O, Erisman B, Galland GR, Mascareñas-Osorio I, Sala E,
 Ezcurra E. Large recovery of fish biomass in a no-take marine reserve.
@@
 \hypertarget{ref-micheli_2012-EU}{}
-9. Micheli F, Saenz-Arroyo A, Greenley A, Vazquez L, Espinoza Montes JA,
+8. Micheli F, Saenz-Arroyo A, Greenley A, Vazquez L, Espinoza Montes JA,
 Rossetto M, et al. Evidence that marine reserves enhance resilience to
@@
 \hypertarget{ref-roberts_2017-J9}{}
-10. Roberts CM, O'Leary BC, McCauley DJ, Cury PM, Duarte CM, Lubchenco
-J, et al. Marine reserves can mitigate and promote adaptation to climate
+9. Roberts CM, O'Leary BC, McCauley DJ, Cury PM, Duarte CM, Lubchenco J,
+et al. Marine reserves can mitigate and promote adaptation to climate
 change. Proc Natl Acad Sci U S A. 2017;114: 6167--6175.
@@
 \hypertarget{ref-munguavega_2015-yg}{}
-11. Munguía-Vega A, Sáenz-Arroyo A, Greenley AP, Espinoza-Montes JA,
+10. Munguía-Vega A, Sáenz-Arroyo A, Greenley AP, Espinoza-Montes JA,
 Palumbi SR, Rossetto M, et al. Marine reserves help preserve genetic
@@
 \hypertarget{ref-edgar_2014-UO}{}
-12. Edgar GJ, Stuart-Smith RD, Willis TJ, Kininmonth S, Baker SC, Banks
+11. Edgar GJ, Stuart-Smith RD, Willis TJ, Kininmonth S, Baker SC, Banks
 S, et al. Global conservation outcomes depend on marine protected areas
@@
 \hypertarget{ref-giakoumi_2017-V2}{}
-13. Giakoumi S, Scianna C, Plass-Johnson J, Micheli F, Grorud-Colvert K,
+12. Giakoumi S, Scianna C, Plass-Johnson J, Micheli F, Grorud-Colvert K,
 Thiriet P, et al. Ecological effects of full and partial protection in
@@
 \hypertarget{ref-sala_2017-69}{}
-14. Sala E, Giakoumi S. No-take marine reserves are the most effective
+13. Sala E, Giakoumi S. No-take marine reserves are the most effective
 protected areas in the ocean. ICES Journal of Marine Science. 2017;
@@
 \hypertarget{ref-betti_2017-lq}{}
-15. Betti F, Bavestrello G, Bo M, Asnaghi V, Chiantore M, Bava S, et al.
+14. Betti F, Bavestrello G, Bo M, Asnaghi V, Chiantore M, Bava S, et al.
 Over 10 years of variation in mediterranean reef benthic communities.
@@
 doi:\href{https://doi.org/10.1111/maec.12439}{10.1111/maec.12439}
+
+\hypertarget{ref-szuwalski_2017-jc}{}
+15. Szuwalski CS, Burgess MG, Costello C, Gaines SD. High fishery
+catches through trophic cascades in china. Proc Natl Acad Sci U S A.
+2017;114: 717--721.
+doi:\href{https://doi.org/10.1073/pnas.1612722114}{10.1073/pnas.1612722114}
 
@@
 
+\hypertarget{ref-lester_2017-nh}{}
+40. Lester S, McDonald G, Clemence M, Dougherty D, Szuwalski C. Impacts
+of turfs and marine reserves on fisheries and conservation goals:
+Theory, empirical evidence, and modeling. BMS. 2017;93: 173--198.
+doi:\href{https://doi.org/10.5343/bms.2015.1083}{10.5343/bms.2015.1083}
+
+\hypertarget{ref-chirico_2017-Rz}{}
+41. Chirico AAD, McClanahan TR, Eklöf JS. Community- and
+government-managed marine protected areas increase fish size, biomass
+and potential value. PLoS ONE. 2017;12: e0182342.
+doi:\href{https://doi.org/10.1371/journal.pone.0182342}{10.1371/journal.pone.0182342}
+
+\hypertarget{ref-rossetto_2015-V0}{}
+42. Rossetto M, Micheli F, Saenz-Arroyo A, Montes JAE, De Leo GA, Rochet
+M-J. No-take marine reserves can enhance population persistence and
+support the fishery of abalone. Can J Fish Aquat Sci. 2015;72:
+1503--1517.
+doi:\href{https://doi.org/10.1139/cjfas-2013-0623}{10.1139/cjfas-2013-0623}
+
 \hypertarget{ref-woodcock_2017-Wm}{}
-40. Woodcock P, O'Leary BC, Kaiser MJ, Pullin AS. Your evidence or mine?
+43. Woodcock P, O'Leary BC, Kaiser MJ, Pullin AS. Your evidence or mine?
 Systematic evaluation of reviews of marine protected area effectiveness.
@@
 
+\hypertarget{ref-carter_2017-Uf}{}
+44. Carter AB, Davies CR, Emslie MJ, Mapstone BD, Russ GR, Tobin AJ, et
+al. Reproductive benefits of no-take marine reserves vary with region
+for an exploited coral reef fish. Sci Rep. 2017;7: 9693.
+doi:\href{https://doi.org/10.1038/s41598-017-10180-w}{10.1038/s41598-017-10180-w}
+
+\hypertarget{ref-villaseorderbez_website_2017-xE}{}
+45. Villaseñor-Derbez JC, Faro C, Wright M, Martínez J. A guide to
+evaluate the effectiveness of no-take marine reserves in mexico
+{[}Internet{]}. 2017. Available:
+\url{https://www.researchgate.net/publication/317840581/_A/_guide/_to/_evaluate/_the/_effectiveness/_of/_no-take/_marine/_reserves/_in/_Mexico}
+
 \hypertarget{ref-ferraro_2006-oW}{}
-41. Ferraro PJ, Pattanayak SK. Money for nothing? A call for empirical
+46. Ferraro PJ, Pattanayak SK. Money for nothing? A call for empirical
 evaluation of biodiversity conservation investments. PLoS Biol. 2006;4:
@@
 \hypertarget{ref-oecd_website_2017-VV}{}
-42. OECD. Prices - inflation (cpi) - oecd data {[}Internet{]}. 2017.
+47. OECD. Prices - inflation (cpi) - oecd data {[}Internet{]}. 2017.
 Available: \url{https://data.oecd.org/price/inflation-cpi.htm}
 
+\hypertarget{ref-sandwich_2014}{}
+48. Zeileis A. Econometric computing with hc and hac covariance matrix
+estimators. Journal of Statistical Software. 2004;11: 1--17.
+
 \hypertarget{ref-espinosaromero_2014-PY}{}
-43. Espinosa-Romero MJ, Rodriguez LF, Weaver AH, Villanueva-Aznar C,
+49. Espinosa-Romero MJ, Rodriguez LF, Weaver AH, Villanueva-Aznar C,
 Torre J. The changing role of ngos in mexican small-scale fisheries:
@@
 \hypertarget{ref-rcore_2017}{}
-44. R Core Team. R: A language and environment for statistical computing
+50. R Core Team. R: A language and environment for statistical computing
 {[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
@@
 \hypertarget{ref-shiny_2017}{}
-45. Chang W, Cheng J, Allaire J, Xie Y, McPherson J. Shiny: Web
+51. Chang W, Cheng J, Allaire J, Xie Y, McPherson J. Shiny: Web
 application framework for r {[}Internet{]}. 2017. Available:
@@
 \hypertarget{ref-shinydashboard_2017}{}
-46. Chang W, Borges Ribeiro B. Shinydashboard: Create dashboards with
+52. Chang W, Borges Ribeiro B. Shinydashboard: Create dashboards with
 'shiny' {[}Internet{]}. 2017. Available:
@@
 \hypertarget{ref-rmarkdown_2017}{}
-47. Allaire J, Cheng J, Xie Y, McPherson J, Chang W, Allen J, et al.
+53. Allaire J, Cheng J, Xie Y, McPherson J, Chang W, Allen J, et al.
 Rmarkdown: Dynamic documents for r {[}Internet{]}. 2017. Available:
@@
 \hypertarget{ref-knitr_2017}{}
-48. Xie Y. Knitr: A general-purpose package for dynamic report
+54. Xie Y. Knitr: A general-purpose package for dynamic report
 generation in r {[}Internet{]}. 2017. Available:
@@
 \hypertarget{ref-mccay_2017-1m}{}
-49. McCay B. Territorial use rights in fisheries of the northern pacific
+55. McCay B. Territorial use rights in fisheries of the northern pacific
 coast of mexico. BMS. 2017;93: 69--81.
@@
 \hypertarget{ref-mccay_2014-CN}{}
-50. McCay BJ, Micheli F, Ponce-Díaz G, Murray G, Shester G,
+56. McCay BJ, Micheli F, Ponce-Díaz G, Murray G, Shester G,
 Ramirez-Sanchez S, et al. Cooperatives, concessions, and co-management
@@
 
+\hypertarget{ref-tmap_2017}{}
+57. Tennekes M. Tmap: Thematic maps {[}Internet{]}. 2017. Available:
+\url{https://CRAN.R-project.org/package=tmap}
+
 \hypertarget{ref-afflerbach_2014-HP}{}
-51. Afflerbach JC, Lester SE, Dougherty DT, Poon SE. A global survey of
+58. Afflerbach JC, Lester SE, Dougherty DT, Poon SE. A global survey of
 -reserves, territorial use rights for fisheries coupled with marine
@@
 
-\hypertarget{ref-lester_2017-nh}{}
-52. Lester S, McDonald G, Clemence M, Dougherty D, Szuwalski C. Impacts
-of turfs and marine reserves on fisheries and conservation goals:
-Theory, empirical evidence, and modeling. BMS. 2017;93: 173--198.
-doi:\href{https://doi.org/10.5343/bms.2015.1083}{10.5343/bms.2015.1083}
-
-\hypertarget{ref-rossetto_2015-V0}{}
-53. Rossetto M, Micheli F, Saenz-Arroyo A, Montes JAE, De Leo GA, Rochet
-M-J. No-take marine reserves can enhance population persistence and
-support the fishery of abalone. Can J Fish Aquat Sci. 2015;72:
-1503--1517.
-doi:\href{https://doi.org/10.1139/cjfas-2013-0623}{10.1139/cjfas-2013-0623}
-
 \hypertarget{ref-suman_2010-ez}{}
-54. Suman CS, Saenz-Arroyo A, Dawson C, Luna MC. Manual de instruccion
+59. Suman CS, Saenz-Arroyo A, Dawson C, Luna MC. Manual de instruccion
 de reef check california: Guia de instruccion para el monitoreo del
@@
 \hypertarget{ref-gutirrez_2011-0U}{}
-55. Gutiérrez NL, Hilborn R, Defeo O. Leadership, social capital and
+60. Gutiérrez NL, Hilborn R, Defeo O. Leadership, social capital and
 incentives promote successful fisheries. Nature. 2011;470: 386--389.
@@
 \hypertarget{ref-finkbeiner_2015-87}{}
-56. Finkbeiner EM, Basurto X. Re-defining co-management to facilitate
+61. Finkbeiner EM, Basurto X. Re-defining co-management to facilitate
 small-scale fisheries reform: An illustration from northwest mexico.
@@
 
+\hypertarget{ref-ball_2009-qi}{}
+62. Ball IR, Possingham HP, Watts ME. Marxan and relatives: Software for
+spatial conservation prioritization. Spatial conservation prioritization
+quantitative methods \& computational tools. United Kingdom: Oxford
+University Press; 2009. pp. 185--195. Available:
+\url{https://espace.library.uq.edu.au/view/UQ:200259}
+
+\hypertarget{ref-padleton_2017-vn}{}
+63. Padleton L, Aghmadia G, Browman H, Thurstand R, Kaplan D, Bartolino
+V. Debating the effectiveness of marine protected areas. ICES Journal of
+Marine Science. 2017;
+doi:\href{https://doi.org/10.1093/icesjms/fsx154}{10.1093/icesjms/fsx154}
+
 \hypertarget{ref-lundquist_2005-OL}{}
-57. Lundquist CJ, Granek EF. Strategies for successful marine
+64. Lundquist CJ, Granek EF. Strategies for successful marine
 conservation: Integrating socioeconomic, political, and scientific

### 2017-12-09 Manuscript.Rmd
incorp-orated Seans comments

@@
 However, these effects are often measured as biological changes within the reserves through time, and many lack a control site for comparison [@betti_2017-lq].
-This approach does not account for other factors (*e.g.* system-level changes in productivity ; @szuwalski_2017-jc) for which one must control [@davies_2017-ml] in order to causally attribute a biological change to the reserve.
+This approach does not account for other factors (*e.g.* system-level changes in productivity caused by predatory release @szuwalski_2017-jc; or favorable environmental conditions @chavez_2003-_m) for which one must control [@davies_2017-ml] in order to causally attribute a biological change to the reserve.
 Other studies have used a control--impact comparison approach that uses control sites but does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
+
 A smaller fraction of studies have used a before--after--control--impact (*i.e.* BACI) design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.
+For example, in ref [@moland_2013-VP] authors use a BACI design and observe increases in lobster catches --a proxy for abundances-- after the implementation of the reserves for protected and control sites.
+However, the temporal changes in the reserve were greater than in the control site, suggesting a positive effect off the reserve on lobster catches.
 But even when proper causal inference can be drawn, results are often different across reserves.
@@
 The purpose of this paper is to describe a user--friendly tool, called “MAREA”, to rigorously systematize the evaluation of marine reserve effectiveness in terms of fisheries and marine conservation.
-The tool is in the form of an open--source application that uses state--of--the--art methods from program evaluation to compare a reserve to control sites along a number of ecological, economic, and governance dimensions.
+The tool is in the form of an open--source application that uses state--of--the--art methods from program evaluation to compare a reserve to control sites along a number of biological, economic, and governance dimensions.
 We first provide a list of commonly stated management objectives and match them to appropriate indicators.
@@
 However, a literature review and discussions with marine reserve researchers suggest that the objectives driving Mexican marine reserve implementation are similar to those in the rest of the world.
-Thus, we group these objectives into seven major categories, which may be applied to marine reserves worldwide.
+Thus, we group these objectives into seven major categories that may be applied to marine reserves worldwide.
 Any given reserve may have been implemented to meet one or more of these.
@@
 Fifteen governance indicators describe the governance structures under which the community operates (*e.g.*, access rights to the fishery, number of fishers, legal recognition of the reserve).
-Some indicators are quantitative and require a numerical entry (*e.g.* Fish biomass) while others are qualitative and rely on a descriptive entry (*e.g.* Reasoning for reserve location).
+Most biological and socioeconomic indicators are quantitative and require a numerical entry (*e.g.* Fish biomass) while all governance indicators, one biological indicator, and one socioeconomic indicator are qualitative and rely on a descriptive entry (*e.g.* Reasoning for reserve location).
 Many of them specifically measure an outcome of the reserve, though some are designed to further the understanding of the mechanisms driving a reserve’s performance.
@@
 Whenever an indicator is applied to “Target species”, it means that the indicator can be used for all species (*e.g.* Fish Biomass) and/or for individual species that are either the conservation target  of the reserve or are of particular economic or ecological interest (*e.g.* Grouper Biomass).
-Finally, indicators B3 and B4 are different in that B3 look only at the density of organisms above size at first maturity (related to reproductive potential), while B4 measures the density of all fish or of a target species.
-Each indicator targets different plausible desired outcomes, like increased reproductive potential (*i.e.* B3; @carter_2017-Uf) or having more fish -regardless of their size- to atract tourism (*i.e.* B4).
+Finally, indicators B3 and B4 are different in that B3 only looks at the density of organisms above size at first maturity (related to reproductive potential), while B4 measures the density of all fish or of a target species.
+Each indicator targets different plausible desired outcomes, like increased reproductive potential (*i.e.* B3; @carter_2017-Uf) or having more fish -regardless of their size- to attract tourism (*i.e.* B4).
 Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with biological and socioeconomic indicators.
-Governance indicators are excluded from Table 2, as we suggest that all of them should be used in an evaluation, regardless of the objectives.
+Governance indicators are excluded from Table 2, as governance indicators are important to consider for every objective.
 
@@
 Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on Table 2.
-Users can also select and un-select indicators based on their interests and data availability by "clicking" on the check-boxes in MAREA.
+Users can also select and deselect indicators based on their interests and data availability by "clicking" on the check-boxes in MAREA.
 The user can then load data on one or more reserves, using standard \*.csv text files; sample datasets are provided within MAREA.
@@
   tm_text("Label") +
-  tm_scale_bar(position = c("left", "bottom"))
+  tm_scale_bar(position = c("left", "bottom")) +
+  tm_grid(n.x = 5, n.y = 5)
 
@@
   tm_shape(controls) +
-  tm_symbols(shape = 22, size = 4, border.col = "black", col = "blue", border.lwd = 1) +
+  tm_symbols(shape = 22, size = 4, border.col = "black", col = "steelblue1", border.lwd = 1) +
   tm_text("Label") +
@@
   tm_compass(position = c("left", "bottom")) +
-  tm_scale_bar(position = c("left", "bottom"))
+  tm_scale_bar(position = c("left", "bottom")) +
+  tm_grid(n.x = 4, n.y = 4)
 
@@
 
-In 2006, the Isla Natividad community implemented two community--based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh; @micheli_2012-EU] after establishing a baseline for the soon--to--be reserves and control sites.
+In 2006, the Isla Natividad community established a biological baseline following the data collection protocol described in this study.
+The community then implemented two community--based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh; @micheli_2012-EU] after establishing a baseline for the soon--to--be reserves and control sites.
 Evidence suggest that these reserves have been effective at enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of high value commercial species such as abalone [@munguavega_2015-yg].
@@
 
-We recognize that the seven objectives and 29 indicators used by MAREA might not fully describe a reserve.
+We recognize that the seven objectives and 29 indicators used by MAREA might not fully describe a reserve in countries other than Mexico.
 In order to ensure the applicability of the tool to reserves in other countries, further testing in other regions should take place.
@@
 1.
-It is not uncommon for control sites or baselines that describe the biological state of the area before the management intervention to be abscent.
-Propperly designing marine reserves by identifying control sites and establishing a baseline before the implementation of the reserve is enough to overcome this issue; reserves for which there is no control site and baseline can not be evaluated with MAREA.
-On the second case, typical underwater surveys require that at least 12 - 16 transects are performed on each site (*i.e.* reserve and control) each year.
+It is not uncommon for control sites or baselines to be absent.
+Properly designing marine reserves by identifying control sites and establishing a baseline before the implementation of the reserve is enough to overcome this issue; reserves for which there is no control site and baseline cannot be evaluated with MAREA.
+Typical underwater surveys require that at least 12 - 16 transects are performed on each site (*i.e.* reserve and control) each year.
 This provides with at least 48 samples (12 samples per site, per year), enough to avoid overfitting Eqn 1.
-On either case, these problems can be easily avoided by accounting for data requirements in the eventual evaluation, an important phase during the design process.
+However, these problems can be easily avoided during the design and implementation phases by anticipating what data will be needed in the eventual evaluation.
 
@@
 Special thanks to the fishers from Isla Natividad, who gathered the data used in this study, and the fishers from El Rosario, who helped us validate our survey and framework, and to Arturo Hernández and Alfonso Romero who provided help with the logistics.
-Finally, we thank the Editor for a professional handling, and two anonymous reviewers for their suggestions, which significantly improved the quality of this paper.
+Finally, we thank the editor and two anonymous reviewers for their suggestions, which significantly improved the quality of this paper.
 

### 2017-12-09 Manuscript.tex
incorp-orated Seans comments

@@
 control site for comparison {[}14{]}. This approach does not account for
-other factors (\emph{e.g.} system-level changes in productivity ;
-{[}15{]}) for which one must control {[}16{]} in order to causally
+other factors (\emph{e.g.} system-level changes in productivity caused
+by predatory release {[}15{]}; or favorable environmental conditions
+{[}16{]}) for which one must control {[}17{]} in order to causally
 attribute a biological change to the reserve. Other studies have used a
 control--impact comparison approach that uses control sites but does not
-address temporal variability {[}4,7,17--19{]}. A smaller fraction of
-studies have used a before--after--control--impact (\emph{i.e.} BACI)
-design comparing reserves to control sites before and after
-implementation {[}4,20,21{]}, which allows the use of causal inference
-techniques that estimate the effect of the reserve. But even when proper
-causal inference can be drawn, results are often different across
+address temporal variability {[}4,7,18--20{]}.
+
+A smaller fraction of studies have used a before--after--control--impact
+(\emph{i.e.} BACI) design comparing reserves to control sites before and
+after implementation {[}4,21,22{]}, which allows the use of causal
+inference techniques that estimate the effect of the reserve. For
+example, in ref {[}21{]} authors use a BACI design and observe increases
+in lobster catches --a proxy for abundances-- after the implementation
+of the reserves for protected and control sites. However, the temporal
+changes in the reserve were greater than in the control site, suggesting
+a positive effect off the reserve on lobster catches. But even when
+proper causal inference can be drawn, results are often different across
 reserves. Effects of reserves on ecological and economic outcomes are
@@
 Standardization of marine reserve evaluation is not new. The IUCN
-framework ``How is your MPA doing?'' {[}22,23{]} provides a
+framework ``How is your MPA doing?'' {[}23,24{]} provides a
 comprehensive list of biological, socioeconomic, and governance
@@
 little guidance about establishing causal inference about the reserve.
-Recent work {[}24{]} integrates these three dimensions via the Social
-Ecological Systems Framework {[}25,26{]} and suggests the use of causal
+Recent work {[}25{]} integrates these three dimensions via the Social
+Ecological Systems Framework {[}26,27{]} and suggests the use of causal
 inference techniques to provide a measure of the effect of conservation
@@
 scalable, and replicable is through Open Science and the development of
-open--access tools {[}27{]}. The Ocean Health Index {[}28,29{]}, for
+open--access tools {[}28{]}. The Ocean Health Index {[}29,30{]}, for
 example, successfully standardized a way to measure the health and
 benefits of the oceans. This approach has been implemented at a global
-scale, but also at country-level {[}30{]}, and regionally {[}31,32{]}.
+scale, but also at country-level {[}31{]}, and regionally {[}32,33{]}.
 Open access tools are not limited to conservation, and have also been
-developed to evaluate fishery performance {[}33,34{]}, design
-territorial use rights for fisheries {[}35{]}, and improve decision
-making in the hydro power industry {[}36{]}.
+developed to evaluate fishery performance {[}34,35{]}, design
+territorial use rights for fisheries {[}36{]}, and improve decision
+making in the hydro power industry {[}37{]}.
 
@@
 methods from program evaluation to compare a reserve to control sites
-along a number of ecological, economic, and governance dimensions. We
+along a number of biological, economic, and governance dimensions. We
 first provide a list of commonly stated management objectives and match
@@
 analyzing these indicators building on causal inference techniques
-{[}20{]}, which help us understand the effect of management
-interventions {[}24,37{]}. To implement the analytical approach, we
+{[}21{]}, which help us understand the effect of management
+interventions {[}25,38{]}. To implement the analytical approach, we
 introduce the Marine Reserve Evaluation Application (MAREA), an open
@@
 the rest of the world. Thus, we group these objectives into seven major
-categories, which may be applied to marine reserves worldwide. Any given
+categories that may be applied to marine reserves worldwide. Any given
 reserve may have been implemented to meet one or more of these. The list
-includes objectives stated in legislation {[}38,39{]} and official
+includes objectives stated in legislation {[}39,40{]} and official
 documents such as the Technical Justification Studies (\emph{Estudios
@@
 identified indicators that were used to measure similar
-objectives{[}3--5,7,11,13,14,17--20,22,23,40--43{]}. A first filter
+objectives{[}3--5,7,11,13,14,18--21,23,24,41--44{]}. A first filter
 eliminated indicators for which baseline data do not typically exist in
@@
 Rosario, Baja California, who provided input. Our final list of
-indicators includes those identified in review works {[}4,43{]}.
+indicators includes those identified in review works {[}4,44{]}.
 
@@
 access rights to the fishery, number of fishers, legal recognition of
-the reserve). Some indicators are quantitative and require a numerical
-entry (\emph{e.g.} Fish biomass) while others are qualitative and rely
-on a descriptive entry (\emph{e.g.} Reasoning for reserve location).
-Many of them specifically measure an outcome of the reserve, though some
-are designed to further the understanding of the mechanisms driving a
-reserve's performance. In that sense, most biological and socioeconomic
-indicators are outcome variables. On the other hand, governance
-indicators are viewed as possible explanatory variables of reserve
-performance. Whenever an indicator is applied to ``Target species'', it
-means that the indicator can be used for all species (\emph{e.g.} Fish
-Biomass) and/or for individual species that are either the conservation
-target of the reserve or are of particular economic or ecological
-interest (\emph{e.g.} Grouper Biomass). Finally, indicators B3 and B4
-are different in that B3 look only at the density of organisms above
-size at first maturity (related to reproductive potential), while B4
-measures the density of all fish or of a target species. Each indicator
-targets different plausible desired outcomes, like increased
-reproductive potential (\emph{i.e.} B3; {[}44{]}) or having more fish
--regardless of their size- to atract tourism (\emph{i.e.} B4). Table 1
-presents the proposed indicators, and Table 2 shows how objectives are
-matched with biological and socioeconomic indicators. Governance
-indicators are excluded from Table 2, as we suggest that all of them
-should be used in an evaluation, regardless of the objectives.
+the reserve). Most biological and socioeconomic indicators are
+quantitative and require a numerical entry (\emph{e.g.} Fish biomass)
+while all governance indicators, one biological indicator, and one
+socioeconomic indicator are qualitative and rely on a descriptive entry
+(\emph{e.g.} Reasoning for reserve location). Many of them specifically
+measure an outcome of the reserve, though some are designed to further
+the understanding of the mechanisms driving a reserve's performance. In
+that sense, most biological and socioeconomic indicators are outcome
+variables. On the other hand, governance indicators are viewed as
+possible explanatory variables of reserve performance. Whenever an
+indicator is applied to ``Target species'', it means that the indicator
+can be used for all species (\emph{e.g.} Fish Biomass) and/or for
+individual species that are either the conservation target of the
+reserve or are of particular economic or ecological interest
+(\emph{e.g.} Grouper Biomass). Finally, indicators B3 and B4 are
+different in that B3 only looks at the density of organisms above size
+at first maturity (related to reproductive potential), while B4 measures
+the density of all fish or of a target species. Each indicator targets
+different plausible desired outcomes, like increased reproductive
+potential (\emph{i.e.} B3; {[}45{]}) or having more fish -regardless of
+their size- to attract tourism (\emph{i.e.} B4). Table 1 presents the
+proposed indicators, and Table 2 shows how objectives are matched with
+biological and socioeconomic indicators. Governance indicators are
+excluded from Table 2, as governance indicators are important to
+consider for every objective.
 
@@
 per transect (Table 1). More information on data collection and
-formatting is provided in a guidebook {[}45{]}, which is available in
+formatting is provided in a guidebook {[}46{]}, which is available in
 English and Spanish in MAREA's interface.
@@
 This sampling design for biological data allows us to use causal
-inference techniques {[}20,46{]} to evaluate the effect of the reserve
+inference techniques {[}21,47{]} to evaluate the effect of the reserve
 on biological indicators. The hypothesis that the indicators will
@@
 and temporal changes in each numeric biological indicator (all but B5)
-using generalized linear models {[}20{]}. To account for variations in
+using generalized linear models {[}21{]}. To account for variations in
 the environment and survey conditions, covariates that are gathered
@@
 sometimes prices for each species. To control for inflation, income is
-adjusted with the country's consumer price index {[}47{]}:
+adjusted with the country's consumer price index {[}48{]}:
 
@@
 ordinary least squares, and calculate heteroskedastic--robust standard
-errors {[}48{]}.
+errors {[}49{]}.
 
@@
 Table). This approach has been proven to successfully evaluate
-governance structures {[}49{]}. Unlike with biological and socioeconomic
+governance structures {[}50{]}. Unlike with biological and socioeconomic
 objectives (see eqs 1 and 3), MAREA does not quantitatively analyze
@@
 
-We developed MAREAin R version 3.4.2 and R Studio 1.1.383 {[}50{]} using
-the Shiny package {[}51{]}, to build an interactive web application
+We developed MAREAin R version 3.4.2 and R Studio 1.1.383 {[}51{]} using
+the Shiny package {[}52{]}, to build an interactive web application
 hosted on an open server; the MAREA app can be accessed at
@@
 MAREA automatically matches to appropriate indicators, based on Table 2.
-Users can also select and un-select indicators based on their interests
+Users can also select and deselect indicators based on their interests
 and data availability by ``clicking'' on the check-boxes in MAREA. The
@@
 The scorecard is produced with functions from the Shinydashboard package
-{[}52{]}. The technical report is produced by a parameterized Rmarkdown
-document {[}53{]} processed by the knitr package {[}54{]}. Another
+{[}53{]}. The technical report is produced by a parameterized Rmarkdown
+document {[}54{]} processed by the knitr package {[}55{]}. Another
 feature of MAREA is that the user can choose to share the data. Once the
@@
 y Pescadores de la Baja California}) that promotes co-management of
-marine resources {[}55,56{]}. Additionally, fishers have Territorial Use
+marine resources {[}56,57{]}. Additionally, fishers have Territorial Use
 Rights for Fisheries (TURFs) that provide them with exclusive access
 rights to exploit the benthic marine resources within a given perimeter
-{[}56{]}.
+{[}57{]}.
 
@@
 www.inegi.org.mx/geo/contenidos/geoestadistica/m\_geoestadistico.aspx)
-and the tmap R package {[}57{]}, respectively.
-
-In 2006, the Isla Natividad community implemented two community--based
-marine reserves within their TURF {[}8,40,58{]} after establishing a
-baseline for the soon--to--be reserves and control sites. Evidence
-suggest that these reserves have been effective at enhancing resilience
-to climate variations {[}8{]} and preserving genetic diversity of high
-value commercial species such as abalone {[}10{]}. These ecological
-benefits have been translated into economic benefits, enhancing
-population persistence and bolstering abalone fisheries {[}42{]}. For
-the purpose of this evaluation, we focused on the ``La Plana / Las
-Cuevas'' marine reserve, located at the southern end of the island (Fig.
-3) and its corresponding control site ``La Dulce / Babencho''.
+and the tmap R package {[}58{]}, respectively.
+
+In 2006, the Isla Natividad community established a biological baseline
+following the data collection protocol described in this study. The
+community then implemented two community--based marine reserves within
+their TURF {[}8,41,59{]} after establishing a baseline for the
+soon--to--be reserves and control sites. Evidence suggest that these
+reserves have been effective at enhancing resilience to climate
+variations {[}8{]} and preserving genetic diversity of high value
+commercial species such as abalone {[}10{]}. These ecological benefits
+have been translated into economic benefits, enhancing population
+persistence and bolstering abalone fisheries {[}43{]}. For the purpose
+of this evaluation, we focused on the ``La Plana / Las Cuevas'' marine
+reserve, located at the southern end of the island (Fig. 3) and its
+corresponding control site ``La Dulce / Babencho''.
 
@@
 information along 30 m transects, with a sampling window of 2 m x 2 m
-following a standardized ReefCheck protocol {[}59{]}. Ecological surveys
+following a standardized ReefCheck protocol {[}60{]}. Ecological surveys
 were performed yearly in each reserve and corresponding control site(s),
@@
 landings and income were aggregated by year and species, and adjusted by
-the Consumer Price Index {[}47{]}. From the nine species available, we
+the Consumer Price Index {[}48{]}. From the nine species available, we
 selected as objective species those that contributed the most (88.27\%)
@@
 suggest that the community is strongly organized, which is a likely
-driver of the successes reported above {[}60{]}. The first point of
+driver of the successes reported above {[}61{]}. The first point of
 success is the existence of a fishing cooperative that is also
@@
 governance structures allow various levels of organization that have
-been shown to foster communication and cooperation {[}49,56{]};
-federations also provide bargain power with governments {[}49,61{]}.
+been shown to foster communication and cooperation {[}50,57{]};
+federations also provide bargain power with governments {[}50,62{]}.
 Access to fishing resources is managed through a TURF, permits, and
-fishing quotas (for some species). McCay {[}55{]} suggests that the TURF
+fishing quotas (for some species). McCay {[}56{]} suggests that the TURF
 promotes a sense of stewardship of their resources and incentivizes
@@
 than to discuss particularities of the marine reserve effectiveness,
-which has been described before {[}8,10,42{]}. We use examples from the
+which has been described before {[}8,10,43{]}. We use examples from the
 case study to build on the utility of MAREA and discuss ways in which
@@
 The causal inference techniques used by MAREA have been suggested
-{[}37,46{]} and used {[}20{]} before in other ad hoc studies. This
+{[}38,47{]} and used {[}21{]} before in other ad hoc studies. This
 approach reduces ambiguity in the interpretation of results. For
@@
 The value of the \(\beta_2\) coefficient represents the difference
-between the temporal trends of the reserve and control sites {[}20{]}.
+between the temporal trends of the reserve and control sites {[}21{]}.
 As exemplified by the case of invertebrate densities, a positive value
@@
 which impede us from using the causal inference techniques employed to
-measure biological changes {[}24{]}. However, we can still extract
+measure biological changes {[}25{]}. However, we can still extract
 useful information from them. Again, by combining results from multiple
@@
 We recognize that the seven objectives and 29 indicators used by MAREA
-might not fully describe a reserve. In order to ensure the applicability
-of the tool to reserves in other countries, further testing in other
-regions should take place. However, the proposed objectives and
-indicators provide a starting point to perform the evaluation, to which
-managers and users can add other indicators (\emph{e.g.} larval
-dispersal or connectivity) that are relevant to their reserve.
-Furthermore, MAREA's value is that it provides a free, simple, and
-replicable way to perform rigorous impact analysis. The tool can easily
-be used by fishers, CSO members, and managers in government agencies,
-providing transparency of the analysis and results. In addition, it can
-empower and enable local managers and fishers to respond to local change
-and adapt by allowing direct and easy access to the information.
+might not fully describe a reserve in countries other than Mexico. In
+order to ensure the applicability of the tool to reserves in other
+countries, further testing in other regions should take place. However,
+the proposed objectives and indicators provide a starting point to
+perform the evaluation, to which managers and users can add other
+indicators (\emph{e.g.} larval dispersal or connectivity) that are
+relevant to their reserve. Furthermore, MAREA's value is that it
+provides a free, simple, and replicable way to perform rigorous impact
+analysis. The tool can easily be used by fishers, CSO members, and
+managers in government agencies, providing transparency of the analysis
+and results. In addition, it can empower and enable local managers and
+fishers to respond to local change and adapt by allowing direct and easy
+access to the information.
 
@@
 coefficients in Eqn. 1. It is not uncommon for control sites or
-baselines that describe the biological state of the area before the
-management intervention to be abscent. Propperly designing marine
-reserves by identifying control sites and establishing a baseline before
-the implementation of the reserve is enough to overcome this issue;
-reserves for which there is no control site and baseline can not be
-evaluated with MAREA. On the second case, typical underwater surveys
-require that at least 12 - 16 transects are performed on each site
-(\emph{i.e.} reserve and control) each year. This provides with at least
-48 samples (12 samples per site, per year), enough to avoid overfitting
-Eqn 1. On either case, these problems can be easily avoided by
-accounting for data requirements in the eventual evaluation, an
-important phase during the design process.
+baselines to be absent. Properly designing marine reserves by
+identifying control sites and establishing a baseline before the
+implementation of the reserve is enough to overcome this issue; reserves
+for which there is no control site and baseline cannot be evaluated with
+MAREA. Typical underwater surveys require that at least 12 - 16
+transects are performed on each site (\emph{i.e.} reserve and control)
+each year. This provides with at least 48 samples (12 samples per site,
+per year), enough to avoid overfitting Eqn 1. However, these problems
+can be easily avoided during the design and implementation phases by
+anticipating what data will be needed in the eventual evaluation.
 
 To the best of our knowledge, MAREA is the first tool designed to
-evaluate marine reserves. Previous work {[}22,24{]} addressed MPA
+evaluate marine reserves. Previous work {[}23,25{]} addressed MPA
 evaluation and provided the foundation for our contribution. However,
@@
 tools that automatize complex calculations can have an important impact
-in management {[}62{]}. The use of open data science enables the
+in management {[}63{]}. The use of open data science enables the
 creation of open-access tools that can address techincal gaps and
-inprove management {[}27{]}.
+inprove management {[}28{]}.
 
 The effectiveness of marine reserves continues to be a matter of debate
-{[}11,43,63{]}. With current targets set to increase ocean protection,
+{[}11,44,64{]}. With current targets set to increase ocean protection,
 it is important that we understand the effects of our interventions
-{[}37{]} so we can better inform management {[}46{]}. It is therefore
+{[}38{]} so we can better inform management {[}47{]}. It is therefore
 important that academics, managers, fishers, and CSOs have access to
@@
 and other Latin American countries, where management agencies are often
-understaffed and underfunded {[}64{]}, or where materials are often not
+understaffed and underfunded {[}65{]}, or where materials are often not
 available in their language. In this context, MAREA provides a simple
@@
 comprehensively presented represents a valuable step towards effective
-evidence---based management {[}46{]}.
+evidence---based management {[}47{]}.
 
@@
 Romero who provided help with the logistics. Finally, we thank the
-Editor for a professional handling, and two anonymous reviewers for
-their suggestions, which significantly improved the quality of this
-paper.
+editor and two anonymous reviewers for their suggestions, which
+significantly improved the quality of this paper.
 
@@
 
+\hypertarget{ref-chavez_2003-_m}{}
+16. Chavez FP, Ryan J, Lluch-Cota SE, Niquen C M. From anchovies to
+sardines and back: Multidecadal change in the pacific ocean. Science.
+2003;299: 217--221.
+doi:\href{https://doi.org/10.1126/science.1075880}{10.1126/science.1075880}
+
 \hypertarget{ref-davies_2017-ml}{}
-16. Davies TK, Mees CC, Milner-Gulland EJ. Use of a counterfactual
+17. Davies TK, Mees CC, Milner-Gulland EJ. Use of a counterfactual
 approach to evaluate the effect of area closures on fishing location in
@@
 \hypertarget{ref-guidetti_2014-8Z}{}
-17. Guidetti P, Baiata P, Ballesteros E, Di Franco A, Hereu B,
+18. Guidetti P, Baiata P, Ballesteros E, Di Franco A, Hereu B,
 Macpherson E, et al. Large-scale assessment of mediterranean marine
@@
 \hypertarget{ref-friedlander_2017-oI}{}
-18. Friedlander AM, Golbuu Y, Ballesteros E, Caselle JE, Gouezo M,
+19. Friedlander AM, Golbuu Y, Ballesteros E, Caselle JE, Gouezo M,
 Olsudong D, et al. Size, age, and habitat determine effectiveness of
@@
 \hypertarget{ref-rodriguez_2017-PD}{}
-19. Rodriguez AG, Fanning LM. Assessing marine protected areas
+20. Rodriguez AG, Fanning LM. Assessing marine protected areas
 effectiveness: A case study with the tobago cays marine park. OJMS.
@@
 \hypertarget{ref-moland_2013-VP}{}
-20. Moland E, Olsen EM, Knutsen H, Garrigou P, Espeland SH, Kleiven AR,
+21. Moland E, Olsen EM, Knutsen H, Garrigou P, Espeland SH, Kleiven AR,
 et al. Lobster and cod benefit from small-scale northern marine
@@
 \hypertarget{ref-soykan_2015-nu}{}
-21. Soykan CU, Lewison RL. Using community-level metrics to monitor the
+22. Soykan CU, Lewison RL. Using community-level metrics to monitor the
 effects of marine protected areas on biodiversity. Conserv Biol.
@@
 \hypertarget{ref-pomeroy_2005-Py}{}
-22. Pomeroy RS, Watson LM, Parks JE, Cid GA. How is your mpa doing? A
+23. Pomeroy RS, Watson LM, Parks JE, Cid GA. How is your mpa doing? A
 methodology for evaluating the management effectiveness of marine
@@
 \hypertarget{ref-pomeroy_2004-23}{}
-23. Pomeroy RS, Parks JE, Watson LM. How is your mpa doing ? A guidebook
+24. Pomeroy RS, Parks JE, Watson LM. How is your mpa doing ? A guidebook
 of natural and social indicators for evaluating marine protected areas
@@
 \hypertarget{ref-mascia_2017-m_}{}
-24. Mascia MB, Fox HE, Glew L, Ahmadia GN, Agrawal A, Barnes M, et al. A
+25. Mascia MB, Fox HE, Glew L, Ahmadia GN, Agrawal A, Barnes M, et al. A
 novel framework for analyzing conservation impacts: Evaluation, theory,
@@
 \hypertarget{ref-ostrom_2009-hg}{}
-25. Ostrom E. A general framework for analyzing sustainability of
+26. Ostrom E. A general framework for analyzing sustainability of
 social-ecological systems. Science. 2009;325: 419--422.
@@
 \hypertarget{ref-basurto_2013-oq}{}
-26. Basurto X, Gelcich S, Ostrom E. The social--ecological system
+27. Basurto X, Gelcich S, Ostrom E. The social--ecological system
 framework as a knowledge classificatory system for benthic small-scale
@@
 \hypertarget{ref-lowndes_2017-xh}{}
-27. Lowndes JSS, Best BD, Scarborough C, Afflerbach JC, Frazier MR,
+28. Lowndes JSS, Best BD, Scarborough C, Afflerbach JC, Frazier MR,
 O'Hara CC, et al. Our path to better science in less time using open
@@
 \hypertarget{ref-halpern_2012-k9}{}
-28. Halpern BS, Longo C, Hardy D, McLeod KL, Samhouri JF, Katona SK, et
+29. Halpern BS, Longo C, Hardy D, McLeod KL, Samhouri JF, Katona SK, et
 al. An index to assess the health and benefits of the global ocean.
@@
 \hypertarget{ref-halpern_2017-Zi}{}
-29. Halpern BS, Frazier M, Afflerbach J, O'Hara C, Katona S, Stewart
+30. Halpern BS, Frazier M, Afflerbach J, O'Hara C, Katona S, Stewart
 Lowndes JS, et al. Drivers and implications of change in global ocean
@@
 \hypertarget{ref-selig_2015-F9}{}
-30. Selig ER, Frazier M, O'Leary JK, Jupiter SD, Halpern BS, Longo C, et
+31. Selig ER, Frazier M, O'Leary JK, Jupiter SD, Halpern BS, Longo C, et
 al. Measuring indicators of ocean health for an island nation: The ocean
@@
 \hypertarget{ref-halpern_2014-lQ}{}
-31. Halpern BS, Longo C, Scarborough C, Hardy D, Best BD, Doney SC, et
+32. Halpern BS, Longo C, Scarborough C, Hardy D, Best BD, Doney SC, et
 al. Assessing the health of the u.S. west coast with a regional-scale
@@
 \hypertarget{ref-elfes_2014-RC}{}
-32. Elfes CT, Longo C, Halpern BS, Hardy D, Scarborough C, Best BD, et
+33. Elfes CT, Longo C, Halpern BS, Hardy D, Scarborough C, Best BD, et
 al. A regional-scale ocean health index for brazil. PLoS ONE. 2014;9:
@@
 \hypertarget{ref-anderson_2015-ND}{}
-33. Anderson JL, Anderson CM, Chu J, Meredith J, Asche F, Sylvia G, et
+34. Anderson JL, Anderson CM, Chu J, Meredith J, Asche F, Sylvia G, et
 al. The fishery performance indicators: A management tool for triple
@@
 \hypertarget{ref-dowling_2016-pO}{}
-34. Dowling N, Wilson J, Rudd M, Babcock E, Caillaux M, Cope J, et al.
+35. Dowling N, Wilson J, Rudd M, Babcock E, Caillaux M, Cope J, et al.
 FishPath: A decision support system for assessing and managing data- and
@@
 \hypertarget{ref-oyanedel_2017-TO}{}
-35. Oyanedel R, Macy Humberstone J, Shattenkirk K, Rodriguez Van-Dyck S,
+36. Oyanedel R, Macy Humberstone J, Shattenkirk K, Rodriguez Van-Dyck S,
 Joye Moyer K, Poon S, et al. A decision support tool for designing
@@
 \hypertarget{ref-vilela_2017-Zo}{}
-36. Vilela T, Reid J. Improving hydropower choices via an online and
+37. Vilela T, Reid J. Improving hydropower choices via an online and
 open access tool. PLoS ONE. 2017;12: e0179393.
@@
 \hypertarget{ref-burgess_2016-HN}{}
-37. Burgess MG, Clemence M, McDermott GR, Costello C, Gaines SD. Five
+38. Burgess MG, Clemence M, McDermott GR, Costello C, Gaines SD. Five
 rules for pragmatic blue growth. Marine Policy. 2016;
@@
 \hypertarget{ref-nom049sagpesc_2014-V6}{}
-38. NOM-049-SAG/PESC. NORMA oficial mexicana nom-049-sag/pesc-2014, que
+39. NOM-049-SAG/PESC. NORMA oficial mexicana nom-049-sag/pesc-2014, que
 determina el procedimiento para establecer zonas de refugio para los
@@
 \hypertarget{ref-lgeepa_2017-jL}{}
-39. LGEEPA. Ley general del equilibrio ecológico y la protección al
+40. LGEEPA. Ley general del equilibrio ecológico y la protección al
 ambiente. DOF. 2017; Available:
@@
 \hypertarget{ref-lester_2017-nh}{}
-40. Lester S, McDonald G, Clemence M, Dougherty D, Szuwalski C. Impacts
+41. Lester S, McDonald G, Clemence M, Dougherty D, Szuwalski C. Impacts
 of turfs and marine reserves on fisheries and conservation goals:
@@
 \hypertarget{ref-chirico_2017-Rz}{}
-41. Chirico AAD, McClanahan TR, Eklöf JS. Community- and
+42. Chirico AAD, McClanahan TR, Eklöf JS. Community- and
 government-managed marine protected areas increase fish size, biomass
@@
 \hypertarget{ref-rossetto_2015-V0}{}
-42. Rossetto M, Micheli F, Saenz-Arroyo A, Montes JAE, De Leo GA, Rochet
+43. Rossetto M, Micheli F, Saenz-Arroyo A, Montes JAE, De Leo GA, Rochet
 M-J. No-take marine reserves can enhance population persistence and
@@
 \hypertarget{ref-woodcock_2017-Wm}{}
-43. Woodcock P, O'Leary BC, Kaiser MJ, Pullin AS. Your evidence or mine?
+44. Woodcock P, O'Leary BC, Kaiser MJ, Pullin AS. Your evidence or mine?
 Systematic evaluation of reviews of marine protected area effectiveness.
@@
 \hypertarget{ref-carter_2017-Uf}{}
-44. Carter AB, Davies CR, Emslie MJ, Mapstone BD, Russ GR, Tobin AJ, et
+45. Carter AB, Davies CR, Emslie MJ, Mapstone BD, Russ GR, Tobin AJ, et
 al. Reproductive benefits of no-take marine reserves vary with region
@@
 \hypertarget{ref-villaseorderbez_website_2017-xE}{}
-45. Villaseñor-Derbez JC, Faro C, Wright M, Martínez J. A guide to
+46. Villaseñor-Derbez JC, Faro C, Wright M, Martínez J. A guide to
 evaluate the effectiveness of no-take marine reserves in mexico
@@
 \hypertarget{ref-ferraro_2006-oW}{}
-46. Ferraro PJ, Pattanayak SK. Money for nothing? A call for empirical
+47. Ferraro PJ, Pattanayak SK. Money for nothing? A call for empirical
 evaluation of biodiversity conservation investments. PLoS Biol. 2006;4:
@@
 \hypertarget{ref-oecd_website_2017-VV}{}
-47. OECD. Prices - inflation (cpi) - oecd data {[}Internet{]}. 2017.
+48. OECD. Prices - inflation (cpi) - oecd data {[}Internet{]}. 2017.
 Available: \url{https://data.oecd.org/price/inflation-cpi.htm}
@@
 \hypertarget{ref-sandwich_2014}{}
-48. Zeileis A. Econometric computing with hc and hac covariance matrix
+49. Zeileis A. Econometric computing with hc and hac covariance matrix
 estimators. Journal of Statistical Software. 2004;11: 1--17.
@@
 \hypertarget{ref-espinosaromero_2014-PY}{}
-49. Espinosa-Romero MJ, Rodriguez LF, Weaver AH, Villanueva-Aznar C,
+50. Espinosa-Romero MJ, Rodriguez LF, Weaver AH, Villanueva-Aznar C,
 Torre J. The changing role of ngos in mexican small-scale fisheries:
@@
 \hypertarget{ref-rcore_2017}{}
-50. R Core Team. R: A language and environment for statistical computing
+51. R Core Team. R: A language and environment for statistical computing
 {[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
@@
 \hypertarget{ref-shiny_2017}{}
-51. Chang W, Cheng J, Allaire J, Xie Y, McPherson J. Shiny: Web
+52. Chang W, Cheng J, Allaire J, Xie Y, McPherson J. Shiny: Web
 application framework for r {[}Internet{]}. 2017. Available:
@@
 \hypertarget{ref-shinydashboard_2017}{}
-52. Chang W, Borges Ribeiro B. Shinydashboard: Create dashboards with
+53. Chang W, Borges Ribeiro B. Shinydashboard: Create dashboards with
 'shiny' {[}Internet{]}. 2017. Available:
@@
 \hypertarget{ref-rmarkdown_2017}{}
-53. Allaire J, Cheng J, Xie Y, McPherson J, Chang W, Allen J, et al.
+54. Allaire J, Cheng J, Xie Y, McPherson J, Chang W, Allen J, et al.
 Rmarkdown: Dynamic documents for r {[}Internet{]}. 2017. Available:
@@
 \hypertarget{ref-knitr_2017}{}
-54. Xie Y. Knitr: A general-purpose package for dynamic report
+55. Xie Y. Knitr: A general-purpose package for dynamic report
 generation in r {[}Internet{]}. 2017. Available:
@@
 \hypertarget{ref-mccay_2017-1m}{}
-55. McCay B. Territorial use rights in fisheries of the northern pacific
+56. McCay B. Territorial use rights in fisheries of the northern pacific
 coast of mexico. BMS. 2017;93: 69--81.
@@
 \hypertarget{ref-mccay_2014-CN}{}
-56. McCay BJ, Micheli F, Ponce-Díaz G, Murray G, Shester G,
+57. McCay BJ, Micheli F, Ponce-Díaz G, Murray G, Shester G,
 Ramirez-Sanchez S, et al. Cooperatives, concessions, and co-management
@@
 \hypertarget{ref-tmap_2017}{}
-57. Tennekes M. Tmap: Thematic maps {[}Internet{]}. 2017. Available:
+58. Tennekes M. Tmap: Thematic maps {[}Internet{]}. 2017. Available:
 \url{https://CRAN.R-project.org/package=tmap}
@@
 \hypertarget{ref-afflerbach_2014-HP}{}
-58. Afflerbach JC, Lester SE, Dougherty DT, Poon SE. A global survey of
+59. Afflerbach JC, Lester SE, Dougherty DT, Poon SE. A global survey of
 -reserves, territorial use rights for fisheries coupled with marine
@@
 \hypertarget{ref-suman_2010-ez}{}
-59. Suman CS, Saenz-Arroyo A, Dawson C, Luna MC. Manual de instruccion
+60. Suman CS, Saenz-Arroyo A, Dawson C, Luna MC. Manual de instruccion
 de reef check california: Guia de instruccion para el monitoreo del
@@
 \hypertarget{ref-gutirrez_2011-0U}{}
-60. Gutiérrez NL, Hilborn R, Defeo O. Leadership, social capital and
+61. Gutiérrez NL, Hilborn R, Defeo O. Leadership, social capital and
 incentives promote successful fisheries. Nature. 2011;470: 386--389.
@@
 \hypertarget{ref-finkbeiner_2015-87}{}
-61. Finkbeiner EM, Basurto X. Re-defining co-management to facilitate
+62. Finkbeiner EM, Basurto X. Re-defining co-management to facilitate
 small-scale fisheries reform: An illustration from northwest mexico.
@@
 \hypertarget{ref-ball_2009-qi}{}
-62. Ball IR, Possingham HP, Watts ME. Marxan and relatives: Software for
+63. Ball IR, Possingham HP, Watts ME. Marxan and relatives: Software for
 spatial conservation prioritization. Spatial conservation prioritization
@@
 \hypertarget{ref-padleton_2017-vn}{}
-63. Padleton L, Aghmadia G, Browman H, Thurstand R, Kaplan D, Bartolino
+64. Padleton L, Aghmadia G, Browman H, Thurstand R, Kaplan D, Bartolino
 V. Debating the effectiveness of marine protected areas. ICES Journal of
@@
 \hypertarget{ref-lundquist_2005-OL}{}
-64. Lundquist CJ, Granek EF. Strategies for successful marine
+65. Lundquist CJ, Granek EF. Strategies for successful marine
 conservation: Integrating socioeconomic, political, and scientific

### 2017-12-15 Manuscript.Rmd
updated manuscript with comments from Melaina

@@
 ---
-title: A user--friendly tool to evaluate the effectiveness of no--take marine reserves
+title: A user-friendly tool to evaluate the effectiveness of no-take marine reserves
 subtitle: Marine reserve evaluation
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = F)
+knitr::opts_chunk$set(echo = F, warning = F, message = F)
 options(knitr.table.format = "latex")
@@
 Its purpose is to evaluate the extent to which any given reserve has achieved its stated objectives.
-MAREA provides specific guidance on data collection and formatting, and then conducts rigorous causal inference analysis based on data input by the user, providing real--time outputs about the effectiveness of the reserve.
-MAREA’s ease of use, standardization of state--of--the--art inference methods, and ability to analyze marine reserve effectiveness across ecological, socioeconomic, and governance objectives could dramatically further our understanding and support of effective marine reserve management.
+MAREA provides specific guidance on data collection and formatting, and then conducts rigorous causal inference analysis based on data input by the user, providing real-time outputs about the effectiveness of the reserve.
+MAREA’s ease of use, standardization of state-of-the-art inference methods, and ability to analyze marine reserve effectiveness across ecological, socioeconomic, and governance objectives could dramatically further our understanding and support of effective marine reserve management.
 
@@
 This approach does not account for other factors (*e.g.* system-level changes in productivity caused by predatory release @szuwalski_2017-jc; or favorable environmental conditions @chavez_2003-_m) for which one must control [@davies_2017-ml] in order to causally attribute a biological change to the reserve.
-Other studies have used a control--impact comparison approach that uses control sites but does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
+Other studies have used a control-impact comparison approach that uses control sites but does not address temporal variability [@guidetti_2014-8Z; @friedlander_2017-oI; @lester_2009-Ks; @aburtooropeza_2011-ya; @rodriguez_2017-PD].
 
-A smaller fraction of studies have used a before--after--control--impact (*i.e.* BACI) design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.
-For example, in ref [@moland_2013-VP] authors use a BACI design and observe increases in lobster catches --a proxy for abundances-- after the implementation of the reserves for protected and control sites.
-However, the temporal changes in the reserve were greater than in the control site, suggesting a positive effect off the reserve on lobster catches.
+A smaller fraction of studies have used a before-after-control-impact (*i.e.* BACI) design comparing reserves to control sites before and after implementation [@moland_2013-VP; @soykan_2015-nu; @lester_2009-Ks], which allows the use of causal inference techniques that estimate the effect of the reserve.
+For example, in ref [@moland_2013-VP] authors use a BACI design and observe increases in lobster catches --a proxy for abundances-- after reserve implementation for protected and control sites.
+However, the temporal changes in the reserve were greater than in the control site, suggesting a positive effect of the reserve on lobster catches.
 But even when proper causal inference can be drawn, results are often different across reserves.
@@
 
-An increasingly popular way to make science accessible, reproducible, scalable, and replicable is through Open Science and the development of open--access tools [@lowndes_2017-xh].
+An increasingly popular way to make science accessible, reproducible, scalable, and replicable is through Open Science and the development of open-access tools [@lowndes_2017-xh].
 The Ocean Health Index [@halpern_2012-k9; @halpern_2017-Zi], for example, successfully standardized a way to measure the health and benefits of the oceans.
@@
 
-The purpose of this paper is to describe a user--friendly tool, called “MAREA”, to rigorously systematize the evaluation of marine reserve effectiveness in terms of fisheries and marine conservation.
-The tool is in the form of an open--source application that uses state--of--the--art methods from program evaluation to compare a reserve to control sites along a number of biological, economic, and governance dimensions.
+The purpose of this paper is to describe a user-friendly tool, called “MAREA”, to rigorously systematize the evaluation of marine reserve effectiveness in terms of fisheries and marine conservation goals.
+The tool is in the form of an open-source application that uses state-of-the-art methods from program evaluation to compare a reserve to control sites along a number of biological, economic, and governance dimensions.
 We first provide a list of commonly stated management objectives and match them to appropriate indicators.
@@
 Alongside, methodologies to analyze these indicators are presented.
-We then describe the development of MAREA and explain how this open--access tool can be used by fishermen, managers, and other stakeholders with little scientific background.
+We then describe the development of MAREA and explain how this tool can be used by fishermen, managers, and other stakeholders with little scientific background.
 Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA to inform management.
@@
 Thus, our focus was on identifying common objectives of marine reserves in Mexico.
-However, a literature review and discussions with marine reserve researchers suggest that the objectives driving Mexican marine reserve implementation are similar to those in the rest of the world.
+However, a literature review and discussions with marine reserve researchers suggested that the objectives driving Mexican marine reserve implementation are similar to those in the rest of the world.
 Thus, we group these objectives into seven major categories that may be applied to marine reserves worldwide.
@@
 Based on these seven objectives, we determined a set of associated indicators to evaluate reserve effectiveness.
-These indicators are specific variables on which data could be collected, and analyzed, to ultimately determine whether the corresponding objective was causally being achieved by the marine reserve.
+These indicators are specific variables on which data could be collected and analyzed, to ultimately determine whether the corresponding objective was causally being achieved by the marine reserve.
 The list of indicators was compiled through a review of scientific literature in which we identified indicators that were used to measure similar objectives[@lester_2017-nh; @sala_2017-69; @chirico_2017-Rz; @edgar_2014-UO; @rodriguez_2017-PD; @rossetto_2015-V0; @betti_2017-lq; @sala_2016-PV; @woodcock_2017-Wm; @friedlander_2017-oI; @moland_2013-VP; @aburtooropeza_2011-ya; @guidetti_2014-8Z; @lester_2008-F_; @lester_2009-Ks; @pomeroy_2005-Py; @pomeroy_2004-23].
@@
 Table 1 presents the proposed indicators, and Table 2 shows how objectives are matched with biological and socioeconomic indicators.
-Governance indicators are excluded from Table 2, as governance indicators are important to consider for every objective.
+Governance indicators are excluded from Table 2, but should be considered for every objective as each serves as a plausible explanatory variable for reserve performance.
+
+\clearpage
+
+**Table 1. List of indicators to evaluate the effectiveness of no-take marine reserves.**
 
@@
 read.csv("./Tables/Table_indicators.csv") %>% 
-  knitr::kable(caption = "List of indicators to evaluate the effectiveness of no-take marine reserves.",
-               col.names = c("Code", "Indicator", "Data type", "Unit")) %>%
+  knitr::kable(col.names = c("Code", "Indicator", "Data type", "Unit")) %>%
   kable_styling(latex_options = c("scale_down")) %>%
@@
   group_rows("Governance", 11, 25) %>% 
-  add_footnote(c("The indicator is also applied to target species"), notation = "symbol")
+  kableExtra::row_spec(0, bold = T)
 ```
+\* Indicates the indicator is applied to target species
 
 \clearpage
+
+**Table 2. Management objectives and respective performance indicators.**
 
@@
   select(1:15) %>% 
-  knitr::kable(caption = "Management objectives and respective performance indicators.",
-               col.names = c("Objective", "B1", "B2", "B3", "B4", "B4*", "B5", "B6", "B7", "B7*", "S1", "S1*", "S2", "S2*", "S3")) %>% 
+  knitr::kable(col.names = c("Objective", "B1", "B2", "B3", "B4", "B4*", "B5", "B6", "B7", "B7*", "S1", "S1*", "S2", "S2*", "S3")) %>% 
   kable_styling(latex_options = c("scale_down"))  %>% 
   column_spec(1, width = "4cm") %>% 
-  add_footnote(c("The indicator is applied to target species"), notation = "symbol")
+  kableExtra::row_spec(0, bold = T)
 ```
+Governance indicators are excluded from the table, but all should be used for any objective.
+
+\* Indicates the indicator is applied to target species
 
@@
 
-Socioeconomic data are often collected by fishers, natural resource management agencies, or CSOs by recording landings, income, and sometimes prices for each species.
+Socioeconomic data are often collected by fishers, natural resource management agencies, or Civil Society Organizations (CSOs) by recording landings, income, and sometimes prices for each species.
 To control for inflation, income is adjusted with the country’s consumer price index [@oecd_website_2017-VV]:
@@
 1 and eq.
-3), we estimate the model coefficient with ordinary least squares, and calculate heteroskedastic–robust standard errors [@sandwich_2014].
+3), we estimate the model coefficients with ordinary least squares, and calculate heteroskedastic–robust standard errors [@sandwich_2014].
 
@@
 
-We developed MAREAin R version 3.4.2 and R Studio 1.1.383 [@rcore_2017] using the Shiny package [@shiny_2017], to build an interactive web application hosted on an open server; the MAREA app can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/).
+We developed MAREA in R version 3.4.2 and R Studio 1.1.383 [@rcore_2017] using the Shiny package [@shiny_2017], to build an interactive web application hosted on an open server; the MAREA app can be accessed at [turfeffect.shinyapps.io/marea](turfeffect.shinyapps.io/marea/).
 While the original version was developed in Spanish because it was aimed for Mexico and other Latin-American countries, all of its content can be translated by a translation widget available within the app.
 
-MAREA is designed as a 6--step process, divided in tabs that appear upon launching the app.
+MAREA is designed as a 6-step process, divided in tabs that appear upon launching the app.
 The first tab introduces the app and summarizes the evaluation process.
@@
 General location of Isla Natividad (left) and map of the island (right).** The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce).
-Shapefiles for Mexican coastline and the US were obtained from the Instituto Nacional de Estadistica, Geografia e Informatica of Mexico (INEGI: www.inegi.org.mx/geo/contenidos/geoestadistica/m_geoestadistico.aspx) and the tmap R package [@tmap_2017], respectively.
+Shapefiles for Mexican coastline and the United States were obtained from the Instituto Nacional de Estadistica, Geografia e Informatica of Mexico (INEGI: www.inegi.org.mx/geo/contenidos/geoestadistica/m_geoestadistico.aspx) and the tmap R package [@tmap_2017], respectively.
 
@@
 In 2006, the Isla Natividad community established a biological baseline following the data collection protocol described in this study.
-The community then implemented two community--based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh; @micheli_2012-EU] after establishing a baseline for the soon--to--be reserves and control sites.
+The community then implemented two community-based marine reserves within their TURF [@afflerbach_2014-HP; @lester_2017-nh; @micheli_2012-EU] after establishing a baseline for the soon-to-be reserves and control sites.
 Evidence suggest that these reserves have been effective at enhancing resilience to climate variations [@micheli_2012-EU] and preserving genetic diversity of high value commercial species such as abalone [@munguavega_2015-yg].
@@
 Socioeconomic data were obtained from the National Commission for Aquaculture and Fisheries (*Comisión Nacional de Acuacultura y Pesca*; CONAPESCA).
-The data contains species--level information on monthly landings and income from nine species from 2000 to 2014.
+The data contains species-level information on monthly landings and income from nine species from 2000 to 2014.
 Data on landings and income were aggregated by year and species, and adjusted by the Consumer Price Index [@oecd_website_2017-VV].
 From the nine species available, we selected as objective species those that contributed the most (88.27%) income from 2000 to 2014: lobster (*Panulirus interruptus*; 71.76%), red sea urchin (*Mesocentrotus franciscanus*; 9.33%), snail (*Megastraea undosa*; 3.93%), and sea cucumber (*Parastichopus parvimensis*; 3.23%).
-Abalone species (*Haliotis fulgens*; 4.52% and *Haliotis corrugata*; 6.16%) were excluded because the cooperative implemented an informal closure of these fisheries in 2010 to allow the population recover.
+Abalone species (*Haliotis fulgens*; 4.52% and *Haliotis corrugata*; 6.16%) were excluded because the cooperative implemented an informal closure of these fisheries in 2010 to allow the population to recover.
 Eliminating all fishing pressure on abalones means that the control site receives (for this species) the same treatment as the reserve.
@@
 
-\clearpage
-
 **Fig.
@@
 
+**Table 3. Summary of average treatment effect of the reserve on biological indicators.**
+
 ```{r table of bio results, results = "asis"}
@@
   select(Indicator, estimate, tscore = statistic) %>% 
-  kable(caption = "Summary of average treatment effect of the reserve on biological indicators.",
-        digits = c(NA, NA, 4),
+  kable(digits = c(NA, NA, 4),
         col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
   kable_styling() %>% 
-  add_footnote(c("Asterisks indicate significance level, with (*) indicating p < 0.1 and (**) p < 0.05."),
-               notation = "symbol")
+  row_spec(0, bold = T)
 ```
+\* Indicate significance level, with (\*) indicating p < 0.1 and (\*\*) p < 0.05.
+
+\clearpage
 
 One of the main objectives of this reserve was to increase landings.
-Results of the socioeconomic indicators show that total landings were, on average, 64.20 metric tonnes higher (*p* > 0.1) after the implementation of the reserves, though this cannot necessarily be interpreted as causal, because it relies entirely on a before--after comparison.
+Results of the socioeconomic indicators show that total landings were, on average, 64.20 metric tonnes higher (*p* > 0.1) after the implementation of the reserves, though this cannot necessarily be interpreted as causal, because it relies entirely on a before-after comparison.
 Total income was $10,344.85 (*p* < 0.05) thousands of Mexican Pesos (K MXP) higher after the implementation of the reserves.
@@
 
+**Table 4. Summary of differences in socioeconomic indicators before and after the implementation of the reserve.**
+
 ```{r table of soc results, results = "asis"}
@@
   select(Indicator, estimate, tscore = statistic) %>% 
-  kable(caption = "Summary of differences in socioeconomic indicators before and after the implementation of the reserve.",
-        digits = c(NA, NA, 4),
+  kable(digits = c(NA, NA, 4),
         col.names = c("Indicator", "Estimate (SD)", "t-score")) %>% 
   kable_styling() %>% 
-  add_footnote(c("Asterisks indicate significance level, with (*) indicating p < 0.1 and (**) p < 0.05."),
-               notation = "symbol")
+  row_spec(0, bold = T)
 ```
+
+\* Indicate significance level, with (\*) indicating p < 0.1 and (\*\*) p < 0.05.
 
@@
 McCay [@mccay_2017-1m] suggests that the TURF promotes a sense of stewardship of their resources and incentivizes sustainable management.
-Together, these structures enabled a participative, bottom–up process during the reserve design phase; opinions of all fishing members ---and often non--fishing community members--- were included.
-Participation of community members in reserve surveillance and yearly monitoring indicate commitment and interest, and allow informal communication of results to un--involved community members.
-Furthermore, the reserve is partially isolated from poaching activity and fishers have internal regulations pertaining to the reserves.
+Together, these structures enabled a participative, bottom–up process during the reserve design phase; opinions of all fishing members --and often non-fishing community members-- were included.
+Participation of community members in reserve surveillance and yearly monitoring indicate commitment and interest, and allow informal communication of results to uninvolved community members.
+Furthermore, the reserve is partially isolated from poaching activity, and fishers have internal regulations pertaining to the reserves.
 The low level of illegal fishing by members of the community and outsiders both inside and outside the reserve represents another indication of effectiveness.
@@
 
+\clearpage
+
+**Table 5. Summary of governance indicators.**
+
 ```{r}
 read.csv("Tables/Governance_data.csv", col.names = c("Indicator", "Description"), stringsAsFactors = F) %>% 
-  kable(caption = "Summary of governance indicators.") %>%
+  kable() %>%
   kable_styling(full_width = F) %>% 
-  column_spec(2, width = "9cm")
+  column_spec(2, width = "9cm") %>% 
+  row_spec(0, bold = T)
 ```
@@
 We have developed and presented an automated approach for evaluating the effectiveness of marine reserves in Mexico, and perhaps around the world.
-Here we highlight MAREA’s utility for evidence–-based management, and comment on a few of its shortcomings.
+Here we highlight MAREA’s utility for evidence-based management, and comment on a few of its shortcomings.
 The findings from Isla Natividad are used purely to validate the relevance of MAREA rather than to discuss particularities of the marine reserve effectiveness, which has been described before [@micheli_2012-EU; @rossetto_2015-V0; @munguavega_2015-yg].
@@
 control site only in 2016) would have identified higher densities inside the reserve, concluding that the reserve increases invertebrate density.
-However, by executing a formal difference--in--differences approach for causal inference, MAREA identifies the changes through time and across sites, and estimates the effect of the reserve on density at $\beta_2 = `r formatC(bio_results_summary$estimate[8], digits = 2, format = "f")`$ (*p* < 0.05).
+However, by executing a formal difference-in-differences approach for causal inference, MAREA identifies the changes through time and across sites, and estimates the effect of the reserve on density at $\beta_2 = `r formatC(bio_results_summary$estimate[8], digits = 2, format = "f")`$ (*p* < 0.05).
 This approach reveals that invertebrate densities decrease in both sites through time, but the decrease is faster for the control site, thus yielding a positive value for $\beta_2$.
@@
 
-MAREA provides in--depth analysis and a convenient snapshot overview of the effect of the reserve, allowing users to rapidly identify trends.
+MAREA provides in-depth analysis and a convenient snapshot overview of the effect of the reserve, allowing users to rapidly identify trends.
 However, users must interpret multiple indicators at a time to better understand the results.
@@
 By making results straightforward to interpret, MAREA may also assist in communication with a broader stakeholder community.
-While stakeholder involvement in the design and implementation phases of marine reserves is important, that may not be sufficient for ensuring long--run buy--in or success.
-The scorecard is easily understandable by experts and non--experts, and can be used as an effective tool for communicating the results of annual evaluations.
+While stakeholder involvement in the design and implementation phases of marine reserves is important, that may not be sufficient for ensuring long-term buy-in or success.
+The scorecard is easily understandable by experts and non-experts, and can be used as an effective tool for communicating the results of annual evaluations.
 Additionally, the technical report can serve as a tool for managers and scientists to rapidly produce and communicate information at a more technical level.
@@
 Properly designing marine reserves by identifying control sites and establishing a baseline before the implementation of the reserve is enough to overcome this issue; reserves for which there is no control site and baseline cannot be evaluated with MAREA.
-Typical underwater surveys require that at least 12 - 16 transects are performed on each site (*i.e.* reserve and control) each year.
-This provides with at least 48 samples (12 samples per site, per year), enough to avoid overfitting Eqn 1.
+Typical underwater surveys require that at least 12 - 16 transects are performed for each site (*i.e.* reserve and control) each year.
+This provides at least 48 samples (12 samples per site, per year), enough to avoid overfitting Eqn 1.
 However, these problems can be easily avoided during the design and implementation phases by anticipating what data will be needed in the eventual evaluation.
@@
 Previous work [@pomeroy_2005-Py; @mascia_2017-m_] addressed MPA evaluation and provided the foundation for our contribution.
-However, these did not intended to create user--friendly tools to aid in the evaluation.
+However, these did not intended to create user-friendly tools to aid in the evaluation.
 As with other conservation management tools, development of tools that automatize complex calculations can have an important impact in management [@ball_2009-qi].
-The use of open data science enables the creation of open-access tools that can address techincal gaps and inprove management [@lowndes_2017-xh].
+The use of open data science enables the creation of open-access tools that can address technical gaps and inprove management [@lowndes_2017-xh].
 

### 2017-12-15 Manuscript.tex
updated manuscript with comments from Melaina

@@
 \hypersetup{unicode=true,
-            pdftitle={A user--friendly tool to evaluate the effectiveness of no--take marine reserves},
+            pdftitle={A user-friendly tool to evaluate the effectiveness of no-take marine reserves},
             pdfborder={0 0 0},
@@
 \setlength{\droptitle}{-2em}
-  \title{A user--friendly tool to evaluate the effectiveness of no--take marine
+  \title{A user-friendly tool to evaluate the effectiveness of no-take marine
 reserves}
@@
 conducts rigorous causal inference analysis based on data input by the
-user, providing real--time outputs about the effectiveness of the
-reserve. MAREA's ease of use, standardization of state--of--the--art
+user, providing real-time outputs about the effectiveness of the
+reserve. MAREA's ease of use, standardization of state-of-the-art
 inference methods, and ability to analyze marine reserve effectiveness
@@
 attribute a biological change to the reserve. Other studies have used a
-control--impact comparison approach that uses control sites but does not
+control-impact comparison approach that uses control sites but does not
 address temporal variability {[}4,7,18--20{]}.
 
-A smaller fraction of studies have used a before--after--control--impact
+A smaller fraction of studies have used a before-after-control-impact
 (\emph{i.e.} BACI) design comparing reserves to control sites before and
@@
 example, in ref {[}21{]} authors use a BACI design and observe increases
-in lobster catches --a proxy for abundances-- after the implementation
-of the reserves for protected and control sites. However, the temporal
+in lobster catches --a proxy for abundances-- after reserve
+implementation for protected and control sites. However, the temporal
 changes in the reserve were greater than in the control site, suggesting
-a positive effect off the reserve on lobster catches. But even when
+a positive effect of the reserve on lobster catches. But even when
 proper causal inference can be drawn, results are often different across
@@
 scalable, and replicable is through Open Science and the development of
-open--access tools {[}28{]}. The Ocean Health Index {[}29,30{]}, for
+open-access tools {[}28{]}. The Ocean Health Index {[}29,30{]}, for
 example, successfully standardized a way to measure the health and
@@
 
-The purpose of this paper is to describe a user--friendly tool, called
+The purpose of this paper is to describe a user-friendly tool, called
 ``MAREA'', to rigorously systematize the evaluation of marine reserve
-effectiveness in terms of fisheries and marine conservation. The tool is
-in the form of an open--source application that uses state--of--the--art
-methods from program evaluation to compare a reserve to control sites
-along a number of biological, economic, and governance dimensions. We
-first provide a list of commonly stated management objectives and match
-them to appropriate indicators. We then develop a simple approach to
-analyzing these indicators building on causal inference techniques
-{[}21{]}, which help us understand the effect of management
-interventions {[}25,38{]}. To implement the analytical approach, we
-introduce the Marine Reserve Evaluation Application (MAREA), an open
-source, web--based tool that automates the framework described in this
-paper and enables its broader use. Finally, we present a case study on
-the evaluation of a marine reserve established by the fishers of Isla
-Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.
+effectiveness in terms of fisheries and marine conservation goals. The
+tool is in the form of an open-source application that uses
+state-of-the-art methods from program evaluation to compare a reserve to
+control sites along a number of biological, economic, and governance
+dimensions. We first provide a list of commonly stated management
+objectives and match them to appropriate indicators. We then develop a
+simple approach to analyzing these indicators building on causal
+inference techniques {[}21{]}, which help us understand the effect of
+management interventions {[}25,38{]}. To implement the analytical
+approach, we introduce the Marine Reserve Evaluation Application
+(MAREA), an open source, web--based tool that automates the framework
+described in this paper and enables its broader use. Finally, we present
+a case study on the evaluation of a marine reserve established by the
+fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential
+of MAREA.
 
@@
 presented. We then describe the development of MAREA and explain how
-this open--access tool can be used by fishermen, managers, and other
-stakeholders with little scientific background. Finally, we provide
-guidelines on how to interpret and use the results and output generated
-by MAREA to inform management.
+this tool can be used by fishermen, managers, and other stakeholders
+with little scientific background. Finally, we provide guidelines on how
+to interpret and use the results and output generated by MAREA to inform
+management.
 
@@
 of marine reserves in Mexico. However, a literature review and
-discussions with marine reserve researchers suggest that the objectives
-driving Mexican marine reserve implementation are similar to those in
-the rest of the world. Thus, we group these objectives into seven major
-categories that may be applied to marine reserves worldwide. Any given
-reserve may have been implemented to meet one or more of these. The list
-includes objectives stated in legislation {[}39,40{]} and official
-documents such as the Technical Justification Studies (\emph{Estudios
-Técnicos Justificativos}), agreements, and decrees associated to these
-areas:
+discussions with marine reserve researchers suggested that the
+objectives driving Mexican marine reserve implementation are similar to
+those in the rest of the world. Thus, we group these objectives into
+seven major categories that may be applied to marine reserves worldwide.
+Any given reserve may have been implemented to meet one or more of
+these. The list includes objectives stated in legislation {[}39,40{]}
+and official documents such as the Technical Justification Studies
+(\emph{Estudios Técnicos Justificativos}), agreements, and decrees
+associated to these areas:
 
@@
 indicators to evaluate reserve effectiveness. These indicators are
-specific variables on which data could be collected, and analyzed, to
+specific variables on which data could be collected and analyzed, to
 ultimately determine whether the corresponding objective was causally
@@
 biological and socioeconomic indicators. Governance indicators are
-excluded from Table 2, as governance indicators are important to
-consider for every objective.
-
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-2}List of indicators to evaluate the effectiveness of no-take marine reserves.}
+excluded from Table 2, but should be considered for every objective as
+each serves as a plausible explanatory variable for reserve performance.
+
+\clearpage
+
+\textbf{Table 1. List of indicators to evaluate the effectiveness of
+no-take marine reserves.}
+
+\begin{table}[H]
 \centering
-\resizebox{\linewidth}{!}{\begin{tabular}[t]{l|l|l|l}
-\hline
-Code & Indicator & Data type & Unit\\
+\resizebox{\linewidth}{!}{\begin{tabular}{l|l|l|l}
+\hline
+\bfseries{Code} & \bfseries{Indicator} & \bfseries{Data type} & \bfseries{Unit}\\
 \hline
@@
 \hline
-\multicolumn{4}{l}{\textsuperscript{*} The indicator is also applied to target species}\\
 \end{tabular}}
@@
 
+* Indicates the indicator is applied to target species
+
 \clearpage
 
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-3}Management objectives and respective performance indicators.}
+\textbf{Table 2. Management objectives and respective performance
+indicators.}
+
+\begin{table}[H]
 \centering
-\resizebox{\linewidth}{!}{\begin{tabular}[t]{>{\raggedright\arraybackslash}p{4cm}|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
-\hline
-Objective & B1 & B2 & B3 & B4 & B4* & B5 & B6 & B7 & B7* & S1 & S1* & S2 & S2* & S3\\
+\resizebox{\linewidth}{!}{\begin{tabular}{>{\raggedright\arraybackslash}p{4cm}|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
+\hline
+\bfseries{Objective} & \bfseries{B1} & \bfseries{B2} & \bfseries{B3} & \bfseries{B4} & \bfseries{B4*} & \bfseries{B5} & \bfseries{B6} & \bfseries{B7} & \bfseries{B7*} & \bfseries{S1} & \bfseries{S1*} & \bfseries{S2} & \bfseries{S2*} & \bfseries{S3}\\
 \hline
@@
 \hline
-\multicolumn{15}{l}{\textsuperscript{*} The indicator is applied to target species}\\
 \end{tabular}}
 \end{table}
+
+Governance indicators are excluded from the table, but all should be
+used for any objective.
+
+* Indicates the indicator is applied to target species
 
@@
 Socioeconomic data are often collected by fishers, natural resource
-management agencies, or CSOs by recording landings, income, and
-sometimes prices for each species. To control for inflation, income is
-adjusted with the country's consumer price index {[}48{]}:
+management agencies, or Civil Society Organizations (CSOs) by recording
+landings, income, and sometimes prices for each species. To control for
+inflation, income is adjusted with the country's consumer price index
+{[}48{]}:
 
@@
 implementation of the reserve and provide valuable input. For both
-models (eq. 1 and eq. 3), we estimate the model coefficient with
+models (eq. 1 and eq. 3), we estimate the model coefficients with
 ordinary least squares, and calculate heteroskedastic--robust standard
@@
 
-We developed MAREAin R version 3.4.2 and R Studio 1.1.383 {[}51{]} using
-the Shiny package {[}52{]}, to build an interactive web application
-hosted on an open server; the MAREA app can be accessed at
+We developed MAREA in R version 3.4.2 and R Studio 1.1.383 {[}51{]}
+using the Shiny package {[}52{]}, to build an interactive web
+application hosted on an open server; the MAREA app can be accessed at
 \href{turfeffect.shinyapps.io/marea/}{turfeffect.shinyapps.io/marea}.
@@
 
-MAREA is designed as a 6--step process, divided in tabs that appear upon
+MAREA is designed as a 6-step process, divided in tabs that appear upon
 launching the app. The first tab introduces the app and summarizes the
@@
 approximate location of control sites is indicated by blue squares (B =
-Babencho, D = La Dulce). Shapefiles for Mexican coastline and the US
-were obtained from the Instituto Nacional de Estadistica, Geografia e
-Informatica of Mexico (INEGI:
+Babencho, D = La Dulce). Shapefiles for Mexican coastline and the United
+States were obtained from the Instituto Nacional de Estadistica,
+Geografia e Informatica of Mexico (INEGI:
 www.inegi.org.mx/geo/contenidos/geoestadistica/m\_geoestadistico.aspx)
@@
 following the data collection protocol described in this study. The
-community then implemented two community--based marine reserves within
+community then implemented two community-based marine reserves within
 their TURF {[}8,41,59{]} after establishing a baseline for the
-soon--to--be reserves and control sites. Evidence suggest that these
+soon-to-be reserves and control sites. Evidence suggest that these
 reserves have been effective at enhancing resilience to climate
@@
 Aquaculture and Fisheries (\emph{Comisión Nacional de Acuacultura y
-Pesca}; CONAPESCA). The data contains species--level information on
+Pesca}; CONAPESCA). The data contains species-level information on
 monthly landings and income from nine species from 2000 to 2014. Data on
@@
 were excluded because the cooperative implemented an informal closure of
-these fisheries in 2010 to allow the population recover. Eliminating all
-fishing pressure on abalones means that the control site receives (for
-this species) the same treatment as the reserve.
+these fisheries in 2010 to allow the population to recover. Eliminating
+all fishing pressure on abalones means that the control site receives
+(for this species) the same treatment as the reserve.
 
@@
 \(\beta_2\) coefficients is provided in Table 3.
-
-\clearpage
 
@@
 
-\begin{table}
-
-\caption{\label{tab:table of bio results}Summary of average treatment effect of the reserve on biological indicators.}
+\textbf{Table 3. Summary of average treatment effect of the reserve on
+biological indicators.}
+
+\begin{table}[H]
 \centering
-\begin{tabular}[t]{l|l|r}
-\hline
-Indicator & Estimate (SD) & t-score\\
+\begin{tabular}{l|l|r}
+\hline
+\bfseries{Indicator} & \bfseries{Estimate (SD)} & \bfseries{t-score}\\
 \hline
@@
 \hline
-\multicolumn{3}{l}{\textsuperscript{*} Asterisks indicate significance level, with (*)}\\
-\multicolumn{3}{l}{indicating p < 0.1 and (**) p < 0.05.}\\
 \end{tabular}
 \end{table}
+
+* Indicate significance level, with (*) indicating p \textless{} 0.1 and
+(**) p \textless{} 0.05.
+
+\clearpage
 
@@
 after the implementation of the reserves, though this cannot necessarily
-be interpreted as causal, because it relies entirely on a before--after
+be interpreted as causal, because it relies entirely on a before-after
 comparison. Total income was \$10,344.85 (\emph{p} \textless{} 0.05)
@@
 
-\begin{table}
-
-\caption{\label{tab:table of soc results}Summary of differences in socioeconomic indicators before and after the implementation of the reserve.}
+\textbf{Table 4. Summary of differences in socioeconomic indicators
+before and after the implementation of the reserve.}
+
+\begin{table}[H]
 \centering
-\begin{tabular}[t]{l|l|r}
-\hline
-Indicator & Estimate (SD) & t-score\\
+\begin{tabular}{l|l|r}
+\hline
+\bfseries{Indicator} & \bfseries{Estimate (SD)} & \bfseries{t-score}\\
 \hline
@@
 \hline
-\multicolumn{3}{l}{\textsuperscript{*} Asterisks indicate significance level, with (*)}\\
-\multicolumn{3}{l}{indicating p < 0.1 and (**) p < 0.05.}\\
 \end{tabular}
 \end{table}
+
+* Indicate significance level, with (*) indicating p \textless{} 0.1 and
+(**) p \textless{} 0.05.
 
@@
 participative, bottom--up process during the reserve design phase;
-opinions of all fishing members ---and often non--fishing community
-members--- were included. Participation of community members in reserve
+opinions of all fishing members --and often non-fishing community
+members-- were included. Participation of community members in reserve
 surveillance and yearly monitoring indicate commitment and interest, and
-allow informal communication of results to un--involved community
-members. Furthermore, the reserve is partially isolated from poaching
-activity and fishers have internal regulations pertaining to the
-reserves. The low level of illegal fishing by members of the community
-and outsiders both inside and outside the reserve represents another
-indication of effectiveness. Governance indicators are summarized in
-Table 5.
-
-\begin{table}
-
-\caption{\label{tab:unnamed-chunk-6}Summary of governance indicators.}
+allow informal communication of results to uninvolved community members.
+Furthermore, the reserve is partially isolated from poaching activity,
+and fishers have internal regulations pertaining to the reserves. The
+low level of illegal fishing by members of the community and outsiders
+both inside and outside the reserve represents another indication of
+effectiveness. Governance indicators are summarized in Table 5.
+
+\clearpage
+
+\textbf{Table 5. Summary of governance indicators.}
+
+\begin{table}[H]
 \centering
-\begin{tabular}[t]{l|>{\raggedright\arraybackslash}p{9cm}}
-\hline
-Indicator & Description\\
+\begin{tabular}{l|>{\raggedright\arraybackslash}p{9cm}}
+\hline
+\bfseries{Indicator} & \bfseries{Description}\\
 \hline
@@
 effectiveness of marine reserves in Mexico, and perhaps around the
-world. Here we highlight MAREA's utility for evidence---based
-management, and comment on a few of its shortcomings. The findings from
-Isla Natividad are used purely to validate the relevance of MAREA rather
-than to discuss particularities of the marine reserve effectiveness,
-which has been described before {[}8,10,43{]}. We use examples from the
-case study to build on the utility of MAREA and discuss ways in which
-results can be interpreted to inform management.
+world. Here we highlight MAREA's utility for evidence-based management,
+and comment on a few of its shortcomings. The findings from Isla
+Natividad are used purely to validate the relevance of MAREA rather than
+to discuss particularities of the marine reserve effectiveness, which
+has been described before {[}8,10,43{]}. We use examples from the case
+study to build on the utility of MAREA and discuss ways in which results
+can be interpreted to inform management.
 
@@
 invertebrate density. However, by executing a formal
-difference--in--differences approach for causal inference, MAREA
+difference-in-differences approach for causal inference, MAREA
 identifies the changes through time and across sites, and estimates the
@@
 
-MAREA provides in--depth analysis and a convenient snapshot overview of
+MAREA provides in-depth analysis and a convenient snapshot overview of
 the effect of the reserve, allowing users to rapidly identify trends.
@@
 involvement in the design and implementation phases of marine reserves
-is important, that may not be sufficient for ensuring long--run buy--in
+is important, that may not be sufficient for ensuring long-term buy-in
 or success. The scorecard is easily understandable by experts and
-non--experts, and can be used as an effective tool for communicating the
+non-experts, and can be used as an effective tool for communicating the
 results of annual evaluations. Additionally, the technical report can
@@
 MAREA. Typical underwater surveys require that at least 12 - 16
-transects are performed on each site (\emph{i.e.} reserve and control)
-each year. This provides with at least 48 samples (12 samples per site,
-per year), enough to avoid overfitting Eqn 1. However, these problems
-can be easily avoided during the design and implementation phases by
+transects are performed for each site (\emph{i.e.} reserve and control)
+each year. This provides at least 48 samples (12 samples per site, per
+year), enough to avoid overfitting Eqn 1. However, these problems can be
+easily avoided during the design and implementation phases by
 anticipating what data will be needed in the eventual evaluation.
@@
 evaluation and provided the foundation for our contribution. However,
-these did not intended to create user--friendly tools to aid in the
+these did not intended to create user-friendly tools to aid in the
 evaluation. As with other conservation management tools, development of
@@
 in management {[}63{]}. The use of open data science enables the
-creation of open-access tools that can address techincal gaps and
+creation of open-access tools that can address technical gaps and
 inprove management {[}28{]}.

### 2017-12-18 Manuscript.Rmd
Added all changes for revised submission

@@
 We first provide a list of commonly stated management objectives and match them to appropriate indicators.
-We then develop a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2016-HN; @mascia_2017-m_].
+We then develop a simple approach to analyzing these indicators building on causal inference techniques [@moland_2013-VP], which help us understand the effect of management interventions [@burgess_2018-HN; @mascia_2017-m_].
 To implement the analytical approach, we introduce the Marine Reserve Evaluation Application (MAREA), an open source, web–based tool that automates the framework described in this paper and enables its broader use.
@@
 
-The causal inference techniques used by MAREA have been suggested [@burgess_2016-HN; @ferraro_2006-oW] and used [@moland_2013-VP] before in other ad hoc studies.
+The causal inference techniques used by MAREA have been suggested [@burgess_2018-HN; @ferraro_2006-oW] and used [@moland_2013-VP] before in other ad hoc studies.
 This approach reduces ambiguity in the interpretation of results.
@@
 The effectiveness of marine reserves continues to be a matter of debate [@woodcock_2017-Wm; @edgar_2014-UO; @padleton_2017-vn].
-With current targets set to increase ocean protection, it is important that we understand the effects of our interventions [@burgess_2016-HN] so we can better inform management [@ferraro_2006-oW].
+With current targets set to increase ocean protection, it is important that we understand the effects of our interventions [@burgess_2018-HN] so we can better inform management [@ferraro_2006-oW].
 It is therefore important that academics, managers, fishers, and CSOs have access to open access tools like MAREA.

### 2017-12-18 Manuscript.tex
Added all changes for revised submission

@@
 1. Pauly D, Watson R, Alder J. Global trends in world fisheries: Impacts
-on marine ecosystems and food security. Philos Trans R Soc Lond, B, Biol
-Sci. 2005;360: 5--12.
+on marine ecosystems and food security. Philosophical Transactions of
+the Royal Society B: Biological Sciences. 2005;360: 5--12.
 doi:\href{https://doi.org/10.1098/rstb.2004.1574}{10.1098/rstb.2004.1574}
@@
 6. Hastings A, Gaines SD, Costello C. Marine reserves solve an important
-bycatch problem in fisheries. Proc Natl Acad Sci U S A. 2017;114:
+bycatch problem in fisheries. Proc Natl Acad Sci USA. 2017;114:
 8927--8934.
@@
 \hypertarget{ref-roberts_2017-J9}{}
-9. Roberts CM, O'Leary BC, McCauley DJ, Cury PM, Duarte CM, Lubchenco J,
+9. Roberts CM, OLeary BC, McCauley DJ, Cury PM, Duarte CM, Lubchenco J,
 et al. Marine reserves can mitigate and promote adaptation to climate
-change. Proc Natl Acad Sci U S A. 2017;114: 6167--6175.
+change. Proc Natl Acad Sci USA. 2017;114: 6167--6175.
 doi:\href{https://doi.org/10.1073/pnas.1701262114}{10.1073/pnas.1701262114}
@@
 14. Betti F, Bavestrello G, Bo M, Asnaghi V, Chiantore M, Bava S, et al.
-Over 10 years of variation in mediterranean reef benthic communities.
+Over 10~years of variation in mediterranean reef benthic communities.
 Marine Ecology. 2017;38: e12439.
@@
 15. Szuwalski CS, Burgess MG, Costello C, Gaines SD. High fishery
-catches through trophic cascades in china. Proc Natl Acad Sci U S A.
+catches through trophic cascades in china. Proc Natl Acad Sci USA.
 2017;114: 717--721.
@@
 protected areas: Inference from an empirical before-after control-impact
-study. Proc Biol Sci. 2013;280: 20122679.
+study. Proceedings of the Royal Society B: Biological Sciences.
+2013;280: 20122679--20122679.
 doi:\href{https://doi.org/10.1098/rspb.2012.2679}{10.1098/rspb.2012.2679}
@@
 28. Lowndes JSS, Best BD, Scarborough C, Afflerbach JC, Frazier MR,
-O'Hara CC, et al. Our path to better science in less time using open
-data science tools. Nat ecol evol. 2017;1: 0160.
+OHara CC, et al. Our path to better science in less time using open data
+science tools. Nat ecol evol. 2017;1: 0160.
 doi:\href{https://doi.org/10.1038/s41559-017-0160}{10.1038/s41559-017-0160}
@@
 \hypertarget{ref-halpern_2017-Zi}{}
-30. Halpern BS, Frazier M, Afflerbach J, O'Hara C, Katona S, Stewart
+30. Halpern BS, Frazier M, Afflerbach J, OHara C, Katona S, Stewart
 Lowndes JS, et al. Drivers and implications of change in global ocean
@@
 
-\hypertarget{ref-burgess_2016-HN}{}
+\hypertarget{ref-burgess_2018-HN}{}
 38. Burgess MG, Clemence M, McDermott GR, Costello C, Gaines SD. Five
-rules for pragmatic blue growth. Marine Policy. 2016;
+rules for pragmatic blue growth. Marine Policy. 2018;87: 331--339.
 doi:\href{https://doi.org/10.1016/j.marpol.2016.12.005}{10.1016/j.marpol.2016.12.005}
@@
 \hypertarget{ref-rossetto_2015-V0}{}
-43. Rossetto M, Micheli F, Saenz-Arroyo A, Montes JAE, De Leo GA, Rochet
-M-J. No-take marine reserves can enhance population persistence and
-support the fishery of abalone. Can J Fish Aquat Sci. 2015;72:
-1503--1517.
+43. Rossetto M, Micheli F, Saenz-Arroyo A, Montes JAE, De Leo GA.
+No-take marine reserves can enhance population persistence and support
+the fishery of abalone. Can J Fish Aquat Sci. 2015;72: 1503--1517.
 doi:\href{https://doi.org/10.1139/cjfas-2013-0623}{10.1139/cjfas-2013-0623}

### 2017-12-20 Manuscript.Rmd
Added Tex file with PLoS template

@@
 - exported-references.bib
-- packages.bib
+# - packages.bib
 csl: plos.csl
@@
 This sampling design for biological data allows us to use causal inference techniques [@moland_2013-VP; @ferraro_2006-oW] to evaluate the effect of the reserve on biological indicators.
-The hypothesis that the indicators will respond to implementation of the reserve is tested by analyzing spatial and temporal changes in each numeric biological indicator (all but B5) using generalized linear models [@moland_2013-VP].
+The hypothesis that the indicators will respond to implementation of the reserve is tested by analyzing spatial and temporal changes in each numeric biological indicator using generalized linear models [@moland_2013-VP].
 To account for variations in the environment and survey conditions, covariates that are gathered during the underwater ecological surveys are included in the difference-in-differences model with form:
@@
 1 and eq.
-3), we estimate the model coefficients with ordinary least squares, and calculate heteroskedastic–robust standard errors [@sandwich_2014].
+3), we estimate the model coefficients with ordinary least squares, and calculate heteroskedastic–robust standard errors [@zeileis_2004-7n].
 
@@
 General location of Isla Natividad (left) and map of the island (right).** The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce).
-Shapefiles for Mexican coastline and the United States were obtained from the Instituto Nacional de Estadistica, Geografia e Informatica of Mexico (INEGI: www.inegi.org.mx/geo/contenidos/geoestadistica/m_geoestadistico.aspx) and the tmap R package [@tmap_2017], respectively.
+Shapefiles for Mexican coastline and the United States were obtained from INEGI @[INEGI] and the tmap R package [@tmap_2017], respectively.
 

### 2017-12-20 plostemplate/ManuscriptPLOS.tex
Added Tex file with PLoS template

--- 2017-12-15 Manuscript.tex
@@
-\documentclass[12pt,]{article}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-\else % if luatex or xelatex
-  \ifxetex
-    \usepackage{mathspec}
-  \else
-    \usepackage{fontspec}
-  \fi
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
+% Template for PLoS
+% Version 3.4 January 2017
+%
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that 
+% no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
+\usepackage{amsmath,amssymb}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% Use Unicode characters when possible
+\usepackage[utf8x]{inputenc}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
 \usepackage{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\usepackage[margin=1in]{geometry}
-\usepackage{hyperref}
-\hypersetup{unicode=true,
-            pdftitle={A user-friendly tool to evaluate the effectiveness of no-take marine reserves},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
 \makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\renewcommand{\@biblabel}[1]{\quad#1.}
 \makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
+
+% Leave date blank
+\date{}
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+\pagestyle{myheadings}
+\pagestyle{fancy}
+\fancyhf{}
+\setlength{\headheight}{27.023pt}
+\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\sf PLOS}
+
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% END MACROS SECTION
+
+\begin{document}
+\vspace*{0.2in}
+
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{A user-friendly tool to evaluate the effectiveness of no-take marine
+reserves} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
 }
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-% Redefines (sub)paragraphs to behave more like sections
-\ifx\paragraph\undefined\else
-\let\oldparagraph\paragraph
-\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-\let\oldsubparagraph\subparagraph
-\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-
-%%% Use protect on footnotes to avoid problems with footnotes in titles
-\let\rmarkdownfootnote\footnote%
-\def\footnote{\protect\rmarkdownfootnote}
-
-%%% Change title format to be more compact
-\usepackage{titling}
-
-% Create subtitle command for use in maketitle
-\newcommand{\subtitle}[1]{
-  \posttitle{
-    \begin{center}\large#1\end{center}
-    }
-}
-
-\setlength{\droptitle}{-2em}
-  \title{A user-friendly tool to evaluate the effectiveness of no-take marine
-reserves}
-  \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
-\subtitle{Marine reserve evaluation}
-  \author{}
-  \preauthor{}\postauthor{}
-  \date{}
-  \predate{}\postdate{}
-
-\usepackage{booktabs}
-\usepackage{longtable}
-\usepackage{array}
-\usepackage{multirow}
-\usepackage[table]{xcolor}
-\usepackage{wrapfig}
-\usepackage{float}
-\usepackage{colortbl}
-\usepackage{pdflscape}
-\usepackage{tabu}
-\usepackage{threeparttable}
-\usepackage{setspace}
-\doublespacing
-\usepackage{lineno}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+Juan Carlos Villaseñor-Derbez\textsuperscript{1\Yinyang*},
+Caio Faro\textsuperscript{1\Yinyang},
+Melaina Wright\textsuperscript{1\Yinyang},
+Jael Martínez\textsuperscript{1\Yinyang},
+Sean Fitzgerald\textsuperscript{1\ddag},
+Stuart Fulton\textsuperscript{2\ddag},
+Maria del Mar Mancha-Cisneros\textsuperscript{3\ddag},
+Gavin McDonald\textsuperscript{1,4,5\ddag},
+Fiorenza Micheli\textsuperscript{6\ddag},
+Alvin Suárez\textsuperscript{2\ddag},
+Jorge Torre\textsuperscript{2\ddag},
+Christopher Costello\textsuperscript{1,4,5\Yinyang}
+\\
+\bigskip
+\textbf{1} Bren School of Environmental Science and Management, University of California Santa Barbara, Santa Barbara, California, United States
+\\
+\textbf{2} Comunidad y Biodiversidad A.C., Calle Isla del Peruano, Guaymas, Sonora, México
+\\
+\textbf{3} School of Life Sciences, Arizona State University, Tempe, Arizona, United States
+\\
+\textbf{4} Sustainable Fisheries Group, University of California Santa Barbara, Santa Barbara, California, United States
+\\
+\textbf{5} Marine Science Institute, University of California
+Santa Barbara, Santa Barbara, California, United States
+\\
+\textbf{6} Hopkins Marine Station and Center for Ocean
+Solutions, Stanford University, Pacific Grove, CA 93950, USA
+\\
+\bigskip
+
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+% 
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+\Yinyang These authors contributed equally to this work.
+
+% Additional Equal Contribution Note
+% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
+\ddag These authors also contributed equally to this work.
+
+% Current address notes
+%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+% \textcurrency b Insert second current address 
+% \textcurrency c Insert third current address
+
+% Deceased author note
+%\dag Deceased
+
+% Group/Consortium Author Note
+%\textpilcrow Membership list can be found in the Acknowledgments section.
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* jvillasenor@bren.ucsb.edu
+
+\end{flushleft}
+% Please keep the abstract below 300 words
+\section*{Abstract}
+Marine reserves are implemented to achieve a variety of objectives, but are seldom rigorously evaluated to determine whether those objectives are met.
+In the rare cases when evaluations do take place, they typically focus on ecological indicators and ignore other relevant objectives such as socioeconomics and governance.
+And regardless of the objectives, the diversity of locations, monitoring protocols, and analysis approaches hinder the ability to compare results across case studies.
+Moreover, analysis and evaluation of reserves is generally conducted by outside researchers, not the reserve managers or users, plausibly thereby hindering effective local management and rapid response to change.
+We present a framework and tool, called ``MAREA'', to overcome these challenges.
+Its purpose is to evaluate the extent to which any given reserve has achieved its stated objectives.
+MAREA provides specific guidance on data collection and formatting, and then conducts rigorous causal inference analysis based on data input by the user, providing real-time outputs about the effectiveness of the reserve.
+MAREA's ease of use, standardization of state-of-the-art inference methods, and ability to analyze marine reserve effectiveness across ecological, socioeconomic, and governance objectives could dramatically further our understanding and support of effective marine reserve management.
+
 \linenumbers
 
-\begin{document}
-\maketitle
-
-Juan Carlos Villaseñor-Derbez\textsuperscript{1¶*}, Caio
-Faro\textsuperscript{1¶}, Melaina Wright\textsuperscript{1¶}, Jael
-Martínez\textsuperscript{1¶}, Sean Fitzgerald\textsuperscript{1\&},
-Stuart Fulton\textsuperscript{2\&}, Maria del Mar
-Mancha-Cisneros\textsuperscript{3\&}, Gavin
-McDonald\textsuperscript{1,4,5\&}, Fiorenza
-Micheli\textsuperscript{6\&}, Alvin Suárez\textsuperscript{2\&}, Jorge
-Torre\textsuperscript{2\&}, Christopher Costello\textsuperscript{1,4,5¶}
-
-\textsuperscript{1} Bren School of Environmental Science and Management,
-University of California Santa Barbara, Santa Barbara, California,
-United States
-
-\textsuperscript{2} Comunidad y Biodiversidad A.C., Calle Isla del
-Peruano, Guaymas, Sonora, México
-
-\textsuperscript{3} School of Life Sciences, Arizona State University,
-Tempe, Arizona, United States
-
-\textsuperscript{4} Sustainable Fisheries Group, University of
-California Santa Barbara, Santa Barbara, California, United States
-
-\textsuperscript{5} Marine Science Institute, University of California
-Santa Barbara, Santa Barbara, California, United States
-
-\textsuperscript{6} Hopkins Marine Station and Center for Ocean
-Solutions, Stanford University, Pacific Grove, CA 93950, USA
-
-*Corresponding author
-
-Email:
-\href{mailto:jvillasenor@bren.ucsb.edu}{\nolinkurl{jvillasenor@bren.ucsb.edu}}
-(JCVD)
-
-¶ These authors contributed equally to this work.
-
-\& These authors also contributed equally to this work.
+\section*{Introduction}
+
+Unsustainable fishing practices threaten biodiversity, conservation, economic and social outcomes \cite{pauly_2005-qV,halpern_2008-dK}.
+Marine Protected Areas (MPAs; and marine reserves, in which all extractive efforts are prohibited) are frequently proposed to aid in the recovery of fish and invertebrate stocks \cite{lester_2008-F_,lester_2009-Ks,sala_2016-PV,hastings_2017-sm} by limiting or restricting fishing effort and gears.
+
+Empirical evidence shows that MPAs increase biomass \cite{aburtooropeza_2011-ya,lester_2009-Ks}, enhance resilience to climatic impacts \cite{micheli_2012-EU,roberts_2017-J9}, and preserve genetic diversity \cite{munguavega_2015-yg}.
+Compared to MPAs that grant partial protection, marine reserves have higher levels of biomass, density, richness, and larger organisms \cite{lester_2008-F_,edgar_2014-UO,giakoumi_2017-V2,sala_2017-69}.
+However, these effects are often measured as biological changes within the reserves through time, and many lack a control site for comparison \cite{betti_2017-lq}.This approach does not account for other factors (\emph{e.g.} system-level changes in productivity caused by predatory release \cite{szuwalski_2017-jc}; or favorable environmental conditions \cite{chavez_2003-mm}) for which one must control \cite{davies_2017-ml} in order to causally attribute a biological change to the reserve.
+Other studies have used a control-impact comparison approach that uses control sites but does not address temporal variability \cite{guidetti_2014-8Z,friedlander_2017-oI,lester_2009-Ks,aburtooropeza_2011-ya,rodriguez_2017-PD}.
+
+A smaller fraction of studies have used a before-after-control-impact (\emph{i.e.} BACI) design comparing reserves to control sites before and after implementation \cite{moland_2013-VP,soykan_2015-nu,lester_2009-Ks}, which allows the use of causal inference techniques that estimate the effect of the reserve.
+For example, in ref \cite{moland_2013-VP} authors use a BACI design and observe increases in lobster catches --a proxy for abundances-- after reserve implementation for protected and control sites.
+However, the temporal changes in the reserve were greater than in the control site, suggesting a positive effect of the reserve on lobster catches.
+But even when proper causal inference can be drawn, results are often different across reserves.
+Effects of reserves on ecological and economic outcomes are highly heterogeneous, and often depend on the specific ecological, economic, and social context.
+
+Standardization of marine reserve evaluation is not new.
+The IUCN framework ``How is your MPA doing?'' \cite{pomeroy_2005-Py,pomeroy_2004-23} provides a comprehensive list of biological, socioeconomic, and governance indicators, and insights into how these may be measured or collected.
+But this framework stops short of analysis, and provides a user with little guidance about establishing causal inference about the reserve.
+Recent work \cite{mascia_2017-m_} integrates these three dimensions via the Social Ecological Systems Framework \cite{ostrom_2009-hg,basurto_2013-oq} and suggests the use of causal inference techniques to provide a measure of the effect of conservation interventions.
+However, neither of these approaches provide a user-friendly tool that ensures replicability and scalability of the analysis, particularly when used by the fishers and decision makers themselves.
+
+An increasingly popular way to make science accessible, reproducible, scalable, and replicable is through Open Science and the development of open-access tools \cite{lowndes_2017-xh}.
+The Ocean Health Index \cite{halpern_2012-k9,halpern_2017-Zi}, for example, successfully standardized a way to measure the health and benefits of the oceans.
+This approach has been implemented at a global scale, but also at country-level \cite{selig_2015-F9}, and regionally \cite{halpern_2014-lQ,elfes_2014-RC}.
+Open access tools are not limited to conservation, and have also been developed to evaluate fishery performance \cite{anderson_2015-ND,dowling_2016-pO}, design territorial use rights for fisheries (TURFs,\cite{oyanedel_2017-TO}), and improve decision making in the hydro power industry \cite{vilela_2017-Zo}.
+
+The purpose of this paper is to describe a user-friendly tool, called ``MAREA'', to rigorously systematize the evaluation of marine reserve effectiveness in terms of fisheries and marine conservation goals.
+The tool is in the form of an open-source application that uses state-of-the-art methods from program evaluation to compare a reserve to control sites along a number of biological, economic, and governance dimensions.
+We first provide a list of commonly stated management objectives and match them to appropriate indicators.
+We then develop a simple approach to analyzing these indicators building on causal inference techniques \cite{moland_2013-VP}, which help us understand the effect of management interventions \cite{burgess_2018-HN,mascia_2017-m_}.
+To implement the analytical approach, we introduce the Marine Reserve Evaluation Application (MAREA), an open source, web–based tool that automates the framework described in this paper and enables its broader use.
+Finally, we present a case study on the evaluation of a marine reserve established by the fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential of MAREA.
+
+\section*{Materials and methods}\label{materials-and-methods}
+
+Here, we describe the proposed framework to evaluate the effectiveness of marine reserves (Fig~\ref{fig1}).
+We explain how management objectives were identified and matched to appropriate indicators that allow the evaluation of the reserves, and provide brief guidelines on data collection.
+Alongside, methodologies to analyze these indicators are presented.
+We then describe the development of MAREA and explain how this tool can be used by fishermen, managers, and other stakeholders with little scientific background.
+Finally, we provide guidelines on how to interpret and use the results and output generated by MAREA to inform management.
+
+\begin{figure}[!h]
+\caption{{\bf Workflow to evaluate the effectiveness of marine reserves.}}
+\label{fig1}
+\end{figure}
+
+\subsection*{Marine Reserve objectives and indicators}\label{marine-reserve-objectives-and-indicators}
+
+Throughout this study, we will refer to the stated goals for which a marine reserve was designed as ``objectives.'' This work was motivated by the case of Mexico, where 39 reserves have been implemented over the past five years to achieve objectives such as increasing productivity in nearby waters or recovery of overexploited species; most of these reserves have never been formally evaluated for effectiveness at meeting those objectives.
+Thus, our focus was on identifying common objectives of marine reserves in Mexico.
+However, a literature review and discussions with marine reserve researchers suggested that the objectives driving Mexican marine reserve implementation are similar to those in the rest of the world.
+Thus, we group these objectives into seven major categories that may be applied to marine reserves worldwide.
+Any given reserve may have been implemented to meet one or more of these.
+The list includes objectives stated in legislation \cite{nom,lgeepa} and official documents such as the Technical Justification Studies (\emph{Estudios Técnicos Justificativos}), agreements, and decrees associated to these areas:
+
+\begin{enumerate}
+  \item Avoid overexploitation
+  \item Conserve species under a special protection regime
+  \item Maintain biological processes (reproduction, recruitment, growth, feeding)
+  \item Improve fishery production in adjacent waters
+  \item Preserve biological diversity and the ecosystem
+  \item Recover overexploited species
+  \item Recover species of economic interest
+\end{enumerate}
+
+Based on these seven objectives, we determined a set of associated indicators to evaluate reserve effectiveness.
+These indicators are specific variables on which data could be collected and analyzed, to ultimately determine whether the corresponding objective was causally being achieved by the marine reserve.
+The list of indicators was compiled through a review of scientific literature in which we identified indicators that were used to measure similar objectives\cite{lester_2017-nh,sala_2017-69,chirico_2017-Rz,edgar_2014-UO,rodriguez_2017-PD,rossetto_2015-V0,betti_2017-lq,sala_2016-PV,woodcock_2017-Wm,friedlander_2017-oI,moland_2013-VP,aburtooropeza_2011-ya,guidetti_2014-8Z,lester_2008-F_,lester_2009-Ks,pomeroy_2005-Py,pomeroy_2004-23}.
+A first filter eliminated indicators for which baseline data do not typically exist in Mexico.
+The preliminary list of indicators was reviewed at a workshop with participation of members from Mexican fishery management agencies and non-government organizations.
+Later, these were presented to fishers from the Ensenada Fishing Cooperative (\emph{S.C.P.P.
+Ensenada}), in El Rosario, Baja California, who provided input.
+Our final list of indicators includes those identified in review works \cite{lester_2009-Ks,woodcock_2017-Wm}.
+
+Indicators are divided into three main categories: biological, socioeconomic, and governance (Table~\ref{table1}).
+The nine biological indicators focus on fish and invertebrate communities that are evaluated using underwater ecological surveys performed inside and outside the reserve (see Data and Analysis section for specific sampling design and methodologies).
+Five socioeconomic indicators reflect the performance of the fishery in terms of landings, income from landings, and availability of alternative livelihoods.
+Fifteen governance indicators describe the governance structures under which the community operates (\emph{e.g.}, access rights to the fishery, number of fishers, legal recognition of the reserve).
+Most biological and socioeconomic indicators are quantitative and require a numerical entry (\emph{e.g.} Fish biomass) while all governance indicators, one biological indicator, and one socioeconomic indicator are qualitative and rely on a descriptive entry (\emph{e.g.} Reasoning for reserve location).
+Many of them specifically measure an outcome of the reserve, though some are designed to further the understanding of the mechanisms driving a reserve's performance.
+In that sense, most biological and socioeconomic indicators are outcome variables.
+On the other hand, governance indicators are viewed as possible explanatory variables of reserve performance.
+Whenever an indicator is applied to ``Target species'', it means that the indicator can be used for all species (\emph{e.g.} Fish Biomass) and/or for individual species that are either the conservation target of the reserve or are of particular economic or ecological interest (\emph{e.g.} Grouper Biomass).
+Finally, indicators B3 and B4 are different in that B3 only looks at the density of organisms above size at first maturity (related to reproductive potential), while B4 measures the density of all fish or of a target species.
+Each indicator targets different plausible desired outcomes, like increased reproductive potential (\emph{i.e.} B3; \cite{carter_2017-Uf}) or having more fish -regardless of their size- to attract tourism (\emph{i.e.} B4).
+Table~\ref{table1} presents the proposed indicators, and Table~\ref{table2} shows how objectives are matched with biological and socioeconomic indicators.
+Governance indicators are excluded from Table~\ref{table2}, but should be considered for every objective as each serves as a plausible explanatory variable for reserve performance.
+
+\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in}
+\centering
+\caption{
+{\bf List of indicators to evaluate the effectiveness of no-take marine reserves.}}
+\resizebox{\linewidth}{!}{\begin{tabular}{l|l|l|l}
+\hline
+\bfseries{Code} & \bfseries{Indicator} & \bfseries{Data type} & \bfseries{Unit}\\
+\hline
+\multicolumn{4}{l}{\textbf{Biological}}\\
+\hline
+\hspace{1em}B1 & Shannon diversity index & Continuous & \\
+\hline
+\hspace{1em}B2 & Species richness & Discrete & Number of species/transect\\
+\hline
+\hspace{1em}B3 & Density of mature organisms & Continuous & Percent\\
+\hline
+\hspace{1em}B4 & Density* & Continuous & Organisms/transect\\
+\hline
+\hspace{1em}B5 & Natural Disturbance & Descriptive & \\
+\hline
+\hspace{1em}B6 & Mean Trophic Level & Continuous & \\
+\hline
+\hspace{1em}B7 & Biomass* & Continuous & kg/transect\\
+\hline
+\multicolumn{4}{l}{\textbf{Socioeconomic}}\\
+\hline
+\hspace{1em}S1 & Total landings* & Continuous & kg\\
+\hline
+\hspace{1em}S2 & Income from total landings* & Continuous & \$\\
+\hline
+\hspace{1em}S3 & Alternative economic opportunities & Ordinal & \\
+\hline
+\multicolumn{4}{l}{\textbf{Governance}}\\
+\hline
+\hspace{1em}G1 & Access to the fishery & Categorical & \\
+\hline
+\hspace{1em}G2 & Number of fishers & Discrete & \\
+\hline
+\hspace{1em}G3 & Legal recognition of reserve & Binary & \\
+\hline
+\hspace{1em}G4 & Reserve type & Descriptive & \\
+\hline
+\hspace{1em}G5 & Illegal harvesting & Ordinal & \\
+\hline
+\hspace{1em}G6 & Management plan & Binary & \\
+\hline
+\hspace{1em}G7 & Reserve enforcement & Descriptive & \\
+\hline
+\hspace{1em}G8 & Size of reserve & Discrete & \\
+\hline
+\hspace{1em}G9 & Reasoning for reserve location & Descriptive & \\
+\hline
+\hspace{1em}G10 & Membership to fisher organizations & Binary & \\
+\hline
+\hspace{1em}G11 & Type of fisheries organizations & Categorical & \\
+\hline
+\hspace{1em}G12 & Representation & Ordinal & \\
+\hline
+\hspace{1em}G13 & Internal Regulation & Binary & \\
+\hline
+\hspace{1em}G14 & Perceived Effectiveness & Categorical & \\
+\hline
+\hspace{1em}G15 & Social Impact of Reserve & Categorical & \\
+\hline
+\end{tabular}}
+\begin{flushleft} * Indicates the indicator is applied to target species
+\end{flushleft}
+\label{table1}
+%\end{adjustwidth}
+\end{table}
+
+\begin{table}[!ht]
+\centering
+\caption{
+{\bf Management objectives and respective performance indicators.}}
+\resizebox{\linewidth}{!}{\begin{tabular}{>{\raggedright\arraybackslash}p{4cm}|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
+\hline
+\bfseries{Objective} & \bfseries{B1} & \bfseries{B2} & \bfseries{B3} & \bfseries{B4} & \bfseries{B4*} & \bfseries{B5} & \bfseries{B6} & \bfseries{B7} & \bfseries{B7*} & \bfseries{S1} & \bfseries{S1*} & \bfseries{S2} & \bfseries{S2*} & \bfseries{S3}\\
+\hline
+Avoid overexploitation &  &  & x & x & x & x & x & x & x & x & x & x & x & x\\
+\hline
+Conserve species under a special protection &  &  & x &  & x & x &  &  & x & x &  & x &  & x\\
+\hline
+Maintain biological process & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
+\hline
+Improve fishery production in nearby waters &  &  &  & x & x & x &  & x & x & x & x & x & x & x\\
+\hline
+Preserve biological diversity and the ecosystem & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
+\hline
+Recover overexploited species &  &  & x &  & x & x &  &  & x &  & x &  & x & x\\
+\hline
+Recover species of economic interest &  &  & x &  & x & x &  &  & x &  & x &  & x & x\\
+\hline
+\end{tabular}}
+\begin{flushleft} Governance indicators are excluded from the table, but all should be
+used for any objective. * Indicates the indicator is applied to target species
+\end{flushleft}
+\label{table2}
+%\end{adjustwidth}
+\end{table}
 
@@
 
-\textbf{Abstract}
-
-Marine reserves are implemented to achieve a variety of objectives, but
-are seldom rigorously evaluated to determine whether those objectives
-are met. In the rare cases when evaluations do take place, they
-typically focus on ecological indicators and ignore other relevant
-objectives such as socioeconomics and governance. And regardless of the
-objectives, the diversity of locations, monitoring protocols, and
-analysis approaches hinder the ability to compare results across case
-studies. Moreover, analysis and evaluation of reserves is generally
-conducted by outside researchers, not the reserve managers or users,
-plausibly thereby hindering effective local management and rapid
-response to change. We present a framework and tool, called ``MAREA'',
-to overcome these challenges. Its purpose is to evaluate the extent to
-which any given reserve has achieved its stated objectives. MAREA
-provides specific guidance on data collection and formatting, and then
-conducts rigorous causal inference analysis based on data input by the
-user, providing real-time outputs about the effectiveness of the
-reserve. MAREA's ease of use, standardization of state-of-the-art
-inference methods, and ability to analyze marine reserve effectiveness
-across ecological, socioeconomic, and governance objectives could
-dramatically further our understanding and support of effective marine
-reserve management.
-
-\section{Introduction}\label{introduction}
-
-Unsustainable fishing practices threaten biodiversity, conservation,
-economic and social outcomes {[}1,2{]}. Marine Protected Areas (MPAs;
-and marine reserves, in which all extractive efforts are prohibited) are
-frequently proposed to aid in the recovery of fish and invertebrate
-stocks {[}3--6{]} by limiting or restricting fishing effort and gears.
-
-Empirical evidence shows that MPAs increase biomass {[}4,7{]}, enhance
-resilience to climatic impacts {[}8,9{]}, and preserve genetic diversity
-{[}10{]}. Compared to MPAs that grant partial protection, marine
-reserves have higher levels of biomass, density, richness, and larger
-organisms {[}3,11--13{]}. However, these effects are often measured as
-biological changes within the reserves through time, and many lack a
-control site for comparison {[}14{]}. This approach does not account for
-other factors (\emph{e.g.} system-level changes in productivity caused
-by predatory release {[}15{]}; or favorable environmental conditions
-{[}16{]}) for which one must control {[}17{]} in order to causally
-attribute a biological change to the reserve. Other studies have used a
-control-impact comparison approach that uses control sites but does not
-address temporal variability {[}4,7,18--20{]}.
-
-A smaller fraction of studies have used a before-after-control-impact
-(\emph{i.e.} BACI) design comparing reserves to control sites before and
-after implementation {[}4,21,22{]}, which allows the use of causal
-inference techniques that estimate the effect of the reserve. For
-example, in ref {[}21{]} authors use a BACI design and observe increases
-in lobster catches --a proxy for abundances-- after reserve
-implementation for protected and control sites. However, the temporal
-changes in the reserve were greater than in the control site, suggesting
-a positive effect of the reserve on lobster catches. But even when
-proper causal inference can be drawn, results are often different across
-reserves. Effects of reserves on ecological and economic outcomes are
-highly heterogeneous, and often depend on the specific ecological,
-economic, and social context.
-
-Standardization of marine reserve evaluation is not new. The IUCN
-framework ``How is your MPA doing?'' {[}23,24{]} provides a
-comprehensive list of biological, socioeconomic, and governance
-indicators, and insights into how these may be measured or collected.
-But this framework stops short of analysis, and provides a user with
-little guidance about establishing causal inference about the reserve.
-Recent work {[}25{]} integrates these three dimensions via the Social
-Ecological Systems Framework {[}26,27{]} and suggests the use of causal
-inference techniques to provide a measure of the effect of conservation
-interventions. However, neither of these approaches provide a
-user-friendly tool that ensures replicability and scalability of the
-analysis, particularly when used by the fishers and decision makers
-themselves.
-
-An increasingly popular way to make science accessible, reproducible,
-scalable, and replicable is through Open Science and the development of
-open-access tools {[}28{]}. The Ocean Health Index {[}29,30{]}, for
-example, successfully standardized a way to measure the health and
-benefits of the oceans. This approach has been implemented at a global
-scale, but also at country-level {[}31{]}, and regionally {[}32,33{]}.
-Open access tools are not limited to conservation, and have also been
-developed to evaluate fishery performance {[}34,35{]}, design
-territorial use rights for fisheries {[}36{]}, and improve decision
-making in the hydro power industry {[}37{]}.
-
-The purpose of this paper is to describe a user-friendly tool, called
-``MAREA'', to rigorously systematize the evaluation of marine reserve
-effectiveness in terms of fisheries and marine conservation goals. The
-tool is in the form of an open-source application that uses
-state-of-the-art methods from program evaluation to compare a reserve to
-control sites along a number of biological, economic, and governance
-dimensions. We first provide a list of commonly stated management
-objectives and match them to appropriate indicators. We then develop a
-simple approach to analyzing these indicators building on causal
-inference techniques {[}21{]}, which help us understand the effect of
-management interventions {[}25,38{]}. To implement the analytical
-approach, we introduce the Marine Reserve Evaluation Application
-(MAREA), an open source, web--based tool that automates the framework
-described in this paper and enables its broader use. Finally, we present
-a case study on the evaluation of a marine reserve established by the
-fishers of Isla Natividad (Mexico) in 2006, to demonstrate the potential
-of MAREA.
-
-\section{Materials and methods}\label{materials-and-methods}
-
-Here, we describe the proposed framework to evaluate the effectiveness
-of marine reserves (Fig. 1). We explain how management objectives were
-identified and matched to appropriate indicators that allow the
-evaluation of the reserves, and provide brief guidelines on data
-collection. Alongside, methodologies to analyze these indicators are
-presented. We then describe the development of MAREA and explain how
-this tool can be used by fishermen, managers, and other stakeholders
-with little scientific background. Finally, we provide guidelines on how
-to interpret and use the results and output generated by MAREA to inform
-management.
-
-\textbf{Fig 1. Workflow to evaluate the effectiveness of marine
-reserves.}
-
-\subsection{Marine Reserve objectives and
-indicators}\label{marine-reserve-objectives-and-indicators}
-
-Throughout this study, we will refer to the stated goals for which a
-marine reserve was designed as ``objectives.'' This work was motivated
-by the case of Mexico, where 39 reserves have been implemented over the
-past five years to achieve objectives such as increasing productivity in
-nearby waters or recovery of overexploited species; most of these
-reserves have never been formally evaluated for effectiveness at meeting
-those objectives. Thus, our focus was on identifying common objectives
-of marine reserves in Mexico. However, a literature review and
-discussions with marine reserve researchers suggested that the
-objectives driving Mexican marine reserve implementation are similar to
-those in the rest of the world. Thus, we group these objectives into
-seven major categories that may be applied to marine reserves worldwide.
-Any given reserve may have been implemented to meet one or more of
-these. The list includes objectives stated in legislation {[}39,40{]}
-and official documents such as the Technical Justification Studies
-(\emph{Estudios Técnicos Justificativos}), agreements, and decrees
-associated to these areas:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  Avoid overexploitation
-\item
-  Conserve species under a special protection regime
-\item
-  Maintain biological processes (reproduction, recruitment, growth,
-  feeding)
-\item
-  Improve fishery production in adjacent waters
-\item
-  Preserve biological diversity and the ecosystem
-\item
-  Recover overexploited species
-\item
-  Recover species of economic interest
-\end{enumerate}
-
-Based on these seven objectives, we determined a set of associated
-indicators to evaluate reserve effectiveness. These indicators are
-specific variables on which data could be collected and analyzed, to
-ultimately determine whether the corresponding objective was causally
-being achieved by the marine reserve. The list of indicators was
-compiled through a review of scientific literature in which we
-identified indicators that were used to measure similar
-objectives{[}3--5,7,11,13,14,18--21,23,24,41--44{]}. A first filter
-eliminated indicators for which baseline data do not typically exist in
-Mexico. The preliminary list of indicators was reviewed at a workshop
-with participation of members from Mexican fishery management agencies
-and non-government organizations. Later, these were presented to fishers
-from the Ensenada Fishing Cooperative (\emph{S.C.P.P. Ensenada}), in El
-Rosario, Baja California, who provided input. Our final list of
-indicators includes those identified in review works {[}4,44{]}.
-
-Indicators are divided into three main categories: biological,
-socioeconomic, and governance (Table 1). The nine biological indicators
-focus on fish and invertebrate communities that are evaluated using
-underwater ecological surveys performed inside and outside the reserve
-(see Data and Analysis section for specific sampling design and
-methodologies). Five socioeconomic indicators reflect the performance of
-the fishery in terms of landings, income from landings, and availability
-of alternative livelihoods. Fifteen governance indicators describe the
-governance structures under which the community operates (\emph{e.g.},
-access rights to the fishery, number of fishers, legal recognition of
-the reserve). Most biological and socioeconomic indicators are
-quantitative and require a numerical entry (\emph{e.g.} Fish biomass)
-while all governance indicators, one biological indicator, and one
-socioeconomic indicator are qualitative and rely on a descriptive entry
-(\emph{e.g.} Reasoning for reserve location). Many of them specifically
-measure an outcome of the reserve, though some are designed to further
-the understanding of the mechanisms driving a reserve's performance. In
-that sense, most biological and socioeconomic indicators are outcome
-variables. On the other hand, governance indicators are viewed as
-possible explanatory variables of reserve performance. Whenever an
-indicator is applied to ``Target species'', it means that the indicator
-can be used for all species (\emph{e.g.} Fish Biomass) and/or for
-individual species that are either the conservation target of the
-reserve or are of particular economic or ecological interest
-(\emph{e.g.} Grouper Biomass). Finally, indicators B3 and B4 are
-different in that B3 only looks at the density of organisms above size
-at first maturity (related to reproductive potential), while B4 measures
-the density of all fish or of a target species. Each indicator targets
-different plausible desired outcomes, like increased reproductive
-potential (\emph{i.e.} B3; {[}45{]}) or having more fish -regardless of
-their size- to attract tourism (\emph{i.e.} B4). Table 1 presents the
-proposed indicators, and Table 2 shows how objectives are matched with
-biological and socioeconomic indicators. Governance indicators are
-excluded from Table 2, but should be considered for every objective as
-each serves as a plausible explanatory variable for reserve performance.
+\subsection*{Data and analyses}\label{data-and-analyses}
+
+In many coastal marine reserves, biological data are often collected via underwater visual censuses as part of a reserve's monitoring program.
+Scientific divers record fish and invertebrate richness and abundances, as well as fish total length along belt transects.
+Ecological surveys are typically performed annually in each reserve and corresponding control site(s), before and after the implementation of the reserve, providing a sampling design that can be used to draw causal inference.
+Control sites are areas where habitat is similar to that of the reserve, but with presence of fishing activity; in principle these are areas that are otherwise observationally identical to the reserve site, but where, for presumably random reasons, a reserve was not implemented.
+While transect dimensions (\emph{i.e.} length and width) and sampling methods might vary from study to study, the general idea remains the same: richness, abundances, and sizes of organisms are recorded in a study--specific standardized way.
+For this reason, MAREA does not assume specific transect dimensions, and pertinent indicators are calculated per transect (Table~\ref{table1}).
+More information on data collection and formatting is provided in a guidebook \cite{villaseorderbez2017-xE}, which is available in English and Spanish in MAREA's interface.
+
+This sampling design for biological data allows us to use causal inference techniques \cite{moland_2013-VP,ferraro_2006-oW} to evaluate the effect of the reserve on biological indicators.
+The hypothesis that the indicators will respond to implementation of the reserve is tested by analyzing spatial and temporal changes in each numeric biological indicator (all but B5) using generalized linear models \cite{moland_2013-VP}.
+To account for variations in the environment and survey conditions, covariates that are gathered during the underwater ecological surveys are included in the difference-in-differences model with form:
+
+\begin{eqnarray}
+\label{eq:difindif}
+I_{i,t,z}=\beta_0 + \sum_{t = 2}^T\gamma_{t}Y_t + \beta_1Z_{i,z} + \beta_2P_{i,t}\times Z_{i,z} + \beta_3T_{i,t,z} + \beta_4V_{i,t,z} + \beta_5D_{i,t,z} + \epsilon_{i,t,z}
+\end{eqnarray}
+
+In this model, \(i\), \(t\), and \(z\) are indices for transect, time, and zone (control or reserve site), respectively.
+This model allows us to estimate the change in an indicator (\(I\)) based on the year (\(Y\)), a dummy variable that indicates treatment (\(Z\); \emph{i.e.} control or reserve), an interaction between a dummy variable that indicates before or after implementation (\(P\)) and treatment (\(Z\)), and covariates such as bottom temperature (\(T\); in °C), horizontal visibility during the survey (\(V\); in m), and depth at which survey was performed (\(D\); in m).
+\(\epsilon\) represents the error term associated to the regression.
+Here, years are modeled as factors, using the first year as the reference level.
+This does not impose a linear structure in the way an indicator changes through time (\emph{i.e.} the change in biomass between 2006 and 2007 does not have to be the same as the change between 2015 and 2016).
+The treatment and implementation variables, modeled as dummy variables, are coded as Control = 0 and Reserve = 1; and Before implementation = 0 and After implementation = 1, respectively.
+
+Socioeconomic data are often collected by fishers, natural resource management agencies, or Civil Society Organizations (CSOs) by recording landings, income, and sometimes prices for each species.
+To control for inflation, income is adjusted with the country's consumer price index \cite{oecd_2017-VV}:
+
+\begin{eqnarray}
+\label{eq:cpi}
+I_t = RI\times \frac{CPI_t}{CPI_T}
+\end{eqnarray}
+
+Where \(I_t\) represents the adjusted income for year \(t\) as the product between the reported income for that year and the ratio between the consumer price index (\(CPI\)) in that year to the most recent year's (\(T\)) CPI.
+Since no control sites are typically available for this data type, numeric socioeconomic indicators (S1 and S2) are evaluated with a simplified version of Eq~(\ref{eq:difindif}):
+
+\begin{eqnarray}
+\label{eq:socio}
+I_{t}=\beta_0 + \beta_1P_{t} + \epsilon_{t}
+\end{eqnarray}
+
+This model does not formally allow for causal inference, but we can still measure changes in mean landings and income before and after the implementation of the reserve and provide valuable input.
+For both models (Eq~(\ref{eq:difindif}) and Eq~(\ref{eq:socio})), we estimate the model coefficients with ordinary least squares, and calculate heteroskedastic--robust standard errors \cite{zeileis_2004-7n}.
+
+While biological and some economic data are regularly collected, governance data are typically not available nor systematically collected by the community or other organizations.
+Therefore, we created a survey specifically designed to collect information needed for the proposed indicators (B5, S3, and G1--G15).
+The survey is included as supplementary material in English (\nameref{S1_Appendix}) and Spanish (\nameref{S2_Appendix}).
+To analyze governance information, we developed a framework based on a literature review of common governance structures and their relation to effectiveness in managing fisheries or marine reserves (\nameref{S3_Table}).
+This approach has been proven to successfully evaluate governance structures \cite{espinosaromero_2014-PY}.
+Unlike with biological and socioeconomic objectives (see Eq~(\ref{eq:difindif}) and Eq~(\ref{eq:socio})), MAREA does not quantitatively analyze governance information.
+Rather, it is presented along with the biological and socioeconomic indicators to provide managers and users with a more complete description of the reserve.
+
+\subsection*{Marine Reserve Evaluation App (MAREA)}\label{marine-reserve-evaluation-app-marea}
+
+We developed MAREA in R version 3.4.2 and R Studio 1.1.383 \cite{rcore_2017} using the Shiny package \cite{shiny_2017}, to build an interactive web application hosted on an open server; the MAREA app can be accessed at \href{turfeffect.shinyapps.io/marea/}{turfeffect.shinyapps.io/marea}.
+While the original version was developed in Spanish because it was aimed for Mexico and other Latin-American countries, all of its content can be translated by a translation widget available within the app.
+
+MAREA is designed as a 6-step process, divided in tabs that appear upon launching the app.
+The first tab introduces the app and summarizes the evaluation process.
+Then, the user selects management objectives, which MAREA automatically matches to appropriate indicators, based on Table~\ref{table2}.
+Users can also select and deselect indicators based on their interests and data availability by ``clicking'' on the check-boxes in MAREA.
+The user can then load data on one or more reserves, using standard *.csv text files; sample datasets are provided within MAREA.
+Once data have been loaded, MAREA identifies all reserves in the data, and lets the user select the reserve to be evaluated.
+At this point, the user can also specify the year of implementation of the reserve, reserve dimensions, and indicate target species that are of particular management interest.
+MAREA provides the user with a section to confirm that all the decisions made leading up to that point are correct.
+Once the user has confirmed all input data, objectives, and other information, MAREA performs the formal program evaluation analyses discussed above.
+For a typical data set, the automated analysis step takes less than one second.
+Finally, the user is taken to the results tab where all results are presented in a simple format.
+The user can also download a more comprehensive technical report produced in *.pdf format.
+
+The first output is a color--coded scorecard intended to provide a general overview of the effectiveness of the reserve.
+The scorecard provides a global score for the reserve, a general score for each category of indicators, and an individual score for each indicator.
+The global and category--level scores are determined by the percentage of positive indicators, overall and for each category, respectively.
+For numeric biological indicators (all but B5), the color is defined by the sign of the interaction term coefficient (\(\beta_2\)) in Eq~(\ref{eq:difindif}).
+For socioeconomic indicators, colors are assigned based on the direction of the slope (\(\beta_1\)) in Eq~(\ref{eq:socio}).
+Red, yellow, and green are used for \(\beta_i<0\), \(\beta_i = 0\), and \(\beta_i>0\), respectively.
+The intensity of the color is defined by the significance of the coefficient, testing the null hypothesis of no change (\emph{i.e.} \(H_0: \beta_i = 0\)) with a Student's t-test.
+Cutoff values are \(p < 0.05\) and \(p < 0.1\).
+Thus, even in a case where \(\beta_i > 0\), if the coefficient is not significant by standard measures (\emph{i.e.
+} \(p>0.1\)), the indicator will be assigned a yellow color.
+A legend (Fig~\ref{fig2}) is provided within the scorecard to aid in the interpretation of these results.
+Governance indicators are represented simply by red or green.
+The color is defined based on what literature shows to be a negative (red) or positive (green) factor for a reserve (\nameref{S3_Table}).
+For example, if the perceived degree of illegal fishing is high, this indicator will be assigned a red color.
+However, due to the nature of some governance indicators, which require the user to provide a narrative, only some indicators are presented in the scorecard (although all are included in the technical report).
+
+\begin{figure}[!h]
+\caption{{\bf Legend used to interpret the scorecard produced by MAREA.}
+Colors indicate direction of change (red = negative; green = positive), and color intensity is given by the statistical significance.}
+\label{fig2}
+\end{figure}
+
+The second output from MAREA is a technical report intended to communicate information and statistical results in a more comprehensive and technical way.
+This report also includes a scorecard as a summary of the results, but provides more information for each indicator.
+For all numeric biological indicators, the report includes a graph of the value of the indicator in the reserve and control sites through time.
+It also provides a regression table that summarizes the value of all coefficients in the regression and their respective robust standard errors.
+The summary table also provides information on model fit (\(R^2\)) and significance of the regression.
+
+The scorecard is produced with functions from the Shinydashboard package \cite{shinydashboard_2017}.
+The technical report is produced by a parameterized Rmarkdown document \cite{rmarkdown_2017} processed by the knitr package \cite{knitr_2017}.
+Another feature of MAREA is that the user can choose to share the data.
+Once the technical report is downloaded, the information on the reserve, its management objectives, and all uploaded data are saved into a central repository.
+These data can be accessed at any time by any person interested in acquiring them at \url{github.com/turfeffect/MAREAdata}.
+
+\subsection*{Case study}\label{case-study}
+
+While MAREA is a general tool that can be easily employed to evaluate the effectiveness of any marine reserve with the required input data, we illustrate its use here by applying it to one marine reserve near Isla Natividad, in Baja California Sur, Mexico.
+Isla Natividad is located 8 Km off the Pacific Coast of the Baja California Peninsula (Fig~\ref{fig3}), where fishers operate under a fishing cooperative (\emph{S.C.P.P.
+Buzos y Pescadores de la Baja California}) that promotes co-management of marine resources \cite{mccay_2017-1m,mccay_2014-CN}.
+Additionally, fishers have Territorial Use Rights for Fisheries (TURFs) that provide them with exclusive access rights to exploit the benthic marine resources within a given perimeter \cite{mccay_2014-CN}.
 
@@
 
-\textbf{Table 1. List of indicators to evaluate the effectiveness of
-no-take marine reserves.}
-
-\begin{table}[H]
+\begin{figure}[!h]
+\caption{{\bf General location of Isla Natividad (left) and map of the island (right).}
+The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce).
+Shapefiles for Mexican coastline and the United States were obtained from INEGI \cite{INEGI} and the tmap R package \cite{tmap_2017}, respectively.}
+\label{fig3}
+\end{figure}
+
+In 2006, the Isla Natividad community established a biological baseline following the data collection protocol described in this study.
+The community then implemented two community-based marine reserves within their TURF \cite{afflerbach_2014-HP,lester_2017-nh,micheli_2012-EU} after establishing a baseline for the soon-to-be reserves and control sites.
+Evidence suggest that these reserves have been effective at enhancing resilience to climate variations \cite{micheli_2012-EU} and preserving genetic diversity of high value commercial species such as abalone \cite{munguavega_2015-yg}.
+These ecological benefits have been translated into economic benefits, enhancing population persistence and bolstering abalone fisheries \cite{rossetto_2015-V0}.
+For the purpose of this evaluation, we focused on the ``La Plana / Las Cuevas'' marine reserve, located at the southern end of the island (Fig~\ref{fig3}) and its corresponding control site ``La Dulce / Babencho''.
+
+The objective of this reserve was to recover species of economic interest --which were overexploited-- and to enhance fishery production in nearby waters.
+Fishers were also interested in preserving biological diversity and the ecosystem.
+Thus, objectives 4---7 were selected.
+Using Table~\ref{table2} to match these objectives with appropriate management indicators, we selected all biological, socioeconomic, and governance indicators included as options in the framework.
+
+Local fishers (who were trained in scientific diving by the CSO Comunidad y Biodiversidad, A.C.
+(COBI; \url{www.cobi.org}), ReefCheck California, and Stanford University) and personnel from these institutions performed SCUBA dives to record fish and invertebrate richness and abundances, as well as fish total length.
+They recorded information along 30 m transects, with a sampling window of 2 m x 2 m following a standardized ReefCheck protocol \cite{suman_2010-ez}.
+Ecological surveys were performed yearly in each reserve and corresponding control site(s), before and after the implementation of the reserve, providing the requisite time series data inside the reserve and for a suitable control site.
+Annual surveys (2006--2016) were carried out in late July -- early August, performing a total of 242 and 245 transects in the reserve site for fish and invertebrate surveys, respectively.
+Similar sampling effort was applied to the control site, with 221 fish and 222 invertebrate transects.
+Between 12 and 27 transects were performed in each site every year.
+
+Socioeconomic data were obtained from the National Commission for Aquaculture and Fisheries (\emph{Comisión Nacional de Acuacultura y Pesca}; CONAPESCA).
+The data contains species-level information on monthly landings and income from nine species from 2000 to 2014.
+Data on landings and income were aggregated by year and species, and adjusted by the Consumer Price Index {[}48{]}.
+From the nine species available, we selected as objective species those that contributed the most (88.27\%) income from 2000 to 2014: lobster (\emph{Panulirus interruptus}; 71.76\%), red sea urchin (\emph{Mesocentrotus franciscanus}; 9.33\%), snail (\emph{Megastraea undosa}; 3.93\%), and sea cucumber (\emph{Parastichopus parvimensis}; 3.23\%).
+Abalone species (\emph{Haliotis fulgens}; 4.52\% and \emph{Haliotis corrugata}; 6.16\%) were excluded because the cooperative implemented an informal closure of these fisheries in 2010 to allow the population to recover.
+Eliminating all fishing pressure on abalones means that the control site receives (for this species) the same treatment as the reserve.
+
+We constructed the governance data based on local knowledge of the area and the community.
+
+\section*{Results from illustrative example}\label{results-from-illustrative-example}
+
+In this section we show the results of the application of MAREA to the La Plana/Las Cuevas marine reserve in Isla Natividad, Mexico.
+These results are intended to highlight the relevance and utility of the MAREA framework and app, which automate the analysis and make it replicable.
+While we highlight some of the general observed trends, we focus on the utility of the tool rather than on the specific effectiveness of this case study marine reserve.
+
+The scorecard (Fig~\ref{fig4}) shows that this reserve achieves a general score of 64\%, suggesting that 64\% of all indicators are positive.
+All category--level scores were also high, with values of 67\%, 60\%, and 71\% positive indicators for biological, socioeconomic and governance, respectively.
+
+\begin{figure}[!h]
+\caption{{\bf Scorecard produced by MAREA for the ``La Plana / Las Cuevas'' marine reserve in Isla Natividad, Mexico.}}
+\label{fig4}
+\end{figure}
+
+Among the biological indicators, the greatest effect of the reserve was observed for snail and sea cucumber densities, with values of \(\beta_2 = 97.17\) (\emph{p} \textless{} 0.05) and \(\beta_2 = 2.31\) (\emph{p} \textless{} 0.05), respectively.
+Fish indicators showed no significant change (\emph{p} \textgreater{} 0.1), with negative trends for Shannon's diversity index and fish species richness and positive trends for density, biomass, and mean trophic level.
+Changes through time for these indicators are presented in Figure~\ref{fig5}, and a summary of \(\beta_2\) coefficients is provided in Table~\ref{table3}.
+
+\begin{figure}[!h]
+\caption{{\bf Plots for values of each biological indicator (y-axis) through time (x-axis).}
+Red and blue correspond to the reserve and control sites, respectively.
+Black lines indicate yearly mean values, and ribbons indicate \(\pm\) 1 standard error.
+Dots are horizontally jittered to aid visualization.
+This figure contains information for fish Shannon's diversity index (a), fish species richness (b), fish density (c), fish trophic level (d), fish biomass (e), invertebrate Shannon's diversity index (f), invertebrate species richness (g), invertebrate density (h), lobster density (i), urchin density (j), snail density (k), and sea cucumber density (l).}
+\label{fig5}
+\end{figure}
+
+\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
-\resizebox{\linewidth}{!}{\begin{tabular}{l|l|l|l}
-\hline
-\bfseries{Code} & \bfseries{Indicator} & \bfseries{Data type} & \bfseries{Unit}\\
-\hline
-\multicolumn{4}{l}{\textbf{Biological}}\\
-\hline
-\hspace{1em}B1 & Shannon diversity index & Continuous & \\
-\hline
-\hspace{1em}B2 & Species richness & Discrete & Number of species/transect\\
-\hline
-\hspace{1em}B3 & Density of mature organisms & Continuous & Percent\\
-\hline
-\hspace{1em}B4 & Density* & Continuous & Organisms/transect\\
-\hline
-\hspace{1em}B5 & Natural Disturbance & Descriptive & \\
-\hline
-\hspace{1em}B6 & Mean Trophic Level & Continuous & \\
-\hline
-\hspace{1em}B7 & Biomass* & Continuous & kg/transect\\
-\hline
-\multicolumn{4}{l}{\textbf{Socioeconomic}}\\
-\hline
-\hspace{1em}S1 & Total landings* & Continuous & kg\\
-\hline
-\hspace{1em}S2 & Income from total landings* & Continuous & \$\\
-\hline
-\hspace{1em}S3 & Alternative economic opportunities & Ordinal & \\
-\hline
-\multicolumn{4}{l}{\textbf{Governance}}\\
-\hline
-\hspace{1em}G1 & Access to the fishery & Categorical & \\
-\hline
-\hspace{1em}G2 & Number of fishers & Discrete & \\
-\hline
-\hspace{1em}G3 & Legal recognition of reserve & Binary & \\
-\hline
-\hspace{1em}G4 & Reserve type & Descriptive & \\
-\hline
-\hspace{1em}G5 & Illegal harvesting & Ordinal & \\
-\hline
-\hspace{1em}G6 & Management plan & Binary & \\
-\hline
-\hspace{1em}G7 & Reserve enforcement & Descriptive & \\
-\hline
-\hspace{1em}G8 & Size of reserve & Discrete & \\
-\hline
-\hspace{1em}G9 & Reasoning for reserve location & Descriptive & \\
-\hline
-\hspace{1em}G10 & Membership to fisher organizations & Binary & \\
-\hline
-\hspace{1em}G11 & Type of fisheries organizations & Categorical & \\
-\hline
-\hspace{1em}G12 & Representation & Ordinal & \\
-\hline
-\hspace{1em}G13 & Internal Regulation & Binary & \\
-\hline
-\hspace{1em}G14 & Perceived Effectiveness & Categorical & \\
-\hline
-\hspace{1em}G15 & Social Impact of Reserve & Categorical & \\
-\hline
-\end{tabular}}
+\caption{
+{\bf Summary of average treatment effect of the reserve on biological indicators.}}
+\begin{tabular}{l|l|r}
+\hline
+\bfseries{Indicator} & \bfseries{Estimate (SD)} & \bfseries{t-score}\\
+\hline
+Shannon fish & -0.22 (0.16) & -1.3969\\
+\hline
+Richness fish & -0.61 (0.43) & -1.4073\\
+\hline
+Density fish & 0.74 (6.15) & 0.1205\\
+\hline
+Trophic fish & 0.00 (0.01) & 0.1399\\
+\hline
+Biomass fish & 0.22 (1.47) & 0.1476\\
+\hline
+Shannon invert & -0.67 (0.22)** & -3.0481\\
+\hline
+Richness invert & -2.71 (0.81)** & -3.3519\\
+\hline
+Density invert & 91.21 (47.11)* & 1.9362\\
+\hline
+Lobster & 7.66 (8.93) & 0.8583\\
+\hline
+Urchin & 2.15 (1.23)* & 1.7425\\
+\hline
+Snail & 97.17 (42.90)** & 2.2652\\
+\hline
+Cucumber & 2.31 (1.17)** & 1.9782\\
+\hline
+\end{tabular}
+\begin{flushleft} * Indicate significance level, with (*) indicating p \textless{} 0.1 and (**) p \textless{} 0.05.
+\end{flushleft}
+\label{table3}
 \end{table}
 
-* Indicates the indicator is applied to target species
-
-\clearpage
-
-\textbf{Table 2. Management objectives and respective performance
-indicators.}
-
-\begin{table}[H]
+One of the main objectives of this reserve was to increase landings.
+Results of the socioeconomic indicators show that total landings were, on average, 64.20 metric tonnes higher (\emph{p} \textgreater{} 0.1) after the implementation of the reserves, though this cannot necessarily be interpreted as causal, because it relies entirely on a before-after comparison.
+Total income was \$10,344.85 (\emph{p} \textless{} 0.05) thousands of Mexican Pesos (K MXP) higher after the implementation of the reserves.
+On average, lobster and sea cucumber landings increased, while urchin and snail landings and income decreased.
+Figure~\ref{fig6} presents the changes in these indicators through time, and Table~\ref{table4} summarizes these results.
+
+\begin{figure}[!h]
+\caption{{\bf Plots for values of each socioeconomic indicator (y-axis) through time (x-axis).}
+Red and blue correspond to before and after the implementation of the reserve, respectively.
+This figure contains information for total landings (a), total income (b), lobster landings (c), urchin landings (d), snail landings (e), sea cucumber landings (f), lobster income (g), urchin income (h), snail income (i), and sea cucumber income (j).}
+\label{fig6}
+\end{figure}
+
+\begin{table}[!ht]
+%\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
-\resizebox{\linewidth}{!}{\begin{tabular}{>{\raggedright\arraybackslash}p{4cm}|l|l|l|l|l|l|l|l|l|l|l|l|l|l}
-\hline
-\bfseries{Objective} & \bfseries{B1} & \bfseries{B2} & \bfseries{B3} & \bfseries{B4} & \bfseries{B4*} & \bfseries{B5} & \bfseries{B6} & \bfseries{B7} & \bfseries{B7*} & \bfseries{S1} & \bfseries{S1*} & \bfseries{S2} & \bfseries{S2*} & \bfseries{S3}\\
-\hline
-Avoid overexploitation &  &  & x & x & x & x & x & x & x & x & x & x & x & x\\
-\hline
-Conserve species under a special protection &  &  & x &  & x & x &  &  & x & x &  & x &  & x\\
-\hline
-Maintain biological process & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
-\hline
-Improve fishery production in nearby waters &  &  &  & x & x & x &  & x & x & x & x & x & x & x\\
-\hline
-Preserve biological diversity and the ecosystem & x & x &  & x &  & x & x & x &  &  &  &  &  & x\\
-\hline
-Recover overexploited species &  &  & x &  & x & x &  &  & x &  & x &  & x & x\\
-\hline
-Recover species of economic interest &  &  & x &  & x & x &  &  & x &  & x &  & x & x\\
-\hline
-\end{tabular}}
+\caption{
+{\bf Summary of differences in socioeconomic indicators before and after the implementation of the reserve.}}
+\begin{tabular}{l|l|r}
+\hline
+\bfseries{Indicator} & \bfseries{Estimate (SD)} & \bfseries{t-score}\\
+\hline
+Landings & 64.20 (90.07) & 0.7127\\
+\hline
+Income & 10344.85 (3982.20)** & 2.5978\\
+\hline
+Lobster landings & 7.37 (13.95) & 0.5281\\
+\hline
+Urchin landings & -30.00 (9.49)** & -3.1620\\
+\hline
+Snail landings & -69.53 (33.82)* & -2.0561\\
+\hline
+Cucumber landings & 9.34 (6.72) & 1.3906\\
+\hline
+Lobster income & 14372.85 (3634.64)** & 3.9544\\
+\hline
+Urchin income & -5800.46 (1867.50)** & -3.1060\\
+\hline
+Snail income & -404.85 (187.07)** & -2.1641\\
+\hline
+Cucumber income & 131.49 (185.66) & 0.7082\\
+\hline
+\end{tabular}
+\begin{flushleft} * Indicate significance level, with (*) indicating p \textless{} 0.1 and (**) p \textless{} 0.05.
+\end{flushleft}
+\label{table4}
 \end{table}
 
-Governance indicators are excluded from the table, but all should be
-used for any objective.
-
-* Indicates the indicator is applied to target species
-
-\subsection{Data and analyses}\label{data-and-analyses}
-
-In many coastal marine reserves, biological data are often collected via
-underwater visual censuses as part of a reserve's monitoring program.
-Scientific divers record fish and invertebrate richness and abundances,
-as well as fish total length along belt transects. Ecological surveys
-are typically performed annually in each reserve and corresponding
-control site(s), before and after the implementation of the reserve,
-providing a sampling design that can be used to draw causal inference.
-Control sites are areas where habitat is similar to that of the reserve,
-but with presence of fishing activity; in principle these are areas that
-are otherwise observationally identical to the reserve site, but where,
-for presumably random reasons, a reserve was not implemented. While
-transect dimensions (\emph{i.e.} length and width) and sampling methods
-might vary from study to study, the general idea remains the same:
-richness, abundances, and sizes of organisms are recorded in a
-study--specific standardized way. For this reason, MAREA does not assume
-specific transect dimensions, and pertinent indicators are calculated
-per transect (Table 1). More information on data collection and
-formatting is provided in a guidebook {[}46{]}, which is available in
-English and Spanish in MAREA's interface.
-
-This sampling design for biological data allows us to use causal
-inference techniques {[}21,47{]} to evaluate the effect of the reserve
-on biological indicators. The hypothesis that the indicators will
-respond to implementation of the reserve is tested by analyzing spatial
-and temporal changes in each numeric biological indicator (all but B5)
-using generalized linear models {[}21{]}. To account for variations in
-the environment and survey conditions, covariates that are gathered
-during the underwater ecological surveys are included in the
-difference-in-differences model with form:
-
-\begin{equation}I_{i,t,z}=\beta_0 + \sum_{t = 2}^T\gamma_{t}Y_t + \beta_1Z_{i,z} + \beta_2P_{i,t}\times Z_{i,z} + \beta_3T_{i,t,z} + \beta_4V_{i,t,z} + \beta_5D_{i,t,z} + \epsilon_{i,t,z}\end{equation}
-
-In this model, \(i\), \(t\), and \(z\) are indices for transect, time,
-and zone (control or reserve site), respectively. This model allows us
-to estimate the change in an indicator (\(I\)) based on the year
-(\(Y\)), a dummy variable that indicates treatment (\(Z\); \emph{i.e.}
-control or reserve), an interaction between a dummy variable that
-indicates before or after implementation (\(P\)) and treatment (\(Z\)),
-and covariates such as bottom temperature (\(T\); in °C), horizontal
-visibility during the survey (\(V\); in m), and depth at which survey
-was performed (\(D\); in m). \(\epsilon\) represents the error term
-associated to the regression. Here, years are modeled as factors, using
-the first year as the reference level. This does not impose a linear
-structure in the way an indicator changes through time (\emph{i.e.} the
-change in biomass between 2006 and 2007 does not have to be the same as
-the change between 2015 and 2016). The treatment and implementation
-variables, modeled as dummy variables, are coded as Control = 0 and
-Reserve = 1; and Before implementation = 0 and After implementation = 1,
-respectively.
-
-Socioeconomic data are often collected by fishers, natural resource
-management agencies, or Civil Society Organizations (CSOs) by recording
-landings, income, and sometimes prices for each species. To control for
-inflation, income is adjusted with the country's consumer price index
-{[}48{]}:
-
-\begin{equation}I_t = RI\times \frac{CPI_t}{CPI_T}\end{equation}
-
-Where \(I_t\) represents the adjusted income for year \(t\) as the
-product between the reported income for that year and the ratio between
-the consumer price index (\(CPI\)) in that year to the most recent
-year's (\(T\)) CPI. Since no control sites are typically available for
-this data type, numeric socioeconomic indicators (S1 and S2) are
-evaluated with a simplified version of eq. 1:
-
-\begin{equation}I_{t}=\beta_0 + \beta_1P_{t} + \epsilon_{t}\end{equation}
-
-While this model does not formally allow for causal inference, we can
-still measure changes in mean landings and income before and after the
-implementation of the reserve and provide valuable input. For both
-models (eq. 1 and eq. 3), we estimate the model coefficients with
-ordinary least squares, and calculate heteroskedastic--robust standard
-errors {[}49{]}.
-
-While biological and some economic data are regularly collected,
-governance data are typically not available nor systematically collected
-by the community or other organizations. Therefore, we created a survey
-specifically designed to collect information needed for the proposed
-indicators (B5, S3, and G1--G15). The survey is included as
-supplementary material in English (S1 Appendix) and Spanish (S2
-Appendix). To analyze governance information, we developed a framework
-based on a literature review of common governance structures and their
-relation to effectiveness in managing fisheries or marine reserves (S3
-Table). This approach has been proven to successfully evaluate
-governance structures {[}50{]}. Unlike with biological and socioeconomic
-objectives (see eqs 1 and 3), MAREA does not quantitatively analyze
-governance information. Rather, it is presented along with the
-biological and socioeconomic indicators to provide managers and users
-with a more complete description of the reserve.
-
-\subsection{Marine Reserve Evaluation App
-(MAREA)}\label{marine-reserve-evaluation-app-marea}
-
-We developed MAREA in R version 3.4.2 and R Studio 1.1.383 {[}51{]}
-using the Shiny package {[}52{]}, to build an interactive web
-application hosted on an open server; the MAREA app can be accessed at
-\href{turfeffect.shinyapps.io/marea/}{turfeffect.shinyapps.io/marea}.
-While the original version was developed in Spanish because it was aimed
-for Mexico and other Latin-American countries, all of its content can be
-translated by a translation widget available within the app.
-
-MAREA is designed as a 6-step process, divided in tabs that appear upon
-launching the app. The first tab introduces the app and summarizes the
-evaluation process. Then, the user selects management objectives, which
-MAREA automatically matches to appropriate indicators, based on Table 2.
-Users can also select and deselect indicators based on their interests
-and data availability by ``clicking'' on the check-boxes in MAREA. The
-user can then load data on one or more reserves, using standard *.csv
-text files; sample datasets are provided within MAREA. Once data have
-been loaded, MAREA identifies all reserves in the data, and lets the
-user select the reserve to be evaluated. At this point, the user can
-also specify the year of implementation of the reserve, reserve
-dimensions, and indicate target species that are of particular
-management interest. MAREA provides the user with a section to confirm
-that all the decisions made leading up to that point are correct. Once
-the user has confirmed all input data, objectives, and other
-information, MAREA performs the formal program evaluation analyses
-discussed above. For a typical data set, the automated analysis step
-takes less than one second. Finally, the user is taken to the results
-tab where all results are presented in a simple format. The user can
-also download a more comprehensive technical report produced in *.pdf
-format.
-
-The first output is a color--coded scorecard intended to provide a
-general overview of the effectiveness of the reserve. The scorecard
-provides a global score for the reserve, a general score for each
-category of indicators, and an individual score for each indicator. The
-global and category--level scores are determined by the percentage of
-positive indicators, overall and for each category, respectively. For
-numeric biological indicators (all but B5), the color is defined by the
-sign of the interaction term coefficient (\(\beta_2\)) in eq. 1. For
-socioeconomic indicators, colors are assigned based on the direction of
-the slope (\(\beta_1\)) in eq. 3. Red, yellow, and green are used for
-\(\beta_i<0\), \(\beta_i = 0\), and \(\beta_i>0\), respectively. The
-intensity of the color is defined by the significance of the
-coefficient, testing the null hypothesis of no change (\emph{i.e.}
-\(H_0: \beta_i = 0\)) with a Student's t-test. Cutoff values are
-\(p < 0.05\) and \(p < 0.1\). Thus, even in a case where
-\(\beta_i > 0\), if the coefficient is not significant by standard
-measures (\emph{i.e. } \(p>0.1\)), the indicator will be assigned a
-yellow color. A legend (Fig. 2) is provided within the scorecard to aid
-in the interpretation of these results. Governance indicators are
-represented simply by red or green. The color is defined based on what
-literature shows to be a negative (red) or positive (green) factor for a
-reserve (S3 Table). For example, if the perceived degree of illegal
-fishing is high, this indicator will be assigned a red color. However,
-due to the nature of some governance indicators, which require the user
-to provide a narrative, only some indicators are presented in the
-scorecard (although all are included in the technical report).
-
-\textbf{Fig 2. Legend used to interpret the scorecard produced by
-MAREA.} Colors indicate direction of change (red = negative; green =
-positive), and color intensity is given by the statistical significance.
-
-The second output from MAREA is a technical report intended to
-communicate information and statistical results in a more comprehensive
-and technical way. This report also includes a scorecard as a summary of
-the results, but provides more information for each indicator. For all
-numeric biological indicators, the report includes a graph of the value
-of the indicator in the reserve and control sites through time. It also
-provides a regression table that summarizes the value of all
-coefficients in the regression and their respective robust standard
-errors. The summary table also provides information on model fit
-(\(R^2\)) and significance of the regression.
-
-The scorecard is produced with functions from the Shinydashboard package
-{[}53{]}. The technical report is produced by a parameterized Rmarkdown
-document {[}54{]} processed by the knitr package {[}55{]}. Another
-feature of MAREA is that the user can choose to share the data. Once the
-technical report is downloaded, the information on the reserve, its
-management objectives, and all uploaded data are saved into a central
-repository. These data can be accessed at any time by any person
-interested in acquiring them at \url{github.com/turfeffect/MAREAdata}.
-
-\subsection{Case study}\label{case-study}
-
-While MAREA is a general tool that can be easily employed to evaluate
-the effectiveness of any marine reserve with the required input data, we
-illustrate its use here by applying it to one marine reserve near Isla
-Natividad, in Baja California Sur, Mexico. Isla Natividad is located 8
-Km off the Pacific Coast of the Baja California Peninsula (Fig. 3),
-where fishers operate under a fishing cooperative (\emph{S.C.P.P. Buzos
-y Pescadores de la Baja California}) that promotes co-management of
-marine resources {[}56,57{]}. Additionally, fishers have Territorial Use
-Rights for Fisheries (TURFs) that provide them with exclusive access
-rights to exploit the benthic marine resources within a given perimeter
-{[}57{]}.
-
-\textbf{Fig 3. General location of Isla Natividad (left) and map of the
-island (right).} The marine reserve polygon is indicated in red, and the
-approximate location of control sites is indicated by blue squares (B =
-Babencho, D = La Dulce). Shapefiles for Mexican coastline and the United
-States were obtained from the Instituto Nacional de Estadistica,
-Geografia e Informatica of Mexico (INEGI:
-www.inegi.org.mx/geo/contenidos/geoestadistica/m\_geoestadistico.aspx)
-and the tmap R package {[}58{]}, respectively.
-
-In 2006, the Isla Natividad community established a biological baseline
-following the data collection protocol described in this study. The
-community then implemented two community-based marine reserves within
-their TURF {[}8,41,59{]} after establishing a baseline for the
-soon-to-be reserves and control sites. Evidence suggest that these
-reserves have been effective at enhancing resilience to climate
-variations {[}8{]} and preserving genetic diversity of high value
-commercial species such as abalone {[}10{]}. These ecological benefits
-have been translated into economic benefits, enhancing population
-persistence and bolstering abalone fisheries {[}43{]}. For the purpose
-of this evaluation, we focused on the ``La Plana / Las Cuevas'' marine
-reserve, located at the southern end of the island (Fig. 3) and its
-corresponding control site ``La Dulce / Babencho''.
-
-The objective of this reserve was to recover species of economic
-interest ----which were overexploited-- and to enhance fishery
-production in nearby waters. Fishers were also interested in preserving
-biological diversity and the ecosystem. Thus, objectives 4---7 were
-selected. Using Table 2 to match these objectives with appropriate
-management indicators, we selected all biological, socioeconomic, and
-governance indicators included as options in the framework.
-
-Local fishers (who were trained in scientific diving by the CSO
-Comunidad y Biodiversidad, A.C. (COBI; \url{www.cobi.org}), ReefCheck
-California, and Stanford University) and personnel from these
-institutions performed SCUBA dives to record fish and invertebrate
-richness and abundances, as well as fish total length. They recorded
-information along 30 m transects, with a sampling window of 2 m x 2 m
-following a standardized ReefCheck protocol {[}60{]}. Ecological surveys
-were performed yearly in each reserve and corresponding control site(s),
-before and after the implementation of the reserve, providing the
-requisite time series data inside the reserve and for a suitable control
-site. Annual surveys (2006--2016) were carried out in late July -- early
-August, performing a total of 242 and 245 transects in the reserve site
-for fish and invertebrate surveys, respectively. Similar sampling effort
-was applied to the control site, with 221 fish and 222 invertebrate
-transects. Between 12 and 27 transects were performed in each site every
-year.
-
-Socioeconomic data were obtained from the National Commission for
-Aquaculture and Fisheries (\emph{Comisión Nacional de Acuacultura y
-Pesca}; CONAPESCA). The data contains species-level information on
-monthly landings and income from nine species from 2000 to 2014. Data on
-landings and income were aggregated by year and species, and adjusted by
-the Consumer Price Index {[}48{]}. From the nine species available, we
-selected as objective species those that contributed the most (88.27\%)
-income from 2000 to 2014: lobster (\emph{Panulirus interruptus};
-71.76\%), red sea urchin (\emph{Mesocentrotus franciscanus}; 9.33\%),
-snail (\emph{Megastraea undosa}; 3.93\%), and sea cucumber
-(\emph{Parastichopus parvimensis}; 3.23\%). Abalone species
-(\emph{Haliotis fulgens}; 4.52\% and \emph{Haliotis corrugata}; 6.16\%)
-were excluded because the cooperative implemented an informal closure of
-these fisheries in 2010 to allow the population to recover. Eliminating
-all fishing pressure on abalones means that the control site receives
-(for this species) the same treatment as the reserve.
-
-We constructed the governance data based on local knowledge of the area
-and the community.
-
-\section{Results from illustrative
-example}\label{results-from-illustrative-example}
-
-In this section we show the results of the application of MAREA to the
-La Plana/Las Cuevas marine reserve in Isla Natividad, Mexico. These
-results are intended to highlight the relevance and utility of the MAREA
-framework and app, which automate the analysis and make it replicable.
-While we highlight some of the general observed trends, we focus on the
-utility of the tool rather than on the specific effectiveness of this
-case study marine reserve.
-
-The scorecard (Fig. 4) shows that this reserve achieves a general score
-of 64\%, suggesting that 64\% of all indicators are positive. All
-category--level scores were also high, with values of 67\%, 60\%, and
-71\% positive indicators for biological, socioeconomic and governance,
-respectively.
-
-\textbf{Fig 4. Scorecard produced by MAREA for the ``La Plana / Las
-Cuevas'' marine reserve in Isla Natividad, Mexico.}
-
-Among the biological indicators, the greatest effect of the reserve was
-observed for snail and sea cucumber densities, with values of
-\(\beta_2 = 97.17\) (\emph{p} \textless{} 0.05) and \(\beta_2 = 2.31\)
-(\emph{p} \textless{} 0.05), respectively. Fish indicators showed no
-significant change (\emph{p} \textgreater{} 0.1), with negative trends
-for Shannon's diversity index and fish species richness and positive
-trends for density, biomass, and mean trophic level. Changes through
-time for these indicators are presented in Figure 5, and a summary of
-\(\beta_2\) coefficients is provided in Table 3.
-
-\textbf{Fig. 5 Plots for values of each biological indicator (y-axis)
-through time (x-axis).} Red and blue correspond to the reserve and
-control sites, respectively. Black lines indicate yearly mean values,
-and ribbons indicate \(\pm\) 1 standard error. Dots are horizontally
-jittered to aid visualization. This figure contains information for fish
-Shannon's diversity index (a), fish species richness (b), fish density
-(c), fish trophic level (d), fish biomass (e), invertebrate Shannon's
-diversity index (f), invertebrate species richness (g), invertebrate
-density (h), lobster density (i), urchin density (j), snail density (k),
-and sea cucumber density (l).
-
-\textbf{Table 3. Summary of average treatment effect of the reserve on
-biological indicators.}
-
-\begin{table}[H]
+Recall that the governance objectives are evaluated based on the institutions present, not on a specific quantitative linkage between governance and biological or economic outcomes.
+Data for this reserve suggest that the community is strongly organized, which is a likely driver of the successes reported above \cite{gutirrez_2011-0U}.
+The first point of success is the existence of a fishing cooperative that is also affiliated with a regional federation of cooperatives.
+These polycentric governance structures allow various levels of organization that have been shown to foster communication and cooperation \cite{mccay_2014-CN,espinosaromero_2014-PY}; federations also provide bargain power with governments \cite{espinosaromero_2014-PY,finkbeiner_2015-87}.
+Access to fishing resources is managed through a TURF, permits, and fishing quotas (for some species).
+It has been suggested that TURFs promotes a sense of stewardship of resources and incentivizes sustainable management \cite{mccay_2017-1m}.
+Together, these structures enabled a participative, bottom–up process during the reserve design phase; opinions of all fishing members --and often non-fishing community members-- were included.
+Participation of community members in reserve surveillance and yearly monitoring indicate commitment and interest, and allow informal communication of results to uninvolved community members.
+Furthermore, the reserve is partially isolated from poaching activity, and fishers have internal regulations pertaining to the reserves.
+The low level of illegal fishing by members of the community and outsiders both inside and outside the reserve is another indication of effectiveness.
+Governance indicators are summarized in Table~\ref{table5}.
+
+\begin{table}[!ht]
+\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
 \centering
-\begin{tabular}{l|l|r}
-\hline
-\bfseries{Indicator} & \bfseries{Estimate (SD)} & \bfseries{t-score}\\
-\hline
-Shannon fish & -0.22 (0.16) & -1.3969\\
-\hline
-Richness fish & -0.61 (0.43) & -1.4073\\
-\hline
-Density fish & 0.74 (6.15) & 0.1205\\
-\hline
-Trophic fish & 0.00 (0.01) & 0.1399\\
-\hline
-Biomass fish & 0.22 (1.47) & 0.1476\\
-\hline
-Shannon invert & -0.67 (0.22)** & -3.0481\\
-\hline
-Richness invert & -2.71 (0.81)** & -3.3519\\
-\hline
-Density invert & 91.21 (47.11)* & 1.9362\\
-\hline
-Lobster & 7.66 (8.93) & 0.8583\\
-\hline
-Urchin & 2.15 (1.23)* & 1.7425\\
-\hline
-Snail & 97.17 (42.90)** & 2.2652\\
-\hline
-Cucumber & 2.31 (1.17)** & 1.9782\\
-\hline
-\end{tabular}
-\end{table}
-
-* Indicate significance level, with (*) indicating p \textless{} 0.1 and
-(**) p \textless{} 0.05.
-
-\clearpage
-
-One of the main objectives of this reserve was to increase landings.
-Results of the socioeconomic indicators show that total landings were,
-on average, 64.20 metric tonnes higher (\emph{p} \textgreater{} 0.1)
-after the implementation of the reserves, though this cannot necessarily
-be interpreted as causal, because it relies entirely on a before-after
-comparison. Total income was \$10,344.85 (\emph{p} \textless{} 0.05)
-thousands of Mexican Pesos (K MXP) higher after the implementation of
-the reserves. On average, lobster and sea cucumber landings increased,
-while urchin and snail landings and income decreased. Figure 6 presents
-the changes in these indicators through time, and Table 4 summarizes
-these results.
-
-\textbf{Fig. 6 Plots for values of each socioeconomic indicator (y-axis)
-through time (x-axis).} Red and blue correspond to before and after the
-implementation of the reserve, respectively. This figure contains
-information for total landings (a), total income (b), lobster landings
-(c), urchin landings (d), snail landings (e), sea cucumber landings (f),
-lobster income (g), urchin income (h), snail income (i), and sea
-cucumber income (j).
-
-\textbf{Table 4. Summary of differences in socioeconomic indicators
-before and after the implementation of the reserve.}
-
-\begin{table}[H]
-\centering
-\begin{tabular}{l|l|r}
-\hline
-\bfseries{Indicator} & \bfseries{Estimate (SD)} & \bfseries{t-score}\\
-\hline
-Landings & 64.20 (90.07) & 0.7127\\
-\hline
-Income & 10344.85 (3982.20)** & 2.5978\\
-\hline
-Lobster landings & 7.37 (13.95) & 0.5281\\
-\hline
-Urchin landings & -30.00 (9.49)** & -3.1620\\
-\hline
-Snail landings & -69.53 (33.82)* & -2.0561\\
-\hline
-Cucumber landings & 9.34 (6.72) & 1.3906\\
-\hline
-Lobster income & 14372.85 (3634.64)** & 3.9544\\
-\hline
-Urchin income & -5800.46 (1867.50)** & -3.1060\\
-\hline
-Snail income & -404.85 (187.07)** & -2.1641\\
-\hline
-Cucumber income & 131.49 (185.66) & 0.7082\\
-\hline
-\end{tabular}
-\end{table}
-
-* Indicate significance level, with (*) indicating p \textless{} 0.1 and
-(**) p \textless{} 0.05.
-
-Recall that the governance objectives are evaluated based on the
-institutions present, not on a specific quantitative linkage between
-governance and biological or economic outcomes. Data for this reserve
-suggest that the community is strongly organized, which is a likely
-driver of the successes reported above {[}61{]}. The first point of
-success is the existence of a fishing cooperative that is also
-affiliated with a regional federation of cooperatives. These polycentric
-governance structures allow various levels of organization that have
-been shown to foster communication and cooperation {[}50,57{]};
-federations also provide bargain power with governments {[}50,62{]}.
-Access to fishing resources is managed through a TURF, permits, and
-fishing quotas (for some species). McCay {[}56{]} suggests that the TURF
-promotes a sense of stewardship of their resources and incentivizes
-sustainable management. Together, these structures enabled a
-participative, bottom--up process during the reserve design phase;
-opinions of all fishing members --and often non-fishing community
-members-- were included. Participation of community members in reserve
-surveillance and yearly monitoring indicate commitment and interest, and
-allow informal communication of results to uninvolved community members.
-Furthermore, the reserve is partially isolated from poaching activity,
-and fishers have internal regulations pertaining to the reserves. The
-low level of illegal fishing by members of the community and outsiders
-both inside and outside the reserve represents another indication of
-effectiveness. Governance indicators are summarized in Table 5.
-
-\clearpage
-
-\textbf{Table 5. Summary of governance indicators.}
-
-\begin{table}[H]
-\centering
+\caption{
+{\bf Summary of governance indicators.}}
 \begin{tabular}{l|>{\raggedright\arraybackslash}p{9cm}}
@@
 \end{tabular}
+\label{table5}
+\end{adjustwidth}
 \end{table}
 
-\clearpage
-
-\section{Discussion}\label{discussion}
-
-We have developed and presented an automated approach for evaluating the
-effectiveness of marine reserves in Mexico, and perhaps around the
-world. Here we highlight MAREA's utility for evidence-based management,
-and comment on a few of its shortcomings. The findings from Isla
-Natividad are used purely to validate the relevance of MAREA rather than
-to discuss particularities of the marine reserve effectiveness, which
-has been described before {[}8,10,43{]}. We use examples from the case
-study to build on the utility of MAREA and discuss ways in which results
-can be interpreted to inform management.
-
-The causal inference techniques used by MAREA have been suggested
-{[}38,47{]} and used {[}21{]} before in other ad hoc studies. This
-approach reduces ambiguity in the interpretation of results. For
-example, invertebrate density decreased through time inside and outside
-of the reserve (Fig. 5h). In this case, a before--after evaluation of
-the reserve (\emph{i.e.} ignoring the control site) would have
-incorrectly concluded that the reserve failed to protect invertebrates.
-On the other hand, a control--impact approach (\emph{i.e.} compare
-reserve vs.~control site only in 2016) would have identified higher
-densities inside the reserve, concluding that the reserve increases
-invertebrate density. However, by executing a formal
-difference-in-differences approach for causal inference, MAREA
-identifies the changes through time and across sites, and estimates the
-effect of the reserve on density at \(\beta_2 = 91.21\) (\emph{p}
-\textless{} 0.05). This approach reveals that invertebrate densities
-decrease in both sites through time, but the decrease is faster for the
-control site, thus yielding a positive value for \(\beta_2\).
-
-The approach used by MAREA to estimate the effect of the reserve on
-biological indicators requires cautious interpretation of the results.
-The value of the \(\beta_2\) coefficient represents the difference
-between the temporal trends of the reserve and control sites {[}21{]}.
-As exemplified by the case of invertebrate densities, a positive value
-(\emph{i.e.} \(\beta_2 > 0\)) does not necessarily indicate an increase
-in the indicator through time, but rather a positive difference with
-respect to the temporal trend of the control site. The inverse occurs
-for negative values of \(\beta_2\).
-
-MAREA provides in-depth analysis and a convenient snapshot overview of
-the effect of the reserve, allowing users to rapidly identify trends.
-However, users must interpret multiple indicators at a time to better
-understand the results. For example, with additional knowledge of local
-environmental variability (\emph{i.e.} indicator B5: Natural
-Disturbance), we can better understand the trends in invertebrate
-densities. As reported before {[}8{]}, hypoxic conditions that have
-occurred in Isla Natividad can cause decreases in invertebrate
-densities, and reserves buffer the negative effect. While MAREA
-automates the analysis and makes results replicable, proper
-interpretation will still depend on the user. Results produced by MAREA
-can only aid in management and decision making when results have been
-correctly interpreted.
-
-Socioeconomic and governance indicators typically lack a control site,
-which impede us from using the causal inference techniques employed to
-measure biological changes {[}25{]}. However, we can still extract
-useful information from them. Again, by combining results from multiple
-indicators, MAREA can provide insights into the effect of the reserve.
-For example, lobster and sea cucumber have shown increases in densities,
-landings, and income. We cannot conclude that landings and income from
-these species have increased due to the reserve, but we can at least
-conclude that landings have not decreased. While further information on
-market behavior of each fishery is needed, these results provide
-insights into the state of the reserve and its associated fisheries.
-
-As for the governance information, it is difficult to establish causal
-links between the state of the reserve and the governance structures
-present in the community. However, providing a single platform
-(\emph{i.e.} scorecard) or document (\emph{i.e.} technical report) where
-biological, socioeconomic, and governance information is comprehensively
-included can aid in management. By using MAREA, this information will be
-reported across reserves in a standardized way, and can help managers
-identify overarching patterns across sites.
-
-By making results straightforward to interpret, MAREA may also assist in
-communication with a broader stakeholder community. While stakeholder
-involvement in the design and implementation phases of marine reserves
-is important, that may not be sufficient for ensuring long-term buy-in
-or success. The scorecard is easily understandable by experts and
-non-experts, and can be used as an effective tool for communicating the
-results of annual evaluations. Additionally, the technical report can
-serve as a tool for managers and scientists to rapidly produce and
-communicate information at a more technical level.
-
-We recognize that the seven objectives and 29 indicators used by MAREA
-might not fully describe a reserve in countries other than Mexico. In
-order to ensure the applicability of the tool to reserves in other
-countries, further testing in other regions should take place. However,
-the proposed objectives and indicators provide a starting point to
-perform the evaluation, to which managers and users can add other
-indicators (\emph{e.g.} larval dispersal or connectivity) that are
-relevant to their reserve. Furthermore, MAREA's value is that it
-provides a free, simple, and replicable way to perform rigorous impact
-analysis. The tool can easily be used by fishers, CSO members, and
-managers in government agencies, providing transparency of the analysis
-and results. In addition, it can empower and enable local managers and
-fishers to respond to local change and adapt by allowing direct and easy
-access to the information.
-
-An evident limitation of MAREA is its dependence on data obtained
-through a BACI design, and the amount of samples needed to estimate
-coefficients in Eqn. 1. It is not uncommon for control sites or
-baselines to be absent. Properly designing marine reserves by
-identifying control sites and establishing a baseline before the
-implementation of the reserve is enough to overcome this issue; reserves
-for which there is no control site and baseline cannot be evaluated with
-MAREA. Typical underwater surveys require that at least 12 - 16
-transects are performed for each site (\emph{i.e.} reserve and control)
-each year. This provides at least 48 samples (12 samples per site, per
-year), enough to avoid overfitting Eqn 1. However, these problems can be
-easily avoided during the design and implementation phases by
-anticipating what data will be needed in the eventual evaluation.
-
-To the best of our knowledge, MAREA is the first tool designed to
-evaluate marine reserves. Previous work {[}23,25{]} addressed MPA
-evaluation and provided the foundation for our contribution. However,
-these did not intended to create user-friendly tools to aid in the
-evaluation. As with other conservation management tools, development of
-tools that automatize complex calculations can have an important impact
-in management {[}63{]}. The use of open data science enables the
-creation of open-access tools that can address technical gaps and
-inprove management {[}28{]}.
-
-The effectiveness of marine reserves continues to be a matter of debate
-{[}11,44,64{]}. With current targets set to increase ocean protection,
-it is important that we understand the effects of our interventions
-{[}38{]} so we can better inform management {[}47{]}. It is therefore
-important that academics, managers, fishers, and CSOs have access to
-open access tools like MAREA. This is particularly relevant for Mexico
-and other Latin American countries, where management agencies are often
-understaffed and underfunded {[}65{]}, or where materials are often not
-available in their language. In this context, MAREA provides a simple
-and replicable way to align management objectives with performance
-indicators. The proposed methodologies, especially the way in which
-biological indicators are evaluated, provide valuable information for
-managers. We acknowledge there is room for improvement in the way in
-which socioeconomic and governance data are analyzed. Despite this,
-providing a unifying platform where all indicators can be analyzed and
-comprehensively presented represents a valuable step towards effective
-evidence---based management {[}47{]}.
-
-The first release of MAREA is now available, and it will continue to be
-developed and maintained to keep up to date with the literature. This
-process will incorporate new features, and enhance current ones, aiming
-to improve user experience and expand the scope of the analysis. Other
-modifications may also include addition of more objectives and
-indicators to ensure applicability in other regions, full translation
-into other languages to avoid any ambiguities introduced via the
-automatic translation, or reporting effects over time in percentages to
-aid interpretation. Yet, we believe that this first release represents a
-major step towards effective, replicable evaluation and management of
-marine reserves.
-
-\section{Acknowledgements}\label{acknowledgements}
-
-We thank Olivier Deschenes and Andrew Plantinga, who provided valuable
-input to design the model that evaluates the biological indicators.
-Special thanks to the fishers from Isla Natividad, who gathered the data
-used in this study, and the fishers from El Rosario, who helped us
-validate our survey and framework, and to Arturo Hernández and Alfonso
-Romero who provided help with the logistics. Finally, we thank the
-editor and two anonymous reviewers for their suggestions, which
-significantly improved the quality of this paper.
-
-\clearpage
-
-\section{References}\label{references}
-
-\hypertarget{refs}{}
-\hypertarget{ref-pauly_2005-qV}{}
-1. Pauly D, Watson R, Alder J. Global trends in world fisheries: Impacts
-on marine ecosystems and food security. Philos Trans R Soc Lond, B, Biol
-Sci. 2005;360: 5--12.
-doi:\href{https://doi.org/10.1098/rstb.2004.1574}{10.1098/rstb.2004.1574}
-
-\hypertarget{ref-halpern_2008-dK}{}
-2. Halpern BS, Walbridge S, Selkoe KA, Kappel CV, Micheli F, D'Agrosa C,
-et al. A global map of human impact on marine ecosystems. Science.
-2008;319: 948--952.
-doi:\href{https://doi.org/10.1126/science.1149345}{10.1126/science.1149345}
-
-\hypertarget{ref-lester_2008-F_}{}
-3. Lester S, Halpern B. Biological responses in marine no-take reserves
-versus partially protected areas. Mar Ecol Prog Ser. 2008;367: 49--56.
-doi:\href{https://doi.org/10.3354/meps07599}{10.3354/meps07599}
-
-\hypertarget{ref-lester_2009-Ks}{}
-4. Lester S, Halpern B, Grorud-Colvert K, Lubchenco J, Ruttenberg B,
-Gaines S, et al. Biological effects within no-take marine reserves: A
-global synthesis. Mar Ecol Prog Ser. 2009;384: 33--46.
-doi:\href{https://doi.org/10.3354/meps08029}{10.3354/meps08029}
-
-\hypertarget{ref-sala_2016-PV}{}
-5. Sala E, Costello C, De Bourbon Parme J, Fiorese M, Heal G, Kelleher
-K, et al. Fish banks: An economic model to scale marine conservation.
-Marine Policy. 2016;73: 154--161.
-doi:\href{https://doi.org/10.1016/j.marpol.2016.07.032}{10.1016/j.marpol.2016.07.032}
-
-\hypertarget{ref-hastings_2017-sm}{}
-6. Hastings A, Gaines SD, Costello C. Marine reserves solve an important
-bycatch problem in fisheries. Proc Natl Acad Sci U S A. 2017;114:
-8927--8934.
-doi:\href{https://doi.org/10.1073/pnas.1705169114}{10.1073/pnas.1705169114}
-
-\hypertarget{ref-aburtooropeza_2011-ya}{}
-7. Aburto-Oropeza O, Erisman B, Galland GR, Mascareñas-Osorio I, Sala E,
-Ezcurra E. Large recovery of fish biomass in a no-take marine reserve.
-PLoS ONE. 2011;6: e23601.
-doi:\href{https://doi.org/10.1371/journal.pone.0023601}{10.1371/journal.pone.0023601}
-
-\hypertarget{ref-micheli_2012-EU}{}
-8. Micheli F, Saenz-Arroyo A, Greenley A, Vazquez L, Espinoza Montes JA,
-Rossetto M, et al. Evidence that marine reserves enhance resilience to
-climatic impacts. PLoS ONE. 2012;7: e40832.
-doi:\href{https://doi.org/10.1371/journal.pone.0040832}{10.1371/journal.pone.0040832}
-
-\hypertarget{ref-roberts_2017-J9}{}
-9. Roberts CM, O'Leary BC, McCauley DJ, Cury PM, Duarte CM, Lubchenco J,
-et al. Marine reserves can mitigate and promote adaptation to climate
-change. Proc Natl Acad Sci U S A. 2017;114: 6167--6175.
-doi:\href{https://doi.org/10.1073/pnas.1701262114}{10.1073/pnas.1701262114}
-
-\hypertarget{ref-munguavega_2015-yg}{}
-10. Munguía-Vega A, Sáenz-Arroyo A, Greenley AP, Espinoza-Montes JA,
-Palumbi SR, Rossetto M, et al. Marine reserves help preserve genetic
-diversity after impacts derived from climate variability: Lessons from
-the pink abalone in baja california. Global Ecology and Conservation.
-2015;4: 264--276.
-doi:\href{https://doi.org/10.1016/j.gecco.2015.07.005}{10.1016/j.gecco.2015.07.005}
-
-\hypertarget{ref-edgar_2014-UO}{}
-11. Edgar GJ, Stuart-Smith RD, Willis TJ, Kininmonth S, Baker SC, Banks
-S, et al. Global conservation outcomes depend on marine protected areas
-with five key features. Nature. 2014;506: 216--220.
-doi:\href{https://doi.org/10.1038/nature13022}{10.1038/nature13022}
-
-\hypertarget{ref-giakoumi_2017-V2}{}
-12. Giakoumi S, Scianna C, Plass-Johnson J, Micheli F, Grorud-Colvert K,
-Thiriet P, et al. Ecological effects of full and partial protection in
-the crowded mediterranean sea: A regional meta-analysis. Sci Rep.
-2017;7: 8940.
-doi:\href{https://doi.org/10.1038/s41598-017-08850-w}{10.1038/s41598-017-08850-w}
-
-\hypertarget{ref-sala_2017-69}{}
-13. Sala E, Giakoumi S. No-take marine reserves are the most effective
-protected areas in the ocean. ICES Journal of Marine Science. 2017;
-doi:\href{https://doi.org/10.1093/icesjms/fsx059}{10.1093/icesjms/fsx059}
-
-\hypertarget{ref-betti_2017-lq}{}
-14. Betti F, Bavestrello G, Bo M, Asnaghi V, Chiantore M, Bava S, et al.
-Over 10 years of variation in mediterranean reef benthic communities.
-Marine Ecology. 2017;38: e12439.
-doi:\href{https://doi.org/10.1111/maec.12439}{10.1111/maec.12439}
-
-\hypertarget{ref-szuwalski_2017-jc}{}
-15. Szuwalski CS, Burgess MG, Costello C, Gaines SD. High fishery
-catches through trophic cascades in china. Proc Natl Acad Sci U S A.
-2017;114: 717--721.
-doi:\href{https://doi.org/10.1073/pnas.1612722114}{10.1073/pnas.1612722114}
-
-\hypertarget{ref-chavez_2003-_m}{}
-16. Chavez FP, Ryan J, Lluch-Cota SE, Niquen C M. From anchovies to
-sardines and back: Multidecadal change in the pacific ocean. Science.
-2003;299: 217--221.
-doi:\href{https://doi.org/10.1126/science.1075880}{10.1126/science.1075880}
-
-\hypertarget{ref-davies_2017-ml}{}
-17. Davies TK, Mees CC, Milner-Gulland EJ. Use of a counterfactual
-approach to evaluate the effect of area closures on fishing location in
-a tropical tuna fishery. PLoS ONE. 2017;12: e0174758.
-doi:\href{https://doi.org/10.1371/journal.pone.0174758}{10.1371/journal.pone.0174758}
-
-\hypertarget{ref-guidetti_2014-8Z}{}
-18. Guidetti P, Baiata P, Ballesteros E, Di Franco A, Hereu B,
-Macpherson E, et al. Large-scale assessment of mediterranean marine
-protected areas effects on fish assemblages. PLoS ONE. 2014;9: e91841.
-doi:\href{https://doi.org/10.1371/journal.pone.0091841}{10.1371/journal.pone.0091841}
-
-\hypertarget{ref-friedlander_2017-oI}{}
-19. Friedlander AM, Golbuu Y, Ballesteros E, Caselle JE, Gouezo M,
-Olsudong D, et al. Size, age, and habitat determine effectiveness of
-palau's marine protected areas. PLoS ONE. 2017;12: e0174787.
-doi:\href{https://doi.org/10.1371/journal.pone.0174787}{10.1371/journal.pone.0174787}
-
-\hypertarget{ref-rodriguez_2017-PD}{}
-20. Rodriguez AG, Fanning LM. Assessing marine protected areas
-effectiveness: A case study with the tobago cays marine park. OJMS.
-2017;07: 379--408.
-doi:\href{https://doi.org/10.4236/ojms.2017.73027}{10.4236/ojms.2017.73027}
-
-\hypertarget{ref-moland_2013-VP}{}
-21. Moland E, Olsen EM, Knutsen H, Garrigou P, Espeland SH, Kleiven AR,
-et al. Lobster and cod benefit from small-scale northern marine
-protected areas: Inference from an empirical before-after control-impact
-study. Proc Biol Sci. 2013;280: 20122679.
-doi:\href{https://doi.org/10.1098/rspb.2012.2679}{10.1098/rspb.2012.2679}
-
-\hypertarget{ref-soykan_2015-nu}{}
-22. Soykan CU, Lewison RL. Using community-level metrics to monitor the
-effects of marine protected areas on biodiversity. Conserv Biol.
-2015;29: 775--783.
-doi:\href{https://doi.org/10.1111/cobi.12445}{10.1111/cobi.12445}
-
-\hypertarget{ref-pomeroy_2005-Py}{}
-23. Pomeroy RS, Watson LM, Parks JE, Cid GA. How is your mpa doing? A
-methodology for evaluating the management effectiveness of marine
-protected areas. Ocean Coast Manag. 2005;48: 485--502.
-doi:\href{https://doi.org/10.1016/j.ocecoaman.2005.05.004}{10.1016/j.ocecoaman.2005.05.004}
-
-\hypertarget{ref-pomeroy_2004-23}{}
-24. Pomeroy RS, Parks JE, Watson LM. How is your mpa doing ? A guidebook
-of natural and social indicators for evaluating marine protected areas
-management effectiveness {[}Internet{]}. IUCN; 2004.
-doi:\href{https://doi.org/10.2305/IUCN.CH.2004.PAPS.1.en}{10.2305/IUCN.CH.2004.PAPS.1.en}
-
-\hypertarget{ref-mascia_2017-m_}{}
-25. Mascia MB, Fox HE, Glew L, Ahmadia GN, Agrawal A, Barnes M, et al. A
-novel framework for analyzing conservation impacts: Evaluation, theory,
-and marine protected areas. Ann N Y Acad Sci. 2017;1399: 93--115.
-doi:\href{https://doi.org/10.1111/nyas.13428}{10.1111/nyas.13428}
-
-\hypertarget{ref-ostrom_2009-hg}{}
-26. Ostrom E. A general framework for analyzing sustainability of
-social-ecological systems. Science. 2009;325: 419--422.
-doi:\href{https://doi.org/10.1126/science.1172133}{10.1126/science.1172133}
-
-\hypertarget{ref-basurto_2013-oq}{}
-27. Basurto X, Gelcich S, Ostrom E. The social--ecological system
-framework as a knowledge classificatory system for benthic small-scale
-fisheries. Global Environmental Change. 2013;23: 1366--1380.
-doi:\href{https://doi.org/10.1016/j.gloenvcha.2013.08.001}{10.1016/j.gloenvcha.2013.08.001}
-
-\hypertarget{ref-lowndes_2017-xh}{}
-28. Lowndes JSS, Best BD, Scarborough C, Afflerbach JC, Frazier MR,
-O'Hara CC, et al. Our path to better science in less time using open
-data science tools. Nat ecol evol. 2017;1: 0160.
-doi:\href{https://doi.org/10.1038/s41559-017-0160}{10.1038/s41559-017-0160}
-
-\hypertarget{ref-halpern_2012-k9}{}
-29. Halpern BS, Longo C, Hardy D, McLeod KL, Samhouri JF, Katona SK, et
-al. An index to assess the health and benefits of the global ocean.
-Nature. 2012;488: 615--620.
-doi:\href{https://doi.org/10.1038/nature11397}{10.1038/nature11397}
-
-\hypertarget{ref-halpern_2017-Zi}{}
-30. Halpern BS, Frazier M, Afflerbach J, O'Hara C, Katona S, Stewart
-Lowndes JS, et al. Drivers and implications of change in global ocean
-health over the past five years. PLoS ONE. 2017;12: e0178267.
-doi:\href{https://doi.org/10.1371/journal.pone.0178267}{10.1371/journal.pone.0178267}
-
-\hypertarget{ref-selig_2015-F9}{}
-31. Selig ER, Frazier M, O'Leary JK, Jupiter SD, Halpern BS, Longo C, et
-al. Measuring indicators of ocean health for an island nation: The ocean
-health index for fiji. Ecosystem Services. 2015;16: 403--412.
-doi:\href{https://doi.org/10.1016/j.ecoser.2014.11.007}{10.1016/j.ecoser.2014.11.007}
-
-\hypertarget{ref-halpern_2014-lQ}{}
-32. Halpern BS, Longo C, Scarborough C, Hardy D, Best BD, Doney SC, et
-al. Assessing the health of the u.S. west coast with a regional-scale
-application of the ocean health index. PLoS ONE. 2014;9: e98995.
-doi:\href{https://doi.org/10.1371/journal.pone.0098995}{10.1371/journal.pone.0098995}
-
-\hypertarget{ref-elfes_2014-RC}{}
-33. Elfes CT, Longo C, Halpern BS, Hardy D, Scarborough C, Best BD, et
-al. A regional-scale ocean health index for brazil. PLoS ONE. 2014;9:
-e92589.
-doi:\href{https://doi.org/10.1371/journal.pone.0092589}{10.1371/journal.pone.0092589}
-
-\hypertarget{ref-anderson_2015-ND}{}
-34. Anderson JL, Anderson CM, Chu J, Meredith J, Asche F, Sylvia G, et
-al. The fishery performance indicators: A management tool for triple
-bottom line outcomes. PLoS ONE. 2015;10: e0122809.
-doi:\href{https://doi.org/10.1371/journal.pone.0122809}{10.1371/journal.pone.0122809}
-
-\hypertarget{ref-dowling_2016-pO}{}
-35. Dowling N, Wilson J, Rudd M, Babcock E, Caillaux M, Cope J, et al.
-FishPath: A decision support system for assessing and managing data- and
-capacity- limited fisheries. In: Quinn II T, Armstrong J, Baker M,
-Heifetz J, Witherell D, editors. Assessing and managing data-limited
-fish stocks. Alaska Sea Grant, University of Alaska Fairbansk; 2016.
-doi:\href{https://doi.org/10.4027/amdlfs.2016.03}{10.4027/amdlfs.2016.03}
-
-\hypertarget{ref-oyanedel_2017-TO}{}
-36. Oyanedel R, Macy Humberstone J, Shattenkirk K, Rodriguez Van-Dyck S,
-Joye Moyer K, Poon S, et al. A decision support tool for designing
-turf-reserves. BMS. 2017;93: 155--172.
-doi:\href{https://doi.org/10.5343/bms.2015.1095}{10.5343/bms.2015.1095}
-
-\hypertarget{ref-vilela_2017-Zo}{}
-37. Vilela T, Reid J. Improving hydropower choices via an online and
-open access tool. PLoS ONE. 2017;12: e0179393.
-doi:\href{https://doi.org/10.1371/journal.pone.0179393}{10.1371/journal.pone.0179393}
-
-\hypertarget{ref-burgess_2016-HN}{}
-38. Burgess MG, Clemence M, McDermott GR, Costello C, Gaines SD. Five
-rules for pragmatic blue growth. Marine Policy. 2016;
-doi:\href{https://doi.org/10.1016/j.marpol.2016.12.005}{10.1016/j.marpol.2016.12.005}
-
-\hypertarget{ref-nom049sagpesc_2014-V6}{}
-39. NOM-049-SAG/PESC. NORMA oficial mexicana nom-049-sag/pesc-2014, que
-determina el procedimiento para establecer zonas de refugio para los
-recursos pesqueros en aguas de jurisdicción federal de los estados
-unidos mexicanos. DOF. 2014;
-
-\hypertarget{ref-lgeepa_2017-jL}{}
-40. LGEEPA. Ley general del equilibrio ecológico y la protección al
-ambiente. DOF. 2017; Available:
-\url{http://www.diputados.gob.mx/LeyesBiblio/pdf/148/_240117.pdf}
-
-\hypertarget{ref-lester_2017-nh}{}
-41. Lester S, McDonald G, Clemence M, Dougherty D, Szuwalski C. Impacts
-of turfs and marine reserves on fisheries and conservation goals:
-Theory, empirical evidence, and modeling. BMS. 2017;93: 173--198.
-doi:\href{https://doi.org/10.5343/bms.2015.1083}{10.5343/bms.2015.1083}
-
-\hypertarget{ref-chirico_2017-Rz}{}
-42. Chirico AAD, McClanahan TR, Eklöf JS. Community- and
-government-managed marine protected areas increase fish size, biomass
-and potential value. PLoS ONE. 2017;12: e0182342.
-doi:\href{https://doi.org/10.1371/journal.pone.0182342}{10.1371/journal.pone.0182342}
-
-\hypertarget{ref-rossetto_2015-V0}{}
-43. Rossetto M, Micheli F, Saenz-Arroyo A, Montes JAE, De Leo GA, Rochet
-M-J. No-take marine reserves can enhance population persistence and
-support the fishery of abalone. Can J Fish Aquat Sci. 2015;72:
-1503--1517.
-doi:\href{https://doi.org/10.1139/cjfas-2013-0623}{10.1139/cjfas-2013-0623}
-
-\hypertarget{ref-woodcock_2017-Wm}{}
-44. Woodcock P, O'Leary BC, Kaiser MJ, Pullin AS. Your evidence or mine?
-Systematic evaluation of reviews of marine protected area effectiveness.
-Fish Fish. 2017;18: 668--681.
-doi:\href{https://doi.org/10.1111/faf.12196}{10.1111/faf.12196}
-
-\hypertarget{ref-carter_2017-Uf}{}
-45. Carter AB, Davies CR, Emslie MJ, Mapstone BD, Russ GR, Tobin AJ, et
-al. Reproductive benefits of no-take marine reserves vary with region
-for an exploited coral reef fish. Sci Rep. 2017;7: 9693.
-doi:\href{https://doi.org/10.1038/s41598-017-10180-w}{10.1038/s41598-017-10180-w}
-
-\hypertarget{ref-villaseorderbez_website_2017-xE}{}
-46. Villaseñor-Derbez JC, Faro C, Wright M, Martínez J. A guide to
-evaluate the effectiveness of no-take marine reserves in mexico
-{[}Internet{]}. 2017. Available:
-\url{https://www.researchgate.net/publication/317840581/_A/_guide/_to/_evaluate/_the/_effectiveness/_of/_no-take/_marine/_reserves/_in/_Mexico}
-
-\hypertarget{ref-ferraro_2006-oW}{}
-47. Ferraro PJ, Pattanayak SK. Money for nothing? A call for empirical
-evaluation of biodiversity conservation investments. PLoS Biol. 2006;4:
-e105.
-doi:\href{https://doi.org/10.1371/journal.pbio.0040105}{10.1371/journal.pbio.0040105}
-
-\hypertarget{ref-oecd_website_2017-VV}{}
-48. OECD. Prices - inflation (cpi) - oecd data {[}Internet{]}. 2017.
-Available: \url{https://data.oecd.org/price/inflation-cpi.htm}
-
-\hypertarget{ref-sandwich_2014}{}
-49. Zeileis A. Econometric computing with hc and hac covariance matrix
-estimators. Journal of Statistical Software. 2004;11: 1--17.
-
-\hypertarget{ref-espinosaromero_2014-PY}{}
-50. Espinosa-Romero MJ, Rodriguez LF, Weaver AH, Villanueva-Aznar C,
-Torre J. The changing role of ngos in mexican small-scale fisheries:
-From environmental conservation to multi-scale governance. Marine
-Policy. 2014;50: 290--299.
-doi:\href{https://doi.org/10.1016/j.marpol.2014.07.005}{10.1016/j.marpol.2014.07.005}
-
-\hypertarget{ref-rcore_2017}{}
-51. R Core Team. R: A language and environment for statistical computing
-{[}Internet{]}. Vienna, Austria: R Foundation for Statistical Computing;
-2017. Available: \url{https://www.R-project.org/}
-
-\hypertarget{ref-shiny_2017}{}
-52. Chang W, Cheng J, Allaire J, Xie Y, McPherson J. Shiny: Web
-application framework for r {[}Internet{]}. 2017. Available:
-\url{https://CRAN.R-project.org/package=shiny}
-
-\hypertarget{ref-shinydashboard_2017}{}
-53. Chang W, Borges Ribeiro B. Shinydashboard: Create dashboards with
-'shiny' {[}Internet{]}. 2017. Available:
-\url{https://CRAN.R-project.org/package=shinydashboard}
-
-\hypertarget{ref-rmarkdown_2017}{}
-54. Allaire J, Cheng J, Xie Y, McPherson J, Chang W, Allen J, et al.
-Rmarkdown: Dynamic documents for r {[}Internet{]}. 2017. Available:
-\url{https://CRAN.R-project.org/package=rmarkdown}
-
-\hypertarget{ref-knitr_2017}{}
-55. Xie Y. Knitr: A general-purpose package for dynamic report
-generation in r {[}Internet{]}. 2017. Available:
-\url{http://yihui.name/knitr/}
-
-\hypertarget{ref-mccay_2017-1m}{}
-56. McCay B. Territorial use rights in fisheries of the northern pacific
-coast of mexico. BMS. 2017;93: 69--81.
-doi:\href{https://doi.org/10.5343/bms.2015.1091}{10.5343/bms.2015.1091}
-
-\hypertarget{ref-mccay_2014-CN}{}
-57. McCay BJ, Micheli F, Ponce-Díaz G, Murray G, Shester G,
-Ramirez-Sanchez S, et al. Cooperatives, concessions, and co-management
-on the pacific coast of mexico. Marine Policy. 2014;44: 49--59.
-doi:\href{https://doi.org/10.1016/j.marpol.2013.08.001}{10.1016/j.marpol.2013.08.001}
-
-\hypertarget{ref-tmap_2017}{}
-58. Tennekes M. Tmap: Thematic maps {[}Internet{]}. 2017. Available:
-\url{https://CRAN.R-project.org/package=tmap}
-
-\hypertarget{ref-afflerbach_2014-HP}{}
-59. Afflerbach JC, Lester SE, Dougherty DT, Poon SE. A global survey of
--reserves, territorial use rights for fisheries coupled with marine
-reserves. Global Ecology and Conservation. 2014;2: 97--106.
-doi:\href{https://doi.org/10.1016/j.gecco.2014.08.001}{10.1016/j.gecco.2014.08.001}
-
-\hypertarget{ref-suman_2010-ez}{}
-60. Suman CS, Saenz-Arroyo A, Dawson C, Luna MC. Manual de instruccion
-de reef check california: Guia de instruccion para el monitoreo del
-bosque de sargazo en la peninsula de baja california. Pacific Palisades,
-CA, USA: Reef Check Foundation; 2010.
-
-\hypertarget{ref-gutirrez_2011-0U}{}
-61. Gutiérrez NL, Hilborn R, Defeo O. Leadership, social capital and
-incentives promote successful fisheries. Nature. 2011;470: 386--389.
-doi:\href{https://doi.org/10.1038/nature09689}{10.1038/nature09689}
-
-\hypertarget{ref-finkbeiner_2015-87}{}
-62. Finkbeiner EM, Basurto X. Re-defining co-management to facilitate
-small-scale fisheries reform: An illustration from northwest mexico.
-Marine Policy. 2015;51: 433--441.
-doi:\href{https://doi.org/10.1016/j.marpol.2014.10.010}{10.1016/j.marpol.2014.10.010}
-
-\hypertarget{ref-ball_2009-qi}{}
-63. Ball IR, Possingham HP, Watts ME. Marxan and relatives: Software for
-spatial conservation prioritization. Spatial conservation prioritization
-quantitative methods \& computational tools. United Kingdom: Oxford
-University Press; 2009. pp. 185--195. Available:
-\url{https://espace.library.uq.edu.au/view/UQ:200259}
-
-\hypertarget{ref-padleton_2017-vn}{}
-64. Padleton L, Aghmadia G, Browman H, Thurstand R, Kaplan D, Bartolino
-V. Debating the effectiveness of marine protected areas. ICES Journal of
-Marine Science. 2017;
-doi:\href{https://doi.org/10.1093/icesjms/fsx154}{10.1093/icesjms/fsx154}
-
-\hypertarget{ref-lundquist_2005-OL}{}
-65. Lundquist CJ, Granek EF. Strategies for successful marine
-conservation: Integrating socioeconomic, political, and scientific
-factors. Conserv Biol. 2005;19: 1771--1778.
-doi:\href{https://doi.org/10.1111/j.1523-1739.2005.00279.x}{10.1111/j.1523-1739.2005.00279.x}
-
-\section{Supporting information}\label{supporting-information}
-
-\textbf{S1 Appendix. Survey to collect governance information from
-fishing communities.} English version
-
-\textbf{S2 Appendix. Survey to collect governance information from
-fishing communities.} Spanish version
-
-\textbf{S3 Table. Assigned values and reasoning of socioeconomic and
-governance indicators used to color-code the scorecard in MAREA}
+\section*{Discussion}\label{discussion}
+
+We have developed and presented an automated approach for evaluating the effectiveness of marine reserves in Mexico, and perhaps around the world.
+Here we highlight MAREA's utility for evidence-based management, and comment on a few of its shortcomings.
+The findings from Isla Natividad are used purely to validate the relevance of MAREA rather than to discuss particularities of the marine reserve effectiveness, which has been described before \cite{micheli_2012-EU,rossetto_2015-V0,munguavega_2015-yg}.
+We use examples from the case study to build on the utility of MAREA and discuss ways in which results can be interpreted to inform management.
+
+The causal inference techniques used by MAREA have been suggested \cite{burgess_2018-HN,ferraro_2006-oW} and used \cite{moland_2013-VP} before in other ad hoc studies.
+This approach reduces ambiguity in the interpretation of results.
+For example, invertebrate density decreased through time inside and outside of the reserve (Figure~\ref{fig5}h).
+In this case, a before--after evaluation of the reserve (\emph{i.e.} ignoring the control site) would have incorrectly concluded that the reserve failed to protect invertebrates.
+On the other hand, a control--impact approach (\emph{i.e.} compare reserve vs.~control site only in 2016) would have identified higher densities inside the reserve, concluding that the reserve increases invertebrate density.
+However, by executing a formal difference-in-differences approach for causal inference, MAREA identifies the changes through time and across sites, and estimates the effect of the reserve on density at \(\beta_2 = 91.21\) (\emph{p} \textless{} 0.05).
+This approach reveals that invertebrate densities decrease in both sites through time, but the decrease is faster for the control site, thus yielding a positive value for \(\beta_2\).
+
+The approach used by MAREA to estimate the effect of the reserve on biological indicators requires cautious interpretation of the results.
+The value of the \(\beta_2\) coefficient represents the difference between the temporal trends of the reserve and control sites \cite{moland_2013-VP}.
+As exemplified by the case of invertebrate densities, a positive value (\emph{i.e.} \(\beta_2 > 0\)) does not necessarily indicate an increase in the indicator through time, but rather a positive difference with respect to the temporal trend of the control site.
+The inverse occurs for negative values of \(\beta_2\).
+
+MAREA provides in-depth analysis and a convenient snapshot overview of the effect of the reserve, allowing users to rapidly identify trends.
+However, users must interpret multiple indicators at a time to better understand the results.
+For example, with additional knowledge of local environmental variability (\emph{i.e.} indicator B5: Natural Disturbance), we can better understand the trends in invertebrate densities.
+As reported before \cite{micheli_2012-EU}, hypoxic conditions that have occurred in Isla Natividad can cause decreases in invertebrate densities, and reserves buffer the negative effect.
+While MAREA automates the analysis and makes results replicable, proper interpretation will still depend on the user.
+Results produced by MAREA can only aid in management and decision making when results have been correctly interpreted.
+
+Socioeconomic and governance indicators typically lack a control site, which impede us from using the causal inference techniques employed to measure biological changes \cite{mascia_2017-m_}.
+However, we can still extract useful information from them.
+Again, by combining results from multiple indicators, MAREA can provide insights into the effect of the reserve.
+For example, lobster and sea cucumber have shown increases in densities, landings, and income.
+We cannot conclude that landings and income from these species have increased due to the reserve, but we can at least conclude that landings have not decreased.
+While further information on market behavior of each fishery is needed, these results provide insights into the state of the reserve and its associated fisheries.
+
+As for the governance information, it is difficult to establish causal links between the state of the reserve and the governance structures present in the community.
+However, providing a single platform (\emph{i.e.} scorecard) or document (\emph{i.e.} technical report) where biological, socioeconomic, and governance information is comprehensively included can aid in management.
+By using MAREA, this information will be reported across reserves in a standardized way, and can help managers identify overarching patterns across sites.
+
+By making results straightforward to interpret, MAREA may also assist in communication with a broader stakeholder community.
+While stakeholder involvement in the design and implementation phases of marine reserves is important, that may not be sufficient for ensuring long-term buy-in or success.
+The scorecard is easily understandable by experts and non-experts, and can be used as an effective tool for communicating the results of annual evaluations.
+Additionally, the technical report can serve as a tool for managers and scientists to rapidly produce and communicate information at a more technical level.
+
+We recognize that the seven objectives and 29 indicators used by MAREA might not fully describe a reserve in countries other than Mexico.
+In order to ensure the applicability of the tool to reserves in other countries, further testing in other regions should take place.
+However, the proposed objectives and indicators provide a starting point to perform the evaluation, to which managers and users can add other indicators (\emph{e.g.} larval dispersal or connectivity) that are relevant to their reserve.
+Furthermore, MAREA's value is that it provides a free, simple, and replicable way to perform rigorous impact analysis.
+The tool can easily be used by fishers, CSO members, and managers in government agencies, providing transparency of the analysis and results.
+In addition, it can empower and enable local managers and fishers to respond to local change and adapt by allowing direct and easy access to the information.
+
+An evident limitation of MAREA is its dependence on data obtained through a BACI design, and the amount of samples needed to estimate coefficients in Eq~(\ref{eq:difindif}).
+It is not uncommon for control sites or baselines to be absent.
+Properly designing marine reserves by identifying control sites and establishing a baseline before the implementation of the reserve is enough to overcome this issue; reserves for which there is no control site and baseline cannot be evaluated with MAREA.
+Typical underwater surveys require that at least 12 - 16 transects are performed for each site (\emph{i.e.} reserve and control) each year.
+This provides at least 48 samples (12 samples per site, per year), enough to avoid overfitting Eq~(\ref{eq:difindif}).
+However, these problems can be easily avoided during the design and implementation phases by anticipating what data will be needed in the eventual evaluation.
+
+To the best of our knowledge, MAREA is the first tool designed to evaluate marine reserves.
+Previous work \cite{pomeroy_2005-Py,mascia_2017-m_} addressed MPA evaluation and provided the foundation for our contribution.
+However, these did not intended to create user-friendly tools to aid in the evaluation.
+Conservation management tools that automatize complex calculations can have an important impact in management \cite{ball_2009}.
+The use of open data science enables the creation of open-access tools that can address technical gaps and inprove management \cite{lowndes_2017-xh}.
+
+The effectiveness of marine reserves continues to be a matter of debate \cite{woodcock_2017-Wm,edgar_2014-UO,padleton_2017-vn}.
+With current targets set to increase ocean protection, it is important that we understand the effects of our interventions \cite{burgess_2018-HN} so we can better inform management \cite{ferraro_2006-oW}.
+It is therefore important that academics, managers, fishers, and CSOs have access to open access tools like MAREA.
+This is particularly relevant for Mexico and other Latin American countries, where management agencies are often understaffed and underfunded \cite{lundquist_2005-OL}, or where materials are often not available in their language.
+In this context, MAREA provides a simple and replicable way to align management objectives with performance indicators.
+The proposed methodologies, especially the way in which biological indicators are evaluated, provide valuable information for managers.
+We acknowledge there is room for improvement in the way in which socioeconomic and governance data are analyzed.
+Despite this, providing a unifying platform where all indicators can be analyzed and comprehensively presented represents a valuable step towards effective evidence-–based management \cite{ferraro_2006-oW}.
+
+The first release of MAREA is now available, and it will continue to be developed and maintained to keep up to date with the literature.
+This process will incorporate new features, and enhance current ones, aiming to improve user experience and expand the scope of the analysis.
+Other modifications may also include addition of more objectives and indicators to ensure applicability in other regions, full translation into other languages to avoid any ambiguities introduced via the automatic translation, or reporting effects over time in percentages to aid interpretation.
+Yet, we believe that this first release represents a major step towards effective, replicable evaluation and management of marine reserves.
+
+\section*{Supporting information}\label{supporting-information}
+
+\paragraph*{S1 Appendix}
+\label{S1_Appendix}
+{\bf Survey to collect governance information from fishing communities.} English version
+
+\paragraph*{S2 Appendix}
+\label{S2_Appendix}
+{\bf Survey to collect governance information from fishing communities.} Spanish version
+
+\paragraph*{S3 Table}
+\label{S3_Table}
+{\bf Assigned values and reasoning of socioeconomic and governance indicators used to color-code the scorecard in MAREA}
+
+\section*{Acknowledgements}\label{acknowledgements}
+
+We thank Olivier Deschenes and Andrew Plantinga, who provided valuable input to design the model that evaluates the biological indicators.
+Special thanks to the fishers from Isla Natividad, who gathered the data used in this study, and the fishers from El Rosario, who helped us validate our survey and framework, and to Arturo Hernández and Alfonso Romero who provided help with the logistics.
+Finally, we thank the editor and two anonymous reviewers for their suggestions, which significantly improved the quality of this paper.
+
+\nolinenumbers
+
+\bibliography{exported-references}
 

### 2017-12-21 Manuscript.Rmd
Updated manuscript formating with LaTeX template

@@
 The global and category–level scores are determined by the percentage of positive indicators, overall and for each category, respectively.
-For numeric biological indicators (all but B5), the color is defined by the sign of the interaction term coefficient ($\beta_2$) in eq.
+For numeric biological indicators, the color is defined by the sign of the interaction term coefficient ($\beta_2$) in eq.
 1.
@@
 General location of Isla Natividad (left) and map of the island (right).** The marine reserve polygon is indicated in red, and the approximate location of control sites is indicated by blue squares (B = Babencho, D = La Dulce).
-Shapefiles for Mexican coastline and the United States were obtained from INEGI @[INEGI] and the tmap R package [@tmap_2017], respectively.
+Shapefiles for Mexican coastline and the United States were obtained from INEGI [@INEGI] and the tmap R package [@tmap_2017], respectively.
 

### 2017-12-21 Manuscript.tex
Updated manuscript formating with LaTeX template

@@
 respond to implementation of the reserve is tested by analyzing spatial
-and temporal changes in each numeric biological indicator (all but B5)
-using generalized linear models {[}21{]}. To account for variations in
-the environment and survey conditions, covariates that are gathered
-during the underwater ecological surveys are included in the
+and temporal changes in each numeric biological indicator using
+generalized linear models {[}21{]}. To account for variations in the
+environment and survey conditions, covariates that are gathered during
+the underwater ecological surveys are included in the
 difference-in-differences model with form:
@@
 positive indicators, overall and for each category, respectively. For
-numeric biological indicators (all but B5), the color is defined by the
-sign of the interaction term coefficient (\(\beta_2\)) in eq. 1. For
-socioeconomic indicators, colors are assigned based on the direction of
-the slope (\(\beta_1\)) in eq. 3. Red, yellow, and green are used for
+numeric biological indicators, the color is defined by the sign of the
+interaction term coefficient (\(\beta_2\)) in eq. 1. For socioeconomic
+indicators, colors are assigned based on the direction of the slope
+(\(\beta_1\)) in eq. 3. Red, yellow, and green are used for
 \(\beta_i<0\), \(\beta_i = 0\), and \(\beta_i>0\), respectively. The
@@
 Babencho, D = La Dulce). Shapefiles for Mexican coastline and the United
-States were obtained from the Instituto Nacional de Estadistica,
-Geografia e Informatica of Mexico (INEGI:
-www.inegi.org.mx/geo/contenidos/geoestadistica/m\_geoestadistico.aspx)
-and the tmap R package {[}58{]}, respectively.
+States were obtained from INEGI {[}58{]} and the tmap R package
+{[}59{]}, respectively.
 
@@
 community then implemented two community-based marine reserves within
-their TURF {[}8,41,59{]} after establishing a baseline for the
+their TURF {[}8,41,60{]} after establishing a baseline for the
 soon-to-be reserves and control sites. Evidence suggest that these
@@
 information along 30 m transects, with a sampling window of 2 m x 2 m
-following a standardized ReefCheck protocol {[}60{]}. Ecological surveys
+following a standardized ReefCheck protocol {[}61{]}. Ecological surveys
 were performed yearly in each reserve and corresponding control site(s),
@@
 suggest that the community is strongly organized, which is a likely
-driver of the successes reported above {[}61{]}. The first point of
+driver of the successes reported above {[}62{]}. The first point of
 success is the existence of a fishing cooperative that is also
@@
 been shown to foster communication and cooperation {[}50,57{]};
-federations also provide bargain power with governments {[}50,62{]}.
+federations also provide bargain power with governments {[}50,63{]}.
 Access to fishing resources is managed through a TURF, permits, and
@@
 tools that automatize complex calculations can have an important impact
-in management {[}63{]}. The use of open data science enables the
+in management {[}64{]}. The use of open data science enables the
 creation of open-access tools that can address technical gaps and
@@
 The effectiveness of marine reserves continues to be a matter of debate
-{[}11,44,64{]}. With current targets set to increase ocean protection,
+{[}11,44,65{]}. With current targets set to increase ocean protection,
 it is important that we understand the effects of our interventions
@@
 and other Latin American countries, where management agencies are often
-understaffed and underfunded {[}65{]}, or where materials are often not
+understaffed and underfunded {[}66{]}, or where materials are often not
 available in their language. In this context, MAREA provides a simple
@@
 \hypertarget{ref-chavez_2003-_m}{}
-16. Chavez FP, Ryan J, Lluch-Cota SE, Niquen C M. From anchovies to
-sardines and back: Multidecadal change in the pacific ocean. Science.
-2003;299: 217--221.
+16. Chavez FP. From anchovies to sardines and back: Multidecadal change
+in the pacific ocean. Science. 2003;299: 217--221.
 doi:\href{https://doi.org/10.1126/science.1075880}{10.1126/science.1075880}
@@
 
-\hypertarget{ref-sandwich_2014}{}
+\hypertarget{ref-zeileis_2004-7n}{}
 49. Zeileis A. Econometric computing with hc and hac covariance matrix
-estimators. Journal of Statistical Software. 2004;11: 1--17.
+estimators. J Stat Softw. 2004;11.
+doi:\href{https://doi.org/10.18637/jss.v011.i10}{10.18637/jss.v011.i10}
 
@@
 
+\hypertarget{ref-INEGI}{}
+58. Estadistica Geografia e Informatica de Mexico IN de. Marco
+geoestadistico nacional. 2017. Available:
+\url{www.inegi.org.mx/geo/contenidos/geoestadistica/m_geoestadistico.aspx}
+
 \hypertarget{ref-tmap_2017}{}
-58. Tennekes M. Tmap: Thematic maps {[}Internet{]}. 2017. Available:
+59. Tennekes M. Tmap: Thematic maps {[}Internet{]}. 2017. Available:
 \url{https://CRAN.R-project.org/package=tmap}
@@
 \hypertarget{ref-afflerbach_2014-HP}{}
-59. Afflerbach JC, Lester SE, Dougherty DT, Poon SE. A global survey of
+60. Afflerbach JC, Lester SE, Dougherty DT, Poon SE. A global survey of
 -reserves, territorial use rights for fisheries coupled with marine
@@
 \hypertarget{ref-suman_2010-ez}{}
-60. Suman CS, Saenz-Arroyo A, Dawson C, Luna MC. Manual de instruccion
+61. Suman CS, Saenz-Arroyo A, Dawson C, Luna MC. Manual de instruccion
 de reef check california: Guia de instruccion para el monitoreo del
@@
 \hypertarget{ref-gutirrez_2011-0U}{}
-61. Gutiérrez NL, Hilborn R, Defeo O. Leadership, social capital and
+62. Gutiérrez NL, Hilborn R, Defeo O. Leadership, social capital and
 incentives promote successful fisheries. Nature. 2011;470: 386--389.
@@
 \hypertarget{ref-finkbeiner_2015-87}{}
-62. Finkbeiner EM, Basurto X. Re-defining co-management to facilitate
+63. Finkbeiner EM, Basurto X. Re-defining co-management to facilitate
 small-scale fisheries reform: An illustration from northwest mexico.
@@
 \hypertarget{ref-ball_2009-qi}{}
-63. Ball IR, Possingham HP, Watts ME. Marxan and relatives: Software for
+64. Ball IR, Possingham HP, Watts ME. Marxan and relatives: Software for
 spatial conservation prioritization. Spatial conservation prioritization
@@
 \hypertarget{ref-padleton_2017-vn}{}
-64. Padleton L, Aghmadia G, Browman H, Thurstand R, Kaplan D, Bartolino
+65. Padleton L, Aghmadia G, Browman H, Thurstand R, Kaplan D, Bartolino
 V. Debating the effectiveness of marine protected areas. ICES Journal of
@@
 \hypertarget{ref-lundquist_2005-OL}{}
-65. Lundquist CJ, Granek EF. Strategies for successful marine
+66. Lundquist CJ, Granek EF. Strategies for successful marine
 conservation: Integrating socioeconomic, political, and scientific

### 2017-12-21 plostemplate/ManuscriptPLOS.tex
Updated manuscript formating with LaTeX template

@@
 
-\bibliography{exported-references}
+% \bibliography{exported-references}
+\begin{thebibliography}{10}
+
+\bibitem{pauly_2005-qV}
+Pauly D, Watson R, Alder J.
+\newblock Global trends in world fisheries: impacts on marine ecosystems and
+  food security.
+\newblock Philosophical Transactions of the Royal Society B: Biological
+  Sciences. 2005;360(1453):5--12.
+\newblock doi:{10.1098/rstb.2004.1574}.
+
+\bibitem{halpern_2008-dK}
+Halpern BS, Walbridge S, Selkoe KA, Kappel CV, Micheli F, D'Agrosa C, et~al.
+\newblock A global map of human impact on marine ecosystems.
+\newblock Science. 2008;319(5865):948--952.
+\newblock doi:{10.1126/science.1149345}.
+
+\bibitem{lester_2008-F_}
+Lester S, Halpern B.
+\newblock Biological responses in marine no-take reserves versus partially
+  protected areas.
+\newblock Mar Ecol Prog Ser. 2008;367:49--56.
+\newblock doi:{10.3354/meps07599}.
+
+\bibitem{lester_2009-Ks}
+Lester S, Halpern B, Grorud-Colvert K, Lubchenco J, Ruttenberg B, Gaines S,
+  et~al.
+\newblock Biological effects within no-take marine reserves: a global
+  synthesis.
+\newblock Mar Ecol Prog Ser. 2009;384:33--46.
+\newblock doi:{10.3354/meps08029}.
+
+\bibitem{sala_2016-PV}
+Sala E, Costello C, De~Bourbon~Parme J, Fiorese M, Heal G, Kelleher K, et~al.
+\newblock Fish banks: An economic model to scale marine conservation.
+\newblock Marine Policy. 2016;73:154--161.
+\newblock doi:{10.1016/j.marpol.2016.07.032}.
+
+\bibitem{hastings_2017-sm}
+Hastings A, Gaines SD, Costello C.
+\newblock Marine reserves solve an important bycatch problem in fisheries.
+\newblock Proc Natl Acad Sci USA. 2017;114(34):8927--8934.
+\newblock doi:{10.1073/pnas.1705169114}.
+
+\bibitem{aburtooropeza_2011-ya}
+Aburto-Oropeza O, Erisman B, Galland GR, Mascareñas-Osorio I, Sala E, Ezcurra
+  E.
+\newblock Large Recovery of Fish Biomass in a No-Take Marine Reserve.
+\newblock PLoS ONE. 2011;6(8):e23601.
+\newblock doi:{10.1371/journal.pone.0023601}.
+
+\bibitem{micheli_2012-EU}
+Micheli F, Saenz-Arroyo A, Greenley A, Vazquez L, Espinoza~Montes JA, Rossetto
+  M, et~al.
+\newblock Evidence that marine reserves enhance resilience to climatic impacts.
+\newblock PLoS ONE. 2012;7(7):e40832.
+\newblock doi:{10.1371/journal.pone.0040832}.
+
+\bibitem{roberts_2017-J9}
+Roberts CM, OLeary BC, McCauley DJ, Cury PM, Duarte CM, Lubchenco J, et~al.
+\newblock Marine reserves can mitigate and promote adaptation to climate
+  change.
+\newblock Proc Natl Acad Sci USA. 2017;114(24):6167--6175.
+\newblock doi:{10.1073/pnas.1701262114}.
+
+\bibitem{munguavega_2015-yg}
+Munguía-Vega A, Sáenz-Arroyo A, Greenley AP, Espinoza-Montes JA, Palumbi SR,
+  Rossetto M, et~al.
+\newblock Marine reserves help preserve genetic diversity after impacts derived
+  from climate variability: Lessons from the pink abalone in Baja California.
+\newblock Global Ecology and Conservation. 2015;4:264--276.
+\newblock doi:{10.1016/j.gecco.2015.07.005}.
+
+\bibitem{edgar_2014-UO}
+Edgar GJ, Stuart-Smith RD, Willis TJ, Kininmonth S, Baker SC, Banks S, et~al.
+\newblock Global conservation outcomes depend on marine protected areas with
+  five key features.
+\newblock Nature. 2014;506(7487):216--220.
+\newblock doi:{10.1038/nature13022}.
+
+\bibitem{giakoumi_2017-V2}
+Giakoumi S, Scianna C, Plass-Johnson J, Micheli F, Grorud-Colvert K, Thiriet P,
+  et~al.
+\newblock Ecological effects of full and partial protection in the crowded
+  Mediterranean Sea: a regional meta-analysis.
+\newblock Sci Rep. 2017;7(1):8940.
+\newblock doi:{10.1038/s41598-017-08850-w}.
+
+\bibitem{sala_2017-69}
+Sala E, Giakoumi S.
+\newblock No-take marine reserves are the most effective protected areas in the
+  ocean.
+\newblock ICES Journal of Marine Science. 2017;doi:{10.1093/icesjms/fsx059}.
+
+\bibitem{betti_2017-lq}
+Betti F, Bavestrello G, Bo M, Asnaghi V, Chiantore M, Bava S, et~al.
+\newblock Over 10 years of variation in Mediterranean reef benthic
+  communities.
+\newblock Marine Ecology. 2017;38(3):e12439.
+\newblock doi:{10.1111/maec.12439}.
+
+\bibitem{szuwalski_2017-jc}
+Szuwalski CS, Burgess MG, Costello C, Gaines SD.
+\newblock High fishery catches through trophic cascades in China.
+\newblock Proc Natl Acad Sci USA. 2017;114(4):717--721.
+\newblock doi:{10.1073/pnas.1612722114}.
+
+\bibitem{chavez_2003-mm}
+Chavez FP.
+\newblock From anchovies to sardines and back: multidecadal change in the
+  pacific ocean.
+\newblock Science. 2003;299(5604):217--221.
+\newblock doi:{10.1126/science.1075880}.
+
+\bibitem{davies_2017-ml}
+Davies TK, Mees CC, Milner-Gulland EJ.
+\newblock Use of a counterfactual approach to evaluate the effect of area
+  closures on fishing location in a tropical tuna fishery.
+\newblock PLoS ONE. 2017;12(3):e0174758.
+\newblock doi:{10.1371/journal.pone.0174758}.
+
+\bibitem{guidetti_2014-8Z}
+Guidetti P, Baiata P, Ballesteros E, Di~Franco A, Hereu B, Macpherson E, et~al.
+\newblock Large-Scale Assessment of Mediterranean Marine Protected Areas
+  Effects on Fish Assemblages.
+\newblock PLoS ONE. 2014;9(4):e91841.
+\newblock doi:{10.1371/journal.pone.0091841}.
+
+\bibitem{friedlander_2017-oI}
+Friedlander AM, Golbuu Y, Ballesteros E, Caselle JE, Gouezo M, Olsudong D,
+  et~al.
+\newblock Size, age, and habitat determine effectiveness of Palau's Marine
+  Protected Areas.
+\newblock PLoS ONE. 2017;12(3):e0174787.
+\newblock doi:{10.1371/journal.pone.0174787}.
+
+\bibitem{rodriguez_2017-PD}
+Rodriguez AG, Fanning LM.
+\newblock Assessing Marine Protected Areas Effectiveness: A Case Study with the
+  Tobago Cays Marine Park.
+\newblock OJMS. 2017;07(03):379--408.
+\newblock doi:{10.4236/ojms.2017.73027}.
+
+\bibitem{moland_2013-VP}
+Moland E, Olsen EM, Knutsen H, Garrigou P, Espeland SH, Kleiven AR, et~al.
+\newblock Lobster and cod benefit from small-scale northern marine protected
+  areas: inference from an empirical before-after control-impact study.
+\newblock Proceedings of the Royal Society B: Biological Sciences.
+  2013;280(1754):20122679--20122679.
+\newblock doi:{10.1098/rspb.2012.2679}.
+
+\bibitem{soykan_2015-nu}
+Soykan CU, Lewison RL.
+\newblock Using community-level metrics to monitor the effects of marine
+  protected areas on biodiversity.
+\newblock Conserv Biol. 2015;29(3):775--783.
+\newblock doi:{10.1111/cobi.12445}.
+
+\bibitem{pomeroy_2005-Py}
+Pomeroy RS, Watson LM, Parks JE, Cid GA.
+\newblock How is your MPA doing? A methodology for evaluating the management
+  effectiveness of marine protected areas.
+\newblock Ocean Coast Manag. 2005;48(7-8):485--502.
+\newblock doi:{10.1016/j.ocecoaman.2005.05.004}.
+
+\bibitem{pomeroy_2004-23}
+Pomeroy RS, Parks JE, Watson LM.
+\newblock How is your MPA doing ? A guidebook of natural and social indicators
+  for evaluating marine protected areas management effectiveness.
+\newblock IUCN; 2004.
+\newblock Available from:
+  \url{http://www.iucn.org/dbtw-wpd/html/PAPS-012/cover.html}.
+
+\bibitem{mascia_2017-m_}
+Mascia MB, Fox HE, Glew L, Ahmadia GN, Agrawal A, Barnes M, et~al.
+\newblock A novel framework for analyzing conservation impacts: evaluation,
+  theory, and marine protected areas.
+\newblock Ann N Y Acad Sci. 2017;1399(1):93--115.
+\newblock doi:{10.1111/nyas.13428}.
+
+\bibitem{ostrom_2009-hg}
+Ostrom E.
+\newblock A General Framework for Analyzing Sustainability of Social-Ecological
+  Systems.
+\newblock Science. 2009;325(5939):419--422.
+\newblock doi:{10.1126/science.1172133}.
+
+\bibitem{basurto_2013-oq}
+Basurto X, Gelcich S, Ostrom E.
+\newblock The social–ecological system framework as a knowledge
+  classificatory system for benthic small-scale fisheries.
+\newblock Global Environmental Change. 2013;23(6):1366--1380.
+\newblock doi:{10.1016/j.gloenvcha.2013.08.001}.
+
+\bibitem{lowndes_2017-xh}
+Lowndes JSS, Best BD, Scarborough C, Afflerbach JC, Frazier MR, OHara CC,
+  et~al.
+\newblock Our path to better science in less time using open data science
+  tools.
+\newblock Nat ecol evol. 2017;1(6):0160.
+\newblock doi:{10.1038/s41559-017-0160}.
+
+\bibitem{halpern_2012-k9}
+Halpern BS, Longo C, Hardy D, McLeod KL, Samhouri JF, Katona SK, et~al.
+\newblock An index to assess the health and benefits of the global ocean.
+\newblock Nature. 2012;488(7413):615--620.
+\newblock doi:{10.1038/nature11397}.
+
+\bibitem{halpern_2017-Zi}
+Halpern BS, Frazier M, Afflerbach J, OHara C, Katona S, Stewart~Lowndes JS,
+  et~al.
+\newblock Drivers and implications of change in global ocean health over the
+  past five years.
+\newblock PLoS ONE. 2017;12(7):e0178267.
+\newblock doi:{10.1371/journal.pone.0178267}.
+
+\bibitem{selig_2015-F9}
+Selig ER, Frazier M, O'Leary JK, Jupiter SD, Halpern BS, Longo C, et~al.
+\newblock Measuring indicators of ocean health for an island nation: The ocean
+  health index for Fiji.
+\newblock Ecosystem Services. 2015;16:403--412.
+\newblock doi:{10.1016/j.ecoser.2014.11.007}.
+
+\bibitem{halpern_2014-lQ}
+Halpern BS, Longo C, Scarborough C, Hardy D, Best BD, Doney SC, et~al.
+\newblock Assessing the Health of the U.S. West Coast with a Regional-Scale
+  Application of the Ocean Health Index.
+\newblock PLoS ONE. 2014;9(6):e98995.
+\newblock doi:{10.1371/journal.pone.0098995}.
+
+\bibitem{elfes_2014-RC}
+Elfes CT, Longo C, Halpern BS, Hardy D, Scarborough C, Best BD, et~al.
+\newblock A Regional-Scale Ocean Health Index for Brazil.
+\newblock PLoS ONE. 2014;9(4):e92589.
+\newblock doi:{10.1371/journal.pone.0092589}.
+
+\bibitem{anderson_2015-ND}
+Anderson JL, Anderson CM, Chu J, Meredith J, Asche F, Sylvia G, et~al.
+\newblock The fishery performance indicators: A management tool for triple
+  bottom line outcomes.
+\newblock PLoS ONE. 2015;10(5):e0122809.
+\newblock doi:{10.1371/journal.pone.0122809}.
+
+\bibitem{dowling_2016-pO}
+Dowling N, Wilson J, Rudd M, Babcock E, Caillaux M, Cope J, et~al.
+\newblock FishPath: A Decision Support System for Assessing and Managing Data-
+  and Capacity- Limited Fisheries.
+\newblock In: Quinn~II T, Armstrong J, Baker M, Heifetz J, Witherell D,
+  editors. Assessing and Managing Data-Limited Fish Stocks. Alaska Sea Grant,
+  University of Alaska Fairbansk; 2016.Available from:
+  \url{https://seagrant.uaf.edu/bookstore/pubs/item.php?id=12335}.
+
+\bibitem{oyanedel_2017-TO}
+Oyanedel R, Macy~Humberstone J, Shattenkirk K, Rodriguez Van-Dyck S, Joye~Moyer
+  K, Poon S, et~al.
+\newblock A decision support tool for designing TURF-reserves.
+\newblock BMS. 2017;93(1):155--172.
+\newblock doi:{10.5343/bms.2015.1095}.
+
+\bibitem{vilela_2017-Zo}
+Vilela T, Reid J.
+\newblock Improving hydropower choices via an online and open access tool.
+\newblock PLoS ONE. 2017;12(6):e0179393.
+\newblock doi:{10.1371/journal.pone.0179393}.
+
+\bibitem{burgess_2018-HN}
+Burgess MG, Clemence M, McDermott GR, Costello C, Gaines SD.
+\newblock Five rules for pragmatic blue growth.
+\newblock Marine Policy. 2018;87:331--339.
+\newblock doi:{10.1016/j.marpol.2016.12.005}.
+
+\bibitem{nom}
+NOM-049-SAG/PESC.
+\newblock NORMA Oficial Mexicana NOM-049-SAG/PESC-2014, Que determina el
+  procedimiento para establecer zonas de refugio para los recursos pesqueros en
+  aguas de jurisdicción federal de los Estados Unidos Mexicanos.
+\newblock DOF. 2014;.
+
+\bibitem{lgeepa}
+LGEEPA.
+\newblock Ley General del Equilibrio Ecológico y la Protección al Ambiente.
+\newblock DOF. 2017;.
+
+\bibitem{lester_2017-nh}
+Lester S, McDonald G, Clemence M, Dougherty D, Szuwalski C.
+\newblock Impacts of TURFs and marine reserves on fisheries and conservation
+  goals: theory, empirical evidence, and modeling.
+\newblock BMS. 2017;93(1):173--198.
+\newblock doi:{10.5343/bms.2015.1083}.
+
+\bibitem{chirico_2017-Rz}
+Chirico AAD, McClanahan TR, Eklöf JS.
+\newblock Community- and government-managed marine protected areas increase
+  fish size, biomass and potential value.
+\newblock PLoS ONE. 2017;12(8):e0182342.
+\newblock doi:{10.1371/journal.pone.0182342}.
+
+\bibitem{rossetto_2015-V0}
+Rossetto M, Micheli F, Saenz-Arroyo A, Montes JAE, De~Leo GA.
+\newblock No-take marine reserves can enhance population persistence and
+  support the fishery of abalone.
+\newblock Can J Fish Aquat Sci. 2015;72(10):1503--1517.
+\newblock doi:{10.1139/cjfas-2013-0623}.
+
+\bibitem{woodcock_2017-Wm}
+Woodcock P, O'Leary BC, Kaiser MJ, Pullin AS.
+\newblock Your evidence or mine? Systematic evaluation of reviews of marine
+  protected area effectiveness.
+\newblock Fish Fish. 2017;18(4):668--681.
+\newblock doi:{10.1111/faf.12196}.
+
+\bibitem{carter_2017-Uf}
+Carter AB, Davies CR, Emslie MJ, Mapstone BD, Russ GR, Tobin AJ, et~al.
+\newblock Reproductive benefits of no-take marine reserves vary with region for
+  an exploited coral reef fish.
+\newblock Sci Rep. 2017;7(1):9693.
+\newblock doi:{10.1038/s41598-017-10180-w}.
+
+\bibitem{villaseorderbez2017-xE}
+Villaseñor-Derbez JC, Faro C, Wright M, Martínez J.
+\newblock A guide to evaluate the effectiveness of no-take marine reserves in
+  Mexico.
+\newblock TURFeffect; 2017.
+\newblock Available from:
+  \url{https://www.researchgate.net/publication/317840581\_A\_guide\_to\_evaluate\_the\_effectiveness\_of\_no-take\_marine\_reserves\_in\_Mexico}.
+
+\bibitem{ferraro_2006-oW}
+Ferraro PJ, Pattanayak SK.
+\newblock Money for nothing? A call for empirical evaluation of biodiversity
+  conservation investments.
+\newblock PLoS Biol. 2006;4(4):e105.
+\newblock doi:{10.1371/journal.pbio.0040105}.
+
+\bibitem{oecd_2017-VV}
+OECD. Inflation {CPI}; 2017.
+\newblock Available from: \url{https://data.oecd.org/price/inflation-cpi.htm}.
+
+\bibitem{zeileis_2004-7n}
+Zeileis A.
+\newblock Econometric Computing with HC and HAC Covariance Matrix Estimators.
+\newblock J Stat Softw. 2004;11(10).
+\newblock doi:{10.18637/jss.v011.i10}.
+
+\bibitem{espinosaromero_2014-PY}
+Espinosa-Romero MJ, Rodriguez LF, Weaver AH, Villanueva-Aznar C, Torre J.
+\newblock The changing role of NGOs in Mexican small-scale fisheries: From
+  environmental conservation to multi-scale governance.
+\newblock Marine Policy. 2014;50:290--299.
+\newblock doi:{10.1016/j.marpol.2014.07.005}.
+
+\bibitem{rcore_2017}
+{R Core Team}. R: A Language and Environment for Statistical Computing; 2017.
+\newblock Available from: \url{https://www.R-project.org/}.
+
+\bibitem{shiny_2017}
+Chang W, Cheng J, Allaire J, Xie Y, McPherson J. shiny: Web Application
+  Framework for R; 2017.
+\newblock Available from: \url{https://CRAN.R-project.org/package=shiny}.
+
+\bibitem{shinydashboard_2017}
+Chang W, {Borges Ribeiro} B. shinydashboard: Create Dashboards with 'Shiny';
+  2017.
+\newblock Available from:
+  \url{https://CRAN.R-project.org/package=shinydashboard}.
+
+\bibitem{rmarkdown_2017}
+Allaire J, Cheng J, Xie Y, McPherson J, Chang W, Allen J, et~al.. rmarkdown:
+  Dynamic Documents for R; 2017.
+\newblock Available from: \url{https://CRAN.R-project.org/package=rmarkdown}.
+
+\bibitem{knitr_2017}
+Xie Y. knitr: A General-Purpose Package for Dynamic Report Generation in R;
+  2017.
+\newblock Available from: \url{http://yihui.name/knitr/}.
+
+\bibitem{mccay_2017-1m}
+McCay B.
+\newblock Territorial use rights in fisheries of the northern Pacific coast of
+  Mexico.
+\newblock BMS. 2017;93(1):69--81.
+\newblock doi:{10.5343/bms.2015.1091}.
+
+\bibitem{mccay_2014-CN}
+McCay BJ, Micheli F, Ponce-Díaz G, Murray G, Shester G, Ramirez-Sanchez S,
+  et~al.
+\newblock Cooperatives, concessions, and co-management on the Pacific coast of
+  Mexico.
+\newblock Marine Policy. 2014;44:49--59.
+\newblock doi:{10.1016/j.marpol.2013.08.001}.
+
+\bibitem{INEGI}
+de~Estadistica Geografia e Informatica~de Mexico IN. Marco Geoestadistico
+  Nacional; 2017.
+\newblock Available from:
+  \url{www.inegi.org.mx/geo/contenidos/geoestadistica/m_geoestadistico.aspx}.
+
+\bibitem{tmap_2017}
+Tennekes M. tmap: Thematic Maps; 2017.
+\newblock Available from: \url{https://CRAN.R-project.org/package=tmap}.
+
+\bibitem{afflerbach_2014-HP}
+Afflerbach JC, Lester SE, Dougherty DT, Poon SE.
+\newblock A global survey of TURF-reserves, Territorial Use Rights for
+  Fisheries coupled with marine reserves.
+\newblock Global Ecology and Conservation. 2014;2:97--106.
+\newblock doi:{10.1016/j.gecco.2014.08.001}.
+
+\bibitem{suman_2010-ez}
+Suman CS, Saenz-Arroyo A, Dawson C, Luna MC.
+\newblock Manual de Instruccion de Reef Check California: Guia de instruccion
+  para el monitoreo del bosque de sargazo en la Peninsula de Baja California.
+\newblock Pacific Palisades, CA, USA: Reef Check Foundation; 2010.
+
+\bibitem{gutirrez_2011-0U}
+Gutiérrez NL, Hilborn R, Defeo O.
+\newblock Leadership, social capital and incentives promote successful
+  fisheries.
+\newblock Nature. 2011;470(7334):386--389.
+\newblock doi:{10.1038/nature09689}.
+
+\bibitem{finkbeiner_2015-87}
+Finkbeiner EM, Basurto X.
+\newblock Re-defining co-management to facilitate small-scale fisheries reform:
+  An illustration from northwest Mexico.
+\newblock Marine Policy. 2015;51:433--441.
+\newblock doi:{10.1016/j.marpol.2014.10.010}.
+
+\bibitem{ball_2009}
+Ball IR, Possingham HP, Watts ME.
+\newblock Marxan and relatives Software for spatial conservation
+  prioritization.
+\newblock In: Moilanen A, Wilson KA, Possingham HP, editors. Spatial
+  conservation prioritization. Quantitative methods and computational tools.
+  United Kingdom: Oxford University Press; 2009. p. 185--195.
+\newblock Available from:
+  \url{https://espace.library.uq.edu.au/view/UQ:200259}.
+
+\bibitem{padleton_2017-vn}
+Padleton LH, Aghmadia GN, Browman HI, Thurstand RH, Kaplan DM, Bartolino V.
+\newblock Debating the effectiveness of marine protected areas.
+\newblock ICES Journal of Marine Science. 2017;doi:{10.1093/icesjms/fsx154}.
+
+\bibitem{lundquist_2005-OL}
+Lundquist CJ, Granek EF.
+\newblock Strategies for successful marine conservation: integrating
+  socioeconomic, political, and scientific factors.
+\newblock Conserv Biol. 2005;19(6):1771--1778.
+\newblock doi:{10.1111/j.1523-1739.2005.00279.x}.
+
+\end{thebibliography}
 

### 2018-01-18 plostemplate/ManuscriptPLOS.tex
Added changes required by editorial team

@@
 Fish indicators showed no significant change (\emph{p} \textgreater{} 0.1), with negative trends for Shannon's diversity index and fish species richness and positive trends for density, biomass, and mean trophic level.
-Changes through time for these indicators are presented in Figure~\ref{fig5}, and a summary of \(\beta_2\) coefficients is provided in Table~\ref{table3}.
+Changes through time for these indicators are presented in Fig~\ref{fig5}, and a summary of \(\beta_2\) coefficients is provided in Table~\ref{table3}.
 
@@
 On average, lobster and sea cucumber landings increased, while urchin and snail landings and income decreased.
-Figure~\ref{fig6} presents the changes in these indicators through time, and Table~\ref{table4} summarizes these results.
+Fig~\ref{fig6} presents the changes in these indicators through time, and Table~\ref{table4} summarizes these results.
 
@@
 This approach reduces ambiguity in the interpretation of results.
-For example, invertebrate density decreased through time inside and outside of the reserve (Figure~\ref{fig5}h).
+For example, invertebrate density decreased through time inside and outside of the reserve (Fig~\ref{fig5}h).
 In this case, a before--after evaluation of the reserve (\emph{i.e.} ignoring the control site) would have incorrectly concluded that the reserve failed to protect invertebrates.
@@
 
-\paragraph*{S3 Table}
-\label{S3_Table}
+\paragraph*{S1 Table}
+\label{S1_Table}
 {\bf Assigned values and reasoning of socioeconomic and governance indicators used to color-code the scorecard in MAREA}
